---
title: "<b>Exam Prep</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: inline
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  fig.asp=.9,
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```


# Stuff we will cover

+ Process

+ Structure of exam

+ Things to take into exam

+ Notes/equation sheet

+ Exam "strategy"

+ Mock answers

---
# Process
+ Arrive in plenty of time

+ There will be direction to room

+ When you go into the room, you will need your student ID on your desk

+ we will come round and check your Notes sheets (see later slide)

+ Invigilators will give you exam instructions

---
# Exam books
+ You will have: 
  + a specific sheet for your MCQ
  + an answer book for all other questions
  + a rough work book ( *if you want to use it* )

+ You will need to fill in all of the above.
  + MCQ will need exam number
  + Answer book will need exam number, exam name, and the questions you have answered

+ When you finish, you leave all of this in the exam hall, including your notes sheet.

---
# Structure of Exam
+ Section A:
  + 20 MCQ
  + Just like quizzes
  + On all topics

+ Section B:
  + 6 questions
  + Varied marks per question
  + Short calculation, definitions, understanding model code, small bits of interpretation

+ Section C:
  + 3 questions (one per section of course)
  + Mixed models, path analysis, data reduction

---
# Things to take to exam
+ Pens (multiple) blue or black ink

+ Pencil (rough work)

+ Ruler
  + If you like, not really needed.
  
+ Calculator
  + Any type
  + No advantage to scientific vs normal
  + **Phone can not be your calculator**

---
# Notes sheet
+ Ideally double side printed.
  + if you cant, either staple back to back, or take 2 sheets
  
+ **BUT**
  + the notes must only be within the box
  
+ **If you have any notes outside the box, you will have your notes taken off of you and will be given a plain equation sheet**

+ You must hand in your notes sheet with your other exam materials.


---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# Questions

---
# Exam Strategy
+ Manage time
  + Section A: 20 marks (20-24mins)
  + Section B: 30 marks (36-40 mins)
  + Section C: 50 marks (1 hour)

+ Section A: MCQ
  + You can answer these quickly at the end
  + Do the ones you know, move on
  + Keep track of how many you have not done, so you can go back to them at the end
  
---
# Exam Strategy
+ Section B: Short answer
  + 6 questions, given time, 5mins per question
  + Calculations **show your working** (marks for this if answer is wrong)
  + Definitions: 1 to 2 sentences
  + Model code: be brief, what does it do, how does it differ etc.

---
# Exam Strategy
+ Section C: Interpretation questions
  + Three questions, 8-10 minutes each
  + Need to be able to interpret the parameters of the models we have looked at.
  + Make comments on these things
  + The questions will list the things we expect comments on
  

---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# Questions

---
# Mock Questions

+ Switch over and talk them over.

---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# Questions
