---
title: "<b> Selection & Assessment </b>"
subtitle: "Current Psychology 1<br><br> "
author: "Tom Booth"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: console
  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")

```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
   base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```

# Acknowledgement
> Materials for the next two weeks have been heavily influenced by the approach and materials of a Dr. David Hughes, Alliance Manchester Business School, University of Manchester.


---
# Topics for today
- Brief introduction to me
- Introduction to organisational psychology
- Background on core ideas from individual differences
- Background on selection and assessment
- Group discussion
- Validity of selection methods
- Adverse impact in selection

---
class: inverse, center, middle

<h2>Part 1: Introduction </h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Key concepts from individual differences </h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Selection and assessment </h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Group discussion </h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity of selection methods </h2>
<h2 style="text-align: left;opacity:0.3;">Part 6: Adverse impact in selection </h2>

---
# Me

+ PhD Organisational Psychology

+ Interests:
  + Org Psych
  + Health
  + Individual differences
  + Quant methods (psychometrics)

---
# Organisational Psychology
+ The scientific study of people in organisations!

+ Roberts (2006) ASTMA model:
  + **A**ttraction, 
  + **S**election, 
  + **T**ransformation, 
  + **M**anipulation, and 
  + **A**ttrition

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction  </h2>
<h2>Part 2: Key concepts from individual differences </h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Selection and assessment </h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Group discussion </h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity of selection methods </h2>
<h2 style="text-align: left;opacity:0.3;">Part 6: Adverse impact in selection </h2>

---
# Core principles of individual differences
+ People vary!

+ Understanding the dimensions and extent of individual variation in a variety of characteristics can help us explain differing outcomes

+ Within "classic" experimental frameworks, individual differences are classed as error to be randomised out.

+ But individual differences have been shown to be useful predictors in certain contexts.
  + Today we will look at how useful they are (or are not) in selection

---
# Trait as a continum 

```{r, echo=FALSE}
sel_a <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/trait_cont.png")
magick::image_scale(sel_a, "x300")
```

---
# Psychometrics
+ Collective term for the development and evaluation of measurement tools for individual differences.

+ Often divided into tests of...

+ **Typical performance** (personality, motivation, well-being)
  + Assessing how likely or frequently people may choose to behave in different ways
  + Often make use of Likert-type rating scales

+ **Maximal performance** (cognitive and mental abilities, knowledge tests)
  + Attempting to assess levels of competence in a given area
  + Right/wrong answers, or speeded performance

+ Wide variety of quantitative tools to analyse the resultant measures
  + Commonly factor analysis or item response theory

---
# Personality traits
+ "How shall a psychological life history be written? What processes and what structure must a full-bodied account of a personality include? How can one detect unifying threads in a life, if they exist?" **(Allport, 1968)**

+ "An individual's characteristic style of behaving,thinking, and feeling"
   + **Funder (2016)** added "together with the psychological mechanisms - hidden or not - behind" it

+ "Those characteristics that account for a person's consistent patterns of feeling, thinking, and behaving" **(Pervin & John, 1999)**

+ "A set of individual differences that are affected by the development of an individual: values, attitudes, personal memories, social relationships, habits, and skills" **(Mischel, et al., 2004)**

---
# Hierarchial Structure

+ It is widely accepted that personality is structured hierarchically.

+ At the top are broad characteristics of traits (domains)
  + These are comprised of narrower traits (facets)
    + Which are made up of items (or even narrower traits - nuances)
    
+ Where do the traits come from?
  + responses to ratings
  + In the earliest research these were adjectives from dictionary
  + In later work, these came from the items in measures
  
---
# Personlity hierarchies
+ There is no single agreed upon hierarchy, but one is dominant **the Five Factor Model**

```{r, echo=FALSE}
big5 <- tibble(
  Domain = c("Neuroticism", "Extraversion", "Openness To Experience", "Agreeableness", "Conscientiousness"),
  Facets = c("Anxiety, Hostility, Depression, Self-consciousness, Impulsiveness, Vulnerability",
             "Warmth, Gregariousness, Assertiveness, Activity, Excitement-Seeking, Positive Emotions",
             "Fantasy, Aesthetics, Feelings, Actions, Ideas, Values",
             "Trust, Straightforwardness, Altruism, Compliance, Modesty, Tender-mindedness",
             "Competence, Order, Dutifulness, Achievement Striving, Self-Discipline, Deliberation")
)

kable(big5) %>%
  kable_styling(.)
```


---
# Cognitive abilities: Definition

*"A very general mental capability that, among other things, involves the ability to reason, plan, solve problems, think abstractly, comprehend complex ideas, learn quickly, and learn from experience. It is not merely book learning, a narrow academic skill, or test-taking smarts. Rather, it reflects a broader and deeper capability for comprehending our surroundings - 'catching on', 'making sense' of things, or 'figuring out what to do.'"* 

(Gottfredson, 1994 consensus statement)

---
# How is it measured?

+ Vocabulary

+ Relations among words
  + Similarities, analogies, opposites

+ Identifying sequence progressions
  + Of numbers, letters, figural configurations

+ Short-term and working memory
  + Lists of digits, unrelated words; keeping track of one thing while doing another

+ Speed of very simple processing
  + Identifying or coding symbols; reaction, inspection times

+ Ability to visualise transformation of shapes and figure

---
# Organising principle (debated)

```{r, echo=FALSE}
sel_b <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/iq.png")
magick::image_scale(sel_b, "x300")
```

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction  </h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Key concepts from individual differences </h2>
<h2>Part 3: Selection and assessment </h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Group discussion </h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity of selection methods </h2>
<h2 style="text-align: left;opacity:0.3;">Part 6: Adverse impact in selection </h2>

---
# What do we mean by selection and assessment?
+ Finding the best candidate for the job

+ We can mean different things by "best"
  + Best in terms of ability, attempting to predict job performance
  + Best in terms of 'organisational fit'

+ As we will see, defining "best" is a key part of the process

---
# Discrimination is necessary

--

+ But we need to be clear what we mean.

--

+ Workplace selection is all about discriminating. 
  + Identifying who is "best" requires us to discriminate candidates

--

+ **BUT** We must ensure that we discriminate in a **FAIR** and **JUSTIFIED** manner. 
  + We need to assess candidates only on job relevant skills and abilities.

--

+ Hence we need to know what these skills and abilities are, and we need to identify assessments tools which can measure them in a fair way to all candidates. 

---
# Selection
+ Process must be legally defensible
  + tools used shown to predict future job performance
  + application of the tools has been consistent to all candidates
  + can be evidenced

+ Employing the 'wrong' person is costly! 
  + Recruitment costs 10-40% of 1st years salary + training costs, lost productivity, knock on effect for colleagues...
  + Human level, firing people is tough

+ Done effectively, good selection can save organisations money. 

---
# Selection Process

```{r, echo=FALSE}
sel_1 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/selection_process.png")
magick::image_scale(sel_1, "x350")
```

---
# Job Analysis (1)
**Purpose is to identify the skills and behaviours needed to be successful in the job**

+ Identifies tasks and responsibilities involved in the job

+ Provides information on skills and characteristics that are important for job success

+ Other factors: 
  + job title, 
  + hours of work, 
  + direct report, 
  + equipment, 
  + pay & benefits etc.


---
# Job Analysis (2)

**How might we do a job analysis?**

.pull-left[
**What we need to know?**

+ Tasks
+ Skills needs
+ outputs
+ Performance Criteria
+ Job context
+ Equipment needs

]

.pull-right[
**How do we gather it?**

+ Current post holders
+ Supervisors
+ Expert job analysts
+ Role documentation
+ Observation
]

---
# Methods for selecting employees

1. References
2. CV
3. Work samples
4. Job knowledge test
5. Assessment centres
6. Structured Interviews
7. Unstructured Interviews
8. Cognitive ability/ aptitude tests
9. Personality 
10. Integrity test

---
# What makes a good assessment?
+ **Reliability**:
  + Precision of measurement.

+ **Validity**
  + Accurate measurement (i.e., measuring what is claimed to be measured) 
  + Appropriate levels of prediction (i.e., predicting future job performance)

+ **Fairness**:
  + Selection methods must not unfairly discriminate against any specific subgroups

+ **Candidate reactions**:
  + Often connected to something termed *face validity*
  + Selection tools might 'put off' certain candidates

+ (For an organisation) Cost and other logistics


---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction </h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Key concepts from individual differences </h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Selection and assessment </h2>
<h2>Part 4: Group discussion </h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity of selection methods </h2>
<h2 style="text-align: left;opacity:0.3;">Part 6: Adverse impact in selection </h2>

---
# Task

- Break off into groups of 4-6
  - Discuss the selection methods introduced in the last section
  - Which are the most common?
  - Identify which you think might predict performance best
  - Identify which you think might have potential issues with bias, adverse impact on candidates etc.

- Will will take 15minutes to discuss in groups, then 15 minutes as a class discussion
  - You do not have to cover all methods
  
---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction </h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Key concepts from individual differences </h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Selection and assessment </h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Group discussion </h2>
<h2>Part 5: Validity of selection methods </h2>
<h2 style="text-align: left;opacity:0.3;">Part 6: Adverse impact in selection </h2>


---
# Evidence for personality

+ Is it useful? Does it predict performance?

+ **Morgeson, Campion, Dipboye, Hollenbeck, Murphy, & Schmitt, (2007)**:
  + Due to the low validity and content of some items, many published self-report personality tests should probably not be used for personnel selection

+ **Ones, Dilchert, Viswesvaran & Judge, (2007)**:
  + Any selection decision that does not take the key personality characteristics of job applicants into account would be deficient.

---
# FFM and work outcomes

+ **Mount & Judge (2001)**: Most comprehensive meta-analysis to-date.

+ Job Performance
  + Conscientiousness: All roles (r = .23)
  + Neuroticism: All roles (r = -.13)
  + Extraversion: Management (r = .21)
  + Openness: Training (r = .33)
  + Agreeableness: Team-working (r = .34)


---
# FFM and leadership
+ **Judge, Bono, Ilies & Gerhardt (2002)**: Meta-analysed 222 correlations from 73 samples
  + Extraversion = .31 
  + Conscientiousness = .28
  + Neuroticism = -.24
  + Openness to Experience = .24
  + Agreeableness = .08


---
# Bandwidth of prediction

```{r, echo=FALSE}
sel_2 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/pers_bandwidth.png")
magick::image_scale(sel_2, "x300")
```

*Findings from Judge et al (2013), Table from Hughes & Batey (2017)*

---
# Some issues

+ **Social desirability**: when faced psychometrics, many people feel 'judged' and so alter their answers accordingly. 

+ This may happen for two main reasons:
  + **Self-deception** - individuals are overly optimistic in their self-perceptions and play down their perceived negative aspects
  + **Impression management** - individuals try to appear 'good' because they fear social disapproval.

---
# What do others think?

```{r, echo=FALSE}
sel_3 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/pers_self_other.png")
magick::image_scale(sel_3, "x300")
```

*Results from Connelly & Ones (2010); Oh, Wang & Mount (2011); summary table from Hughes & Batey (2015)*

---
# Cognitive ability: Predictive validity (1)
+ General mental ability (GMA) is one of the best predictors of job performance (r=.51) exceeded only by work samples. (Robertson & Smith, 2001).  

+ In Europe the validity of GMA appears to be higher at .62. (Salgado, Anderson, Moscoso, Bertua & Fruyt, 2003).  


---
# Cognitive ability: Predictive validity (2)

```{r, echo=FALSE}
magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/cog_perf.png")
```

*Bertua, Anderson & Salgado (2005)*

---
# Cognitive ability: Predictive validity (3)

```{r, echo=FALSE}
magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/cog_training.png")
```

*Bertua, Anderson & Salgado (2005)*

---
# Cognitive ability: Predictive validity (4)

*".intuition may tell [people] that personality and other non-cognitive traits are more important than GMA (Hunter & Schmidt, 1996). However... Causal analyses of the determinants of job performance show that the major effect of GMA is on the acquisition of job knowledge: People who are higher in GMA acquire more job knowledge and acquire it faster. The amount of job-related knowledge required on even less complex jobs is much greater than is generally realized. Higher levels of job knowledge lead to higher levels of job performance. Viewed negatively, not knowing what one should be doing--or even not knowing all that one should about what one should be doing--is detrimental to job performance. In addition, knowing what one should be doing and how to do it depends strongly on GMA."* 

(Schmidt & Hunter, 2004)


---
# Big picture?

```{r, echo=FALSE}
sel_4 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/SH_summary.png")
magick::image_scale(sel_4, "x350")
```

*From Schmidt & Hunter (1998)*

---
# Schmidt, Hunter & Oh (2008)

```{r, echo=FALSE}
sel_5 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/trinity.png")
magick::image_scale(sel_5, "x300")
```

+ 1,577 studies
+ corrections for host of statistical artefacts
+ Personality adds 5-8%

---
# Arnold et al., (2005)

```{r, echo=FALSE}
sel_6 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/select_summary.png")
magick::image_scale(sel_6, "x350")
```

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction </h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Key concepts from individual differences </h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Selection and assessment </h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Group discussion </h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity of selection methods </h2>
<h2>Part 6: Adverse impact in selection </h2>

---
# What do we mean by adverse impact (AI)?

*Broadly speaking, adverse impact has been defined as subgroup differences in selection rates (e.g., hiring, licensure and certification, college admissions) that disadvantage subgroups protected under Title VII of the 1964 Civil Rights Act. Protected subgroups are defined on the basis of a number of demographics, including race, sex, age, religion, and national origin (Uniform Guidelines, 1978).*

(Outtz & Newman, )

---
# Formal definition of AI

$$AI = \frac{SR_1}{SR_2}$$

+ AI = adverse impact
+ $SR_1$ = selection ratio for minority group
+ $SR_2$ = selection ratio for majority group

$$SR = \frac{\text{Number of people hired from a group}}{\text{Total number of applicants from group}}$$
+ AI values of 1.0 are optimal (identical selection ratio)
  + In the US, AI of 0.80 is used as a marker when considering compliance (Roth, Bobko, & Switzer, 2006)

---
# Why is this considered a major issue

+ The definitions concern selection rates, so we may ask why this is a problem

+ It becomes a potential issue when we also note that many of the tools used for selection (particularly cognitive tests), have been shown to have mean score level differences across groups.
  + The magnitude of these differences is much debated
  + As are the reason why such differences might be observed

---
# Outtz & Newman

```{r, echo=FALSE}
ai_1 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/ai_i.png")
magick::image_scale(ai_1, "x350")
```

--- 
# Outtz & Newman
+ Based on meta-analytic estimates (Newman, Hanges, & Outtz, 2007), they argue that:

  1. The semi-partial correlation between race and cognitive test scores (controlling) for job performance is 0.42
  2. Other estimates put component a from previous slide at 13% 
  3. If we translate this into odds ratios, this would lead to a white applicant with equal expected job performance being 7.4 times more likely to be hired.

+ Important to note these are one set of estimates, and are also dependent on the proportion of applicants from different groups, but it is illustrative of how important the issue could be.

---
# Theory of AI (Outtz & Newman)

```{r, echo=FALSE}
ai_2 <- magick::image_read("C:/Work/Teaching/psychstats/dapr3/dapr3_lectures/pathfa_lectures/ai_2.png")
magick::image_scale(ai_2, "x350")
```

---
# Reducing adverse impact
+ Suggestions range from..
  + the simple - use multiple measures
  + to the extraordinarily complex major societal interventions in key areas around education and opportunity

+ Outtz & Newman argue that $r_{PRIV}$ should be a focal parameter in psychometric measure development for selection.
  + The closer a measure is to performance on the job, the better, as this should reduce $r_{PRIV}$ by making more test variance job related

+ Aguinis & Smith (2007): Use data on test bias explicitly in the decision making process.

---
# Summary
+ Today we have outlined:
  + Background to individual differences in context of selection
  + Discussed a variety of selection tools
  + Compared the predictive validity of these tools
  + Defined adverse impact
  + Discussed adverse impact with respect to various selection tools

---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# That's all for this week
