---
title: "<b>WEEK 2<br>Path Mediation</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      navigation:
        scroll: false
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()

library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(lavaan)

knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.asp=.7
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="70%","overflow-y"="scroll"))
)
```

# Learning Objectives
1. Understand the purpose of mediation models and the conceptual challenges
2. Be able to describe direct, indirect and total effects in a mediation model.
3. Estimate and interpret a mediation model using `lavaan`

---
class: inverse, center, middle

<h2>Part 1: Introduction to mediation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Direct, indirect and total effects</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Estimating mediation in `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reporting</h2>

---
# Mediation

- Is when a predictor X, has an effect on an outcome Y, via a mediating variable M

- The mediator **transmits** the effect of X  to Y 

- Examples of mediation hypotheses:
    - Conscientiousness (X) affects health (Y) via health behaviors (M)
    - Conduct problems (X) increase the risk of depression (Y) via peer problems (M)
    - Attitudes to smoking (X) predict intentions to smoke (M) which in turn predicts smoking behavior (Y) 
    - An intervention (X) to reduce youth crime (Y) works by increasing youth self-control (M)


---
# Traditional approaches
- Traditional approaches to mediation fit a series of linear models, and interpretations were based comparing across models.
  - A classic you may see in papers will refer to Baron and Kenny

- These approaches suffer from:
  - Low power
  - Very cumbersome for multiple mediators, predictors, or outcomes
  - You don't get significance tests for many of the things we are interested in when we have mediation questions

- Much better way: **path mediation model**

 
---
# Visualising a mediation model

.pull-left[
- (Right) Diagram of a simple mediation models

- Conscientiousness (X) affects health (Y) via health behaviors (M)

]

.pull-right[
```{r, echo=FALSE}
magick::image_read("figs/basic_mediation.png")
```
]


---
# Cautions regarding path mediation
- We are going to talk about path mediation models, but these should only really be used with longitudinal data.

- Consider our example:
  - Conscientiousness (X) affects health (Y) via health behaviors (M)

- This only really makes sense if we say:
  - We believe people have varying levels of Conscientiousness
    - This will lead them to behave in specific health related ways
      - And these behaviors will subsequently have a health impact

- This happens over time. So to test it, we need longitudinal data.
  - X time 1
  - M time 2
  - Y time 3

---
# Cautions: Indistinguishable models

- Of course it is possible to do on cross-sectional data, but there is a big conceptual problem.
  - We are modelling correlations.
  - When we only have cross-sectional data, we have multiple **indistinguishable** models
  - So there is **nothing** to demonstrate one model is better than another.


---
# Cautions: Indistinguishable models

```{r, echo=FALSE}
magick::image_read("figs/med versus confounding.png")
```

---
# Mediation...not to be confused with moderation
- Mediation is commonly confused with **moderation** 

- Moderation is when a moderator z modifies the effect of X on Y
  - e.g., the effect of X on Y is stronger at higher levels of Z
  - Also known as an **interaction** between X and Z

- Examples of moderation could be:
    - An intervention (X) works better to reduce bullying (Y) at older ages (Z) of school pupil
    - The relation between stress (X) and depression (Y) is lower for those scoring higher on spirituality (Z) 

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 1

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to mediation</h2>
<h2>Part 2: Direct, indirect and total effects</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Estimating mediation in `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reporting</h2>  



---
# Total, direct and indirect effects in path mediation

- The overall effect of a predictor on an outcome is known as the **total**

- The total effect comprises two sources of influence of X on Y, namely the **indirect effects** and **direct effects**


$$Total = Indirect + Direct$$
- The indirect effects of X on Y are those transmitted via the mediator

- The direct effect of X on Y is the remaining effect of X on Y



---
# Visualing direct and indirect effects in mediation

```{r, echo=FALSE}
magick::image_read("figs/indirect_direct.png")
```

---
# Estimating an indirect effect

.pull-left[
- To actually calculate the value of an indirect effect, we multiply together the paths x->M and M->Y

- We can show this more clearly if we label the paths in our diagram

$$Total = a*b + c$$

]


.pull-right[
```{r, echo=FALSE}
magick::image_read("figs/path_label.png")
```
]


---
# Interpreting the total, direct, and indirect effect coefficients
- The total effect can be interpreted as the **unit increase in Y expected to occur when X increases by one unit**

- The indirect effect can be interpreted as the **unit increase in Y expected to occur via M when X increases by one unit**

- The direct effect can be interpreted as the **unit increase in Y expected to occur with a unit increase in X over and above the increase transmitted by M**

- **Note**: 'direct' effect may not actually be direct - it may be acting via other mediators not included in our model


---
# Testing mediation
- Demonstrating mediation will usually rely on a number of things:

1. Evaluating the significance of the indirect, direct and total effects
2. Considering the proportion of the total effect which is due to the indirect path


$$ProportionMediated = \frac{indirect}{total}$$ 

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 2

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to mediation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Direct, indirect and total effects</h2>
<h2>Part 3: Estimating mediation in `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reporting</h2>

---
# Testing a path mediation model in lavaan

- Specification
  - Create a lavaan syntax object

- Estimation
  - Estimate the model using e.g., maximum likelihood estimation

- Evaluation/interpretation
  - Inspect the model to judge how good it is
  - Interpret the parameter estimates


---
# Example

+ Researchers are interested in looking at self-reported subjective well-being in the workplace.

+ They believe that abusive leadership behaviours will result in more interpersonal aggression at work, and subsequently reduced well-being.

+ Key question: Does interpersonal aggression mediate the effect of abusive leadership behaviour on psychological well-being, over and above the effects of sleep and exercise?

+ They want to control for well known health variables which impact subjective well-being, so also measure hours of exercise, and hours of sleep.


```{r, echo=FALSE}
# parameters
n <- 550
b1 <- 0.4
b2 <- 0.38
b3 <- 0.01
b4 <- 0.22
b5 <- 0.18
var1 <- 1
var2 <- 1
var3 <- 1
res1 <- 0.16
res2 <- 0.225

# Sim data
set.seed(7284)

leader <- rnorm(n, mean=0, sd = sqrt(var1))
sleep <- rnorm(n, mean=0, sd = sqrt(var2))
exercise <- rnorm(n, mean=0, sd = sqrt(var3))
aggression <- b1*leader + rnorm(n, sd = sqrt(res1))
swb <- b3*leader + b2*aggression + b4*exercise + b5*sleep + rnorm(n, sd = sqrt(res2))

# Create data set
leader_dat <- tibble(
  ID = paste("participant", 1:550, sep = ""),
  leader = round(leader,2),
  sleep = sleep,
  exercise = exercise,
  aggression = round(aggression,2),
  swb = round(swb,2)
) %>%
  mutate(
    sleep = round(sleep+abs(min(sleep)), 1) +1.5,
    exercise = round(exercise + abs(min(exercise)))
    
  )
```

---
# Example model

```{r, echo=FALSE}
magick::image_read("figs/example_l7.png")
```

---
# Data

```{r}
slice(leader_dat, 1:10)
```


---
# Basic model code
+ Does interpersonal aggression mediate the effect of abusive leadership behaviour on psychological well-being, over and above the effects of sleep and exercise?

```{r}
model1<-'aggression ~ leader  # aggression (M) predicted by leader abusive behaviour (X)
         swb ~ aggression     # well-being (Y) predicted by aggression (M)
         swb ~ leader         # well-being (Y) predicted by leader abusive behaviour (X): direct effect
         swb ~ exercise + sleep # covariates
'     
```

+ Note we could combine all the predictors of `swb` into a single line.
  + It is split here for ease of reading
  
---
# Coding Indirect and Total Effects
- To calculate the indirect effect of X on Y in path mediation, we need to create some new parameters

- First we label those from the simple model:
  - $a$ = the regression coefficient for M~X
  - $b$ = the regression coefficient for Y~M
  - $c$ = the regression coefficient for M~X

- We then use := operator to create a new parameter
  - Name appears on the left (here `ind` and `tot`), and the calculation on the right

```{r}
model1<-'aggression ~ a*leader  # aggression (M) predicted by leader abusive behaviour (X)
         swb ~ b*aggression     # well-being (Y) predicted by aggression (M)
         swb ~ c*leader         # well-being (Y) predicted by leader abusive behaviour (X): direct effect
         swb ~ exercise + sleep # covariates
         
ind := a*b
tot := (a*b)+c
'     
```


---
# Estimating the model

```{r}
model1.est<-sem(model1, data=leader_dat)
```

- This is very straight-forward
  - As we have noted we can generally rely on the defaults for basic path models

---
# Model Evaluation

+ Typically we want to see:
  + Model estimates
  + Model fit
  + Standardized solutions
  + Possibly modification indices
  
+ We get those by:

```{r, eval=FALSE}
summary(model1.est, 
        fit.measures = T, 
        std=T, 
        modindices = T)
```

---
# Model Output

.scroll-output[
```{r}
summary(model1.est)
```
]

---
# Things to note from the model output (1)
+ All effects other than the direct effect of abusive leadership on well-being are significant.

+ We have positive degrees of freedom so we can assess model fit.


.scroll-output[
```{r}
summary(model1.est, 
        fit.measures = T)
```
]

---
# Statistical significance of the indirect effects
- Default method of assessing the statistical significance of indirect effects assume normal sampling distribution

- May not hold for indirect effects which are the product of regression coefficients

- Instead we can use **bootstrapping**
  - Allows 95% confidence intervals (CIs) to be computed
  - If 95% CI includes 0, the indirect effect is not significant at alpha=.05
    
---
# Bootstapped CIs for indirect effect in lavaan


- Run the model:

```{r}
model1.est<-sem(model1, data=leader_dat, se='bootstrap') #we add the argument se='bootstrap'
```

- View the output with CI's

.scroll-output[
```{r indirect effect with bootstrapped CIs output, eval=FALSE}
summary(model1.est, ci=T) # we add the argument ci=T to see the confidence intervals in the output
```
]

---
# Bootstrap CI output

.scroll-output[
```{r }
summary(model1.est, ci=T) # we add the argument ci=T to see the confidence intervals in the output
```
]


---
# Standardised parameters
- As with other statistical analyses, if our units of measurement do not have easy interpretations, it may be beneficial to standardized results. 

- Standardized parameters can be obtained using:

```{r, eval=FALSE}
summary(model1.est, ci=T, std=T)
```

---
# Standardised parameters
.scroll-output[
```{r stand param, echo=FALSE}
summary(model1.est, ci=T, std=T)
```
]


---
# What if my model doesn't fit?
+ In this case our model fits well. But what if it doesn't?

+ First, we should not draw substantive conclusion if fit is poor. And we could stop here.

+ If we want to understand why the fit is poor, we can look at modification indices.

.scroll-output[
```{r}
modindices(model1.est)
```

]

---
# Making model modifications
- You **may** want to make some modifications to your initially hypothesized model 
  - non-significant paths that you want to trim
  - include some additional paths not initially included

- As soon as we make a modification, we are no longer testing a model in a confirmatory way.
  - Our analysis switches to being led by the data, not the theory. 
  - This is why it is generally not preferred.

- If we do:
  - Model modifications should be substantively as well as statistically justifiable
  - You must be aware of the possibility that you are capitalizing on chance
  - You should aim to replicate the modifications in independent data

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 3

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to mediation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Direct, indirect and total effects</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Estimating mediation in `lavaan`</h2>
<h2>Part 4: Reporting</h2>

---
# Reporting path mediation models 

- Methods/ Analysis Strategy
  - The model being tested
  - e.g. 'Y was regressed on both X and M and M was regressed on X'
  - The estimator used (e.g., maximum likelihood estimation)
  - The method used to test the significance of indirect effects ('bootstrapped 95% confidence intervals')

- Results
  - Model fit (for over-identified models)
  - The parameter estimates for the path mediation  and their statistical significance
  - Can be useful to present these in a SEM diagram
  - The diagrams from R not considered 'publication quality' draw in powerpoint or similar


---
# Reporting path mediation models - example of SEM diagram with results

- Include the key parameter estimates

- Indicate statistically significant paths (e.g. with an '*')

- Include a figure note that explains how statistically significant paths (and at what level) are signified 

---
# Example Diagram

```{r, echo=FALSE}
magick::image_read("figs/example_l7_param.png")
```

---
# Visualising the model 
- There are a number of packages in R that will produce path diagrams.

- The default presentation of these diagrams is often not clear.
  - And it can take some time to master refining them

- All diagrams in this presentation were made in powerpoint and saved as image files
  - I would strongly advocate this approach.
  

---
# Reporting path mediation models - the indirect effects

- Results
 - The coefficient for the indirect effect and the bootstrapped 95% confidence intervals.
 
> The indirect effect of abusive leadership on well-being via workplace aggression was significant ( $\beta$ = 0.16, bootstrap 95% CI [0.12, 0.20])
 
 - Common to also report **proportion mediation**
  
- However, important to be aware of limitations:
  - Big proportion mediation possible when total effect is small - makes effect seem more impressive
  - Small proportion mediation even when total effect is big - can underplay importance of effect
  - Should be interpreted in context of total effect

- Tricky interpretation if there are a mix of negative and positive effects involved

---
## Other path analysis models
- Path mediation models are a common application of path models
  - But they are just one example

- Anything that can be expressed in terms of regressions between observed variables can be tested as a path model
  - Can include ordinal or binary variables
  - Can include moderation

- Other common path analysis models include:
    - Auto-regressive models for longitudinal data
    - Cross-lagged panel models for longitudinal data


---
# Path analysis summary
- Path analysis can be used to fit sets of regression models
  - Common path analysis model is the path mediation model
  - But very flexible huge range of models that can be tested

- In R, path analysis can be done using the `sem()` function in `lavaan`

- Need to be aware that we aren't *testing* causality but assuming it


---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# End

