---
title: "week 3"
author: "uoepsystat"
date: "9/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(sjPlot)
```


# Levels

## Example 1: Intervention study
A research team conducts an intervention study on exercise. They want to know if total hours exercise increases whether someone uses an in gym personal trainer, or has one-to-one sessions from home. They randomly assign 100 people to each condition. They measure number of hours exercised one week pre, and two weeks post, a 3 week training intervention period. They also measure a set of demographics.

1. What is the outcome?
  total hours of exercise 
2. What are the predictors?
  timeperiod, in-gym vs at-home, covariates such as demographics
3. What aspect of the study creates dependent structure?
  repeated measures for each participant
4. Due to (3) what are the potential random effects?
  some participants will exercise more, some less.
  participants vary in the amount they generally exercise
  participants may change differently pre/post 

```{r eval=FALSE}
lmer(total_hours_exercise ~ 1 + age + timeperiod * intervention +
       (1 + timeperiod | participants)
     )
```


## Example 2: Longitudinal study
A research team is interested in change in aggressive behaviour across adolescents. They measure aggression using a questionnaire measure every year from age 7 to age 17. The children in the study come from different areas of the same city.

1. What is the outcome?
  aggressive behaviour questionnaire
2. What are the predictors?
  age
3. in what way are our observations grouped?
  multiple observations per child
  multiple children per area of city
4. Due to (3) what are the potential random effects?
  overall aggressive behaviour varies by both participant and by area
  effect of age on aggressive behaviour varying by participant and by area
  

```{r eval=FALSE}
lmer(aggression ~ 1 + age +
       (1 + age | cityarea/child))
lmer(aggression ~ 1 + age +
       (1 | cityarea) +
       (1 + age | cityarea:child)
     )

```


# applet

http://mfviz.com/hierarchical-models/  

# Intervention study

The data used for this worked example are simulated to represent data from 50 participants, each measured at 3 different time-points (pre, during, and post) on a measure of stress. Half of the participants are in the control group, and half are in the treatment group (e.g. half received some cognitive behavioural therapy (CBT), half did not). 

```{r}
set.seed(983)
library(tidyverse)
simMIX <- tibble(
  ppt = factor(rep(paste("ID", 1:50, sep=""),each=3)),
  ppt_int = rep(rnorm(50,0,15), each=3), 
  stress = round(
    c(rnorm(75,c(50,45,42),sd=c(5, 6, 5)),
      rnorm(75,c(55,30,20),sd=c(6, 10, 5))) + ppt_int,
    0),
  time = as_factor(rep(c("Pre", "During", "Post"), each=1,50)),
  group = as_factor(c(rep("Control", 75), rep("Treatment", 75)))
) %>% select(-ppt_int)
```

```{r}
simMIX
```

## plotting - stat_summary

```{r}
simMIX %>%
ggplot(., aes(x = time, y = stress))+
  geom_line(aes(group=ppt), alpha=.2)+
  facet_wrap(~group)+
  stat_summary(geom="pointrange")
```


## fitting models

The data used for this worked example are simulated to represent data from 50 participants, each measured at 3 different time-points (pre, during, and post) on a measure of stress. Half of the participants are in the control group, and half are in the treatment group (e.g. half received some cognitive behavioural therapy (CBT), half did not). 

```{r}
m1 <- lmer(stress ~ time * group + 
       (1 | ppt),
     data = simMIX)
summary(m1)

library(broom.mixed)
augment(m1) %>%
  ggplot(.,aes(x=time))+
  geom_line(aes(y = stress, group=ppt),alpha=.2)+
  geom_line(aes(y = .fitted, group=ppt))+
  facet_wrap(~group)

ggplot(augment(m1), aes(x=.fitted,y=stress))+geom_point()

sjPlot::plot_model(m1,type="int")
```

