---
title: "<b>WEEK 4<br>Centering Predictors<br>Generalisations</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "Josiah King"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: "AY 2021-2022"
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  code_font_size = "0.7rem",
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```


---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: EXAMPLE 1</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: EXAMPLE 2</h2>

---
# EXAMPLE

- experience sampling study
- negative affect ~ stress
- within / between 


```{r eval=F, echo=F}
filepath <- "https://quantdev.ssri.psu.edu/sites/qdev/files/AMIBshare_persons_2019_0501.csv"
#read in the .csv file using the url() function
AMIB_persons <- read.csv(file=url(filepath),header=TRUE)
#subsetting to variables of interest
AMIB_persons <- AMIB_persons[ ,c("id","bfi_n")]
#set filepath for data file
filepath <- "https://quantdev.ssri.psu.edu/sites/qdev/files/AMIBshare_daily_2019_0501.csv"
#read in the .csv file using the url() function
AMIB_daily <- read.csv(file=url(filepath),header=TRUE)
#subsetting to variables of interest
AMIB_daily <- AMIB_daily[ ,c("id","day","negaff","pss")]
#reverse coding the pss variable into a new stress variable
AMIB_daily$stress <- 4 - AMIB_daily$pss
#describing new variable
describe(AMIB_daily$stress)
#histogram
ggplot(data=AMIB_daily, aes(x=stress)) +
  geom_histogram(fill="white", color="black",bins=20) +
  labs(x = "Stress (high = more stressed)")
AMIB_imeans <- ddply(AMIB_daily, "id", summarize,
                       stress_trait = mean(stress, na.rm=TRUE),
                       negaff_trait = mean(negaff, na.rm=TRUE))
describe(AMIB_imeans)
#merging into person-level file
AMIB_persons <- merge(AMIB_persons, AMIB_imeans, by="id")   
#make centered versions of the person-level scores
AMIB_persons$bfi_n_c <- scale(AMIB_persons$bfi_n,center=TRUE,scale=FALSE)
AMIB_persons$stress_trait_c <- scale(AMIB_persons$stress_trait,center=TRUE,scale=FALSE)
#describe person-level data
describe(AMIB_persons)
#merging person-level data into daily data
daily_long <- merge(AMIB_daily,AMIB_persons,by="id")
#calculating state variables
daily_long$stress_state <- daily_long$stress - daily_long$stress_trait
daily_long$negaff_state <- daily_long$negaff - daily_long$negaff_trait
#describing data
describe(daily_long)
```


```{r eval=F, echo=F}
daily_long %>%
ggplot(., aes(x=stress,y=negaff))+
  geom_point()+
  geom_smooth(aes(group=id), se=F, method="lm", col="black",lwd=.5)

lmer(negaff ~ stress + (1+stress|id), daily_long) %>% summary

lmer(negaff_state~ stress + (1+stress|id), daily_long) %>%
  broom.mixed::augment() %>% 
  ggplot(.,aes(x=stress,y=.fitted,group=id))+
  geom_line(alpha=.2)

lmer(negaff_state ~ stress_state * bfi_n_c + (1+stress_state|id), daily_long) %>%
  broom.mixed::augment() %>% 
  ggplot(.,aes(x=bfi_n_c,y=.fitted,group=id))+
  geom_line(alpha=.2)

```

---

a write up/paper?

Ram, N., Conroy, D. E., Pincus, A. L., Hyde, A. L., & Molloy, L. E. (2012). Tethering theory to method: Using measures of intraindividual variability to operationalize individuals’ dynamic characteristics. In G. Hancock & J. Harring (Eds.), Advances in longitudinal methods in the social and behavioral sciences (pp. 81-110). New York: Information Age.





---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 2

---
class: inverse, center, middle



---
# End

