---
title: "01 playground"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)

source('simfns.R')  # load simdat functions
```


# week 1 examples

## lifesat ~ treatment + day + (1 + day | ppt)

lifesat: continuous, measured after practicing each method daily for a week
treatment: journal, meditation (btwn)
day 

```{r}
half_ppts <- 18
journal <- sim_int_1grp_1pred(seed=2, b0 = 1, b1=.1, N = half_ppts*7, n_groups = half_ppts) |>
  mutate(
    condition = 'journal',
    ppt_id = paste0('j', g)
  ) |>
  select(-y_bin, -g) 
meditate <- sim_int_1grp_1pred(seed=4, b0 = 1, b1=.5, N = half_ppts*7, n_groups = half_ppts) |>
  mutate(
    condition = 'meditate',
    ppt_id = paste0('m', g)
  ) |>
  select(-y_bin, -g)

lifesat_dat <- bind_rows(journal, meditate) |>
  rename(day = x1) |>
  mutate(lifesat = datawizard::rescale(y, to = c(15, 75))) |>
  select(-y)
```

### simple lm

btwn ppt: day 7 only

```{r}
lifesat_dat_7 <- lifesat_dat |>
  filter(day == 7) |>
  select(-day)

lifesat_dat_7 |>
  ggplot(aes(x = condition, y = lifesat)) +
  geom_violin()
```

what categ vars in dataset?

```{r}
lifesat_dat_7
```

- condition
- ppt_id


what categ vars not an outcome or a predictor?

- ppt_id



what levels?

```{r}
lifesat_dat_7$ppt_id
```

is each level associated with more than one data point?

```{r}
lifesat_dat_7 |>
  group_by(ppt_id) |>
  count()
```

nope, appears only 1x.



### lmm

Good for Week 1 and Week 2

```{r}
lifesat_dat |>
  ggplot(aes(x = day, y = lifesat, colour = condition)) +
  geom_point() +
  facet_wrap(~ ppt_id)
```

what categ vars in dataset?

- condition
- ppt_id


what categ vars not an outcome or a predictor?

- ppt_id


what levels?

```{r}
lifesat_dat$ppt_id
```

is each level associated with more than one data point?

```{r}
lifesat_dat
```

one way: just look at the df. the answer is yes bc j1 appears more than 1x, j2 appears more than 1x, etc.

another way, more efficient: use tidyverse to count how many times each ppt label appears.

```{r}
lifesat_dat |>
  group_by(ppt_id) |>
  count()
```




## attentiveness ~ hour + (1 + hour | child)

Good for Week 1 and Week 2


children's attentiveness at school across the day

```{r}
attent <- sim_int_1grp_1pred(seed=1, b0 = 2, b1=-.2, N = 140) |>
  rename(
    hour = x1,
    child = g,
    attentiveness = y
  ) |>
  select(-y_bin) |>
  mutate(
    hour = hour - 1,
    child = factor(child),
    attentiveness_z = attentiveness - mean(attentiveness)
  )
 
attent |>
  ggplot(aes(x = hour, y = attentiveness_z)) +
  geom_point() +
  facet_wrap(~ child, labeller = label_both)
```


What categorical variables does your data contain?

- child


How many times does each each level of this variable appear in the data?

```{r}
attent |>
  group_by(child) |>
  count()
```

each child contributes seven data points to the data set.
so the data has a grouping by child.


## correct ~ trial_type + (1 + trial_type | ppt) + (1 | item)

Good for Week 1 and Week 2

How does people's accuracy on an experimental task differ between conditions, and how is this difference affected by people's DASS scores?

```{r}
dass <- read_csv('../data/dass.csv')
dass |> head(20)
```


what categ vars in dataset?

- ppt_id
- target_id
- condition
- correct


what categ vars not an outcome or a predictor?

- ppt_id
- target_id 


for each categ var, what levels does it have?

```{r}
dass$ppt_id |> unique()
dass$target_id |> unique()
```

is each of those levels associated with more than one data point?

```{r}
dass |>
  group_by(ppt_id) |>
  count()
```

yes

```{r}
dass |>
  group_by(target_id) |>
  count()
```

yes.

# week 1 lab examples

## y ~ x + (1 + x | ppts)

```{r}
d1 <- simcross(seed=1, b3 = 3, n_itms = 6, n_ppts = 12, n_obs_per = 3) |>
  select(ppts, y, x) |>
  mutate(y = datawizard::rescale(y, to = c(0, 12)))

d1 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(ppts))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```

## y ~ x 

```{r}
d2 <- simcross(seed=2, b3 = 3, n_itms = 2, n_ppts = 4, n_obs_per = 24) |>
  select(y, x) |>
  rowid_to_column() |>
  rename(ppt_id = rowid) |>
  mutate(
    y = datawizard::rescale(y, to = c(0, 48)),
  )

d2
```


## y ~ x

```{r}
d3 <- MASS::mvrnorm(64, mu = c(0, 0), Sigma = matrix(c(5, 1, 1, 7), nrow=2, byrow=T)) |>
  as_tibble() |>
  rowid_to_column() |>
  rename(
    y = V1,
    x = V2,
    book = rowid
  )

d3 

d3 |>
  ggplot(aes(x = x, y = y)) + geom_point()
```


## y ~ x + (1 + x | ppts) + (1 | itms)

```{r}
d4 <- simcross(seed=4, b0 = 20, b2 = 10, n_itms = 20) |>
  select(ppts, itms, y, wi_ppts_btw_itms) |>
  rename(x = wi_ppts_btw_itms)

d4 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(itms))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```

```{r}
d4 |>
  group_by(ppts) |>
  count()

d4 |>
  group_by(itms) |>
  count()
```

^ grouping structures for both ppts and items

```{r}
d4 |>
  group_by(ppts, x) |>
  count()
```

- each ppt sees both vals of x, so rd slope is poss

```{r}
d4 |>
  group_by(itms, x) |>
  count()
```

- each item does NOT appear w both vals of x, so rd slope is not poss



## y ~ x + (1 | itms)


```{r}
d5 <- simcross(seed=5, b0 = -10, b2 = -5) |>
  select(itms, y, wi_ppts_btw_itms) |>
  rename(x = wi_ppts_btw_itms)

d5 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(itms))) +
  geom_point()
```

- by-item grouping struc bc rep'd obs of ea item
- but no slopes possible by item over cond x, bc not every item appears in every cond x

```{r}
d5 |>
  group_by(itms, x) |>
  count()
```


## y ~ x + (1 + x | ppts) + (1 + x | itms)

```{r}
d6 <- simcross(seed=6, b0 = 4, b3 = 2, n_itms = 6, n_obs_per = 10) |>
  select(ppts, itms, y, x) |>
  mutate(
    # x = datawizard::rescale(x, to = c(pi, pi*3)),
    y = abs(log(y))
  )
  # rename(cond = wi_ppts_btw_itms)

d6 |>
  ggplot(aes(x = x, y = y, colour = factor(itms))) +
  facet_wrap(~ ppts) +
  geom_point() +
  theme(legend.position = 'none')
```



## y ~ x + z + (1 + x + z | ppts) + (1 + x | itms)

```{r}
d7 <- simcross(seed=7, b0 = 0, b3 = 0, n_itms = 2, n_obs_per = 6) |>
  select(ppts, itms, y, x, wi_ppts_btw_itms) |>
  rename(z = wi_ppts_btw_itms)

d7
```


```{r}
d7 |>
  ggplot(aes(x = factor(z), y = y, colour = factor(itms))) +
  facet_wrap(~ ppts, labeller = label_both) +
  geom_point() +
  theme(legend.position = 'none')
```


## y ~ x + (1 + x | ppts) + (1 | itms)

```{r}
d8 <- simcross(seed=8, b0 = 0, b2 = -10, n_ppts = 4, n_itms = 8) |>
  select(ppts, itms, y, wi_ppts_btw_itms) |>
  rename(x = wi_ppts_btw_itms)

d8 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(itms))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```



## y ~ x + (1 | itms)

```{r}
d9 <- simcross(seed=9, b0 = -10, b2 = -5, n_itms = 50, n_obs_per = 5) |>
  select(itms, y, x) |>
  mutate(
    x = sample(x),
    y = abs(y)
    )

d9
```


```{r}
d9 |>
  ggplot(aes(x = x, y = y, colour = factor(itms))) +
  geom_point() +
  theme(legend.position = 'none')
```




## y ~ x + (1 + x | ppts)

```{r}
d10 <- simcross(seed=10, b3 = 2, n_itms = 2, n_ppts = 6, n_obs_per = 8) |>
  select(ppts, y, wi_itms_wi_ppts) |>
  rename(x = wi_itms_wi_ppts)

d10 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(ppts))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```
