---
title: "02 playground"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)
library(lme4)

source('simfns.R')  # load simdat functions
```

# One idea: erm_belief

```{r}
df <- read_csv("https://uoepsy.github.io/data/erm_belief.csv")

# make trial_n numeric
df$trial_n <- as.numeric(gsub("trial_","",df$trial_n))
# relevel condition:
df$condition <- factor(df$condition, levels=c("fluent","disfluent"))
```


```{r eval=F}
# failed to converge
mod1 <- lmer(
  belief ~ trial_n + condition + 
    (1 + trial_n + condition | ppt) + 
    (1 + condition | statement), 
  data = df
)

# converges, and vcov matrix doesn't look degenerate
mod2 <- lmer(
  belief ~ trial_n + condition + 
    (1 + condition | ppt) + 
    (1 + condition | statement), 
  data = df
)

summary(mod2)
```

Don't use trial_n, just use condition.

```{r}
mod3 <- lmer(
  belief ~ condition +
    (1 + condition | ppt) +
    (1 + condition | statement),
  data = df
)
summary(mod3)
```


fully crossed (yay)

RESUME






# OLD idea: Test enhanced learning

```{r}
load(url("https://uoepsy.github.io/data/testenhancedlearning.RData"))
```


```{r}
glimpse(tel)
```

Get rid of "Eskimo" (not a nice word):

```{r}
tel <- tel |>
  filter(Test_word != 'Eskimo') |>
  mutate(Delay = ifelse(Delay == 'min', 'Minute', 'Week'))
```


```{r}
tel |>
  ggplot(aes(x = Group, y = Correct, colour = Delay)) +
  geom_violin() +
  stat_summary(geom = 'pointrange')
```

Use correctness as mean for simulating test score data (rescaled to 0–100):

```{r}
set.seed(1)
tel <- tel |>
  rowwise() |>
  mutate(
    TestScore = rnorm(n=1, mean = Correct, sd = 0.5),
  ) |>
  ungroup() |>
  mutate(
    TestScore = round(datawizard::rescale(TestScore, to = c(0, 100)))
  )
  
tel |>
  ggplot(aes(x = Group, y = TestScore, colour = Delay)) +
  geom_violin(alpha = 0.1) +
  geom_jitter(alpha = 0.1) +
  facet_wrap(~ Delay) +
  stat_summary(geom = 'pointrange', colour = 'black')
```

## Week 2 analysis: no `Group` dimension

```{r}
tel |>
  ggplot(aes(x = Delay, y = TestScore)) +
  geom_boxplot()
```


```{r}
lmer(
  TestScore ~ Delay + (1 + Delay | Subject_ID),
  data = tel
)
```

hmmm ...


### Max mod with no Delay

`Testscore ~ Group + (1 | Subject_ID) + (1 + Group | Test_word)`

```{r}
mdl_group_max <- lmer(
  TestScore ~ Group +
    (1 | Subject_ID) +
    (1 + Group | Test_word),
  data = tel
)
summary(mdl_group_max)
```

Correlation of –1, so the vcov mtx is degenerate.
Will need to simplify – BUT save that for week 4.





## Maximal model analysis (week 3 or 4)

```{r}
# singular fit
m1 <- lmer(
  TestScore ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay * Group | Test_word),
  data = tel
)

# singular fit
m2 <- lmer(
  TestScore ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay + Group | Test_word),
  data = tel
)

# no singular fit!
m3 <- lmer(
  TestScore ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay | Test_word),
  data = tel
)

summary(m3)
```


## Grouping variables


```{r}
tel |>
  group_by(Test_word, Subject_ID) |>
  count()
```


I thiiiink I might still stick with the TEL one, just bc then when we get to the maximal model bit, I can  be like "ah hah there is also an interaction, now what" 




# OLD idea for running example for lec 2

latin square, condition fully crossed w ppts and w items

The story: masked priming, primed with a related word vs. when primed with a distractor

```{r eval=F}
n_ppts  <- 4
n_items <- 20

dat <- simcross(
  seed=1,
  b0=2, 
  b1=1,  # wi_itms_wi_ppts
  b2=1,  # wi_ppts_btw_itms
  b3=2,  # x or xcat
  n_ppts=n_ppts, n_itms=n_items,
  n_obs_per=2
) |>
  select(ppts, itms, xcat, y) |>
  rename(
    ppt_id = ppts,
    item_id = itms,
    priming_cond = xcat,
    logRT = y
  ) |>
  mutate(
    priming_cond = ifelse(priming_cond == 'a', 'related_prime', 'distractor_prime'),
    ppt_id = paste0('ppt', ppt_id),
    item_id = paste0('item', item_id),
  )

dat |>
  ggplot(aes(x = priming_cond, y = logRT)) +
  geom_violin() +
  geom_jitter () +
  facet_wrap(~ ppt_id)
```


Identify grouping structure:

- we are looking only at categorical variables
- we want to see: what categorical variables have values that are associated with more than one data point?

```{r eval=F}
dat |> head()
```

what cat vars do we have? (not including the outcome, if it's binary)
- ppt_id
- item_id
- priming_cond

we'll go through each of these.

assuming data is set up for analysis using lm, with one column per variable and one row per observation, then we should just be able to count how many times each value shows up

```{r eval=F}
dat |>
  group_by(ppt_id) |>
  count()
```

can sense check by visually inspecting df. "ppt1" shows up with a bunch of different logRTs.

```{r eval=F}
dat |>
  group_by(item_id) |>
  count()
```

each item also appears multiple times.

```{r eval=F}
dat |>
  group_by(priming_cond) |>
  count()
```

and each priming condition shows up multiple times.

So we have three groups that each contribute multiple data points:

- `ppt_id`
- `item_id`
- `priming_cond`
 