---
title: "02 playground"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)
library(lme4)

theme_set(theme_bw())

source('simfns.R')  # load simdat functions
```



# Implicit association test

<https://psych.wisc.edu/Brauer/BrauerLab/index.php/implicit-bias/>

condition fully crossed w ppts and w items

```{r}
n_ppts  <- 100
n_items <- 32

set.seed(1)
implicit_data <- simcross(
  seed=1,
  b0=2, 
  b1=1,  # wi_itms_wi_ppts
  b2=1,  # wi_ppts_btw_itms
  b3=2,  # x or xcat
  n_ppts=n_ppts, n_itms=n_items,
  n_obs_per=2
) |>
  select(ppts, itms, xcat, y) |>
  rename(
    ppt_id = ppts,
    item_id = itms,
    pairing = xcat,
    logRT = y
  ) |>
  mutate(
    pairing = factor(ifelse(pairing == 'a', 'Associated', 'Unassociated'), levels = c('Unassociated', 'Associated')),
    ppt_id = paste0('ppt', ppt_id),
    item_id = paste0('item', item_id),
    logRT = datawizard::rescale(logRT, to = c(0.28, 8.62))
  )

write_csv(implicit_data, '../data/iat.csv')
```


```{r}
implicit_data |>
  ggplot(aes(x = pairing, y = logRT)) +
  geom_violin() +
  geom_jitter () +
  NULL
  # facet_wrap(~ ppt_id)
```

```{r}
# stats::xtabs(~ ppt_id + item_id + pairing, data = implicit_data)
```

```{r}
m <- lmer(
  logRT ~ pairing + 
    # (1 | ppt_id),
    # (1 + pairing | ppt_id),
    (1 + pairing | ppt_id) + (1 + pairing | item_id),
  data = implicit_data
)
summary(m)
```

```{r}
ranef(m)$ppt_id |>
  arrange(desc(pairingAssociated))
```


```{r}
dotplot.ranef.mer(ranef(m))$ppt_id
dotplot.ranef.mer(ranef(m))$item_id
```

## plot by-ppt data

```{r}
ppts_to_plot <- c(56, 14, 33, 24, 44, 31, 67, 97, 99)

p_implicit_ppts <- implicit_data |>
  mutate(
    pairing_num = ifelse(pairing == 'Unassociated', 0, 1)
  ) |>
  filter(ppt_id %in% paste0('ppt', ppts_to_plot)) |>
  ggplot(aes(x = pairing_num, y = logRT)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  stat_summary(fun = mean, geom = 'point') +
  stat_summary(aes(group = ppt_id), fun = mean, geom = 'line') +
  facet_wrap(~ ppt_id) +
  scale_x_continuous(breaks = c(0, 1), limits = c(-0.3, 1.3), labels = c('Unassociated', 'Associated')) +
  labs(x = 'Topic pairing') +
  theme(
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  )

p_implicit_ppts
```


## simple LM

```{r}
implicit_lm <- lm(logRT ~ pairing, data = implicit_data)
summary(implicit_lm)
```

```{r}
coef(implicit_lm)[['(Intercept)']]
coef(implicit_lm)[['pairingAssociated']]
```

```{r}
p_implicit_ppts +
  geom_abline(
    intercept = coef(implicit_lm)[['(Intercept)']],
    slope = coef(implicit_lm)[['pairingAssociated']],
    colour = 'red'
  ) 
```

wrong bc 

- errors are no longer independent
- ignores that ppts have diff effects


## diff LMs to each ppt

NOT WHAT YOU SHOULD DO

```{r}
implicit_lmlist <- lmList(
  logRT ~ pairing | ppt_id,
  data = implicit_data
)

implicit_lmlist
```

```{r}
indiv_ppt_coefs <- coef(implicit_lmlist) |>
  rownames_to_column(var = 'ppt_id')

p_implicit_ppts +
  geom_abline(
    data = filter(indiv_ppt_coefs, ppt_id %in% paste0('ppt', ppts_to_plot)),
    aes(intercept = `(Intercept)`, slope = pairingAssociated, colour = ppt_id),
  )  +
  theme(legend.position = 'none')
```


wrong bc

- we're still collapsing/aggregating over items
- goal: look at ALL sources of variability, not just by ppt but also by item

here's what we actually do! the current state of the art in psych and adjacent fields.

## intro intercept adjustments

- in between fitting one single LM and fitting one indiv LM per ppt.
- crucial twist: a new `+ (1 | ppt_id)` thing in the model formula
	- this adjusts the intercept estimate by some amount for each subj
	- if ppt has smaller intercept than avg, then int adjustment neg; if higher, then int adjustment pos

```{r}
library(lme4)

implicit_lmm_int <- lmer(
  logRT ~ pairing + (1 | ppt_id),
  data = implicit_data
)

summary(implicit_lmm_int)
```

- ^ new stuff: residual variance sigma, and one more source of variance: SD of ppt's intercept adjustments. this represents how variable ppts are around fixed intercept

```{r}
ranef(implicit_lmm_int)$ppt_id
```

```{r}
dotplot.ranef.mer(ranef(implicit_lmm_int))$ppt_id
```

- ppt_56 most above average intercept
- ppt_24 most below average intercept
 
 
Actual line parameters:

- every ppt has diff intercept (fixed intercept  + each ppt's intercept adjustment)
- every ppt has same slope
 
```{r}
random_int_coefs <- coef(implicit_lmm_int)$ppt_id |>
  rownames_to_column(var = 'ppt_id')
random_int_coefs
```
 
```{r}
p_implicit_ppts +
  geom_abline(
    data = filter(random_int_coefs, ppt_id %in% paste0('ppt', ppts_to_plot)),
    aes(intercept = `(Intercept)`, slope = pairingAssociated, colour = ppt_id),
  )  +
  theme(legend.position = 'none')
```
  
 
<!-- - show mathematically as term being added on to intercept -->
<!-- 		- compare simple and new models: -->
<!-- 			- simple lm intercept: beta0. -->
<!-- 			- varying-intercepts mixed model intercept: beta0 + u0i for each ppt i -->


 
## intro slp adj

```{r}
implicit_int_slp <- lmer(
  logRT ~ pairing + (1 + pairing | ppt_id),
  data = implicit_data
)

summary(implicit_int_slp)
```

```{r}
random_int_slp_coefs <- coef(implicit_int_slp)$ppt_id |>
  rownames_to_column(var = 'ppt_id')
random_int_slp_coefs
```


```{r}
p_implicit_ppts +
  geom_abline(
    data = filter(random_int_slp_coefs, ppt_id %in% paste0('ppt', ppts_to_plot)),
    aes(intercept = `(Intercept)`, slope = pairingAssociated, colour = ppt_id),
  )  +
  theme(legend.position = 'none')
```


not yet the final form, because there's more than one grouping variable, and we're not adjusting for different effects for each item...

```{r}
items_to_plot <- c(1, 2, 4, 13, 6, 8, 14, 27, 32)

p_implicit_items <- implicit_data |>
  mutate(
    pairing_num = ifelse(pairing == 'Unassociated', 0, 1)
  ) |>
  filter(item_id %in% paste0('item', items_to_plot)) |>
  ggplot(aes(x = pairing_num, y = logRT)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  stat_summary(fun = mean, geom = 'point') +
  stat_summary(aes(group = item_id), fun = mean, geom = 'line') +
  facet_wrap(~ item_id) +
  scale_x_continuous(breaks = c(0, 1), limits = c(-0.3, 1.3), labels = c('Unassociated', 'Associated')) +
  labs(x = 'Topic pairing') +
  theme(
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  )

p_implicit_items
```




```{r}
p_implicit_items +
  geom_abline(
    intercept = fixef(implicit_int_slp)[['(Intercept)']],
    slope = fixef(implicit_int_slp)[['pairingAssociated']],
    colour = 'red'
  ) 
```



## Week 3: intro to >1 ranefs

```{r}
implicit_maximal_model <- lmer(
  logRT ~ pairing + (1 + pairing | ppt_id) + (1 + pairing | item_id),
  data = implicit_data
)

summary(implicit_maximal_model)
```

```{r}
dotplot.ranef.mer(ranef(implicit_maximal_model))$ppt_id
dotplot.ranef.mer(ranef(implicit_maximal_model))$item_id
```


```{r}
maximal_ppt_coefs <- coef(implicit_maximal_model)$ppt_id |>
  rownames_to_column(var = 'ppt_id')
maximal_ppt_coefs

maximal_item_coefs <- coef(implicit_maximal_model)$item_id |>
  rownames_to_column(var = 'item_id')
maximal_item_coefs
```

Plot with adjustments by ppt:

```{r}
p_implicit_ppts +
  geom_abline(
    data = filter(maximal_ppt_coefs, ppt_id %in% paste0('ppt', ppts_to_plot)),
    aes(intercept = `(Intercept)`, slope = pairingAssociated, colour = ppt_id),
  )  +
  theme(legend.position = 'none')
```

^ notice how the lines fit even better now that we've taken the by-item variability into acct!




Plot with adjustments by item:

```{r}
p_implicit_items +
  geom_abline(
    data = filter(maximal_item_coefs, item_id %in% paste0('item', items_to_plot)),
    aes(intercept = `(Intercept)`, slope = pairingAssociated, colour = item_id),
  )  +
  theme(legend.position = 'none')
```





----

# LEGACY

# DOES NOT CONVERGE: erm_belief

MWE

```{r}
belief_data <- read_csv("https://uoepsy.github.io/data/erm_belief.csv") |>
  mutate(
    fluent = ifelse(condition == 'disfluent', 'No', 'Yes'),
    fluent = factor(fluent, levels = c('Yes', 'No')),
  )
 
belief_mod <- lmer(
  # max mod of belief ~ fluent + (1 + fluent | ppt) + (1 + fluent | statement) works no problem
  belief ~ fluent + (1 + fluent || ppt),
  data = belief_data
)
 
summary(belief_mod)
```


```{r}
belief_data <- read_csv("https://uoepsy.github.io/data/erm_belief.csv") |>
  mutate(
    fluent = ifelse(condition == 'disfluent', 'No', 'Yes'),
    fluent = factor(fluent, levels = c('Yes', 'No')),
    fluent_num = ifelse(fluent == 'Yes', 0, 1)
  ) |>
  select(ppt, sentence, statement, fluent, fluent_num, belief, trial_n) |>
  arrange(ppt, sentence)

belief_data
```


```{r}
m <- lmer(
  belief ~ fluent 
    + (1 + fluent | ppt)
    # + (1 + fluent || statement)
  ,
  data = belief_data,
  control = lmerControl(optimizer = 'bobyqa')
)

summary(m)
```


fully crossed (yay)

```{r}
belief_mod <- lmer(
  belief ~ trial_n + (1 + trial_n || ppt),
  data = belief_data |> mutate(trial_n=parse_number(trial_n))
)
summary(belief_mod)
```


## plot data

By ppt, aggregating over statement:

```{r}
p_belief_all_ppts <- belief_data |>
  ggplot(aes(x = fluent_num, y = belief)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  stat_summary(fun = mean, geom = 'point') +
  stat_summary(aes(group = ppt), fun = mean, geom = 'line') +
  facet_wrap(~ ppt) +
  scale_x_continuous(breaks = c(0, 1), limits = c(-0.3, 1.3), labels = c('No', 'Yes')) +
  labs(x = 'fluent?') +
  theme(
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  )

p_belief_all_ppts
```


By statement, aggregating over ppt:

```{r eval=F}
belief_data |>
  ggplot(aes(x = fluent, y = belief)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  stat_summary(fun = mean, geom = 'point') +
  stat_summary(aes(group = statement), fun = mean, geom = 'line') +
  facet_wrap(~ statement) +
  theme(
    strip.background = element_blank()
  )
```


Zoom in on some particular ppts with different patterns:

```{r}
ppts_to_plot <- c(4, 7, 20, 22, 25)

p_belief_3ppts <- belief_data |>
  filter(ppt %in% paste0('ppt_', ppts_to_plot)) |>
  ggplot(aes(x = fluent_num, y = belief)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  stat_summary(fun = mean, geom = 'point') +
  stat_summary(aes(group = ppt), fun = mean, geom = 'line') +
  facet_wrap(~ ppt) +
  scale_x_continuous(breaks = c(0, 1), limits = c(-0.3, 1.3), labels = c('No', 'Yes')) +
  labs(x = 'fluent?') +
  theme(
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  )

p_belief_3ppts
```


## simple LM

```{r}
belief_lm <- lm(belief ~ fluent, data = belief_data)
summary(belief_lm)
```

```{r}
coef(belief_lm)[['(Intercept)']]
coef(belief_lm)[['fluentNo']]
```

```{r}
p_belief_all_ppts +
  geom_abline(
    intercept = coef(belief_lm)[['(Intercept)']],
    slope = coef(belief_lm)[['fluentNo']],
    colour = 'red'
  ) 

p_belief_3ppts +
  geom_abline(
    intercept = coef(belief_lm)[['(Intercept)']],
    slope = coef(belief_lm)[['fluentNo']],
    colour = 'red'
  ) 
```

wrong bc 

- errors are no longer independent
- ignores that ppts have diff effects


## diff LMs to each ppt

NOT WHAT YOU SHOULD DO

```{r}
belief_lmlist <- lmList(
  belief ~ fluent | ppt,
  data = belief_data
)

belief_lmlist
```

```{r}
indiv_ppt_coefs <- coef(belief_lmlist) |>
  rownames_to_column(var = 'ppt')

# names(indiv_ppt_params) <- c('ppt', 'intercept', 'slope')
```

```{r}
p_belief_all_ppts +
  geom_abline(
    data = indiv_ppt_coefs,
    aes(intercept = `(Intercept)`, slope = fluentNo, colour = ppt),
  )  +
  theme(legend.position = 'none')
```

```{r}
p_belief_3ppts +
  geom_abline(
    data = filter(indiv_ppt_coefs, ppt %in% paste0('ppt_', ppts_to_plot)),
    aes(intercept = `(Intercept)`, slope = fluentNo, colour = ppt),
  )  +
  theme(legend.position = 'none')
```


wrong bc

- we're still collapsing/aggregating over statements
- goal: look at ALL sources of variability, not just by ppt but also by statement

here's what we actually do! the current state of the art in psych and adjacent fields.

## intro intercept adjustments

- in between fitting one single LM and fitting one indiv LM per ppt.
- crucial twist: a new `+ (1 | ppt)` thing in the model formula
	- this adjusts the intercept estimate by some amount for each subj
	- if ppt has smaller intercept than avg, then int adjustment neg; if higher, then int adjustment pos

```{r}
library(lme4)

belief_lmm_int <- lmer(
  belief ~ fluent + (1 | ppt),
  data = belief_data
)

summary(belief_lmm_int)
```

- ^ new stuff: residual variance sigma, and one more source of variance: SD of ppt's intercept adjustments. this represents how variable ppts are around fixed intercept

```{r}
ranef(belief_lmm_int)$ppt
```

```{r}
dotplot.ranef.mer(ranef(belief_lmm_int))$ppt
```


- ppt_22 most above average intercept
- ppt_4 most below average intercept
 
 
Actual line parameters:

- every ppt has diff intercept (fixed intercept of 49.30 points + each ppt's intercept adjustment)
- every ppt has same slope
 
```{r}
random_int_coefs <- coef(belief_lmm_int)$ppt |>
  rownames_to_column(var = 'ppt')
random_int_coefs
```
 
 
```{r}
p_belief_all_ppts +
  geom_abline(
    data = random_int_coefs,
    aes(intercept = `(Intercept)`, slope = fluentNo, colour = ppt),
  )  +
  theme(legend.position = 'none')
```

  
```{r}
p_belief_3ppts +
  geom_abline(
    data = filter(random_int_coefs, ppt %in% paste0('ppt_', ppts_to_plot)),
    aes(intercept = `(Intercept)`, slope = fluentNo, colour = ppt),
  )  +
  theme(legend.position = 'none')
```
  
 
<!-- - show mathematically as term being added on to intercept -->
<!-- 		- compare simple and new models: -->
<!-- 			- simple lm intercept: beta0. -->
<!-- 			- varying-intercepts mixed model intercept: beta0 + u0i for each ppt i -->



## intro slp adjustments

need to allow the slopes to vary as well.

- add another source of variance: slopes varying by subj
- "varying intercepts + varying slopes by subject model"
<!-- - use (1 + x || subj), no correl, for now. "just trust me for now, I will explain what that is about soon" -->


```{r}
belief_lmm_int_slp <- lmer(
  belief ~ fluent + (1 + fluent | ppt),
  data = belief_data,
)

summary(belief_lmm_int_slp)
```

```{r}
random_int_slp_coefs <- coef(belief_lmm_int_slp)$ppt |>
  rownames_to_column(var = 'ppt')
random_int_slp_coefs
```

```{r}
dotplot.ranef.mer(ranef(belief_lmm_int_slp))$ppt
```

 
```{r}
p_belief_all_ppts +
  geom_abline(
    data = random_int_slp_coefs,
    aes(intercept = `(Intercept)`, slope = fluentNo, colour = ppt),
  )  +
  theme(legend.position = 'none')
```

  
```{r}
p_belief_3ppts +
  geom_abline(
    data = filter(random_int_slp_coefs, ppt %in% paste0('ppt_', ppts_to_plot)),
    aes(intercept = `(Intercept)`, slope = fluentNo, colour = ppt),
  )  +
  theme(legend.position = 'none')
```
 
cry!


<!-- --- -->

<!-- JK comments from 13.05.26 on why uncorrelated ranefs look the way they do for categ preds. -->

<!-- if you uncorrelate intercepts and slopes, then it essentially creates a second grouping var (which is just the same groups as the first grouping var). so it's like saying:  -->
<!--  G1==G2 -->
<!-- intercept varies by grouping G1 -->
<!-- slope of x varies by grouping G2 -->
<!-- so if you had:  -->
<!-- belief_mod <- lmer( -->

<!--   belief ~ trial_n + ( + trial_n || ppt), -->

<!--   data = belief_data |> mutate(trial_n=parse_number(trial_n)) -->

<!-- ) -->

<!-- summary(belief_mod) -->
<!-- then it just uncorrelates the two things by creating a second grouping to separate them out.  -->

<!-- but....  -->
<!-- with a categorical predictor, this is more confusing. it keep the correlations within the levels of the predictor.  -->

<!-- so what your model is showing is:  -->
<!-- ppt variability at intercept -->
<!-- ppt variability at fluentYes -->
<!-- ppt variability at fluentNo -->
<!-- cor between 2 and 3 -->
<!-- the intercept is the same as the variability at fluentYes, so you've kind of got that in there twice.  -->

<!-- what i think you want is:  -->
<!-- belief_mod <- lmer( -->

<!--   belief ~ fluent + (0 + fluent | ppt), -->

<!--   data = belief_data -->

<!-- ) -->

<!-- VarCorr(belief_mod) -->
<!-- # ppt SD in fluentYes condition - 0.78 -->

<!-- # ppt variance in fluentNo condition - 0.44 -->
<!-- as opposed to:  -->
<!-- belief_mod <- lmer( -->

<!--   belief ~ fluent + (1 + fluent | ppt), -->

<!--   data = belief_data -->

<!-- ) -->

<!-- VarCorr(belief_mod) -->

<!-- # ppt SD in fluentYes condition - 0.78 -->

<!-- # ppt variance in fluentNo-fluentYes difference - 0.34 -->



<!-- but in that you'll still have the within-predictor correlations!! -->

<!-- so... -->

<!-- (suspensful..) -->

<!-- categorical effects on the RHS!  -->
<!-- belief_mod <- lmer( -->

<!--   belief ~ fluent + (1 | ppt) + (1 | ppt:fluent), -->

<!--   data = belief_data -->

<!-- ) -->


<!-- only it doesn't converge in this case  -->
 

----


# OLD idea: Test enhanced learning


```{r}
load(url("https://uoepsy.github.io/data/testenhancedlearning.RData"))
```


```{r}
glimpse(tel)
```

Get rid of "Eskimo" (not a nice word):

```{r}
tel <- tel |>
  filter(Test_word != 'Eskimo') |>
  mutate(Delay = ifelse(Delay == 'min', 'Minute', 'Week'))
```


```{r}
tel |>
  ggplot(aes(x = Group, y = Correct, colour = Delay)) +
  geom_violin() +
  stat_summary(geom = 'pointrange')
```

Use correctness as mean for simulating test score data (rescaled to 0–100):

```{r}
set.seed(1)
tel <- tel |>
  rowwise() |>
  mutate(
    TestScore = rnorm(n=1, mean = Correct, sd = 0.5),
  ) |>
  ungroup() |>
  mutate(
    TestScore = round(datawizard::rescale(TestScore, to = c(0, 100)))
  )
  
tel |>
  ggplot(aes(x = Group, y = TestScore, colour = Delay)) +
  geom_violin(alpha = 0.1) +
  geom_jitter(alpha = 0.1) +
  facet_wrap(~ Delay) +
  stat_summary(geom = 'pointrange', colour = 'black')
```

## Week 2 analysis: no `Group` dimension

```{r}
tel |>
  ggplot(aes(x = Delay, y = TestScore)) +
  geom_boxplot()
```


```{r}
lmer(
  TestScore ~ Delay + (1 + Delay | Subject_ID),
  data = tel
)
```

hmmm ...


### Max mod with no Delay

`Testscore ~ Group + (1 | Subject_ID) + (1 + Group | Test_word)`

```{r}
mdl_group_max <- lmer(
  TestScore ~ Group +
    (1 | Subject_ID) +
    (1 + Group | Test_word),
  data = tel
)
summary(mdl_group_max)
```

Correlation of –1, so the vcov mtx is degenerate.
Will need to simplify – BUT save that for week 4.





## Maximal model analysis (week 3 or 4)

```{r}
# singular fit
m1 <- lmer(
  TestScore ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay * Group | Test_word),
  data = tel
)

# singular fit
m2 <- lmer(
  TestScore ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay + Group | Test_word),
  data = tel
)

# no singular fit!
m3 <- lmer(
  TestScore ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay | Test_word),
  data = tel
)

summary(m3)
```


## Grouping variables


```{r}
tel |>
  group_by(Test_word, Subject_ID) |>
  count()
```


I thiiiink I might still stick with the TEL one, just bc then when we get to the maximal model bit, I can  be like "ah hah there is also an interaction, now what" 


