---
title: "03 playground"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(lme4)
library(lmerTest)
```

# plotting estims

## setup

```{r}
implicit_data <- read_csv('../data/iat.csv') |>
  mutate(pairing = factor(pairing, levels = c('Unassociated', 'Associated')))
```

```{r}
implicit_full_lmm <- lmer(
  logRT ~ pairing + 
    (1 + pairing | ppt_id) + 
    (1 + pairing | item_id),
  data = implicit_data
)
```

```{r}
summary(implicit_full_lmm)
```


## plot fixef

```{r}
library(effects)

effect(term = "pairing", mod = implicit_full_lmm) |>
  as.data.frame() |>
  ggplot(aes(x = pairing, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) +
  labs(
    y = 'Log RT',
    x = 'Topic pairing'
  )
```


## plot ranef lines

```{r}
library(broom.mixed)

augment(implicit_full_lmm)
```


```{r}
augment(implicit_full_lmm) |>
  ggplot(aes(
    x = pairing, 
    y = .fitted, 
    group = interaction(ppt_id, item_id))
  ) +
  geom_line(alpha = 0.02)
```


## overlay ranefs and fixefs

```{r}
# Store the fitted estimates for the fixed effects
eff_df <- effect(term = "pairing", mod = implicit_full_lmm) |>
  as.data.frame()

# Create the ggplot based on `augment()`
augment(implicit_full_lmm) |>
  
  # Both augment() and eff_df have the same variable called `pairing` for the x axis,
  # so we can define that x axis for the whole ggplot:
  ggplot(aes(x = pairing)) +
  
  # Create the group-level lines with the variable names specific to augment()
  geom_line(
    aes(y = .fitted, group = interaction(ppt_id, item_id)),
    alpha = 0.02
  ) +
  
  # Drawing on data from eff_df, which has different column names,
  # create the fixed effect points, line, and 95% CI bars
  geom_point(data = eff_df, aes(y = fit)) +
  geom_errorbar(data = eff_df, aes(ymin = lower, ymax = upper), width = 0) +
  geom_line(data = eff_df, aes(y = fit, group = 1)) +
  
  # Cosmetics: adjust axis labels
  labs(
    x = 'Topic pairing',
    y = 'Log RT'
  ) 
```



# identifying possible ranefs

## erm_belief

```{r}
belief_data <- read_csv("https://uoepsy.github.io/data/erm_belief.csv") |>
  mutate(
    hesitation = factor(ifelse(condition == 'disfluent', 'Yes', 'No')),
  ) |>
  select(ppt, sentence, statement, hesitation, belief) |>
  arrange(ppt, sentence)
```

- "Water makes different sounds depending on its temperature"
- "It's impossible to hum while holding your nose" 
- "A hashtag is technically called an octothorp"  
- "The largest snowflake was bigger than most pizzas"
- "Pigs don't sweat"


```{r}
belief_data |>
  ggplot(aes(x = hesitation, y = belief, colour = hesitation, fill = hesitation)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5, size = 3) +
  stat_summary(geom = 'point', fun = mean, colour = 'black', size = 5) +
  theme(
    legend.position = 'none'
  )
```

### Step 1: Predictors / fixed effects formula

RQ: Do hesitation markers like "um"/"erm" make true statements sound less believable?

- Outcome: `belief`
- Predictor: `hesitation`

Model formula: `belief ~ hesitation`


### Step 2: Grouping variables

```{r}
names(belief_data)
```

- no need to check `belief` because (a) it's continuous (b) it's the outcome

```{r}
belief_data |> group_by(ppt) |> count()  # yes
belief_data |> group_by(sentence) |> count()  # yes
belief_data |> group_by(statement) |> count()  # yes (same as `sentence`)
belief_data |> group_by(hesitation) |> count()  # yes
```

### Step 3: DGP + Random intercepts

- `ppt`
  - RQ doesn't specify that we manipulate or control for specific people in our analysis
  - if we re-ran the study, we could look at different people
  - therefore random variability
- `sentence`
  - RQ doesn't specify that we manipulate or control for specific sentences/statements
  - if we re-ran the study, we could use other true statements
  - therefore random variability
- `statement`
  - equivalent to `sentence`, will only use one
- `hesitation`
  - RQ specifies that we want to manipulate the hesitation markers in the sentence presentation
  - if we re-ran the study, we would use exactly these two conditions again
  - therefore manipulated and reproducible variability

two grouping variables contribute random variability: `ppt` and `sentence`.
therefore our model needs a random intercept for each one.

so at minimum, our model formula is now: `belief ~ hesitation + (1 | ppt) + (1 | sentence)`


### Step 4: Random slopes

For each random intercept: do at least some levels of that grouping variable appear with >1 value of `hesitation`?

(draw the little table?)


**ppt:**

```{r}
stats::xtabs(~ ppt + hesitation, data = belief_data)
```

yes! so we can have random slopes over `hesitation` by `ppt`.

we upgrade `(1 | ppt)` to `(1 + hesitation | ppt)`.


**sentence**

```{r}
stats::xtabs(~ sentence + hesitation, data = belief_data)
```

yes! 
we upgrade `(1 | sentence)` to `(1 + hesitation | sentence)`.


### model with all possible random effects

`belief ~ hesitation + (1 + hesitation | ppt) + (1 + hesitation | statement)`


```{r}
m <- lmer(
  belief ~ hesitation 
    + (1 + hesitation | ppt)
    + (1 + hesitation | statement)
  ,
  data = belief_data
)

# summary(m)
```


## TEL

```{r}
load(url("https://uoepsy.github.io/data/testenhancedlearning.RData"))
tel_data <- tel |>
  filter(Test_word != 'Eskimo')
```

```{r}
tel_data |>
  ggplot(aes(x = Delay, y = Correct, colour = Group)) +
  stat_summary(geom = 'pointrange', fun.data = mean_se) +
  labs(
    y = 'Proportion correct'
  )
```


### Step 1: Predictors / fixed effects formula

two groups of 25 participants were presented with material to learn. One group studied the material twice (`StudyStudy`), the other group studied the material once then did a test (`StudyTest`). Recall was tested immediately (one minute) after the learning session and one week later. The recall tests were composed of 175 items identified by a keyword (`Test_word`).

RQ: Will the `StudyStudy` group perform better on the immediate test (one minute delay), but the `StudyTest` group perform better on the follow-up test (one-week delay)?

```{r}
names(tel_data)
```

Outcome: `Correct`
Predictors: `Delay`, `Group`, and their interaction

Formula:

`Correct ~ Delay * Group`


### Step 2: Grouping variables

- don't need to check `Correct` bc that's the outcome
- don't need to check `Rtime` bc it's continuous

```{r}
tel_data |> group_by(Subject_ID) |> count()  # yes
tel_data |> group_by(Group) |> count()  # yes
tel_data |> group_by(Delay) |> count()  # yes
tel_data |> group_by(Test_word) |> count()  # yes
```



### Step 3: DGP + random intercepts

- `Subject_ID`: random
- `Group`: manipulated
- `Delay`: manipulated
- `Test_word`: random

At minimum:

`Correct ~ Delay * Group + (1 | Subject_ID) + (1 | Test_word)`


### Step 4: Random slopes

For each random intercept: do at least some levels of that grouping variable appear with >1 value of each predictor?

(draw the little table?)

**Subject_ID x Delay:**

```{r}
stats::xtabs(~ Subject_ID + Delay, data = tel_data)
```

yes

**Subject_ID x Group:**

```{r}
stats::xtabs(~ Subject_ID + Group, data = tel_data)
```

No!

So, for `Subject_ID`:

`(1 + Delay | Subject_ID)`

**Test_word x Delay:**

```{r}
stats::xtabs(~ Test_word + Delay, data = tel_data)
```

yes

**Test_word x Group:**

```{r}
stats::xtabs(~ Test_word + Group, data = tel_data)
```

yes

bc both, we can also include the ixn between them too!

`(1 + Delay * Group | Test_word)`

### model with all poss ranefs

`Correct ~ Delay * Group + (1 + Delay | Subject_ID) + (1 + Delay * Group | Test_word)`


