---
title: "04 playground"
output: html_document
---

```{r setup}
library(tidyverse)
library(lme4)
library(lmerTest)
library(HLMdiag)
```


```{r}
load(url("https://uoepsy.github.io/data/testenhancedlearning.RData"))

ppts_to_keep <- c(
  paste0('StudyStudy_', LETTERS[1:10]),
  paste0('StudyTest_', LETTERS[1:10])
)

set.seed(1)
tel_data <- tel |>
  filter(
    Test_word != 'Eskimo',
    Subject_ID %in% ppts_to_keep
  ) |>
  rowwise() |>
  mutate(
    TestScore = rnorm(n=1, mean = Correct, sd = 0.5),
  ) |>
  ungroup() |>
  mutate(
    TestScore = case_when(
      Group == 'StudyStudy' & Delay == 'min'  ~ TestScore + 0.2,
      Group == 'StudyStudy' & Delay == 'week' ~ TestScore - 0.5,
      Group == 'StudyTest'  & Delay == 'min'  ~ TestScore + 0.2,
      Group == 'StudyTest'  & Delay == 'week' ~ TestScore,
    ),
    TestScore = round(datawizard::rescale(TestScore, to = c(0, 100)))
  )
 
rm(tel)

tel_data |>
  ggplot(aes(x = Delay, y = TestScore, colour = Group)) +
  geom_violin(alpha = 0.2) +
  geom_jitter(alpha = 0.2) +
  facet_wrap(~ Group) +
  stat_summary(geom = 'pointrange', colour = 'black')
```


# TEL TestScore lm

## simplify models

if so, will have to update lec 3 about this as well

```{r eval=F}
# singular fit
m1 <- lmer(
  TestScore ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay * Group | Test_word),
  data = tel_data
)

# singular fit
m2 <- lmer(
  TestScore ~ Delay * Group +
    (1 + Delay | Subject_ID) +
    (1 + Delay + Group | Test_word),
  data = tel_data
)

m3 <- lmer(
  TestScore ~ Delay * Group +
    (1 + Delay | Subject_ID) +
    (1 + Delay | Test_word),
  data = tel_data
)

# singular fit
m4 <- lmer(
  TestScore ~ Delay * Group +
    (1 | Subject_ID) +
    (1 + Delay | Test_word),
  data = tel_data
)
```


```{r}
# no singual fit
m5 <- lmer(
  TestScore ~ Delay * Group +
    (1 + Delay | Subject_ID) +
    (1 | Test_word),
  data = tel_data
)

summary(m5)
```






## check assump

### linearity

```{r}
plot(
  m5, 
  type=c(
    "p",      # includes points representing Pearson residuals
    "smooth"  # includes smoothed line representing mean of residuals
  )
)
```

two weird clouds, kind of bimodal distribution of test scores evidently, but fine

horizontal enough!


### normdist errors

```{r}
qqnorm(resid(m5))
qqline(resid(m5))
```

pretty darn good


### equal varc errors

```{r}
plot(
  m5, 
  type=c(
    "p",      # includes points representing Pearson residuals
    "smooth"  # includes smoothed line representing mean of residuals
  )
)
```

vertical spread of errors is OK
no fan or funnel shapes, which is what we generally would be worried about

fine.


### normdist ranefs

#### by-ppt int adjs

```{r}
qqnorm(ranef(m5)$Subject_ID[,1]) 
qqline(ranef(m5)$Subject_ID[,1])
```

not a great match ... positive adjustments are closer to zero than would be ideal, and negative adjustments are farther from zero than would be ideal

so in this case, would be something to report and just say that we gotta take the SD with a grain of salt


#### by-ppt slp adjs

```{r}
qqnorm(ranef(m5)$Subject_ID[,2]) 
qqline(ranef(m5)$Subject_ID[,2])
```

better ish fit



#### by-word int adjs


```{r}
qqnorm(ranef(m5)$Test_word[,1]) 
qqline(ranef(m5)$Test_word[,1])
```

again, some more negative residuals at the lower end and some also smaller residuals at the pos end
but w/e it's close enough, not too severe



## checking influential obs and groups

```{r eval=F}
inf_obs  <- hlm_influence(model = m5, level = 1, approx = TRUE)
inf_subj <- hlm_influence(model = m5, level = "Subject_ID", approx = TRUE)
inf_word <- hlm_influence(model = m5, level = "Test_word", approx = TRUE)

#save to rda file
save(inf_obs,  file = "../data/m5_inf_obs.rda")
save(inf_subj, file = "../data/m5_inf_subj.rda")
save(inf_word, file = "../data/m5_inf_word.rda")
```

```{r}
load("../data/m5_inf_obs.rda")
load("../data/m5_inf_subj.rda")
load("../data/m5_inf_word.rda")
```

```{r}
dotplot_diag(inf_obs$cooksd, cutoff = "internal")
dotplot_diag(inf_subj$cooksd, index = inf_subj$Subject_ID, cutoff = "internal")
dotplot_diag(inf_word$cooksd, index = inf_word$Test_word, cutoff = "internal")
```

```{r eval=F}
inf_word$Test_word[137]
inf_word$Test_word[61]
```


### sensitivity analysis

<https://uoepsy.github.io/dapr3/2627/flashcards/block1/influence.html> 

Check on the influence of each of these words individually

#### noseless

```{r}
m5_noseless <- lmer(
  TestScore ~ Delay * Group +
    (1 + Delay | Subject_ID) +
    (1 | Test_word),
  data = tel_data |> filter(Test_word != "nose")
)
```


```{r}
summary(m5)
```

```{r}
summary(m5_noseless)
```

#### saddleless

```{r}
m5_saddleless <- lmer(
  TestScore ~ Delay * Group +
    (1 + Delay | Subject_ID) +
    (1 | Test_word),
  data = tel_data |> filter(Test_word != "saddle")
)
```

```{r}
summary(m5_saddleless)
```




---

# TEL GLM – nah

nah bc

- GLM assumptions are different
- the GLM actually fits out of the gate, don't need to simplify ranefs


```{r eval=F}
# takes a while
m1 <- lmer(
  Correct ~ Delay * Group + 
    (1 + Delay | Subject_ID) +
    (1 + Delay * Group | Test_word),
  data = tel_data,
)

summary(m1)
```

