---
title: "05 playground"
output: html_document
---

```{r setup}
library(tidyverse)
library(lme4)
library(lmerTest)
library(HLMdiag)
library(effects)
```


# Routine and emotional dysregulation in children

## playin around

```{r}
df <- read_csv("https://uoepsy.github.io/data/crqeds.csv")
```

RQ: Are children with more day-to-day routine better at regulating their emotions, while controlling for the type of school children attend (as a proxy for SES)?

```{r}
df |>
  ggplot(aes(x = CRQ, y = EDS, colour = schooltype)) +
  geom_point()
```


```{r}
mod1 <- lmer(EDS ~ schooltype + CRQ + 
       (1 + CRQ | schoolid), 
       data = df)

summary(mod1)
```


## phase 1 

### Load required R packages and read in data

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(stats)
library(HLMdiag)

eds_data <- read_csv("https://uoepsy.github.io/data/crqeds.csv")
glimpse(eds_data)
```


### Identify fixefs based on RQ

RQ: Are children with more day-to-day routine better at regulating their emotions, while controlling for the type of school children attend (as a proxy for SES)?

Outcome: `EDS` (emotional dysregulation score)
Predictors: `CRQ` (childhood routines questionnaire) + `schooltype` (no interaction bc we're just controlling for this, not looking for how assocs may change)

Fixef model formula: `EDS ~ CRQ + schooltype`



### Plot fixef variables all together

```{r}
eds_data |>
  ggplot(aes(x = CRQ, y = EDS, colour = schooltype)) +
  geom_point()
```


### Set up categorical (fixed) predictors

```{r}
eds_data <- eds_data |>
  mutate(
    schooltype = factor(schooltype)
  )

contrasts(eds_data$schooltype)
```

Ref level of `schooltype`: `Private`


### Set up continuous predictors

```{r}
eds_data$CRQ
```

where does zero fall relative to these values?

zero IS a possible value for the CRQ (it ranges 0 to 7, and 0 means no routine), but we don't have it in our dataset.

so there's a reasonable motivation to mean-centre CRQ (make the mean = 0), but also reasonable motivation to keep CRQ as-is (model makes estimates at 0 = people with no routine)


### Identify grouping variables that contribute random variability (these will be random intercepts)

```{r}
glimpse(eds_data)
```

Only remaining var: `schoolid`

```{r}
eds_data |>
  group_by(schoolid) |>
  count()
```

Yes, mult children from ea school.
So `schoolid` is a grouping variable.

Does it contribute controlled / reproducible variability or random variability?

If we did this study again, we could sample diff schools.
And we want our results to generalise across schools.
So it contributes random variability.

That means our model will have random intercepts by `schoolid`.

Model formula so far: `EDS ~ CRQ + schooltype + (1 | schoolid)`


### Identify what random slopes are possible in maximal model

Do at least some levels of each grouping variable appear with more than one value of each predictor?

- Do at least some levels of `schoolid` appear with more than one value of `CRQ`?
- Do at least some levels of `schoolid` appear with more than one value of `schooltype`?



First: Do at least some levels of `schoolid` appear with more than one value of `CRQ`?
Yes.

```{r}
xtabs(~ schoolid + CRQ, data = eds_data)
```


```{r}
# a trick from the flash cards to count all the non-zero entries in each row:
(xtabs(~ schoolid + CRQ, data = eds_data)  != 0) |>
  rowSums()
```


So, we can have random slopes over `CRQ` for each `schoolid`.


Second: Do at least some levels of `schoolid` appear with more than one value of `schooltype`?

```{r}
xtabs(~ schoolid + schooltype, data = eds_data)
```

No! A school is either Private or State.

So our maximal model formula is:

`EDS ~ CRQ + schooltype + (1 | schoolid)`



## phase 2

### fit maximal model

```{r}
eds_mod <- lmer(
  EDS ~ schooltype + CRQ + (1 + CRQ | schoolid), 
  data = eds_data
)
# no warnings, but let's check variance components to make sure we're ok
VarCorr(eds_mod)
```
Good! model converges + needs no massaging.


## phase 3

### Check assumptions

```{r}
plot(eds_mod, 
     type=c(
       "p",      # includes points representing Pearson residuals
       "smooth"  # includes smoothed line representing mean of residuals
       )
    )
```

Could be some deviation from linearity, but oh well.
satisfied it's linear enough that our model will be okay.

Equal variance looks OK.



```{r}
qqnorm(resid(eds_mod)); qqline(resid(eds_mod))
```

oo yes, it's weird at the extremes.
that's because our outcome variable of EDS is bounded between 1 and 6, which is not something that a linear model can technically handle.

In worlds beyond DAPR, there are ways to model this kind of bounded data properly.
But a lot of scientists will still analyse it like this, even though the assumptions aren't totally supported.
So if we accept this assumption, which we will have to in order to use the linear modelling machinery, then we just live with this.



Normality of ranefs:

```{r}
ranef(eds_mod)$schoolid[,1]  # int
ranef(eds_mod)$schoolid[,2]  # slp
```


```{r}
qqnorm(ranef(eds_mod)$schoolid[,1]); qqline(ranef(eds_mod)$schoolid[,1])
```

Some divergence—it's not as nice as it could be.
But not as bad as it could be either.




```{r}
qqnorm(ranef(eds_mod)$schoolid[,2]); qqline(ranef(eds_mod)$schoolid[,2])
```

Better, except for one of the slope adjustments being more negative than we'd expect!
Oh well, not bad overall.


### Check influential observations/schools

```{r}
inf_obs  <- hlm_influence(model = eds_mod, level = 1, approx = TRUE)
inf_schoolid <- hlm_influence(model = eds_mod, level = "schoolid", approx = TRUE)
```

```{r}
dotplot_diag(inf_obs$cooksd, cutoff = "internal")
```

```{r}
dotplot_diag(inf_schoolid$cooksd, index = inf_schoolid$schoolid, cutoff = "internal")
```


worth doing sensitivity analysis for 108 and 168, potentially, but can set that aside for reasons of time.


#### Sensitivity analysis 168 feat bobyqa

```{r}
# failed to converge
eds_mod_no168 <- lmer(
  EDS ~ schooltype + CRQ + (1 + CRQ | schoolid), 
  data = eds_data[-168,]
)

# add in bobyqa
eds_mod_no168 <- lmer(
  EDS ~ schooltype + CRQ + (1 + CRQ | schoolid), 
  data = eds_data[-168,],
  control = lmerControl(optimizer = "bobyqa")  # add in this line
)

# to compare to main model, re-fit that model with bobyqa too
# (we want only one thing to be diff between the models we compare)
eds_mod_bobyqa <- lmer(
  EDS ~ schooltype + CRQ + (1 + CRQ | schoolid), 
  data = eds_data,
  control = lmerControl(optimizer = "bobyqa")  # add in this line
)
```

```{r}
summary(eds_mod_bobyqa)
summary(eds_mod_no168)
```


no big diff.


#### Sensitivity analysis 108

```{r}
eds_mod_no108 <- lmer(
  EDS ~ schooltype + CRQ + (1 + CRQ | schoolid), 
  data = eds_data[-108,]
)

summary(eds_mod_no108)
```

no difference


### plot model estimates

```{r}
effect(term = c("CRQ", 'schooltype'), mod = eds_mod) |>
  as.data.frame()
```

CRQ is spaced kind of weird, so let's give it a diff vector of CRQ values

```{r message=F}
effect(term = c("CRQ", 'schooltype'), mod = eds_mod, xlevels = list(CRQ = 0:7)) |>
  as.data.frame() |>
  ggplot(aes(x = CRQ, y = fit)) +
  geom_line(aes(colour = schooltype)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = schooltype), alpha = 0.3) +
  labs(
    y = 'Emotion Dysregulation Score (EDS)',
    x = 'Childhood Routines Questionnaire (CRQ) score'
  )
```

## interpret fixefs

```{r}
# summary(eds_mod)
fixef(eds_mod) |> round(2)
```


`(Intercept)`:
The average estimated emotional dysregulation for pupils in private schools with CRQ = 0 (that is, zero routine) is 4.53 points.

```{r message = F, eval = F}
effect(term = c("CRQ", 'schooltype'), mod = eds_mod, xlevels = list(CRQ = 0:7)) |>
  as.data.frame() |>
  ggplot(aes(x = CRQ, y = fit)) +
  geom_line(aes(colour = schooltype)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = schooltype), alpha = 0.3) +
  labs(
    y = 'Emotion Dysregulation Score (EDS)',
    x = 'Childhood Routines Questionnaire (CRQ) score'
  ) +
  geom_point(x = 0, y = 4.53, size = 5)  # plot the intercept
```


`schooltypeState`:
Holding CRQ constant (ie, for any value of CRQ), moving from private schools to state schools is associated with a decrease of EDS of 0.17.
That is, dysregulation decreases, or regulation increases.
This difference is not significantly different from zero.

`CRQ`:
Holding school type constant (ie, for any value of school type), an increase of one CRQ point is associated with a decrease of EDS of 0.05.
That is, dysregulation decreases, or regulation increases.
This difference is also not significantly different from zero.


## interpret ranefs relative to fixefs

```{r}
VarCorr(eds_mod)
```

The by-school intercept adjustments have a SD of 0.46 points.
Thus the model suggests that 95% of the adjusted school-level intercepts fall between 4.53 – 2 * 0.46 = 3.61 and 4.53 + 2 * 0.46 = 5.45 EDS points.

The by-school adjustments to the slope over `CRQ` have a SD of 0.14 points.
Thus the model suggests that 95% of the adjusted school-level slopes over `CRQ` fall between –0.17 – 2 * 0.14 = –0.45 and –0.17 + 2 * 0.14 = 0.11.
Thus most schools show a negative association between CRQ and EDS (meaning that as routine increases, dysregulation decreases—in line with the hypothesis), but some schools are also estimated to show a positive association (not in line with the hypothesis).

The intercept and slope adjustments have a fairly large negative correlation of –0.75.
This suggests that schools that tend to have students with higher emotional dysregulation scores (big intercept) tend to have a more negative adjustment to the slope of `CRQ`.
Since this fixed slope is already negative, the more negative adjustments suggest that the effect of `CRQ` tends to be stronger in schools with pupils having higher emotional dysregulation.

```{r}
dotplot.ranef.mer(ranef(eds_mod))$schoolid
```



## Advanced: plot the random effect lines

to sense check our interpretation of the negative correlation

```{r}
library(broom.mixed)

augment(eds_mod) |>
  ggplot(aes(x = CRQ, y = .fitted, group = schoolid, colour = schooltype)) +
  geom_line()
```




# OLD


# stress and CBD drinks

## playin around


```{r}
unique(df$pid)
```


```{r}
df <- read_csv("https://uoepsy.github.io/data/stressweek1.csv")
# re-center 'day' so the intercept is day 1
# df$day <- df$day-1

glimpse(df)
```


```{r}
staff_gender <- tibble(
  pid = unique(df$pid), 
  gender = c( 'F', 'M', 'M', 'M', 'F', 'M', 'M', 'M', 'M', 'F', 'F', 'F', 'M', 'M', 'F', 'M', 'M', 'F', 'F', 'M' ) 
)

df <- df |>
  merge(staff_gender, by = 'pid')
```




RQ: Does consuming CBD drinks affect Psych staff members' stress levels across the working week, when controlling for staff members' gender?

```{r}
mod1 <- lmer(stress ~ 1 + day * CBD + gender+ 
               (1 + day | pid), 
             data = df)

summary(mod1)
```




# Needless Game Violence

## playin around

```{r}
ngv <- read_csv('https://uoepsy.github.io/data/NGV.csv')

ngv |>
  ggplot(aes(x = mode, y = NGV, fill = mode, colour = mode)) +
  geom_violin(alpha = 0.5) + 
  geom_jitter(alpha = 0.5, size = 3) +
  stat_summary(fun = mean, geom = 'point', colour = 'black', size = 5) +
  facet_wrap(~ character) +
  theme(legend.position = 'none')
```

```{r}
m <- lmer(NGV ~ character * mode + (1 + character | PID) +  (1 + mode | level), data = ngv)
summary(m)
```


RQ: **How is the realism of games associated with unnecessarily violent gameplay? How is this association impacted by the mode of play (on a screen vs. a virtual reality (VR) headset)?**


## phase 1 

### Load required R packages and read in data

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(HLMdiag)

ngv <- read_csv('https://uoepsy.github.io/data/NGV.csv')
glimpse(ngv)
```


### Identify fixefs based on RQ

RQ: **How is the realism of games associated with unnecessarily violent gameplay? How is this association impacted by the mode of play (on a screen vs. a virtual reality (VR) headset)?**

It is possible to answer these qs with a single model, so that's what we should do.

Outcome: `NGV` score
Predictors: `character`, `mode`, and their interaction

Model formula: `NGV ~ character * mode`


### Plot fixef variables all together

```{r}
ngv |>
  ggplot(aes(x = mode, y = NGV, fill = mode, colour = mode)) +
  geom_violin(alpha = 0.5) + 
  geom_jitter(alpha = 0.5, size = 3) +
  stat_summary(fun = mean, geom = 'point', colour = 'black', size = 5) +
  facet_wrap(~ character) +
  theme(legend.position = 'none')
```

### Set up categorical (fixed) predictors

```{r}
ngv <- ngv |>
  mutate(
    mode = factor(mode),
    character = factor(character)
  )

contrasts(ngv$mode)
contrasts(ngv$character)
```

Ref level of `mode`: `screen`
Ref level of `character`: `realistic`:


### Identify grouping variables that contribute random variability (these will be random intercepts)

```{r}
glimpse(ngv)
```

Remaining categorical or ordinal variables:
- `PID`
- `level`
- `P` psychopathy
- `N` narcissism
- `M` machiavellianism


Think about data generating process: 

- the dark triad traits are plausible covariates – controlled variability
- `PID` and `level` are random variability ... hmm but `level` could also be controlled, do we want htings to generalise across all levels? i guess so. i guess that's why.


### Identify what random slopes are possible in maximal model












## phase 2


## phase 3


