---
title: "01 playground"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)

source('simfns.R')  # load simdat functions
```


# main running example for lec 2

latin square, condition fully crossed w ppts and w items

The story: masked priming, primed with a related word vs. when primed with a distractor

```{r}
n_ppts  <- 4
n_items <- 20

dat <- simcross(
  seed=1,
  b0=2, 
  b1=1,  # wi_itms_wi_ppts
  b2=1,  # wi_ppts_btw_itms
  b3=2,  # x or xcat
  n_ppts=n_ppts, n_itms=n_items,
  n_obs_per=2
) |>
  select(ppts, itms, xcat, y) |>
  rename(
    ppt_id = ppts,
    item_id = itms,
    priming_cond = xcat,
    logRT = y
  ) |>
  mutate(
    priming_cond = ifelse(priming_cond == 'a', 'related_prime', 'distractor_prime'),
    ppt_id = paste0('ppt', ppt_id),
    item_id = paste0('item', item_id),
  )

dat |>
  ggplot(aes(x = priming_cond, y = logRT)) +
  geom_violin() +
  geom_jitter () +
  facet_wrap(~ ppt_id)
```



Identify grouping structure:

- we are looking only at categorical variables
- we want to see: what categorical variables have values that are associated with more than one data point?

```{r}
dat |> head()
```

what cat vars do we have? (not including the outcome, if it's binary)
- ppt_id
- item_id
- priming_cond

we'll go through each of these.

assuming data is set up for analysis using lm, with one column per variable and one row per observation, then we should just be able to count how many times each value shows up

```{r}
dat |>
  group_by(ppt_id) |>
  count()
```

can sense check by visually inspecting df. "ppt1" shows up with a bunch of different logRTs.

```{r}
dat |>
  group_by(item_id) |>
  count()
```

each item also appears multiple times.

```{r}
dat |>
  group_by(priming_cond) |>
  count()
```

and each priming condition shows up multiple times.

So we have three groups that each contribute multiple data points:

- `ppt_id`
- `item_id`
- `priming_cond`
 

how do these groups relate to one another?

```{r}
dat |>
  group_by(ppt_id, item_id, priming_cond) |>
  count()
```

each ppt sees each item twice, once in the distractor prime condition and once in the related prime condition.



# week 1 examples

## lifesat ~ treatment + day + (1 + day | ppt)

Good for Week 1 and Week 2

lifesat: continuous, measured after practicing each method daily for a week
treatment: journal, meditation (btwn)
day 

```{r}
half_ppts <- 18
journal <- sim_int_1grp_1pred(seed=2, b0 = 1, b1=.1, N = half_ppts*7, n_groups = half_ppts) |>
  mutate(
    condition = 'journal',
    ppt_id = paste0('j', g)
  ) |>
  select(-y_bin, -g) 
meditate <- sim_int_1grp_1pred(seed=4, b0 = 1, b1=.5, N = half_ppts*7, n_groups = half_ppts) |>
  mutate(
    condition = 'meditate',
    ppt_id = paste0('m', g)
  ) |>
  select(-y_bin, -g)

lifesat_dat <- bind_rows(journal, meditate) |>
  rename(day = x1) |>
  mutate(lifesat = datawizard::rescale(y, to = c(15, 75))) |>
  select(-y)

lifesat_dat |>
  ggplot(aes(x = day, y = lifesat, colour = condition)) +
  geom_point() +
  facet_wrap(~ ppt_id)
```

what categ vars in dataset?

- condition
- ppt_id


what categ vars not an outcome or a predictor?

- ppt_id


what levels?

```{r}
lifesat_dat$ppt_id
```

is each level associated with more than one data point?

```{r}
lifesat_dat
```

one way: just look at the df. the answer is yes bc j1 appears more than 1x, j2 appears more than 1x, etc.

another way, more efficient: use tidyverse to count how many times each ppt label appears.

```{r}
lifesat_dat |>
  group_by(ppt_id) |>
  count()
```




### btwn ppt: day 7 only

```{r}
lifesat_dat_7 <- lifesat_dat |>
  filter(day == 7) |>
  select(-day)

lifesat_dat_7 |>
  ggplot(aes(x = condition, y = lifesat)) +
  geom_violin()
```


what categ vars in dataset?

```{r}
lifesat_dat_7
```

- condition
- ppt_id


what categ vars not an outcome or a predictor?

- ppt_id



what levels?

```{r}
lifesat_dat_7$ppt_id
```

is each level associated with more than one data point?

```{r}
lifesat_dat_7 |>
  group_by(ppt_id) |>
  count()
```

nope, appears only 1x.


## attentiveness ~ hour + (1 + hour | child)

Good for Week 1 and Week 2


children's attentiveness at school across the day

```{r}
attent <- sim_int_1grp_1pred(seed=1, b0 = 2, b1=-.2, N = 140) |>
  rename(
    hour = x1,
    child = g,
    attentiveness = y
  ) |>
  select(-y_bin) |>
  mutate(
    hour = hour - 1,
    child = factor(child),
    attentiveness_z = attentiveness - mean(attentiveness)
  )
 
attent |>
  ggplot(aes(x = hour, y = attentiveness_z)) +
  geom_point() +
  facet_wrap(~ child, labeller = label_both)
```


What categorical variables does your data contain?

- child


How many times does each each level of this variable appear in the data?

```{r}
attent |>
  group_by(child) |>
  count()
```

each child contributes seven data points to the data set.
so the data has a grouping by child.


## correct ~ trial_type + (1 + trial_type | ppt) + (1 | item)

Good for Week 1 and Week 2

How does people's accuracy on an experimental task differ between conditions, and how is this difference affected by people's DASS scores?

```{r}
dass <- read_csv('../data/dass.csv')
dass |> head(20)
```


what categ vars in dataset?

- ppt_id
- target_id
- condition
- correct


what categ vars not an outcome or a predictor?

- ppt_id
- target_id 


for each categ var, what levels does it have?

```{r}
dass$ppt_id |> unique()
dass$target_id |> unique()
```

is each of those levels associated with more than one data point?

```{r}
dass |>
  group_by(ppt_id) |>
  count()
```

yes

```{r}
dass |>
  group_by(target_id) |>
  count()
```

yes.

# week 1 lab examples

## y ~ x + (1 + x | ppts)

```{r}
d1 <- simcross(seed=1, b3 = 3, n_itms = 6, n_ppts = 12, n_obs_per = 3) |>
  select(ppts, y, x) |>
  mutate(y = datawizard::rescale(y, to = c(0, 12)))

d1 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(ppts))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```

## y ~ x 

```{r}
d2 <- simcross(seed=2, b3 = 3, n_itms = 2, n_ppts = 4, n_obs_per = 24) |>
  select(y, x) |>
  rowid_to_column() |>
  rename(ppt_id = rowid) |>
  mutate(
    y = datawizard::rescale(y, to = c(0, 48)),
  )

d2
```


## y ~ x

```{r}
d3 <- MASS::mvrnorm(64, mu = c(0, 0), Sigma = matrix(c(5, 1, 1, 7), nrow=2, byrow=T)) |>
  as_tibble() |>
  rowid_to_column() |>
  rename(
    y = V1,
    x = V2,
    book = rowid
  )

d3 

d3 |>
  ggplot(aes(x = x, y = y)) + geom_point()
```


## y ~ x + (1 + x | ppts) + (1 | itms)

```{r}
d4 <- simcross(seed=4, b0 = 20, b2 = 10, n_itms = 20) |>
  select(ppts, itms, y, wi_ppts_btw_itms) |>
  rename(x = wi_ppts_btw_itms)

d4 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(itms))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```

```{r}
d4 |>
  group_by(ppts) |>
  count()

d4 |>
  group_by(itms) |>
  count()
```

^ grouping structures for both ppts and items

```{r}
d4 |>
  group_by(ppts, x) |>
  count()
```

- each ppt sees both vals of x, so rd slope is poss

```{r}
d4 |>
  group_by(itms, x) |>
  count()
```

- each item does NOT appear w both vals of x, so rd slope is not poss



## y ~ x + (1 | itms)


```{r}
d5 <- simcross(seed=5, b0 = -10, b2 = -5) |>
  select(itms, y, wi_ppts_btw_itms) |>
  rename(x = wi_ppts_btw_itms)

d5 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(itms))) +
  geom_point()
```

- by-item grouping struc bc rep'd obs of ea item
- but no slopes possible by item over cond, bc not every item appears in every cond

```{r}
d5 |>
  group_by(itms, cond) |>
  count()
```


## y ~ x + (1 + x | ppts) + (1 + x | itms)

```{r}
d6 <- simcross(seed=6, b0 = 4, b3 = 2, n_itms = 6, n_obs_per = 10) |>
  select(ppts, itms, y, x) |>
  mutate(
    # x = datawizard::rescale(x, to = c(pi, pi*3)),
    y = abs(log(y))
  )
  # rename(cond = wi_ppts_btw_itms)

d6 |>
  ggplot(aes(x = x, y = y, colour = factor(itms))) +
  facet_wrap(~ ppts) +
  geom_point() +
  theme(legend.position = 'none')
```



## y ~ x + z + (1 + x + z | ppts) + (1 + x | itms)

```{r}
d7 <- simcross(seed=7, b0 = 0, b3 = 0, n_itms = 2, n_obs_per = 6) |>
  select(ppts, itms, y, x, wi_ppts_btw_itms) |>
  rename(z = wi_ppts_btw_itms)

d7
```


```{r}
d7 |>
  ggplot(aes(x = factor(z), y = y, colour = factor(itms))) +
  facet_wrap(~ ppts, labeller = label_both) +
  geom_point() +
  theme(legend.position = 'none')
```


## y ~ x + (1 + x | ppts) + (1 | itms)

```{r}
d8 <- simcross(seed=8, b0 = 0, b2 = -10, n_ppts = 4, n_itms = 8) |>
  select(ppts, itms, y, wi_ppts_btw_itms) |>
  rename(x = wi_ppts_btw_itms)

d8 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(itms))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```



## y ~ x + (1 | itms)

```{r}
d9 <- simcross(seed=9, b0 = -10, b2 = -5, n_itms = 50, n_obs_per = 5) |>
  select(itms, y, x) |>
  mutate(
    x = sample(x),
    y = abs(y)
    )

d9
```


```{r}
d9 |>
  ggplot(aes(x = x, y = y, colour = factor(itms))) +
  geom_point() +
  theme(legend.position = 'none')
```




## y ~ x + (1 + x | ppts)

```{r}
d10 <- simcross(seed=10, b3 = 2, n_itms = 2, n_ppts = 6, n_obs_per = 8) |>
  select(ppts, y, wi_itms_wi_ppts) |>
  rename(x = wi_itms_wi_ppts)

d10 |>
  ggplot(aes(x = factor(x), y = y, colour = factor(ppts))) +
  facet_wrap(~ppts) +
  geom_point() +
  theme(legend.position = 'none')
```
