---
title: "01 playground"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)
library(simglm)
```


# group structure sim functions

- when a bunch of data comes from
  - same ppt
  - same school
  - same item
  
  
```{r varying int one group}
sim_int_1grp_1pred <- function(
    seed=NULL,
    b0=0, b1=1,
    sigma2_u0=1,
    sigma=1,
    N=200, n_groups=20,
    eq=TRUE
){
  #'  Quickly sim data for a basic model y~1+x+(1|g)
  #' @param seed random seed to use
  #' @param b0 fixed intercept
  #' @param b1 fixed slope
  #' @param sigma2_u0 random intercept variance
  #' @param sigma residual sd
  #' @param N how many obs
  #' @param n_groups how many groups
  #' @param eq equality of x between groups (if true, categ x1, else zscore)
  
  if(!is.null(seed)){
    set.seed(seed)
  }
  
  g = rep(1:n_groups, e = N/n_groups)
  
  if(eq){
    x1 = rep(1:(N/n_groups), n_groups)
  } else {
    x1 = rnorm(N)
  }
  
  # intercept adjustments
  u0 = rnorm(n_groups, mean = 0, sd = sqrt(sigma2_u0))[g]
  
  lp = (b0 + u0) + (b1 * x1)
  
  y = rnorm(N, mean = lp, sd = sigma)             # continuous target
  y_bin = rbinom(N, size = 1, prob = plogis(lp))  # binary target
  
  data.frame(x1, g=factor(g), y, y_bin)
}

sim_int_1grp_1pred(seed = 1, N = 6, n_groups = 2, eq = T)
```


```{r varying int and slp one group}
sim_intslp_1grp_1pred <- function(
    seed=NULL,
    b0=0, b1=1,
    sigma2_u0=1, sigma2_u1=1,
    rho=0,
    sigma=1,
    N=200, n_groups=20,
    eq=TRUE
){
  #'  Quickly sim data for a basic model y~1+x+(1+x|g)
  #' @param seed random seed to use
  #' @param b0 fixed intercept
  #' @param b1 fixed slope
  #' @param sigma2_u0 random intercept variance
  #' @param sigma2_u1 random slope var
  #' @param rho int-slope correl
  #' @param sigma residual SD
  #' @param eq equality of x between groups (if true, categ x1, else zscore)
  
  if(!is.null(seed)){
    set.seed(seed)
  }
  
  g = rep(1:n_groups, e = N/n_groups)
  
  if(eq){
    x1 = rep(1:(N/n_groups), n_groups)
  } else {
    x1 = rnorm(N)
  }
  
  Gcov = Matrix::nearPD(matrix(c(
    sigma2_u0,
    rho*sigma2_u0*sigma2_u1,
    rho*sigma2_u0*sigma2_u1,
    sigma2_u1
  ),nrow=2))$mat
  
  res = MASS::mvrnorm(n_groups, mu=c(0,0), Sigma=Gcov)
  u0  = res[,1][g]  # random intercepts
  u1  = res[,2][g]  # random slopes
  
  lp = (b0 + u0) + (b1 + u1) * x1
  
  y = rnorm(N, mean = lp, sd = sigma)             # continuous target
  y_bin = rbinom(N, size = 1, prob = plogis(lp))  # binary target
  
  data.frame(x1, g=factor(g), y, y_bin)
}

sim_intslp_1grp_1pred(seed = 2, N = 8, n_groups = 2, eq = T)
```
  
  
```{r varying int two grps crossed}
sim_int_2grp_1pred <- function(
    seed=NULL,
    b0=0, b1=1,
    sigma2_u0=1,
    sigma2_w0=1,
    sigma=1,
    N=200,
    n_groups_u=20,
    n_groups_w=10,
    eq=TRUE
){
  #' Quickly sim data for a basic crossed ranef model y~1+x+(1|g)+(1|h)
  #' @param seed random seed to use
  #' @param b0 fixed intercept
  #' @param b1 fixed slope
  #' @param sigma2_u0 random intercept variance for group g
  #' @param sigma2_w0 random intercept variance for group h
  #' @param sigma residual SD
  #' @param N how many obs (multiple of n_groups_u AND n_groups_w)
  #' @param n_groups_u how many groups of type g
  #' @param n_groups_w how many groups of type h
  #' @param eq equality of x between groups (if true, categ x1, else zscore)
  
  if(!is.null(seed)){
    set.seed(seed)
  }
  
  g = rep(1:n_groups_u, e = N/n_groups_u)
  h = rep(1:n_groups_w, e = N/n_groups_w)
  
  if(eq){
    x1 = rep(1:(N/(n_groups_u*n_groups_w)), (n_groups_u*n_groups_w))
  } else {
    x1 = rnorm(N)
  }
  
  # intercept adjustments
  u0 = rnorm(n_groups_u, mean = 0, sd = sqrt(sigma2_u0))[g]
  w0 = rnorm(n_groups_w, mean = 0, sd = sqrt(sigma2_w0))[h]

  lp = (b0 + u0 + w0) + (b1 * x1)

  y = rnorm(N, mean = lp, sd = sigma)             # continuous target
  y_bin = rbinom(N, size = 1, prob = plogis(lp))  # binary target

  data.frame(x1, g=factor(g), h=factor(h), y, y_bin)
}

sim_int_2grp_1pred(seed = 1, eq = T)
```


TODO FIXME: the function below is not working.
there's some excessive repetition that's not quite right.
try jo's simcrossed stuff.

```{r varying int + slp two groups crossed}
sim_intslp_2grp_1pred <- function(
    seed=NULL,
    b0=0, b1=1,
    sigma2_u0=1,
    sigma2_u1=1,
    sigma2_w0=1,
    sigma2_w1=1,
    rho_u = 0,
    rho_w = 0,
    sigma=1,
    N=200,
    n_groups_u=20,
    n_groups_w=10,
    eq=TRUE
){
  #' Quickly sim data for a basic crossed ranef model y~1+x+(1|g)+(1|h)
  #' @param seed random seed to use
  #' @param b0 fixed intercept
  #' @param b1 fixed slope
  #' @param sigma2_u0 random intercept variance for group g
  #' @param sigma2_w0 random intercept variance for group h
  #' @param sigma2_u1 random slope variance for group g
  #' @param sigma2_w1 random slope variance for group h
  #' @param rho_u correl btwn int, slope for group g
  #' @param rho_w correl btwn int, slope for group h
  #' @param sigma residual SD
  #' @param N how many obs (multiple of n_groups_u AND n_groups_w)
  #' @param n_groups_u how many groups of type g
  #' @param n_groups_w how many groups of type h
  #' @param eq equality of x between groups (if true, categ x1, else zscore)
  
  if(!is.null(seed)){
    set.seed(seed)
  }
  
  n_reps <- (N/n_groups_u)/n_groups_w
  print(n_reps)
  
  g = rep(1:n_groups_u, e = N/n_groups_u)
  h = rep(rep(1:n_groups_w, e = 2), N/n_groups_w)
  
  # print(g)
  # print(h)
  
  if(eq){
    x1 = rep(1:(N/(n_groups_u*n_groups_w)), (n_groups_u*n_groups_w))
  } else {
    x1 = rnorm(N)
  }
  
  vcov_u = Matrix::nearPD(matrix(c(
    sigma2_u0,
    rho_u*sigma2_u0*sigma2_u1,
    rho_u*sigma2_u0*sigma2_u1,
    sigma2_u1
  ),nrow=2))$mat
  
  vcov_w = Matrix::nearPD(matrix(c(
    sigma2_w0,
    rho_w*sigma2_w0*sigma2_w1,
    rho_w*sigma2_w0*sigma2_w1,
    sigma2_w1
  ),nrow=2))$mat
  
  res_u = MASS::mvrnorm(n_groups_u, mu=c(0,0), Sigma=vcov_u)
  res_w = MASS::mvrnorm(n_groups_w, mu=c(0,0), Sigma=vcov_w)
  
  u0  = res_u[,1][g]  # random intercepts for group g
  u1  = res_u[,2][g]  # random slopes for group g
  
  w0  = res_w[,1][h]  # random intercepts for group h
  w1  = res_w[,2][h]  # random slopes for group h
  
  lp = (b0 + u0 + w0) + (b1 + u1 + w1) * x1

  y = rnorm(N, mean = lp, sd = sigma)             # continuous target
  y_bin = rbinom(N, size = 1, prob = plogis(lp))  # binary target

  data.frame(x1, g=factor(g), h=factor(h), y, y_bin)
}

sim_intslp_2grp_1pred(seed = 1, N = 64, n_groups_u = 4, n_groups_w = 8, eq = T)
```


^ if want btwn effects, then can just filter some combinations out!


```{r nested}
n_highergroup <- 10
N <- 200

df <- do.call(rbind,
  lapply(rnorm(n_highergroup), \(x) sim_intslp_1grp_1pred(b0=x, N = N))
)
df$highergroup = rep(1:n_highergroup,e=N)
 
lme4::lmer(y~1+x1+(1|highergroup)+(1+x1|highergroup:g),df)
```


# simdat for lec, labs

## ppts w/i, conds btwn: main running example for lec 2

```{r}
n_ppts  <- 4
n_items <- 8
n_conds <- 4

dat <- sim_intslp_2grp_1pred(
  seed = 1, 
  N = n_ppts * n_items * n_conds,
  n_groups_u = n_ppts,
  n_groups_w = n_items,
  eq = T
) |> rename(
    cond = x1,
    ppt_id = g,
    item_id = h
  ) |>
  arrange(ppt_id)

dat
```


```{r}
# make item_id btwn conds, not within, so filter out half.
# eg, if five items per condition: cond 1 has items 1:5, cond 2 has items 6:10
dat <- dat |>
  filter(
    (cond == 1) & (item_id %in% 1:(n_items/2)) |
    (cond == 2) & (item_id %in% ((n_items/2)+1):n_items)
  )

dat
```

why does ppt 1 never see cond 2? 


Cosmetic tweaks:

```{r}

```













# graveyard

simglm can't force_equal w/i ppt, rip

```{r eval=F}
set.seed(321) 

sim_arguments <- list(
  formula = y ~ 1 + sleep + (1 + sleep | ppt),
  reg_weights = c(0.5, 0.3),
  fixed = list(
    # weight = list(var_type = 'continuous', mean = 180, sd = 30),
    # age = list(var_type = 'ordinal', levels = 30:60),
    sleep = list(
      var_type = 'factor', 
      levels = c('before', 'after'),
      force_equal = TRUE
    )
  ),
  randomeffect = list(
    int_ppt = list(variance = 8, var_level = 2),
    sleep_ppt = list(variance = 3, var_level = 2)
    ),
  sample_size = list(
    level1 = 2,   # how many obs per ppt
    level2 = 3   # how many ppts
  )
  # # the below args for binary outcome
  # outcome_type = 'binary',
  # model_fit = list(
  #   model_function = 'glm',
  #   family = binomial
  # )
)

nested_data <- simulate_fixed(data = NULL, sim_arguments) |>
  simulate_randomeffect(sim_arguments) |>
  simulate_error(sim_arguments) |>
  generate_response(sim_arguments)

nested_data
```
