---
title: "<b>WEEK 2<br>Path Mediation</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      navigation:
        scroll: false
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.asp=.7
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="70%","overflow-y"="scroll"))
)
```

# Learning Objectives
1. Understand the purpose of mediation models and the conceptual challenges
2. Be able to describe direct, indirect and total effects in a mediation model.
3. Estimate and interpret a mediation model using `lavaan`

---
class: inverse, center, middle

<h2>Part 1: Introduction to mediation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Direct, indirect and total effects</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Estimating mediation in `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reporting</h2>

---
# Mediation

- Is when a predictor X, has an effect on an outcome Y, via a mediating variable M

- The mediator **transmits** the effect of X  to Y 

- Examples of mediation hypotheses:
    - Conscientiousness (X) affects health (Y) via health behaviors (M)
    - Conduct problems (X) increase the risk of depression (Y) via peer problems (M)
    - Attitudes to smoking (X) predict intentions to smoke (M) which in turn predicts smoking behavior (Y) 
    - An intervention (X) to reduce youth crime (Y) works by increasing youth self-control (M)
    
---
# Visualising a mediation model

.pull-left[
- In a SEM diagram we can represent mediation as:
]

.pull-right[
```{r, echo=FALSE}
magick::image_read("Mediation diagram basic.png")
```
]

---
# Mediation...not to be confused with moderation
- Mediation is commonly confused with **moderation** 

- Moderation is when a moderator z modifies the effect of X on Y
  - e.g., the effect of X on Y is stronger at higher levels of Z
  - Also known as an **interaction** between X and Z

- Examples of moderation could be:
    - An intervention (X) works better to reduce bullying (Y) at older ages (Z) of school pupil
    - The relation between stress (X) and depression (Y) is lower for those scoring higher on spirituality (Z) 

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 1

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to mediation</h2>
<h2>Part 2: Direct, indirect and total effects</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Estimating mediation in `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reporting</h2>  

---
# Direct and indirect effects in mediation
- We seldom hypothesize that a mediator completely explains the relation between X and Y

- More commonly, we expect both **indirect effects** and **direct effects** of X on Y
  - The indirect effects of X on Y are those transmitted via the mediator
  - The direct effect of X on Y is the remaining effect of X on Y


---
# Visualing direct and indirect effects in mediation

```{r, echo=FALSE}
magick::image_read("Mediation diagram basic indirect.png")
```

---
# Testing mediation

.pull-left[
- Traditionally, mediation was tested using a series of separate linear models:
  1. Y~X  
  2. Y~X+M  
  3. M~X

- May see this referred to as the Baron and Kenny approach.

]

.pull-right[
```{r, echo=FALSE}
magick::image_read("trad med.png")
```

]



---
# Traditional methods for mediation
.pull-left[
- The three regression models:
  1. Y~X  
  2. Y~X+M  
  3. M~X
    
]

.pull-right[
- Model 1 estimates the overall effect of X on Y
- Model 2 estimates the partial effects of X and M on Y
- Model 3 estimates the effect of X on M

- If the following conditions were met, mediation was assumed to hold:
  - The effect of X on Y (eq.1) is significant
  - The effect of X on M (eq.3) is significant
  - The effect of X on Y becomes reduced when M is added into the model (eq.2)
]


---
# Limitations of traditional methods for mediation
- Low power

- Very cumbersome for multiple mediators, predictors, or outcomes

- You don't get an estimate of the magnitude of the indirect effect

- Much better way: **path mediation model**

---
# BREAK QUIZ
- Quiz question:
    - Which of these hypotheses is a mediation hypothesis?
      - 1) Vocabulary development in childhood follows a non-linear trajectory
      - 2) The effects of conscientiousness on academic achievement are stronger at low levels of cognitive ability
      - 3) Poverty affects child behavior problems through increasing parental stress
      - 4) Earlier pubertal onset increases the risk of antisocial behavior only in girls and not boys
      
---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 2

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to mediation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Direct, indirect and total effects</h2>
<h2>Part 3: Estimating mediation in `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reporting</h2>

---
# WELCOME BACK

- Welcome back!
- The answer to the quiz question is...
   - Which of these hypotheses is a mediation hypothesis?
      - 1) Vocabulary development in childhood follows a non-linear trajectory
      - 2) The effects of conscientiousness on academic achievement are stronger at low levels of cognitive ability
      - 3) **Poverty affects child behavior problems through increasing parental stress**
      - 4) Earlier pubertal onset increases the risk of antisocial behavior only in girls and not boys


---
# Testing a path mediation model in lavaan

- Specification
  - Create a lavaan syntax object

- Estimation
  - Estimate the model using e.g., maximum likelihood estimation

- Evaluation/interpretation
  - Inspect the model to judge how good it is
  - Interpret the parameter estimates


---
# Example
.pull-left[
- Does peer rejection mediate the association between aggression and depression?

```{r peer rejection example simulate data, include=F}
library(lavaan)
agg.PR.dep<-'Dep~0.3*PR
             Dep~0.2*Agg
             PR~0.5*Agg'

agg.data2<-simulateData(model=agg.PR.dep, standardized = T, model.type='sem')
write.csv(agg.data2, "~/agg-data.csv")
```
]

.pull-right[

```{r, echo=FALSE}
magick::image_read("Mediation diagram example.png")
```

]


---
# The data
```{r example 1 descriptives}
slice(agg.data2, 1:10)
  
```


---
# Mediation Example

- Does peer rejection mediate the association between aggression and depression?

```{r}
model1<-'Dep ~ PR      # Depression predicted by peer rejection
         Dep ~ Agg     # Depression predicted by aggression (the direct effect)
         PR ~ Agg      # Peer rejection predicted by aggression
'     
```

- Estimate the model 

```{r}
model1.est<-sem(model1, data=agg.data2)
```


---
# Model Ouput

+ Typically we want to see:
  + Model estimates
  + Model fit
  + Standardized solutions
  + Possibly modification indices
  
+ We get those by:

```{r, eval=FALSE}
summary(model1.est, 
        fit.measures = T, 
        std=T, 
        modindices = T)
```


---
# The model output

.scroll-output[
```{r}
summary(model1.est)
```
]

---
# Things to note from the model output

- All three regressions paths are statistically significant

- The model is **just-identified**
  - The degrees of freedom are equal to 0
  - The model fit cannot be tested
  - The model fit statistics (TLI, CFI, RMSEA, SRMR) all suggest perfect fit but this is meaningless

---
# Visualising the model using `semPaths()`

- We can use semPaths() from the semPlot package to help us visualize the model
    - Shows the parameter estimates within an SEM diagram

```{r, eval=FALSE}
library(semPlot)
semPaths(model1.est, what='est')
```

---
# Visualising the model using `semPaths()`

```{r, echo=FALSE}
library(semPlot)
semPaths(model1.est, what='est')
```


---
# Calculating the indirect effects 

.pull-left[
- To calculate the indirect effect of X on Y in path mediation, we need to create some new parameters

- The indirect effect of X on Y via M is:
  - $a*b$
  - $a$ = the regression coefficient for M~X
  - $b$ = the regression coefficient for Y~M
]

.pull-right[
```{r, echo=FALSE}
magick::image_read("Mediation diagram example a b.png")
```

]

---
# Calculating indirect effects in lavaan

.pull-left[
- To calculate the indirect effect of X on Y in lavaan we:

- Use parameter labels 'a' and 'b' to label the relevant paths
  - a is for the effect of X on M
  - b is for the effect of M on Y

- Use the ':=' operator to create a new parameter 'ind'
  - 'ind' represents our indirect effect
]

.pull-right[
```{r indirect effects}
model1<-'Dep~b*PR        
         Dep~Agg     
         PR~a*Agg
         ind:=a*b
         '
```
]

---
# Indirect effects in the output

.scroll-output[
```{r indirect effects output}

model1.est<-sem(model1, data=agg.data2)
summary(model1.est)
```

]

---
# Statistical significance of the indirect effects
- Default method of assessing the statistical significance of indirect effects assume normal sampling distribution

- May not hold for indirect effects which are the product of regression coefficients

- Instead we can use **bootstrapping**
  - Allows 95% confidence intervals (CIs) to be computed
  - If 95% CI includes 0, the indirect effect is not significant at alpha=.05
    
---
# Bootstapped CIs for indirect effect in lavaan

```{r indirect effects with bootstrapped CIs}
  model1<-'Dep~b*PR          
           Dep~Agg     
           PR~a*Agg      
ind:=a*b'    

model1.est<-sem(model1, data=agg.data2, se='bootstrap') #we add the argument se='bootstrap'


```

---
# Output for bootstrapped CIs 

.scroll-output[
```{r indirect effect with bootstrapped CIs output}
summary(model1.est, ci=T) # we add the argument ci=T to see the confidence intervals in the output
```
]

---
# Total effects in path mediation

- As well as the direct and indirect effect, it is often of interest to know the **total** effect of X on Y

$$Total = Indirect + Direct$$

---
# Total effects in path mediation

$$Total = a*b + c$$

```{r, echo=FALSE}
magick::image_read("Mediation diagram example a b c.png")
```

---
# Total effect in lavaan

```{r total effects}
  model1<-'Dep~b*PR          
           Dep~c*Agg         # we add the label c for our direct effect    
           PR~a*Agg      
ind:=a*b
total:=a*b+c                 # we add a new parameter for the total effect'    

model1.est<-sem(model1, data=agg.data2, se='bootstrap') #we add the argument se='bootstrap'

```

---
# Total effect in lavaan output

.scroll-output[
```{r total effect output}

summary(model1.est, ci=T)

```
]

---
# Why code the total effect in lavaan?

- We could have just added up the coefficients for the direct and indirect effects

- By coding it in lavaan, however, we can assess the statistical significance of the total effect

- Useful because sometimes the direct and indirect effects are not individually significant but the total effect is
  - May be especially relevant in cases where there are many mediators of small effect

---
# Interpreting the total, direct, and indirect effect coefficients
- The total effect can be interpreted as the **unit increase in Y expected to occur when X increases by one unit**

- The indirect effect can be interpreted as the **unit increase in Y expected to occur via M when X increases by one unit**

- The direct effect can be interpreted as the **unit increase in Y expected to occur with a unit increase in X over and above the increase transmitted by M**

- **Note**: 'direct' effect may not actually be direct - it may be acting via other mediators not included in our model

---
# Standardised parameters
- As with CFA models, standardized parameters can be obtained using:

```{r, eval=FALSE}

summary(model1.est, ci=T, std=T)
```

---
# Standardised parameters
.scroll-output[
```{r stand param, echo=FALSE}

summary(model1.est, ci=T, std=T)
```
]

---
# BREAK QUIZ

- Time for a pause
- Quiz question
    - If the effect of X on M is b=.30 and the effect of M on Y is b=.10, what is the indirect effect of X on Y?
      - 1) b=.40
      - 2) b=.03
      - 3) b=.30
      - 4) b=.10
      

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 3

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to mediation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Direct, indirect and total effects</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Estimating mediation in `lavaan`</h2>
<h2>Part 4: Reporting</h2>



---
# Welcome back
- The answer to the quiz question is...
- Quiz question
    - If the effect of X on M is b=.30 and the effect of M on Y is b=.10, what is the indirect effect of X on Y?
      - 1) b=.40
      - 2) **b=.03**
      - 3) b=.30
      - 4) b=.10

---
# Reporting path mediation models 

- Methods/ Analysis Strategy
  - The model being tested
  - e.g. 'Y was regressed on both X and M and M was regressed on X'
  - The estimator used (e.g., maximum likelihood estimation)
  - The method used to test the significance of indirect effects ('bootstrapped 95% confidence intervals')

- Results
  - Model fit (for over-identified models)
  - The parameter estimates for the path mediation  and their statistical significance
  - Can be useful to present these in a SEM diagram
  - The diagrams from R not considered 'publication quality' draw in powerpoint or similar

---
# Reporting path mediation models - example of SEM diagram with results

.pull-left[
- Include the key parameter estimates

- Indicate statistically significant paths (e.g. with an '*')

- Include a figure note that explains how statistically significant paths (and at what level) are signified 

]

.pull-right[

```{r, echo=FALSE}
magick::image_read("med reporting.png")
```

]

---
# Reporting path mediation models - the indirect effects

- Results
 - The coefficient for the indirect effect and the bootstrapped 95% confidence intervals
 - Common to also report **proportion mediation**:
  
$$\frac{indirect}{total}$$ 
  
- However, important to be aware of limitations:
  - Big proportion mediation possible when total effect is small - makes effect seem more impressive
  - Small proportion mediation even when total effect is big - can underplay importance of effect
  - Should be interpreted in context of total effect

- Tricky interpretation if there are a mix of negative and positive effects involved


---
# Extensions of path mediation models
- We can extend our path mediation model in various ways:
  - Several mediators in sequence or parallel
  - Multiple outcomes
  - Multiple predictors
  - Multiple groups (e.g., comparing direct and indirect effects across males and females)
  - Add covariates to adjust for potential confounders


---
# Example: Multiple mediation model
.pull-left[
```{r mult med}
model2<-'Dep~b2*Aca  
         Aca~a2*Agg
         Dep~b1*PR
         PR~a1*Agg
         Dep~c*Agg  

        ind1:=a1*b1
        ind2:=a2*b2
        
        total=a1*b1+a2*b2+c
'
```
]

.pull-right[

```{r, echo=FALSE}
magick::image_read("Mediation diagram example multiple mediation.png")
```

]

---
## Other path analysis models
- Path mediation models are a common application of path models
  - But they are just one example

- Anything that can be expressed in terms of regressions between observed variables can be tested as a path model
  - Can include ordinal or binary variables
  - Can include moderation

- Other common path analysis models include:
    - Auto-regressive models for longitudinal data
    - Cross-lagged panel models for longitudinal data

---
# Making model modifications
- You **may** want to make some modifications to your initially hypothesized model 
  - non-significant paths that you want to trim
  - include some additional paths not initially included

- Remember that this now moves us into exploratory territory where:
  - Model modifications should be substantively as well as statistically justifiable
  - You must be aware of the possibility that you are capitalizing on chance
  - You should aim to replicate the modifications in independent data

---
# Cautions regarding path analysis models
- **Assumption** that the paths represent causal effects is only an assumption
  - Especially if using cross-sectional data
  - Mediation models should ideally be estimated on longitudinal data.
    - X time 1
    - M time 2
    - Y time 3

- The parameters are only accurate if the model is correctly specified

---
# Cautions: Indistinguishable models

```{r, echo=FALSE}
magick::image_read("med versus confounding.png")
```

---
# Measurement error in path analysis
- Path analysis models use observed variables
  - Assumes no measurement error in these variables

- Path coefficients likely to be attenuated due to unmodeled measurement error

- Structural equation models solve this issue
  - They are path analysis models where the paths are between latent rather than observed variables
  - ...very brief comment on this in the final week

---
# Path analysis summary
- Path analysis can be used to fit sets of regression models
  - Common path analysis model is the path mediation model
  - But very flexible huge range of models that can be tested

- In R, path analysis can be done using the `sem()` function in `lavaan`

- Need to be aware that we aren't *testing* causality but assuming it


---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# Bonus `laavan`

---
# 2+ category nominal predictors

+ To include these variables, we need to create the 0/1 dummy coded variables:

```{r}
tib1 <- tibble(
  city = c("Bristol","Bristol","Birmingham","Birmingham", "London", "London"),
  cityd1 = ifelse(city =="Bristol", 1, 0),
  cityd2 = ifelse(city == "Birmingham", 1, 0)
)

tib1

```

+`cityd1` and `cityd2` can then be added to a `lavaan` model syntax.

---
# Interactions

+ Here we need to create the product variable in our data set:

```{r}
tib2 <- tibble(
  pred1 = rnorm(100, 15, 2),
  pred2 = rnorm(100, 10, 4),
  pred1z = scale(pred1)[,1],
  pred2z = scale(pred2)[,1],
  prod12 = pred1z*pred2z
) 

slice(tib2, 1:3)
```



---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# End

