---
title: "Tests"
---




```{r setup, include=FALSE}
source('assets/setup.R')
library(xaringanExtra)
xaringanExtra::use_panelset()
```

:::lo
learning obj
:::

:::imp
important
:::

:::sticky
sticky
:::

:::div-dot
:::

:::rtip
r tips
:::

:::statbox
statbox
:::

:::int
interprtation
interprtation
interprtation
:::

`r qbegin()`
question  
what is your name?  
what is your favourite colour?  
`r qend()`
`r solbegin(show=TRUE,toggle=TRUE)`
solution  

hello 
```{r}
2+2
```


`r solend()`

`r optbegin("hello my optional friend")`
it's nice to see you again
`r optend()`

::::panelset
:::panel
this is not a panel
:::
:::panel
this is a panel
:::
:::panel
this is a panel
:::

::::

::: {.callout-note collapse="true"}
Note that there are five types of callouts, including:
`note`, `warning`, `important`, `tip`, and `caution`.
:::

::: {.callout-tip}
## Tip with Title

This is an example of a callout with a title.
:::



::: {.callout-caution collapse="true"}
## Expand To Learn About Collapse

This is an example of a 'folded' caution callout that can be expanded by the user. You can use `collapse="true"` to collapse it by default or `collapse="false"` to make a collapsible callout that is expanded by default.
:::


