---
title: 'Analysis Example: Rpt & Mixed ANOVA'
code-fold: true
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
---

```{r setup, include=F}
library(tidyverse)
knitr::opts_chunk$set(message = F, warning = F, fig.align = 'center')
```

:::imp
This is optional for the DAPR3 course, but may be useful for your dissertations should your field/supervisor prefer the ANOVA framework to that of the linear model. 

This walks briefly through these models with the __ez__ package. There are many other packages available, and many good tutorials online should you desire extra resources in the future:

- https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r
- https://www.r-bloggers.com/2021/04/repeated-measures-of-anova-in-r-complete-tutorial/
- https://stats.idre.ucla.edu/r/seminars/repeated-measures-analysis-with-r/ 
- https://www.datanovia.com/en/lessons/mixed-anova-in-r/  

:::


:::frame
__Data: Audio interference in executive functioning__  


```{r}
#| echo: false
#| include: false
# SDMT ~ audio*headph + (1 + audio | ppt)
set.seed(24)
n_groups = 24
N = n_groups*3*5
g = rep(1:n_groups, e = N/n_groups)

w = rep(rep(letters[1:3],5),n_groups)
w1 = model.matrix(lm(rnorm(N)~w))[,2]
w2 = model.matrix(lm(rnorm(N)~w))[,3]

b = rep(0:1, e = N/2)

re0 = rnorm(n_groups, sd = 2)[g]
re_w1  = rnorm(n_groups, sd = 1)[g]
re_w2  = rnorm(n_groups, sd = 1)[g]

lp = (0 + re0) + 
  (2)*b + 
  (0 + re_w1)*w1 +
  (-2 + re_w2)*w2 + 
  (1)*b*w1 +
  (-2)*b*w2
  
y = rnorm(N, mean = lp, sd = 1.5) # create a continuous target variable

df <- data.frame(w, g=factor(g),b, y)
head(df)
with(df,boxplot(y~interaction(w,b)))

library(tidyverse)
df %>% transmute(
  PID = paste0("PPT_",formatC(g,width=2,flag=0)),
  audio = fct_recode(factor(w),
                     no_audio = "a",
                     white_noise = "b",
                     music = "c"),
  headphones = fct_recode(factor(b),
                          speakers = "0",
                          anc_headphones = "1"),
  SDMT = pmax(0,round(35 + scale(y)[,1]*12))
) %>% arrange(PID,audio,headphones) -> ef_music


ef_music <- ef_music %>% group_by(PID) %>%
  mutate(trial_n = paste0("Trial_",formatC(sample(1:15),width=2,flag=0))) %>%
  arrange(PID,trial_n) %>% ungroup()

ef_sdmt <- ef_music %>% select(PID, trial_n, SDMT) %>%
  pivot_wider(values_from=SDMT,names_from=trial_n)

# for(i in 1:38){
#   ef_sdmt[sample(1:24,1), sample(2:15,1)] <- NA
# }

ef_music <- ef_music %>% select(PID, trial_n, audio, headphones) %>%
  arrange(PID,audio)

# write_csv(ef_music,"../../data/ef_music.csv")
# write_csv(ef_sdmt,"../../data/ef_sdmt.csv")

# executive functioning

# library(lme4)
# m = lmer(SDMT ~ 1 + audio * headphones + (1 + audio | PID), ef_music)
# summary(m)
# interactions::cat_plot(m,pred="audio",modx="headphones")
```

This data is from a simulated study that aims to investigate the following research questions: 

> How do different types of audio interfere with executive functioning, and does this interference differ depending upon whether or not noise-cancelling headphones are used? 

`r length(unique(ef_music$PID))` healthy volunteers each completed the Symbol Digit Modalities Test (SDMT) - a commonly used test to assess processing speed and motor speed - a total of 15 times. During the tests, participants listened to either no audio (5 tests), white noise (5 tests) or classical music (5 tests). Half the participants listened via active-noise-cancelling headphones, and the other half listened via speakers in the room.  

The data is in stored in two separate files - the research administering the tests recorded the SDMT score in one spreadsheet, while details of the audio used in the experiment are held in a separate sheet

1. Information about the audio condition for each trial of each participant is stored in __.csv__ format at [https://uoepsy.github.io/data/ef_music.csv](https://uoepsy.github.io/data/ef_music.csv). The data is in long format (1 row per participant-trial).  
```{r echo=FALSE}
tibble(
  variable = names(ef_music),
  description = c("Participant ID",
                  "Trial Number (1-15)",
                  "Audio heard during the test ('no_audio', 'white_noise','music')","Whether the participant listened via speakers in the room or via noise cancelling headphones")
  ) %>% knitr::kable()
```

2. Information on participants' Symbol Digit Modalities Test (SDMT) for each trial is stored in **.xlsx** format at [https://uoepsy.github.io/data/ef_sdmt.xlsx](https://uoepsy.github.io/data/ef_sdmt.xlsx). The data is in wide format (1 row per participant, 1 column per trial).   
```{r echo=FALSE}
tibble(
  variable = c(names(ef_sdmt)[c(1:4)],"...","...",names(ef_sdmt)[16]),
  description = c("Participant ID",paste0("SDMT score in trial ",c(1,2,3,"...","...",15)))
) %>% knitr::kable()
```


:::

The code below will read in both datasets and join them for you:

```{r message=F,warning=F}
library(tidyverse)
library(readxl)
download.file(url = "https://uoepsy.github.io/data/ef_sdmt.xlsx",
              destfile = "ef_sdmt.xlsx",
              mode = "wb")
efdata <- 
  left_join(
    read_csv("https://uoepsy.github.io/data/ef_music.csv"),
    read_xlsx("ef_sdmt.xlsx") %>%
      pivot_longer(Trial_01:Trial_15, names_to = "trial_n", values_to = "SDMT")
  )
```



# One-Way Repeated Measures ANOVA

For a repeated measures ANOVA, we have one independent variable that is _within_ group.  
This would be appropriate if our research question were the following:    

> How do different types of audio interfere with executive functioning?  

Mapping this to the variables in our dataset, our model is going to be `SDMT ~ audio`, and we want to account for `PID` differences. So for now we will ignore the `headphones` variable.  
```{r}
head(efdata)
```


The easiest way to conduct a repeated measures ANOVA in R is to use the __ez__ package, which comes with some handy functions to visualise the experimental design.  
We can see from below that every participant completed 5 trials for each type of audio interference:
```{r}
library(ez)
ezDesign(data = efdata, x = audio, y = PID)
```

The `ezANOVA()` function takes a few arguments. 

The ones you will need for this are:

- *data* the name of the dataframe
- *dv* the column name for the dependent variable
- *wid* the column name for the participant id variable
- *within* the column name(s) for the predictor variable(s) that vary within participants 
- *between* the column name(s) for any predictor variable(s) that vary between participants

Fit a repeated measures ANOVA to examine the effect of the audio type on SDMT:  

```{r}
ezANOVA(data = efdata, dv = SDMT, wid = PID, within = audio)
```


# Mixed ANOVA 

Mixed ANOVA can be used to investigate effects of independent variables that are at two different levels,  i.e. some are *within* clusters and some are *between*.  

> Does the effect of audio interference on executive functioning differ depending upon whether or not noise-cancelling headphones are used? 

Look at the two lines below. Can you work out what the plots will look like _before_ you run them? 

```{r eval=F}
ezDesign(data = efdata, x = headphones, y = PID)
ezDesign(data = efdata, x = headphones, y = audio)
```

Participants 1-20 are in one condition, and 21-40 are in another.  
This should look like a two big blocks on the diagonal. 
```{r}
ezDesign(data = efdata, x = headphones, y = PID)
```
In each condition, all different types of audio were observed in the same number of trials.
This should be a full grid:
```{r}
ezDesign(data = efdata, x = headphones, y = audio)
```

Fit a mixed ANOVA to examine the interaction between audio and headphone use on SDMT:  
```{r}
ezANOVA(data = efdata, dv = SDMT, wid = PID, within = audio, between = headphones)
```

The __ez__ package also contains some easy plotting functions for factorial experiments, such as `ezPlot()`. It takes similar arguments to the `ezANOVA()` function. 

- look up the help documentation for `ezPlot()`. 
- let's use `ezPlot()` to make a nice plot

```{r}
ezPlot(data = efdata, dv = SDMT, 
       wid = PID, within = audio, between = headphones,
       x = audio, split = headphones)
```


::: {.callout-note collapse="true"}
#### The same thing in lmer

```{r}
library(lme4)
library(lmerTest)
mod <- lmer(SDMT ~ 1 + headphones * audio + (1 + audio | PID), 
            data = efdata)
anova(mod, type="III")
```

```{r}
library(sjPlot)
plot_model(mod, type="eff", terms=c("audio","headphones"))
```

:::



<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>

