---
title: 'Analysis Example: Intervention'
output:
  html_document:
    code_folding: show
    theme: flatly
    toc: yes
    toc_depth: 3
    toc_float: yes
    css: assets/style-labs.css
---
```{r setup, include=F}
knitr::opts_chunk$set(message = F, warning = F, fig.align = 'center')
```

:::frame

Each of these pages provides an analysis run through for a different type of design. Each document is structured in the same way:  

  - First the data is introduced. For the purpose of these tutorials, I will only use examples where the data can be shared - either because it is from an open access publication, or because it is unpublished or simulated. 
  - Second, the structure of the data is discussed so that we can more easily see what data structure the design creates, and how this aligns to the variables in the data.
  - Third, we translate the research questions into formal equations, and then `lmer()` code. 
  - Finally, we will follow those through for our example data.

:::
 
# Overview 

# Data structure

# Model specification

## Equations


# Analysis

```{r}
library(tidyverse)
library(lme4)
library(kableExtra)
```


## Tidy data
```{r}
simMIX <- tibble(
  ID = factor(rep(paste("ID", 1:50, sep=""),each=3)) ,
  Stress = round(c(rnorm(75,c(40,45,42),sd=c(4, 6, 5)),rnorm(75,c(90,60,20),sd=c(6, 10, 5))),0),
  Time = factor(rep(c("Pre", "During", "Post"), each=1,50)),
  Group = factor(c(rep("Control", 75), rep("Treatment", 75)))
)

simMIX[c(1:3, 148:150),]
```

## Describe

We now have two factors of interest, scores over `Time` and by `Group`. So we do our descriptive statistics by multiple grouping factors.

```{r}
sumMIX <- 
  simMIX %>%
  group_by(Time, Group) %>%
  summarise(
    N = n_distinct(ID),
    Mean = round(mean(Stress, na.rm=T),2),
    SD = round(sd(Stress, na.rm=T),2)
    )
```

```{r}
kable(sumMIX) %>%
  kable_styling("striped")
```

## Visualize Data

We can use a similar violin plot to visualise our data.

```{r warning=FALSE, message=FALSE}
simMIX %>% 
  ggplot(aes(Time, Stress, color=Group, fill=Group))  + 
  geom_violin(alpha = .25) + 
  geom_dotplot(binaxis='y', stackdir='center', position=position_dodge(1),
               dotsize = 0.5) +
  scale_x_discrete(limits = c("Pre", "During", "Post")) +
  labs(x="Time", y = "Stress Score", 
       title="Stress over Time by Treat Vs Control", color="Condition", fill="Condition")
```

It is useful to note that the violin plot contains much of the same information that a more traditional boxplot would have, but provides more information on the actual distribution with groups as it draws on density plots.

What we can see here is that over time, the control group shows little change, but the treatment group shows distinct declines.

## Run models
Base model:

```{r}
m0 <- lmer(Stress ~ 1 +
             (1 | ID), data = simMIX)
summary(m0)
```

Main effects: 
```{r}
m1 <- lmer(Stress ~ 1 + Time + Group +
             (1 | ID), data = simMIX)
summary(m1)
```

Interaction:
```{r}
m2 <- lmer(Stress ~ 1 + Time + Group + Time*Group +
             (1 | ID), data = simMIX)
summary(m2)
```

Compare all models
```{r}
anova(m0,m1,m2)
```


As we did with the simple repeated measures, we can also compare the sums of squares breakdown for the LMM (`m2`) by calling `anova()` on the `lmer()` model.

First with `aov()`:

```{r}
m3 <- aov(Stress ~ Time + Group + Time*Group +
            Error(ID/Time), data = simMIX)
summary(m3)
```

And then summarise:

```{r}
anova(m2)
```


## Check model

## Visualise model

## Interpret model
