---
title: "<b>WEEK 5<br>Exploratory Factor Analysis 2</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```

# Learning Objectives
1. Understand and apply criteria for identifying good EFA solutions
2. Apply and interpret factor congruence for testing the replicability of factor solution
3. Estimate different measures of reliability
4. Compute and save factor scores from EFA solutions. 

---
class: inverse, center, middle

<h2>Part 1: Evaluating and modifying a solution</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Factor congruence</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Factor Scores</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reliability</h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity</h2>



---
# Practical Steps
1. Check the appropriateness of the data and decide of the appropriate estimator.
2. Decide which methods to use to select a number of factors.
3. Decide conceptually whether to apply rotation and how to do so.
4. Decide on the criteria to assess and modify your solution.
5. Run the analysis.
6. Evaluate the solution (apply 4)
7. Select a final solution and interpret the model, labelling the factors. 
8. Report your results.

---
# Evaluating results
+ Good idea to start by examining how much variance each factor
  accounts for and the total amount of variance

+ You interpret the meaning of the factors based on the size and the sign of the loadings that you deem to be "salient"
  + What's salient is defined by research question
  + In personality, and most other, research, salient loadings are those $\ge |.3| or |.4|$


---
# Check results
+ Need to also look for signs of trouble

+ Heywood cases
  + If present, something is **wrong**; you should not trust these results
  + Try different rotation, eliminate item, rethink whether factor analysis is right "tool"

+ Are there items that do not have any salient loadings?
  + Could signal a problem item, which should be removed
  + Could signal presence of another factor; can pursue this when revising the questionnaire
  
+ Do some items have multiple salient loadings (cross-loadings)?
  + Indicated by the item complexity values.

+ Do any factors load on only two or three items?
  + Minimum number of items should = 3
  + May have over-extracted
  + May be that you're trying to measure too many things with too few items

---
# Good list of criteria

1. All factors load on 3+ items at salient level.
2. All items have at least one loading above salient cut off.
3. No heywood cases
4. Complex items removed in accordance with goals.
5. Solution accounts for an acceptable level of variance given goals.
6. Item content of factors is coherent and substantively meaningful.

---
# Check results

**Remember**:
If you deleted one or more items, you **must** re-run your factor analysis
starting at trying to determine how many factors you should extract


**Most important is this**: 
If one or more factors don't make sense, then either your items are
bad, your theory is bad, your analysis is bad, or all three are bad!



---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 1

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Evaluating and modifying a solution</h2>
<h2>Part 2: Factor congruence</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Factor Scores</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reliability</h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity</h2>

---
# Replicability
+ After conducting a factor analysis and developing a questionnaire, it's a good idea to test whether it replicates

+ One way to do this is to see whether similar factors appear when similar data are collected
  + Examples: Arguably the Big Five (some caveats)
  + The "positive manifold" of mental abilities 
  
+ Another way is to test this formally by collecting data on another sample.
  + Or split one large sample into two (exploratory vs confirmatory)

+ Based on these two samples, we can then use a number of approaches:
  + Compute **congruence coefficients** between the factors
	+ 'Targeted' rotations that try to rotated one set of factors towards another set
	+ **Confirmatory factor analysis** where you specify what factors load on what items


---
# Congruence Coefficients
+ Congruence coefficients, or Tucker’s Congruence Coefficients, are essentially the correlations between vectors of factor loadings across samples.

$$
r_c = \frac{\Sigma{x_iy_i}}{\sqrt{\sum x_i^2\sum y_i^2}}
$$

+ To calculate congruence:
  + Run the factor model in sample 1.
  + Run the factor model in sample 2
    + Ensuring the same items are included, same number of factors specified etc.
  + Calculate congruence (very simple in R).

---
# Congruence Coefficients R

```{r, warning=FALSE, message=FALSE}
library(psych)
library(tidyverse)
bfi <- na.omit(bfi) #drop missing data for ease

expl <- bfi %>% 
  sample_frac(.5) # randomly select one half

conf <- anti_join(bfi, expl) # select the non-matching cases

res1 <- fa(expl[1:25], nfactors = 5, rotate = "oblimin") # run EFA on expl 
res2 <- fa(conf[1:25], nfactors = 5, rotate = "oblimin") # run same analysis on conf

fa.congruence(res1, res2) # test the congruence

```


---
# Congruence Coefficients
+ Lorenzo-Seva & ten Berge (2006) discuss nice properties of the Tucker coefficient:
  + It measures similarity independent of the mean size of the loadings.
  + It is insensitive to a change in the sign of any pair of loadings.

+ MacCallum et al. (1999) suggest the following criteria following Tucker:
  + < 0.68 = terrible
  + 0.68 to 0.82 = poor
  + 0.82 to 0.92 = borderline
  + 0.92 to 0.98 = good
  + 0.98 to 1.00 = excellent


---

# Confirmatory factor analysis
+ In EFA, all factors load on all items
  + These loadings, as you have seen, are purely data driven
  + However, if we have idea about which items should group, we may want to test this explicitly


+ In CFA, we specify a model and test how well it fits the data
  + We specify a model by indicating what loadings we believe will be zero
  + We then try to reject this model


+ CFA is powerful and can be used for other purposes.
  + Test whether people 'use' a test in the same way across samples
  + Control for measurement error
  + Separate our method and trait variance
  
---
.pull-left[
<center>
# Exploratory
</center>
\begin{bmatrix}
\lambda_{11} & \lambda_{12} \\
\lambda_{21} & \lambda_{22} \\
\lambda_{31} & \lambda_{32} \\
\lambda_{41} & \lambda_{42} \\
\lambda_{51} & \lambda_{52} \\
\lambda_{61} & \lambda_{62} \\
\lambda_{71} & \lambda_{72} \\
\lambda_{81} & \lambda_{82}
\end{bmatrix}
]

.pull-right[
<center>
# Confirmatory
</center>
\begin{bmatrix}
0 & \lambda_{12} \\
0 & \lambda_{22} \\
0 & \lambda_{32} \\
0 & \lambda_{42} \\
\lambda_{51} & 0 \\
\lambda_{61} & 0 \\
\lambda_{71} & 0 \\
\lambda_{81} & 0
\end{bmatrix}
]

+ If the correlations derived from the CFA don't match the correlations from your data, it *probably* means that parameters set to zero should not have been set to zero

---

# Confirmatory factor analysis
+ Requires strong theory

+ Requires that the constructs you're studying exhibit simple structure, i.e., *not* personality
  
+ Never conduct a CFA on the same data you used for your EFA
  + Inflates the Type I error rate (as other statistical methods)

+ If you have an interest in CFA, feel free to email me.

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 2

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Evaluating and modifying a solution</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Factor congruence</h2>
<h2>Part 3: Factor Scores</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reliability</h2>
<h2 style="text-align: left;opacity:0.3;">Part 5:  Validity</h2>


---
# Factor scores

+ Sometimes EFA is an end in and of itself: research question may concern the structure of a set of items

+ However, you usually want to "do something" with your factors
  + Test whether your construct is related to other constructs in ways that you would predict
  + Test whether your construct is related systematically to other variables, including those that you manipulate
 
+ This pursuit was key to the resurgence of trait theory

+ To do these things, you need variables that represent what you've measured

---
# Factor scores
+ Unit-weighting
  + Sum raw scores on the observed variables which have primary loadings on each factor
  + Which items to sum is a matter of defining what loadings are salient
  + Need to reverse score items with negative loadings

+ Regression-based methods
  + Thurstone or Thompson method
	  + Ordinary least squares approach
	  + Computes scores from observed item correlations and loadings
   + Bartlett method focuses on minimizing sums of squares for unique factors
   
+ How do they fare?
  + Simple, produce estimates that are highly correlated with more exact techniques (especially true for unit-weighting)
  + Negative consequence of unit-weighting or regression method is they can produce correlated scores for orthogonal factors

---

# Factor scores
+ Anderson-Rubin and Ten Berge methods
  + Negative consequence of unit-weighting or regression method is they can produce correlated scores for orthogonal factors
  + The Anderson-Rubin method preserves orthogonality of factors
  + The Ten Berge method generalizes the Anderson-Rubin method to preserve correlations (or lack thereof) between factors

---

# Choosing factor scores
+ Simple sum scores (unit-weighting) require strict properties in the data, but these are rarely tested, and do not often hold

+ If the goal is to try to find higher-order factors, the correlations between the scores are important, and so ten Berge scores are preferable
  + Not doing this can lead to really biased results
  + An alternative is to factor the Phi matrix

+ Alternative is to use structural equations modeling, which includes a measurement component (a CFA) and a structural component (regression)
  + Doesn't require you to compute factor scores
  + Requires good theory of measurement and structure
  + If your constructs don't approximate simple structure, you may have to turn to other alternatives
  
---

# Sample size
+ In the past, rules of thumb have guided sample size decisions for factor analysis (reviewed by MacCallum et al., 1999)
  + Rules based on minimum number of participants: 100, 200, 250, 500
  + Rules based on the participant-to-item (*N*:*p*) ratio: 3 to 6, 5, 10


+ MacCallum et al. (1999) tested what was important using simulated datasets
  + Tested how well sample sizes of 60, 100, 200, and 400 would recover factors from each of nine population correlation 
  + The populations varied in two ways
	  + Items to factor (*p*:*m*) ratio (10:3, 20:7, 20:3)
	  + Communalities: low (.2, .3, .4), wide (.2 to .8 in .1 increments), high (.6, .7, .8)

---
# Sample size
+ As MacCallum et al. hypothesized, the crucial determinants of minimum sample size was not the number of items, but communalities and *p*:*m*
	  + Fewer subjects were needed if communalities were wide or high
	  + Fewer subjects were needed if *p*:*m* was high
	  + Communalities were even more important when *p*:*m* = 20:7 (an interaction effect)

+ Subsequent studies support these findings; all reject that the
  *N*:*p* ratio should be used

---
# Sample size
+ Thus, when planning a study you should do the following to determine your minimum sample size
  + Think of how many factors you expect and get many items measuring each
  + Use pilot data and previous studies to make an "educated guess" about what communalities you're going to expect
  

---

# GIGO
+ Make sure to check the quality of your data

+ PCA and factor analysis cannot turn bad data into good data

+ 'Garbage in, garbage out'


---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 3

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Evaluating and modifying a solution</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Factor congruence</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Reliability</h2>
<h2>Part 4: Reliability</h2>
<h2 style="text-align: left;opacity:0.3;">Part 5: Validity</h2>


---

# Acknowledgement
+ Slides *heavily* influenced by work of Bill Revelle and his [book on
  psychometrics](http://www.personality-project.org/r/book/)

---

# Measurement
+ The aim of measurement is to develop and use measures of constructs to test psychological theories
  
  
+ 'Classical test theory' describes data from any measure as a combination of
  + The signal of the construct, or the 'true score'
  + Noise or 'error', that is, measure of other, unintended things
  
$$\text{Observed score} = \text{True score} + \text{Error}$$


+ This should remind you of the factor analysis formula that you learned in the last lecture
  
---

# True score theory
+ If we assume of our test that:
	1. It measures some ability or trait
	2. In the world, there is a "true" value or score for this test
       for each individual


+ Then the reliability of the test is a measure of how well it
  reflects the true score
  
---
# Parallel tests
+ Charles Spearman was the first to note that, under certain assumptions (parallelism, tau equivalence, congeneric tests), the correlations between two **parallel tests** of the same construct provided an estimate of reliability

+ Where do parallel tests come from?
  + Previous, older definitions of "parallel tests" were somewhat abstract
  
+ Parallel tests can come from several sources
  + Items (alternate forms, split-half, internal consistency)
  + Time tests were administered (test-retest)
  + Raters

---
# Alternate forms reliability
+ Correlation between two variants of a test
  + Same items in different order (randomization of stimuli)
  + Tests with similar, but not identical, content, e.g., tests with a fixed number of basic addition, subtraction, division, and multiplication problems
  + Ideally, alternate tests have equal means and variances

+ Assumption is that, if the tests are perfectly reliable, then they should correlate perfectly
  + They won't
  + To the extent that they don't, we have a measure of reliability
  
---
# Alternate forms reliability
+ Alternate forms can be an expensive (time, and money) way to assess reliability, but is probably a gold standard.

+ Developing alternate forms is becoming much easier:

  1. Write a large item bank
  2. Get many respondents
  3. Use statistical tools (item-response theory) to assess how "difficult" items are,
  4. Create your tests with items matched on difficulty

---
# Split-half reliability
+ This measure of reliable indicates how internally consistent the test is
  1. Split test into a pair of equal subsets of $n$ items
  2. Score the two subsets
  3. Correlate these scores
  
+ With an increasing number of items, the number of possible splits gets very large, the relationship is:

$$\frac{n!}{2\left(\frac{n}{2}\right)!^{2}}$$


---

# Cronbach's alpha
+ If we take the idea of correlating subsets of items to its logical conclusion
  + Split-half reliability is a special case of reliability among all test items
  + The best known estimate of this is formula 2 from Cronbach (1951)
  
$$\alpha=\frac{n}{n-1}\left(1-\frac{\sum\limits_{i=1}^n V_{i}}{V_{t}}\right)$$

+ $V_{t}$ is the variance of test scores (total variance)

+ $V_{i}$ is the variance of the $n$ item scores after they've been weighted (error variance)

---

# Cronbach's alpha
+ It does **not** indicate whether items measure one unidimensional construct
  + This is clear when one recognizes that Cronbach's alpha increases as you add items, so it can be high even if there's no underlying factor
  + Known as 'Spearman-Brown prophecy formula'
  
$$\rho^{*}_{xx'}=\frac{n\rho_{xx'}}{1+(n-1)\rho_{xx'}}$$

+ $\rho^{*}_{xx'}$ is the predicted (prophesized) Cronbach's alpha

+ $\rho_{xx'}$ is the original Cronbach's alpha

+ $n$ ratio of number of new to old measures, so, e.g., if $n=2$, the new test has twice as many items; can also be a fraction

---
# McDonald's omega
+ Any item may measure
  + A "general" factors that load on all items
  + A "group" or "specific" factor that loads on a subset of items

+ Given this, we can derive two internal consistency measures
  + Omega hierarchical $(\omega_{h})$, the proportion of item variance that is general
  + Omega total $(\omega_{t})$, the total proportion of reliable item variance

+ Estimating these values
  + Use `omega` function in the psych package
  + Use CFA and compute it manually
	

---
# Test-retest reliability
+ Correlation between tests taken at 2+ points in time (think again about parallel, tau equivalent, and congeneric reliability)

+ Corner-stone of test assessment and appears in many test manuals, but poses some tricky questions
  + What's the appropriate time between when measures are taken?
  + How stable should the construct be if we are to consider it a trait?

+ **Remember**: Even if you have within-individual changes in mean scores, so long as rank ordering is consistent, correlations can stay high


---
# Interrater reliability
+ Ask a set of judges to rate a set of targets
  + Get friends to rate the personality of a family member
  + Get zoo keepers to rate the subjective well-being of an animal
  

+ We can determine how consistent raters are by means of intraclass correlation coefficients
  + How reliable are their individual estimates
  + How reliable is the average estimate based on the judges' ratings

---

# Intraclass correlations
+ Splits variance of a set of ratings into multiple components
  + Variance between subjects (across targets)
  + Variance within subjects (across raters, same target)
  + Variance due to raters (across targets, same rater)


+ Depending on what we want to know and the design of our study, we
  can calculate intraclass correlations from these variance components
  
---

# Intraclass correlations

| ICC | Description                                                                                                                                         |
|:----|:----------------------------------------------------------------------------------------------------------------------------------------------------|
| 1,1 | Targets rated by a different set of randomly selected raters and reliability is based on one measurement                                            |
| 1,k | As (1,1), but with reliability calculated as the average of k raters’ measurements                                                                  |
| 2,1 | Each target measured by each rater. Raters are considered representative of larger pool of raters. Reliability calculated from a single measurement |
| 2,k | As (2,1), but with reliability calculated as the average of k raters’ measurements                                                                  |
| 3,1 | Each target measured by each rater. Raters only raters of interest. Reliability calculated from a single measurement                                |
| 3,k | As (3,1), but with reliability calculated as the average of k raters’ measurements                                                                  |
<p align='right'>Shrout and Fleiss (1979)</p>


---

# Uses of reliability
+ Good to know how reliable a measure is
  + Implications for validity (will discuss it soon)
  + Also allows us to 'correct for attenuation'
  
$$r^{*}_{xy}=\frac{r_{xy}}{\sqrt{\rho^{2}_{\theta x}\rho^{2}_{\theta y}}}$$

$r^{*}_{xy}$ is the correlation between $x$ and $y$ after correcting for attenuation

$r_{xy}$ is the correlation before correcting for attenuation

$\rho^{2}_{\theta x}$ is the reliability of $x$

$\rho^{2}_{\theta y}$ is the reliability of $y$

---
# Reliability in R

| Reliability over | Estimate        | Functions in R |
|:-----------------|:----------------|:---------------|
| Forms            | Alternate forms | cor            |
| Time             | Test-retest     | cor            |
|                  |                 | rptR           |
| Split-half       | Random split    | splitHalf      |
|                  | Worst split     | splitHalf      |
|                  | Best split      | splitHalf      |
| Items            | General factor  | omega          |
|                  | Average         | alpha          |
| Raters           | All variants    | ICC            |


---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 4

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Evaluating and modifying a solution</h2>
<h2 style="text-align: left;opacity:0.3;">Factor congruence</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Factor scores</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Reliability</h2>
<h2>Part 5: Validity</h2>


---

# Validity
> Validity refers to the degree to which evidence and theory support
> the interpretations of test scores for proposed uses of
> tests. Validity is, therefore, the most fundamental consideration in
> developing tests and evaluating tests. The process of validation
> involves accumulating relevant evidence to provide a sound scientific
> basis for the proposed score interpretations. It is the
> interpretations of the test scores for the proposed uses that are
> valuated, not the test itself.
<p align='right'>Standard for Educational and Psychological Testing</p>

---

# Debates about the definition
*[W]hether a test really measures what it purports to measure* (Kelley,
1927)

*[H]ow well a test does the job it is employed to do. The same may be
used for ... different purposes and its validity may be high for
one, moderate for another and low for a third* (Cureton, 1951)

*Validity is "an integrated evaluative judgment of the degree to
which empirical evidence and theoretical rationales support the
adequacy and appropriateness of inferences and actions based on test
scores or other modes of assessment* (Messick, 1989)

*A test is valid for measuring an attribute if (a) the attribute
exists and (b) variations in the attribute causally produce
variation in the measurement outcomes* (Borsboon et al., 2004)

*[V]alidity means that the information yielded by a test is
appropriate, meaningful, and useful for decision making -- the purpose
of mental measurement* (Osterlind, 2010)

---

# Evidence for validity
+ Debates about how to define validity lead to questions about what
  constitutes evidence for validity
  

+ Sources of evidence align to what may be viewed as "classical"
  concepts reported in textbooks, studies, and test manuals

---

# Evidence related to content
+ **Content validity**
  + A test should contain only content relevant to the intended construct
  + It should measure what it was intended to measure
  
  
+ **Face validity**
  + i.e., for those taking the test, does the test "appear to" measure what it was designed to measure?

---

# Evidence related to the scale
+ Do the items measure a single "intended" construct?


+ Factor analysis only provides very limited information towards this


+ How else might we assess it?

--

  + Construct validity, or convergent and discriminant validity (Cronbach & Meehl, 1955)
  
---

# Relationships with other constructs
+ **Convergent**: 
  + Measure should have high correlations with other measures of the same construct

+ **Discriminant**: 
  + Measure should have low correlations with measures of different constructs

+ **Nomological Net**
  + Measure should have expected patterns (positive/negative) correlations with different sets of constructs
	 + Also, some measures should vary depending on manipulations, 
	 + e.g., a measure of "stress" should be higher among students who are told that a test is "high stakes" than among students told that a test is "low stakes"
	 
---

# Relationships with other constructs
+ Consider relations in terms of temporal sequence.

+ **Concurrent validity**: Correlations with contemporaneous measures
  + Neuroticism and subjective well-being
  + Extraversion and leadership

+ **Predictive validity**: Related to expected future outcomes
  + IQ and health
  + Agreeableness and future income

---

# Evidence related to response processes
+ Discussed in a paper by Karabenick et al. (2007)

+ Not commonly considered in validation studies
  + Is how people "process" the items belonging to a scale the way we think they ought to?
	+ Do tests of intelligence engage problem-solving behaviors?
	+ Do extraversion items lead people to reflect on related past behaviors?

> Karabenick, S. A., Woolley, M. E., Friedel, J. M., Ammon, B. V., Blazevski, J., Bonney, C. R., ... & Kelly, K. L. (2007). Cognitive processing of self-report items in educational research: Do they think what we mean?. Educational Psychologist, 42(3), 139-151.

---

# Evidence related to consequences
+ Perhaps most controversial aspect of current validity discussions


+ Should potential consequences of test use be considered part of the
  evidence for test's validity?


+ Important questions for the use of tests
  + Is my measure systematically biased or fair for all groups of test
  takers?
  + Does bias have social ramifications?

---
# Relationship between reliability and validity
+ Reliability: relation of true score with observed score

+ Validity: correlations with other measures play a key role

+ Logically, a score or measure cannot correlate with anything more than it correlates with itself, so reliability is the limit on validity

---
# Importance of test reliability and validity
+ Fundamental first step in measurement
  + If we cannot measure well variables of interest, then we cannot study them
  + Large, tricky problem in psychology; many variables/constructs are not directly accessible


+ Important for later research
  + Poor reliability and validity may lead to erroneous conclusions due to measurement problems
  + If we know reliability of test, we can sometimes make adjustments (correction for attenuation)
  + Not "glamorous" research, but can be interesting in and of itself

---
# Where can you find this information?
+ Test manuals


+ Papers describing new tests and papers investigating exisitng
  measures in different groups, languages, contexts, etc.
  + *Assessment*
  + *Psychological Assessment*
  + *European Journal of Psychological Assessment*
  + *Organisational Research Methods*
  + Personality journals
  
  
+ Papers describing new ways to establish reliability, validity, etc.
  + *Methdology*
  + *Psychometrika*
  + *Journal of Educational Statistics*
  
  
---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# End

