---
title: "<b>WEEK 4<br>Exploratory Factor Analysis 1</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```

# Learning Objectives
1. Understand the conceptual difference between PCA and EFA
2. Understand the strengths and weaknesses of different methods to estimate the optimal number of factors
3. Understand the purpose of factor rotation and the difference between orthogonal and oblique rotation.
4. Run and interpret EFA analyses in R.

---
class: inverse, center, middle

<h2>Part 1: Introduction to EFA</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Number of factors problem</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Factor rotation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Example and interpretation</h2>
---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 1

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to EFA</h2>
<h2>Part 2: Number of factors problem</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Factor rotation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Example and interpretation</h2>

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 2

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to EFA</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Number of factors problem</h2>
<h2>Part 3: Factor rotation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Example and interpretation</h2>

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 3

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction to EFA</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Number of factors problem</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Factor rotation</h2>
<h2>Part 4: Example and interpretation</h2>

---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# End

