---
title: "<b>WEEK 1<br>Path Analysis</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```

# Weeks 7 to 11 Overview
- Section introduction (w7)
- Path analysis (w7)
- Path mediation (w8)
- Data Reduction:
  - Principal Components Analysis (w9)
  - Exploratory Factor Analysis (w10 & 11)
- Where next? (w11)

---
# Learning Objectives
1. Understand the core principle of path modelling
2. Be able to use `lavaan` to estimate linear models and simple path models
3. Interpret the output from a `lavaan` model.

---
class: inverse, center, middle

<h2>Part 1: Introduction and Motivation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Introducing `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Model Estimation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Model Evaluation</h2>
---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 1

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction and Motivation</h2>
<h2>Part 2: Introducing `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Model Estimation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Model Evaluation</h2>

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 2

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction and Motivation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Introducing `lavaan`</h2>
<h2>Part 3: Model Estimation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Model Evaluation</h2>

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 3

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction and Motivation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Introducing `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Model Estimation</h2>
<h2>Part 4: Model Evaluation</h2>

---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# End

