---
title: "<b>WEEK 1<br>Path Analysis</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```

# Learning Objectives
1. Understand the core principle of path modelling
2. Be able to use `lavaan` to estimate linear models and simple path models
3. Interpret the output from a `lavaan` model.

---
class: inverse, center, middle

<h2>Part 1: Introduction and Motivation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Introducing `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Model Estimation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Model Evaluation</h2>

---
# Weeks 7 to 11 Overview
- Section introduction (w7)
- Path analysis (w7)
- Path mediation (w8)
- Data Reduction:
  - Principal Components Analysis (w9)
  - Exploratory Factor Analysis (w10 & 11)
- Where next? (w11)

---
# What issue do these methods solve?
- Path models
  - Sometimes we have more than one variable that needs to be treated as an outcome/dependent variable
  - We cant do this in a linear model.
  - **A path model allows us to test several linear models together as a set**
  - Good way to learn basics of *structural equation modelling*
  
- Data reduction
  - Psychology uses many surveys and psychometric tools
  - Here we asked lots of questions we believe relate to some construct
  - We need a way to:
    - Check the relationships between each question
    - Produce plausible scores that represent this construct

- We will start with path models...

---
# Diagramatic Conventions

.pull-left[

- Some conventions
  - Square = observed/measured
  - Circle/elipse = latent/unobserved
  - Two-headed arrow = covariance
  - Single headed arrow = regression path

]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("Conventions.pdf",pages = 1)
```
]

---
# Terminology
- A couple of distinctions are also useful here.

- Broadly, variables can be categorised as either exogenous or endogenous.

- **Exogenous:** are essentially independent variables. 
    - Only have directed arrows going out.

- **Endogenous:** are dependent variables in at least one part of the model. 
    - They have directed arrows going in.
    - In a linear model there is only one endogenous variable, but in a path model we can have multiple.
    - They also have an associated residual variance.
      - Just like in a `lm`
      - If something predicts (explains variance) a variable, there will be something left unexplained

- So how does this relate to practical research problem?

---
# A hypothetical example
- Suppose we are interested in how Neuroticism predicts psychological well-being and physical health outcomes.
    - Neuroticism measured by a questionnaire with 5 items (5-point scale).
    - Well-being is measured by a questionnaire with 5 items (7-point scale).
    - Physical health is measured based on BMI, V02 max, and presence or absence of cancer (binary).

- How do we test our model?

---
# A hypothetical example

.pull-left[
- **Approach 1**: Aggregate everything into composite scores and use 2 regression models.

]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("Simple1.pdf",pages = 1)

```

```{r echo=FALSE}
magick::image_read_pdf("Simple2.pdf",pages = 1)
```
]

---
# A hypothetical example

.pull-left[
- **Approach 2**: Aggregate everything, and use a path model to simultaneously estimate model with 2 outcomes.

]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("TwoDV.pdf",pages = 1)
```
]

---
# A hypothetical example

.pull-left[
- **Approach 3**: Use a mix of latent and composite variables to simultaneously estimate model with 2 outcomes.
]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("SEM1.pdf",pages = 1)
```
]



---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 1

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction and Motivation</h2>
<h2>Part 2: Introducing `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Model Estimation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Model Evaluation</h2>

---
# `lavaan`
- The package we will use to fit our path models is called `lavaan`.

- Using `lavaan` requires us to write...
  - code to specify our model
  - code to run the model

- This is because we (a) generally use path models for slightly bigger models, and (b) have lots more options when running our model than in a `lm` 

---
# Model statements in `lavaan`

.pull-left[

- When a variable is observed, we use the name it has in our data set.
  - Here `X1`, `X2`, `X3`, `Y1`

- When a variable is latent (more on this in the data reduction section) we give it a new name.
  - Here `Latent`

- To specify a covariance, we use `~~`

- To specify a regression path, we use `~`

]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("Conventions.pdf",pages = 1)
```
]


---
# `lavaan` model code: Approach 1

.pull-left[
- **Approach 1**: Aggregate everything into composite scores and use 2 regression models.

```{r, eval=FALSE}
a1a = '
wb ~ n
'
```

```{r, eval=FALSE}
a1b = '
health ~ n
'
```

- But this gains us nothing

]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("Simple1.pdf",pages = 1)

```

```{r echo=FALSE}
magick::image_read_pdf("Simple2.pdf",pages = 1)
```
]

---
# `lavaan` model code: Approach 2

.pull-left[
- **Approach 2**: Aggregate everything, and use a path model to simultaneously estimate model with 2 outcomes.

```{r, eval=FALSE}
a2 = '
wb ~ n
health ~ n
wb ~~ health
'
```


]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("TwoDV.pdf",pages = 1)
```
]

---
# `lavaan` model code: Approach 1

.pull-left[
- **Approach 3**: Use a mix of latent and composite variables to simultaneously estimate model with 2 outcomes.
  - This one is just for fun (we wont get this far on this course)
  - I will include some examples of this in the final week for anyone interested

```{r, eval=FALSE}
a3 <- '
n =~ n1 + n2 + n3 + n4 + n5
wb =~ wb1 + wb2 + wb3 + wb4 + wb5
health ~ n
wb ~ n
'
```

]

.pull-right[
```{r echo=FALSE}
magick::image_read_pdf("SEM1.pdf",pages = 1)
```
]

---
# Running a `lavaan` model
- Once we have our model statement, we then need to run our model.
  - There are a number of functions to do this, we will only use `sem()`

```{r, eval=FALSE}
library(lavaan)
m1 <- sem(model, # your model statement
          orderd = c(), # if variables are ordered categories list them
          estimator ="ml" , # name of the estimation method you wish to use
          missing = "" , # name of the missing data method you wish to use
          data = tbl) # your data set
```

- `lavaan` has sensible defaults, meaning most of the time you will only need to state you model and data.

- There is **lots** of information on using `lavaan` with lots of examples [on-line](https://lavaan.ugent.be/)

---
# Viewing the results
- Lastly, we need to use a `summary` function (like in `lm` and `glm`) to see results. 

```{r, eval=FALSE}
summary(m1, # name given to our results object
        fit.measures = T, # model fit information
        standardized = T # provides standardized coefficients
        )
```


---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 2

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction and Motivation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Introducing `lavaan`</h2>
<h2>Part 3: Model Estimation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 4: Model Evaluation</h2>

---
class: inverse, center, middle, animated, rotateInDownLeft

# End of Part 3

---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Introduction and Motivation</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Introducing `lavaan`</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Model Estimation</h2>
<h2>Part 4: Model Evaluation</h2>

---
class: extra, inverse, center, middle, animated, rotateInDownLeft

# End

