---
title: "Data for Lab 04"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(HLMdiag)
```

all data drawn from <https://uoepsy.github.io/lmm/00_datasets>


# clothing

```{r}
clothing <- read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")
glimpse(clothing)
```


## get model to converge

```{r}
clothing_m1 <- lmer(
  purch_rating ~ price * condition + (1 + price | ppt), 
  data = clothing
)
```

> Show in New Window
Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
  Model failed to converge with max|grad| = 8.80186 (tol = 0.002, component 1)
  See ?lme4::convergence and ?lme4::troubleshooting.
Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
  Model is nearly unidentifiable: very large eigenvalue
 - Rescale variables?
 
Try diff optimiser: still wants things rescaled.
 
```{r}
clothing_m2 <- lmer(
  purch_rating ~ price * condition + (1 + price | ppt), 
  data = clothing,
  control = lmerControl(optimizer = "bobyqa")  # add in this line
)
```
 
 > Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
  Model is nearly unidentifiable: very large eigenvalue
 - Rescale variables?
 
 

```{r}
clothing <- clothing |>
  mutate(
    price_z = (price - mean(price)) / sd(price)
  )

# check that z-scoring worked the way we want
clothing$price_z |> mean()
clothing$price_z |> sd()
```


```{r}
clothing_m3 <- lmer(
  purch_rating ~ price_z * condition + (1 + price_z | ppt), 
  data = clothing
)

summary(clothing_m3)
```


## assump

### Linearity of association

```{r}
plot(clothing_m3, 
     type=c(
       "p",      # includes points representing Pearson residuals
       "smooth"  # includes smoothed line representing mean of residuals
       )
    )
```

line is flat = looks good and linear



### Indep of errors

we fit the model with the maximal ranef structure so we're good to go


### Normality of errors

```{r}
qqnorm(resid(clothing_m3))
qqline(resid(clothing_m3))
```

great

### Equal variance of errors

```{r}
plot(clothing_m3, 
     type=c(
       "p",      # includes points representing Pearson residuals
       "smooth"  # includes smoothed line representing mean of residuals
       )
    )
```

nice even cloud, hurray


### Normality of random effects

#### by-ppt intercept adjustments

```{r}
qqnorm(ranef(clothing_m3)$ppt[,1]); qqline(ranef(clothing_m3)$ppt[,1])
```

some divergence at extremes but ok


#### by-ppt slope over `price_z` adjustments

```{r}
qqnorm(ranef(clothing_m3)$ppt[,2]); qqline(ranef(clothing_m3)$ppt[,2])
```

good match


## influence diags

### influence of each observation

```{r}
inf_obs <- hlm_influence(model = clothing_m3, level = 1, approx = TRUE)
```

```{r}
dotplot_diag(inf_obs$cooksd, cutoff = "internal")
```

```{r}
inf_obs[561,]
```

```{r}
clothing[561,]

clothing[-561,]
```



#### sensitivity analysis


```{r}
clothing_m3_no561 <- lmer(
  purch_rating ~ price_z * condition + (1 + price_z | ppt),
  data = clothing[-561,]
)
```

```{r}
summary(clothing_m3)
```

```{r}
summary(clothing_m3_no561)
```


### influence of each ppt

```{r}
inf_ppt <- hlm_influence(model = clothing_m3, level = 'ppt', approx = TRUE)
```

```{r}
dotplot_diag(inf_ppt$cooksd, index = inf_ppt$ppt, cutoff = "internal")
```




# monkeystatus - GLM CHALLENGE


```{r}
monkeystatus <- read_csv('https://uoepsy.github.io/data/msmr_monkeystatus.csv')
monkeystatus |>
  glimpse()
```

`solved ~ status + difficulty + (1 + difficulty | monkeyID)`


### model

```{r eval=F}
monkeystatus_m1 <- glmer(
  solved ~ status + difficulty + (1 + difficulty | monkeyID),
  data = monkeystatus, 
  family = binomial
)
summary(monkeystatus_m1)
```

