---
title: "Data for Lab 01"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)
library(lme4)
```

all data drawn from <https://uoepsy.github.io/lmm/00_datasets>


## clothing

```{r}
clothing <- read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")
glimpse(clothing)
```

From last lab, we know that the grouping variables in `clothing` and the kind of variability they contribute are as follows:

- `price` contributes reproducible variability (manipulated as part of RQ)
- `ppt` contributes random variability
- `condition` contributes reproducible variability (manipulated as part of RQ)





### model

```{r eval=F}
#scale price to help convergence
#change 1 in price is now change of £10
df$price <- df$price/10

mod1 <- lmer(purch_rating ~ price*condition + 
       (1+price|ppt), 
       data = df)
  
summary(mod1)
```


## monkeystatus


```{r}
monkeystatus <- read_csv('https://uoepsy.github.io/data/msmr_monkeystatus.csv')
monkeystatus |>
  glimpse()
```

### status -- rep

```{r}
monkeystatus |>
  group_by(status) |>
  count()
```

^ grouping variable

```{r}
monkeystatus |>
  group_by(status) |>
  n_groups()
```

Three groups!

### difficulty -- rep (ctrl)

```{r}
monkeystatus |>
  group_by(difficulty) |>
  count()
```

^ grouping variable

```{r}
monkeystatus |>
  group_by(difficulty) |>
  n_groups()
```

### monkeyID -- rd

```{r}
monkeystatus |>
  group_by(monkeyID) |>
  count()
```

^ grouping variable

```{r}
monkeystatus |>
  group_by(monkeyID) |>
  n_groups()
```


### model

```{r eval=F}
mmod <- lme4::glmer(solved ~ difficulty + status + 
      (1 + difficulty | monkeyID),
      data = monkeystatus, family = binomial)
summary(mmod)
```


## laughs

```{r}
laughs <- read_csv('https://uoepsy.github.io/data/lmm_laughs.csv')
```


```{r}
glimpse(laughs)
```

### ppt -- rd

```{r}
laughs |>
  group_by(ppt) |>
  count()

laughs |>
  group_by(ppt) |>
  n_groups()
```


^ random variability


### joke_label -- rd

```{r}
laughs |>
  group_by(joke_label) |>
  count()

laughs |>
  group_by(joke_label) |>
  n_groups()
```

^ random variability

### joke_id -- rd

```{r}
laughs |>
  group_by(joke_id) |>
  count()

laughs |>
  group_by(joke_id) |>
  n_groups()
```

^ random variability

### delivery -- rep

```{r}
laughs |>
  group_by(delivery) |>
  count()

laughs |>
  group_by(delivery) |>
  n_groups()
```

^ reproducible variability

### model

```{r eval=F}
mod <- lmer(rating ~ delivery + 
              (1 + delivery | joke_id) +
              (1 + delivery | ppt), data = laughs)
```



