---
title: "Data for Lab 01"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)
```

all data drawn from <https://uoepsy.github.io/lmm/00_datasets>


## clothing

<!-- i like this one but never used it. it's based on a UG dissertation project from ages ago. fairly bog standard experiment with just person level ranefs -->

<!-- **Does clothing seem more attractive to shoppers when it is viewed on a model, and is this dependent on item price?** 30 participants were presented with a set of pictures of items of clothing, and rated each item how likely they were to buy it. Each participant saw 20 items, ranging in price from £5 to £100. 15 participants saw these items worn by a model, while the other 15 saw the items against a white background.   -->

<!-- Data are available at [https://uoepsy.github.io/data/dapr3_mannequin.csv](https://uoepsy.github.io/data/dapr3_mannequin.csv) -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- d3 <- read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv") -->
<!-- tibble( -->
<!--   variable=names(d3), -->
<!--   description=c("Purchase Rating (sliding scale 0 to 100, with higher ratings indicating greater perceived likelihood of purchase)", -->
<!--                 "Price presented with item (range £5 to £100)", -->
<!--                 "Participant Identifier", -->
<!--                 "Whether items are seen on a model or on a white background") -->
<!-- ) |> gt::gt() -->

<!-- ``` -->


```{r}
mannequin <- read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")
glimpse(mannequin)
```

### price -- rep

```{r}
mannequin |> group_by(price) |> count()
mannequin |> group_by(price) |> n_groups()
```

### ppt -- rd

```{r}
mannequin |> group_by(ppt) |> count()
mannequin |> group_by(ppt) |> n_groups()
```



### model

```{r eval=F}
#scale price to help convergence
#change 1 in price is now change of £10
df$price <- df$price/10

mod1 <- lmer(purch_rating ~ price*condition + 
       (1+price|ppt), 
       data = df)
  
summary(mod1)
```


## monkeystatus

<!-- Our primate researchers have been busy collecting more data. They have given a sample of Rhesus Macaques various problems to solve in order to receive treats. Troops of Macaques have a complex social structure, but adult monkeys tend can be loosely categorised as having either a "dominant" or "subordinate" status. The monkeys in our sample are either adolescent monkeys, subordinate adults, or dominant adults. Each monkey attempted various problems before they got bored/distracted/full of treats. Each problems were classed as either "easy" or "difficult", and the researchers recorded whether or not the monkey solved each problem.  -->

<!-- We're interested in **how the social status of monkeys is associated with the ability to solve problems, while controlling for the difficulty of the problem.** -->

<!-- Data are available at [https://uoepsy.github.io/data/msmr_monkeystatus.csv](https://uoepsy.github.io/data/msmr_monkeystatus.csv){target="_blank"}.   -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- tibble(variable= -->
<!--          names(read_csv("https://uoepsy.github.io/data/msmr_monkeystatus.csv")), -->
<!--        description = c( -->
<!--          "Social Status of monkey (adolescent, subordinate adult, or dominant adult)", -->
<!--          "Problem difficulty ('easy' vs 'difficult')", -->
<!--          "Monkey Name", -->
<!--          "Whether or not the problem was successfully solved by the monkey") -->
<!-- ) |> gt::gt() -->
<!-- ``` -->


```{r}
monkeystatus <- read_csv('https://uoepsy.github.io/data/msmr_monkeystatus.csv')
monkeystatus |>
  glimpse()
```

### status -- rep

```{r}
monkeystatus |>
  group_by(status) |>
  count()
```

^ grouping variable

```{r}
monkeystatus |>
  group_by(status) |>
  n_groups()
```

Three groups!

### difficulty -- rep (ctrl)

```{r}
monkeystatus |>
  group_by(difficulty) |>
  count()
```

^ grouping variable

```{r}
monkeystatus |>
  group_by(difficulty) |>
  n_groups()
```

### monkeyID -- rd

```{r}
monkeystatus |>
  group_by(monkeyID) |>
  count()
```

^ grouping variable

```{r}
monkeystatus |>
  group_by(monkeyID) |>
  n_groups()
```


### model

```{r eval=F}
mmod <- lme4::glmer(solved ~ difficulty + status + 
      (1 + difficulty | monkeyID),
      data = monkeystatus, family = binomial)
summary(mmod)
```


## laughs

```{r}
laughs <- read_csv('https://uoepsy.github.io/data/lmm_laughs.csv')
```


```{r}
glimpse(laughs)
```

### ppt -- rd

```{r}
laughs |>
  group_by(ppt) |>
  count()

laughs |>
  group_by(ppt) |>
  n_groups()
```


^ random variability


### joke_label -- rd

```{r}
laughs |>
  group_by(joke_label) |>
  count()

laughs |>
  group_by(joke_label) |>
  n_groups()
```

^ random variability

### joke_id -- rd

```{r}
laughs |>
  group_by(joke_id) |>
  count()

laughs |>
  group_by(joke_id) |>
  n_groups()
```

^ random variability

### delivery -- rep

```{r}
laughs |>
  group_by(delivery) |>
  count()

laughs |>
  group_by(delivery) |>
  n_groups()
```

^ reproducible variability

### model

```{r eval=F}
mod <- lmer(rating ~ delivery + 
              (1 + delivery | joke_id) +
              (1 + delivery | ppt), data = laughs)
```





-----

## OLD: lmm_mindfulmindful (last – some vars not relev)

this is typically my go to for longitudinal example. cog ~ age*condition + (1+age|person)


<!-- A study is interested in examining **whether engaging in mindfulness can prevent cognitive mindful in older adults**. They recruit a sample of 20 participants at age 60, and administer the Addenbrooke's Cognitive Examination (ACE) every 2 years (until participants were aged 78). Half of the participants complete weekly mindfulness sessions, while the remaining participants did not.    -->


<!-- Data are available at [https://uoepsy.github.io/data/lmm_mindfulmindful.csv](https://uoepsy.github.io/data/lmm_mindfuldecline.csv){target="_blank"}.   -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- mmd <- read_csv("https://uoepsy.github.io/data/lmm_mindfuldecline.csv") -->
<!-- tibble( -->
<!--   variable=names(mmd), -->
<!--   description=c( -->
<!--     "Site Identifier", -->
<!--     "Participant Identifier", -->
<!--                 "Whether the participant engages in mindfulness or not (control/mindfulness)", -->
<!--                 "Study Visit Number (1 - 10)", -->
<!--                 "Age (in years) at study visit", -->
<!--                 "Addenbrooke's Cognitive Examination Score. Scores can range from 0 to 100", -->
<!--                 "Clinical diagnosis of cognitive impairment ('imp' = impaired, 'unimp' = unimpaired)") -->
<!-- ) |> gt::gt() -->
<!-- ``` -->

```{r}
library(tidyverse)
mindful <- read_csv('https://uoepsy.github.io/data/lmm_mindful.csv')
glimpse(mindful)
```


### sitename -- not grouping var

```{r}
mindful |> group_by(sitename) |> count()
```

excellent, only one value. so not a grouping variable.


### ppt -- rd

```{r}
mindful |> group_by(ppt) |> count()
```


### condition -- rep

```{r}
mindful |> group_by(condition) |> count()
```

### visit -- rep (not involved in RQ tho!)

```{r}
mindful |> group_by(visit) |> count()
```



### imp -- rep (not involved in RQ tho!)


```{r}
mindful |> group_by(imp) |> count()
```

### age -- rep

```{r}
mindful |> group_by(age) |> count()
```




### model


if we don't recenter age, then the intercept variability `(1|ppt)` is estimated way back 60 years before we collected any data. The slopes will basically perfectly predicted these intercept differences, so the model will be singular.  
re-centering will help.  

```{r}
#| eval: false
#recenter age
df$ageC <- df$age-60

mod1 <- lmer(ACE ~ 1 + ageC * condition + 
               (1 + ageC | ppt), 
             data = df)

summary(mod1)
```


## CBD

```{r}
cbd <- read_csv('https://uoepsy.github.io/data/stressweek1.csv')
```
```{r}
cbd |> group_by(dept) |> count()
cbd |> group_by(measure) |> count()
```

Grouping vars:

```{r}
cbd |> group_by(pid) |> count()  # rd
cbd |> group_by(CBD) |> count()  # repr
cbd |> group_by(day) |> count()  # repr
```



## OLD: drivingmusic (order after laughs or monkeystatus)

a good one for making them initially think "music|person" should work (because it's within person) but the model becomes unidentifiable (only one observation of each music for each person, so too many ranefs)

<!-- These data are simulated to represent data from a fake experiment, in which participants were asked to drive around a route in a 30mph zone. Each participant completed the route 3 times (i.e. "repeated measures"), but each time they were listening to different audio (either speech, classical music or rap music). Their average speed across the route was recorded.  -->
<!-- This is a fairly simple design, that we might use to ask **"how is the type of audio being listened to associated with driving speeds?"**  -->

<!-- The data are available at [https://uoepsy.github.io/data/drivingmusicwithin.csv](https://uoepsy.github.io/data/drivingmusicwithin.csv).   -->

<!-- ```{r echo=FALSE} -->
<!-- tibble( -->
<!--   variable = names(read_csv("https://uoepsy.github.io/data/drivingmusicwithin.csv")), -->
<!--   description = c("Participant Identifier","Avg Speed Driven on Route (mph)","Music listened to while driving (classical music / rap music / spoken word)") -->
<!-- ) |> gt::gt() -->
<!-- ``` -->

### pid -- rd

```{r}
drivingmusic <- read_csv('https://uoepsy.github.io/data/drivingmusicwithin.csv')

drivingmusic |> group_by(pid) |> count()
drivingmusic |> group_by(pid) |> n_groups()
```


### music -- rep

```{r}
drivingmusic |> group_by(music) |> count()
drivingmusic |> group_by(music) |> n_groups()
```


### model

Temptation is to fit the below, but it won't work, because each pid has only 1 obs for each music, so we're overfitting

```{r}
#| eval: false
#| error: true
#| warning: true
#| message: true
mod1 <- lmer(speed ~ music + 
               (1 + music | pid), 
             data = df)
```

```{r}
#| eval: false
mod1 <- lmer(speed ~ music + 
               (1 | pid), 
             data = df)

summary(mod1)
```
