---
title: "Data for Lab 02"
output: 
  html_notebook:
    toc: true
---

```{r}
library(tidyverse)
library(lme4)
```

all data drawn from <https://uoepsy.github.io/lmm/00_datasets>


## clothing

```{r}
clothing <- read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")
glimpse(clothing)
```

From the last lab, here's what we know about `clothing`:

- Outcome: `purch_rating`
- Predictors: `price`, `condition` (and their interaction)
- Grouping variables ...
  - ... that contribute reproducible variability: `price`, `condition`
  - ... that contribute random variability: `ppt`
  
### fixef formula
  
Q: Write the R formula for the fixed effects part of this model (aka the outcome + predictors).

```
purch_rating ~ price * condition
```

### distinct vals of preds

Q: Do at least some levels of each randomly varying grouping variable appear with more than one distinct value of each predictor?

For predictor `price`: YES!

```{r}
stats::xtabs(~ ppt + price, data = clothing)  # wide
clothing |> group_by(ppt) |> count(price)     # tall
```

For predictor `condition`: NO!

```{r}
stats::xtabs(~ ppt + condition, data = clothing)  # wide – zeroes!
clothing |> group_by(ppt) |> count(condition)     # tall – not as easy
```


### slopes

Q: 

a. Based on the results from previous question, think about what random slopes are possible in this dataset.
We know that every randomly varying grouping variable *requires* a random intercept.
For each random intercept, can we *additionally* include random slopes over the predictor(s)?

b. Write the full R formula for this model, including the fixed effects and the random effects.

a: 
- for `price`, yes, bc each ppt sees more than one price. 
- for `condition`, no, because each ppt sees only one condition and not both.

b: `purch_rating ~ price * condition + (1 + price | ppt)`



### model

```{r eval=F}
#scale price to help convergence
#change 1 in price is now change of £10
df$price <- df$price/10

mod1 <- lmer(purch_rating ~ price*condition + 
       (1+price|ppt), 
       data = df)
  
summary(mod1)
```


## monkeystatus


```{r}
monkeystatus <- read_csv('https://uoepsy.github.io/data/msmr_monkeystatus.csv')
monkeystatus |>
  glimpse()
```


From the last lab, here's what we know about `monkeystatus`:

- Outcome: `solved`
- Predictors: `status`, `difficulty` (no interaction because `difficulty` is just a covariate to control for)
- Grouping variables ...
  - ... that contribute reproducible variability: `status`, `difficulty`
  - ... that contribute random variability: `monkeyID`
  
  
### fixef formula
  
Q: Write the R formula for the fixed effects part of this model (aka the outcome + predictors).

```
solved ~ status + difficulty
```

### distinct vals of preds

Q: Do at least some levels of each randomly varying grouping variable appear with more than one distinct value of each predictor?

For predictor `status`: no! each monkey has only one status.

```{r}
stats::xtabs(~ monkeyID + status, data = monkeystatus)  # wide
```

For `difficulty`? 
Yes! At least some monkeys see both problem types.
(It's fine that and Nadheera Richard both only see one, we have enough data from all the other monkeys that see both)

```{r}
stats::xtabs(~ monkeyID + difficulty, data = monkeystatus)
```


### slopes

a. Based on the results from previous question, think about what random slopes are possible in this dataset.
We know that every randomly varying grouping variable *requires* a random intercept.
For each random intercept, can we *additionally* include random slopes over the predictor(s)?

b. Write the full R formula for this model, including the fixed effects and the random effects.

a:

- rd slope over `status` by `monkeyID`? no
- rd slope over `difficulty` by `monkeyID`? yes

b: `solved ~ status + difficulty + (1 + difficulty | monkeyID)`


### model

```{r eval=F}
mmod <- lme4::glmer(solved ~ difficulty + status + 
      (1 + difficulty | monkeyID),
      data = monkeystatus, family = binomial)
summary(mmod)
```


## laughs

```{r}
laughs <- read_csv('https://uoepsy.github.io/data/lmm_laughs.csv')
```


```{r}
glimpse(laughs)
```


From the last lab, here's what we know about `laughs`:

- Outcome: `rating`
- Predictor: `delivery`
- Grouping variables ...
  - ... that contribute reproducible variability: `delivery`
  - ... that contribute random variability: `ppt` and `joke_label`/`joke_id` (two forms of the same information, for simplicity we'll use `joke_id`)
  
  
### fixef formula
  
Q: Write the R formula for the fixed effects part of this model (aka the outcome + predictors).

```
rating ~ delivery
```

### distinct vals of preds

Q: Do at least some levels of each randomly varying grouping variable appear with more than one distinct value of each predictor?

For `ppt`, for predictor `delivery`: yes!

```{r}
stats::xtabs(~ ppt + delivery, data = laughs)
```


For `joke_id`, for predictor `delivery`: yes!

```{r}
stats::xtabs(~ joke_id + delivery, data = laughs)
```


### slopes

Q: 

a. Based on the results from previous question, think about what random slopes are possible in this dataset.
We know that every randomly varying grouping variable *requires* a random intercept.
For each random intercept, can we *additionally* include random slopes over the predictor(s)?

b. Write the full R formula for this model, including the fixed effects and the random effects.

a: yes, rd slps for delivery for both ppt and joke

b: `rating ~ delivery + (1 + delivery | ppt) + (1 + delivery | joke_id)`


### model

```{r eval=F}
mod <- lmer(rating ~ delivery + 
              (1 + delivery | joke_id) +
              (1 + delivery | ppt), data = laughs)
```


## plot

final q (challenge): choose your favourite of the three datasets and practice getting the code from the flashcards to compute group-level descriptives + making a group-level plot.





