---
title: "Grouping variables"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


## What are grouping variables?

**Grouping variables are ordinal or categorical variables whose values appear more than once in a dataset.**
In other words, when multiple observations in a dataset are related to the same value of an ordinal or categorical variable, then you're dealing with grouping variables.

For example:

- The same child is observed at a few different times of day, so our data contains multiple observations from that child.
- The same experimental participant sees all conditions of an experiment, so our data contains multiple observations from that participant.
- The same experimental condition is seen by many participants, so our data contains multiple observations of that experimental condition.
- The same test question is seen by every student, so our data contains multiple observations from that test question.
- The same student sees many test questions, so our data contains multiple observations from that student.


## Why are grouping variables important?

In DAPR2, you learned that linear models make a number of assumptions ([revisit last year's DAPR2 flashcards here](https://uoepsy.github.io/dapr2/2425/labs/1_b4_reading.html#model-assumptions){target="_blank"}).

One of those assumptions was the **assumption of independence of errors.**
For this assumption to be met, every observation must be independent of every other observation.
But when (for example) our data contains multiple observations from the same child, those data points are **not independent, because they all come from the same child.**

In general, when our data contains grouping structures, the data points associated with each member of each group are not independent.
Our analysis will violate the assumption of independence of errors unless we include the grouping variables in our model.

Depending on the [data generating process](dgp.html){target="_blank"}, we model grouping variables either as [fixed effects](fixef.html){target="_blank"} or as [random effects](ranef.html){target="_blank"}.


## How do we identify grouping variables?

We can use the `tidyverse` functions [`group_by()`](https://dplyr.tidyverse.org/articles/grouping.html?q=group_by){target="_blank"} and [`count()`](https://dplyr.tidyverse.org/reference/count.html?q=count){target="_blank"} to look for variables whose values appear more than once in a dataset.

For example, in a dataset on workplace pride in different departments of the UK civil service, let's look at the variable `department_name`.

```{r}
workpride <- read_csv('https://uoepsy.github.io/data/lmm_jsup.csv')

workpride |>
  group_by(department_name) |>
  count() 
```

Each department is associated with multiple observations.
The specific number of observations is different between departments (some have 5, some have 45), but that doesn't matter.
What matters is that **there are departments that are linked to more than one observation.**
Because the count values in the `n` column are greater than 1, `department_name` is a grouping variable.

::: {.callout-caution collapse="true"}
## What if some levels have counts of 1?

Sometimes you might end up with variables where *some* levels appear multiple times, but *other* levels appear only once.
Even if this is the case, the variable is still considered a grouping  variable.

For example, the `starwars` dataset (which comes with `tidyverse`) tells us a bit about several Star Wars characters:

```{r}
starwars |>
  head()
```

In the `starwars` dataset, is `species` a grouping variable?

```{r}
starwars |>
  group_by(species) |>
  count()
```

Yes, `species` is a grouping variable, because *at least some* levels of `species` group together more than one item (e.g., the dataset contains six droids, three Gungans).
:::


**Note: Counting how many times each level of the variable appears will only work if data is "tidy".**
All the data that we'll use in DAPR3 is tidy, so for this course, the method will always work.


::: {.callout-caution collapse="true"}
## What is tidy data?


![Stylized text providing an overview of Tidy Data. The top reads “Tidy data is a standard way of mapping the meaning of a dataset to its structure. - Hadley Wickham.” On the left reads “In tidy data: each variable forms a column; each observation forms a row; each cell is a single measurement.” There is an example table on the lower right with columns ‘id’, ‘name’ and ‘color’ with observations for different cats, illustrating tidy data structure.](../assets/allisonhorst_tidydata_1.jpg){fig-align="center"}
*Graphic by [Allison Horst](https://github.com/allisonhorst/stats-illustrations/blob/main/rstats-artwork/tidydata_1.jpg), CC BY 4.0*


According to [tidyr's Tidy data vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html), in tidy data:

1. Each variable is a column; each column is a variable.
2. Each observation is a row; each row is an observation.
3. Each value is a cell; each cell is a single value.

**Here's an example of data that's tidy:**

```{r}
iris |> 
  head() |>
  gt::gt()
```



We know this data is tidy because:

1. Each column is a variable: length and width of sepal and petal, as well as the species of iris.
2. Each row contains information gathered for one specific iris: one observation.
3. Each cell contains a single value of the given variable for the given observation.


**Here's an example of data that's messy:**

```{r}
relig_income |>   # this dataset comes with tidyverse
  head() |>
  gt::gt()
```

We know this data is messy because:

1. Each column is not a distinct variable; each column header actually contains a value of a variable representing how much money the household makes.
2. Each row is not a distinct observation (which would correspond to one household), but a collection of frequency values for a given religion.
3. Each cell does contain a single value, a count of how many households of the given religion have the given income.

Here's a tidy version of that same data:

```{r}
relig_income |>
  pivot_longer(-religion, names_to = "income", values_to = "frequency") |>
  head() |>
  gt::gt()
```


:::



::: {.callout-caution collapse="true"}
## What if I gather data that *isn't* tidy?

It's very likely that when you gather data yourself, it will be messy in one way or another.

Before you can plot and analyse your data, you'll need to tidy it.

Check out tidyr's [Tidy data vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).
It walks you through five common ways in which data can be messy and how to fix each one.
(The `relig_income` example from the rabbit hole above is from that vignette!)
:::



## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO

