---
title: "Grouping variables"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


## What are grouping variables?

lorem ipsum


## How to identify grouping variables

lorem ipsum


**Note: This method only works if data is structured to be "tidy".**
All the data that we'll use in DAPR3 is tidy, so for this course, the method will always work.


::: {.callout-caution collapse="true"}
## What is tidy data?


![Stylized text providing an overview of Tidy Data. The top reads “Tidy data is a standard way of mapping the meaning of a dataset to its structure. - Hadley Wickham.” On the left reads “In tidy data: each variable forms a column; each observation forms a row; each cell is a single measurement.” There is an example table on the lower right with columns ‘id’, ‘name’ and ‘color’ with observations for different cats, illustrating tidy data structure.](../assets/allisonhorst_tidydata_1.jpg){fig-align="center"}
*Graphic by [Allison Horst](https://github.com/allisonhorst/stats-illustrations/blob/main/rstats-artwork/tidydata_1.jpg), CC BY 4.0*


According to [tidyr's Tidy data vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html), in tidy data:

1. Each variable is a column; each column is a variable.
2. Each observation is a row; each row is an observation.
3. Each value is a cell; each cell is a single value.

**Here's an example of data that's tidy:**

```{r}
iris |> 
  head() |>
  gt::gt()
```



We know this data is tidy because:

1. Each column is a variable: length and width of sepal and petal, as well as the species of iris.
2. Each row contains information gathered for one specific iris: one observation.
3. Each cell contains a single value of the given variable for the given observation.


**Here's an example of data that's messy:**

```{r}
relig_income |>   # this dataset comes with tidyverse
  head() |>
  gt::gt()
```

We know this data is messy because:

1. Each column is not a distinct variable; each column header actually contains a value of a variable representing how much money the household makes.
2. Each row is not a distinct observation (which would correspond to one household), but a collection of frequency values for a given religion.
3. Each cell does contain a single value, a count of how many households of the given religion have the given income.

Here's a tidy version of that same data:

```{r}
relig_income |>
  pivot_longer(-religion, names_to = "income", values_to = "frequency") |>
  head() |>
  gt::gt()
```




:::



::: {.callout-caution collapse="true"}
## What if I gather data that *isn't* tidy?

It's very likely that when you gather data yourself, it will be messy in one way or another.

Before you can plot and analyse your data, you'll need to tidy it.

Check out tidyr's [Tidy data vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).
It walks you through five common ways in which data can be messy and how to fix each one.
(The `relig_income` example from the rabbit hole above is from that vignette!)
:::

