---
title: "Identify possible random effects"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

## The process in words

1. Use your RQ to figure out what your model's outcome and predictors are (i.e., your model's [fixed effects](fixef.html){target="_blank"}).

1. [Identify the grouping variables](identify-group.html) in your dataset.

1. Think about the [data generating process](dgp.html) and figure out which grouping variables contribute random / non-manipulated / non-controlled variability to your data.
**For every randomly-varying grouping variable, your model must contain a random intercept for that grouping variable.**

1. Refer back to all predictors you identified in Step 1. 
For each randomly-varying grouping variable you have identified, do at least some of its levels appear with more than one value of each predictor?
**If yes—if some levels of the grouping variable appear with more than one value of a predictor—then (in addition to the random intercepts) your model can _also_ contain a random slope over that predictor for that grouping variable.**


## The process in a flowchart

[TODO ranef flowchart]


## Example 1: Life satisfaction in Scotland

These data come from 112 people across 12 different Scottish dwellings (cities and towns). Information is captured on their ages and a measure of life satisfaction. The researchers are interested in **if there is an association between age and life-satisfaction**. 

Data are available at [https://uoepsy.github.io/data/lmm_lifesatscot.csv](https://uoepsy.github.io/data/lmm_lifesatscot.csv){target="_blank"}.

```{r}
#| echo: false
df <- read_csv("https://uoepsy.github.io/data/lmm_lifesatscot.csv")
tibble(
  variable = names(df),
  description = c("Age (years)","Life Satisfaction score","Dwelling (town/city in Scotland)","Size of Dwelling (> or <100k people)")
) |> gt::gt()
```

```{r}
lifesatscot <- read_csv("https://uoepsy.github.io/data/lmm_lifesatscot.csv")
```

### 1. Predictors

The RQ asks whether there's an association betwen `age` and `lifesat`.
The model's outcome is `lifesat` and the predictor is `age`.

The RQ doesn't mention `size`, so we will not consider it a predictor.
(In principle, it looks like a plausible covariate, that is, a way to control for how populous each area is.)


### 2. Grouping variables

The grouping variables are

- `age` (treated as ordinal because each value is spaced five years apart)
- `dwelling` (categorical)
- `size` (categorical)

```{r}
# age:
lifesatscot |> group_by(age) |> count()
```

```{r}
# dwelling:
lifesatscot |> group_by(dwelling) |> count()
```

```{r}
# size:
lifesatscot |> group_by(size) |> count()
```


### 3. Random intercepts

Thinking about what kind of variability each grouping variable contributes:

- `age` contributes **reproducible (manipulated) variability:** the researchers have gathered data from a bunch of different ages. We are interested in drawing conclusions specifically about the ages analysed.
- `dwelling` contributes **random variability:** the RQ doesn't specify that we manipulate or control for the dwelling place. And we want our results to be able to generalise beyond these twelve dwelling places.
- `size` contributes **reproducible (controlled) variability:** the researchers have recorded how big each dwelling is in case this information might otherwise confound the association between `lifesat` and `age`.

**Because `dwelling` contributes random variability, our model formula must at least include a random intercept by `dwelling`, which looks like `(1 | dwelling)`.**


### 4. Random slopes

Our model's only predictor is `age`, and our only randomly-varying grouping variable is `dwelling`.

Do at least some levels of `dwelling` appear with more than one value of `age`?

```{r}
stats::xtabs(~ dwelling + age, data = lifesatscot)
```

Yes: in every `dwelling`, there is more than one value that's not a zero.

**Therefore we can have random slopes over `age` by `dwelling`: instead of just `(1 | dwelling)`, our model will include `(1 + age | dwelling)`.**


:::{.callout-caution collapse='true'}
### Automatically count the non-zero values for each `dwelling`

It might be nice to get R to count how many non-zero values there are for each level of `dwelling`.
This way we can make extra sure that the numbers are bigger than 1.

First we check whether each cell's value is different from zero:

```{r}
stats::xtabs(~ dwelling + age, data = lifesatscot) != 0
```

And we can pipe this table into `rowSums()` to add up the number of `TRUE` values per row.
(This works because underlyingly, R stores `TRUE` as the number 1 and `FALSE` as the number 0.)

```{r}
(stats::xtabs(~ dwelling + age, data = lifesatscot) != 0) |>
  rowSums()
```

So in Aberdeen, for example, we have five distinct values of `age`, and in Kirkcaldy, we have two.
All of these numbers are bigger than 1, so we know we can have random slopes over `age` by `dwelling`.

:::




## Example 2: CBT and stress

These data are simulated to represent data from 50 participants, each measured at three different time-points (pre, during, and post) on a measure of stress. Participants were randomly allocated such that half received some cognitive behavioural therapy (CBT) treatment, and half did not. This study is interested in assessing **whether the two groups (control vs treatment) differ in changes in stress across the three time points**. 

The data are available at [https://uoepsy.github.io/data/stressint.csv](https://uoepsy.github.io/data/stressint.csv).  

```{r echo=FALSE}
tibble(
  variable = names(read_csv("https://uoepsy.github.io/data/stressint.csv")),
  description = c("Participant Identifier","Stress (range 0 to 100)","Time (Pre/During/Post)",
                  "Whether participant is in the CBT group or control group")
) |> gt::gt()
```

```{r}
stress <- read_csv("https://uoepsy.github.io/data/stressint.csv")
```


### 1. Predictors

The RQ asks whether control vs. treatment groups differ in how their stress changes across the three time points.
(Alarm bells for an interaction model should be ringing!)

The outcome variable is `stress`.
The predictors are `time` and `group`.


### 2. Grouping variables

The grouping variables are

- `ppt` (categorical)
- `time` (categorical)
- `group` (categorical)


```{r}
# ppt:
stress |> group_by(ppt) |> count()
```

```{r}
# time:
stress |> group_by(time) |> count()
```

```{r}
# group:
stress |> group_by(group) |> count()
```


### 3. Random intercepts

Thinking about what kind of variability each grouping variable contributes:

- `ppt` contributes **random variability:** we're not interested in these participants specifically; we want our conclusions to be able to generalise across different people than the ones we've studied here.
If we ran this study again, we would recruit different participants.
- `time` contributes **reproducible variability:** we're interested in drawing conclusions specifically about these three time points. 
If we ran this study again, we would keep the `time` variable the same.
- `group` contributes **reproducible variability:** the design manipulates whether people are in the control group or treatment (CBT) group, and we care about these levels specifically.
If we ran this study again, we would keep the `group` variable the same.

**Because `ppt` contributes random variability, our model formula must at least include a random intercept by `ppt`, which looks like `(1 | ppt)`.**

### 4. Random slopes

There's one randomly-varying grouping variable, `ppt`, and two predictors, `time` and `group`.
We'll check each predictor in turn.

Do at least some levels of `ppt` appear with more than one value of `time`?

```{r}
stats::xtabs(~ ppt + time, data = stress)
```

Yes: every `ppt` has one data point for every `time`.
**Therefore in principle, we can have random slopes over `time` by `ppt`: instead of just `(1 | ppt)`, we can write `(1 + time | ppt)`.**

(We'll see in the [troubleshooting flash card](troubleshoot.html){target="_blank"} that although *in principle* these random slopes are possible, in practice the model won't always manage to fit them.)


:::{.callout-caution collapse='true'}
### Automatically count the non-zero `time` values for each `ppt`

```{r}
(stats::xtabs(~ ppt + time, data = stress) != 0 ) |>
  rowSums()
```
Because these values are greater than 1, random slopes over `time` by `ppt` are possible in principle.


:::

Now: do at least some levels of `ppt` appear with more than one value of `group`, the other predictor?

```{r}
stats::xtabs(~ ppt + group, data = stress)
```

No!
Each participant is in only one group (control or treatment), not both.
**Therefore we cannot include random slopes over `group` by `ppt`. Our random effect for `ppt` remains `(1 + time | ppt)`.**


:::{.callout-caution collapse='true'}
### Automatically count the non-zero `group` values for each `ppt`

```{r}
(stats::xtabs(~ ppt + group, data = stress) != 0 ) |>
  rowSums()
```

Because these values are all equal to 1, random slopes over `group` by `ppt` are not possible.

:::


## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
