---
title: "Detect influential points and groups"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(lme4)
```


After fitting our LMM, we can check whether any individual points or any particular levels of our grouping variables are unduly influencing our model's estimates.


## Example data and model: University job satisfaction

Let's suppose we are studying employee job satisfaction at the university, and we want to estimate the **association between pay-scale and job satisfaction, controlling for the NSS rating of departments**.   

We have 399 employees from 25 different departments, and we got them to fill in a job satisfaction questionnaire, and got information on what their payscale was.
We have also taken information from the National Student Survey on the level of student satisfaction for each department.  

Each datapoint here represents an individual employee, and these employees are grouped into departments.  

Data are available at [https://uoepsy.github.io/data/msmr_nssjobsat.csv](https://uoepsy.github.io/data/msmr_nssjobsat.csv).  

```{r}
#| echo: false
tibble(variable=names(read_csv("https://uoepsy.github.io/data/msmr_nssjobsat.csv")),
       description = c(
         "National Student Satisfaction rating for the department",
         "Department name",
         "Pay scale of employee",
         "Job satisfaction of employee",
         "Binary question of whether the employee considered themselves to be satisfied with their work (1) or not (0)"
       )
) %>% gt::gt()
```

```{r}
jobsat <- read_csv("https://uoepsy.github.io/data/msmr_nssjobsat.csv")
```


Here's the [maximal model](maxmodel.html) for this data:

```{r}
jsmod <- lmer(
  jobsat ~ 1 + payscale + NSSrating + (1 + payscale | dept), 
  data = jobsat
)
```


::: {.callout-caution collapse="true"}
## Mathematical model formulation

In the following, $n$ represents the $n$th observation, that is, the $n$th row in the dataframe.
The notation $\text{dept}[n]$ identifies the department of the observation in row $n$.
For example, $\text{dept}[5]$ (the department of the fifth observation) is African Studies.

<br>

\begin{align}
\text{jobsat}_{n} &= 
(\beta_0 + u_{\text{dept}[n],0}) + 
((\beta_1 + u_{\text{dept}[n],1}) \cdot \text{payscale}_n) + 
(\beta_2 \cdot \text{NSSrating}_{\text{dept}[n]}) +
\epsilon_{n} \\

\pmatrix{u_{\text{dept}[n],0} \\ u_{\text{dept}[n],1}} &\sim \text{Normal}
\begin{bmatrix}
\pmatrix{0 \\ 0},
\pmatrix{
\sigma_{u_{\text{dept}[n],0}}^2  & 
\rho_u \sigma_{u_{\text{dept}[n],0}} \sigma_{u_{\text{dept}[n],1}} \\ 
\rho_u \sigma_{u_{\text{dept}[n],0}} \sigma_{u_{\text{dept}[n],1}} &
\sigma_{u_{\text{dept}[n],1}}^2}
\end{bmatrix} \\

\epsilon_{} &\sim \text{Normal}(0, \sigma) \\
\end{align}

```
$$
\begin{align}
\text{jobsat}_{n} &= 
(\beta_0 + u_{\text{dept}[n],0}) + 
((\beta_1 + u_{\text{dept}[n],1}) \cdot \text{payscale}_n) + 
(\beta_2 \cdot \text{NSSrating}_{\text{dept}[n]}) +
\epsilon_{n} \\

\pmatrix{u_{\text{dept}[n],0} \\ u_{\text{dept}[n],1}} &\sim \text{Normal}
\begin{bmatrix}
\pmatrix{0 \\ 0},
\pmatrix{
\sigma_{u_{\text{dept}[n],0}}^2  & 
\rho_u \sigma_{u_{\text{dept}[n],0}} \sigma_{u_{\text{dept}[n],1}} \\ 
\rho_u \sigma_{u_{\text{dept}[n],0}} \sigma_{u_{\text{dept}[n],1}} &
\sigma_{u_{\text{dept}[n],1}}^2}
\end{bmatrix} \\

\epsilon_{} &\sim \text{Normal}(0, \sigma) \\
\end{align}
$$
```
<!-- Notice that in the mathematical model formulation, we can see the assumptions that our model makes about what values are sampled from normal distributions! -->

:::









## Detecting influential points and clusters

There are two main packages for examining influence in multilevel models, __HLMdiag__ and __influence.ME__ that work by going through and deleting each observation/cluster and examining how much things change.
__HLMdiag__ works for `lmer()` but not `glmer()`, but provides a one-step approximation of the influence, which is often quicker than the full refitting process.
The __influence.ME__ package will be slower, but it works for both `lmer()` and `glmer()`.    


### HLMdiag

We use this package by creating an object using `hlm_influence()`, specifying the level at which we want to examine influence.  
In this case, `level = 1` means we are looking at the influence of individual observations, and `level = "dept"` specifies that we are looking at the influence of the levels in the `dept` variable.  
The `approx = TRUE` part means that we are asking for the approximate calculations of influence measures, rather than asking for it to actually iteratively delete each observation and re-fit the model.  

```{r}
library(HLMdiag)

inf1 <- hlm_influence(model = jsmod, level = 1, approx = TRUE)
inf2 <- hlm_influence(model = jsmod, level = "dept", approx = TRUE)
```

We can then plot metrics such as the Cook's Distance for each of these.
Note that the "internal" cutoff here adds a red line to the plots, and is simply calculated as the 3rd quartile plus 3 times the IQR, thereby capturing a relative measure of outlyingness.  
As with the measures of influence for `lm()`, such cutoffs are somewhat arbitrary, and so should be interpreted with caution.  

```{r}
dotplot_diag(inf1$cooksd, cutoff = "internal")

dotplot_diag(inf2$cooksd,  index = inf2$dept, cutoff = "internal")
```


### influence.ME

We use this package by creating an object using `influence()`, specifying the level at which we want to examine influence.  
In this case, `obs = TRUE` means we are looking at the influence of individual observations, and `group = "dept"` specifies that we are looking at the influence of the levels in the `dept` variable.  

```{r}
library(influence.ME)
inf1 <- influence(model = jsmod, obs = TRUE)
inf2 <- influence(model = jsmod, group = "dept")
```

We can then plot metrics such as the Cook's Distance for each of these:

```{r}
plot(inf1, which = "cook", sort = TRUE)
plot(inf2, which = "cook", sort = TRUE)
```

## Sensitivity analysis

To conduct a sensitivity analysis of our model's robustness to the inclusion of such influential observations, we would simply fit our model with and without that observation/cluster, and examine if/how our conclusions would change.  
For instance, to refit our model without employees of the 'Nursing Studies' department: 

```{r}
jsmod_NS <- lmer(
  jobsat ~ 1 + payscale + NSSrating +  (1 + payscale| dept), 
  data = jobsat |> filter(dept != "Nursing Studies")
)
```

```{r}
summary(jsmod)
summary(jsmod_NS)
```

The effect size for `NSSrating` becomes a bit smaller once we exclude Nursing Studies.
But the pattern of significance does not change.
So we would conclude that our results are not being driven by the data from the Nursing Studies department.



## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
