---
title: "Detect influential points and groups"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(lme4)
```


After fitting our LMM, we can check whether any individual points or any particular levels of our grouping variables are unduly influencing our model's estimates.


## Example: University job satisfaction

Let's suppose we are studying employee job satisfaction at the university, and we want to estimate the **association between pay-scale and job satisfaction, controlling for the NSS rating of departments**.   

We have 399 employees from 25 different departments, and we got them to fill in a job satisfaction questionnaire, and got information on what their payscale was.
We have also taken information from the National Student Survey on the level of student satisfaction for each department.  

Each datapoint here represents an individual employee, and these employees are grouped into departments.  

```{r}
#| echo: false
tibble(variable=names(read_csv("https://uoepsy.github.io/data/msmr_nssjobsat.csv")),
       description = c(
         "National Student Satisfaction rating for the department",
         "Department name",
         "Pay scale of employee",
         "Job satisfaction of employee",
         "Binary question of whether the employee considered themselves to be satisfied with their work (1) or not (0)"
       )
) %>% gt::gt()
```

```{r}
jobsat <- read_csv("https://uoepsy.github.io/data/msmr_nssjobsat.csv")
```


Here's the [maximal model](maxmodel.html) for this data:

```{r}
jobsat_mod <- lmerTest::lmer(
  jobsat ~ 1 + payscale + NSSrating + (1 + payscale | dept), 
  data = jobsat
)
```

We are predicting employee's job satisfaction as a function of their pay scale and the department's NSS rating.
We're also including by-department adjustments to the fixed intercept and to the slope over `payscale`.

To see where the `lmerTest::lmer()` bit comes from, see [Get p-values for LMM coefficients](pvals.html).


## Detecting influential points and groups

In simple LMs, we only looked out for influential points.
In LMMs, we should also be on the lookout for influential *groups*, that is, particular levels of a [grouping variable](groupstruc.html) which influence the model's estimates.

There are two main R packages for examining influence in multilevel models: `HLMdiag` and `influence.ME`.
We'll focus on using `HLMdiag` because it's faster than `influence.ME`, though note that if you are fitting a logistic mixed model with random effects (i.e., if you need to use `glmer()`), you'll need to use `influence.ME` instead (see the rabbit hole below).


:::{.callout-caution collapse='true'}
#### `HLMdiag` vs. `influence.ME`

`HLMdiag`:

- works only for `lmer()`, not `glmer()`
- runs more quickly than `influence.ME` because it only approximates the diagnostic measures instead of dropping each observation/level and re-fitting the model completely


`influence.ME`:

- works both for `lmer()` and `glmer()`
- runs more slowly than `HLMdiag`

:::


`HLMdiag` gives us a function called `hlm_influence()`.
To check observation-level influence, we use the argument `level = 1`, or to check group-level influence for the `dept` grouping variable, we write `level = 'dept'`.
Finally, we write `approx = TRUE` to specify that we're only looking for approximate calculations of the influence measures (approximating the measures is why `HLMdiag` is faster than `influence.ME`).

```{r}
library(HLMdiag)

inf_obs  <- hlm_influence(model = jobsat_mod, level = 1, approx = TRUE)
inf_dept <- hlm_influence(model = jobsat_mod, level = "dept", approx = TRUE)
```

The resulting data frames contain a lot of different influence measures.
For example, here are the approximated Cook's Distances of the first six observations:

```{r}
inf_obs$cooksd |> head()
```

And here are the approximated Cook's Distances of the first six `dept`s:

```{r}
inf_dept$cooksd |> head()
```

We can also plot these metrics.

Here's the Cook's Distance for all observations, including a cutoff (the red line) at the third quartile + three times the inter-quartile range of the observed Cook's Distances.
This is a somewhat arbitrary threshold—to really check the impact of these observations, you'd want to run a sensitivity analysis (see below).

```{r}
dotplot_diag(inf_obs$cooksd, cutoff = "internal")
```


And here is the Cook's Distance for each department:

```{r}
dotplot_diag(inf_dept$cooksd, index = inf_dept$dept, cutoff = "internal")
```

The sensitivity analysis below will double-check on the Nursing Studies group to see whether it might be unduly influencing the model's results.


:::{.callout-caution collapse='true'}
#### Examples with `influence.ME`

The `influence.ME` library gives us a function called `influence()`.
To check observation-level influence, we specify `obs = TRUE`, or to check group-level influence for `dept`, we specify `group = "dept"`.

`influence()` does a bit more work behind the scenes, which is what makes it slower: it iteratively deletes each observation or each group, re-fits the model, and compares the complete model to the one with the observation/group deleted.

```{r}
library(influence.ME)

inf_obs  <- influence(model = jobsat_mod, obs = TRUE)
inf_dept <- influence(model = jobsat_mod, group = "dept")
```


Now we can use `plot()` to show influence measures like Cook's Distance.

The Cook's Distance for each observation (the y axis labels are the index of each observation, but so squished together that you can't really read them):

```{r}
plot(inf_obs, which = "cook", sort = TRUE)
```

For each `dept`:

```{r}
plot(inf_dept, which = "cook", sort = TRUE)
```

Again, Nursing Studies appears to have a stronger influence on the model than the other departments.
:::

## Sensitivity analysis

A sensitivity analysis involves fitting our model with and without a given observation/group and examining whether/how our conclusions change.

Because Nursing Studies appears to have fairly strong influence on `jobsat_mod`, let's try fitting another version of this model that doesn't include the Nursing Studies data.

```{r}
jobsat_mod_NS <- lmerTest::lmer(
  jobsat ~ 1 + payscale + NSSrating +  (1 + payscale| dept), 
  data = jobsat |> filter(dept != "Nursing Studies")
)
```


And we'll compare the coefficients and pattern of significance in the two models.

:::: {.columns}

::: {.column width="45%"}
WITH Nursing Studies:

```{r}
summary(jobsat_mod)
```
:::

::: {.column width="10%"}
:::

::: {.column width="45%"}
WITHOUT Nursing Studies:

```{r}
summary(jobsat_mod_NS)
```
:::

::::


The effect size for `NSSrating` becomes a bit smaller once we exclude Nursing Studies.
But at $\alpha = .05$ (the standard significance level in Psychology and related fields), the pattern of significance does not change.
So the sensitivity analysis allows us to conclude that the model's estimates aren't being driven purely by the data from the Nursing Studies department.


## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
