---
title: "Change optimiser"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```

If a [maximal model](maxmodel.html) will not converge, then sometimes all you need to do is change the optimiser (see [Troubleshoot](troubleshoot.html)).

Optimisers change how the model estimates the parameters.
An optimiser that often works well is `bobyqa`:

```{r eval=F}
lmer(
  y ~ 1 + x * z + (1 + x * z | g), 
  data = df, 
  control = lmerControl(optimizer = "bobyqa")  # add in this line
)
```

(For logistic regression models with `glmer()`, write `glmerControl(optimizer = "bobyqa")` instead.)

If `bobyqa` still isn't helping your model converge, then you can use the function `allFit()` to try all available optimisers for a given model.
It will be slow, but it'll help you identify which optimisers work (if any), or whether you'll have to move on to simplify your model.

```{r eval = F}
allopts <- allFit(myfittedmodel)
summary(allopts)
```


If no optimisers help your model converge, then you will need to [simplify](simplify.html) the model's random effect structure.


## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
