---
title: "Plot random effects"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(lme4)
```

To explore the patterns that your model has identified, it's often useful to plot the random effects, that is, the group-level adjustments to the fixed effect parameters.

This kind of plot is often nice to include in the appendix of a report or (mini)dissertation.

For the kind of plot that's best for the Results section, see [Plot model-fitted values](plot-fitted.html).



## Example: Life satisfaction in Scotland

These data come from 112 people across 12 different Scottish dwellings (cities and towns). Information is captured on their ages and a measure of life satisfaction. The researchers are interested in **if there is an association between age and life-satisfaction**. 

Data are available at [https://uoepsy.github.io/data/lmm_lifesatscot.csv](https://uoepsy.github.io/data/lmm_lifesatscot.csv){target="_blank"}.

```{r}
#| echo: false
df <- read_csv("https://uoepsy.github.io/data/lmm_lifesatscot.csv")
tibble(
  variable = names(df),
  description = c("Age (years)","Life Satisfaction score","Dwelling (town/city in Scotland)","Size of Dwelling (> or <100k people)")
) |> gt::gt()
```

```{r}
lifesatscot <- read_csv("https://uoepsy.github.io/data/lmm_lifesatscot.csv")
```


Our model is the following (to see how we figured out this random effect structure, see [Identify possible random effects](identify-ranef.html)).

```{r}
lifesat_mod <- lmer(
  lifesat ~ age + (1 + age | dwelling),
  data = lifesatscot
)
```


## Plot group-level adjustments

The function `dotplot.ranef.mer()` will take the model's group-level adjustments (extracted via `ranef()`; see [Extract estimates from a fitted model](extract-ranefs.html)) and plot them.

```{r}
dotplot.ranef.mer(ranef(lifesat_mod))$dwelling
```

In the plot above, the `age` adjustments all appear tightly clustered around 0.
This is because their magnitude (i.e., the size of the numbers) is a lot smaller than the magnitude of the intercept adjustments.

To "zoom in" on the `age` adjustments for each dwelling, we can add a bit of extra code that allows the scales of each subplot to vary:

```{r}
dotplot.ranef.mer(
  ranef(lifesat_mod),
  scales = list(x = list(relation = 'free'))  # allow scales of subplots to vary
)$dwelling
```

Much better!


## Reading random effects plots

In general:

- Each subplot/panel/facet contains the adjustments to one fixed effect parameter.
- Each "row" represents one level of the grouping variable.
- The dots represent the estimated adjustment values for each dwelling place for each fixed effect.
- The black error bars represent the 95% confidence interval for each adjustment value.


Things to notice about this plot in particular:

- Dwellings with negative intercept adjustments tend to have positive adjustments to the slope over `age`. This suggests that the intercept and slope adjustments for each dwelling are **negatively correlated**. Let's check:

```{r}
VarCorr(lifesat_mod)
```

Yes indeed: the `Corr` column shows a correlation between intercept and slope of –0.87 (see [Interpret LMM summary > The random effects](interpret-lmm.html#the-random-effects) and [Interpret LMM summary > Interpreting the numbers: Correlation](interpret-lmm.html#interpreting-the-numbers-correlation)).




## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
