---
title: "Plot group-structured data"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

When you choose a way to plot your data, you are choosing the story you want your data to tell.
If you're telling a story about how groups within your data vary, then it's useful for your plots to show your data's grouping structure.
Here are a couple ways to do that.

## The example data sets

### Example 1: Life satisfaction in Scotland

These data come from 112 people across 12 different Scottish dwellings (cities and towns). Information is captured on their ages and a measure of life satisfaction. The researchers are interested in **if there is an association between age and life-satisfaction**. 

Data are available at [https://uoepsy.github.io/data/lmm_lifesatscot.csv](https://uoepsy.github.io/data/lmm_lifesatscot.csv){target="_blank"}.

```{r}
#| echo: false
df <- read_csv("https://uoepsy.github.io/data/lmm_lifesatscot.csv")
tibble(
  variable = names(df),
  description = c("Age (years)","Life Satisfaction score","Dwelling (town/city in Scotland)","Size of Dwelling (> or <100k people)")
) |> gt::gt()
```

```{r}
lifesatscot <- read_csv("https://uoepsy.github.io/data/lmm_lifesatscot.csv")
```

In this dataset, the predictor `age` is continuous and varies within the [grouping variable](groupstruc.html) `dwelling`.


### Example 2: Hesitation and believability

These data are simulated to represent data from 30 participants who took part in an experiment designed to investigate **whether fluency of speech influences how believable an utterance is perceived to be**.  

Each participant listened to the same 20 statements, with 10 being presented in fluent speech, and 10 being presented with a disfluency (an "erm, ..."). Fluency of the statements was counterbalanced such that 15 participants heard statements 1 to 10 as fluent and 11 to 20 as disfluent, and the remaining 15 participants heard statements 1 to 10 as disfluent, and 11 to 20 as fluent. The order of the statements presented to each participant was random. Participants rated each statement on how believable it is on a scale of 0 to 100.  

Data are available at [https://uoepsy.github.io/data/erm_belief.csv](https://uoepsy.github.io/data/erm_belief.csv). 

```{r echo=FALSE}
tibble(
  variable = names(read_csv("https://uoepsy.github.io/data/erm_belief.csv")),
  description = c("Participant Identifier","Trial number", "Statement identifier", "Condition (fluent v disfluent)", "belief rating (0-100)", "Statement")
) |> gt::gt()
```

```{r}
belief <- read_csv("https://uoepsy.github.io/data/erm_belief.csv")
```

In this dataset, the predictor `condition` is categorical and varies within the [grouping variable](groupstruc.html) `ppt`.



## Facetting by group

By including `+ facet_wrap(~ VARIABLE)` in your ggplot code, you can make separate little plots for each level of `VARIABLE`.

### Example 1: `lifesatscot`

We want to visualise the association between `age` and `lifesat` in each Scottish `dwelling`.
Each dot represents data from one person from the given area.

```{r}
lifesatscot |>
  ggplot(aes(x = age, y = lifesat)) +
  geom_point() +
  facet_wrap(~ dwelling)
```


::: {.callout-caution collapse="true"}
#### Add colour by dwelling `size`

```{r}
lifesatscot |>
  ggplot(aes(x = age, y = lifesat, colour = size)) +
  geom_point() +
  facet_wrap(~ dwelling)
```

:::



### Example 2: `belief`

We want to visualise the association between `condition` and `belief` for each `ppt` in the experiment.
The violins represent the density of `belief` data for each person in each condition, the grey dots represent individual observations from the given participant for the given condition, and the black dots represent the condition means for each participant.

```{r fig.width = 9, fig.height = 10}
belief |>
  ggplot(aes(x = condition, y = belief)) +
  geom_violin() +
  geom_jitter(alpha = 0.2) +
  stat_summary(fun = mean, geom = 'point') +
  facet_wrap(~ ppt)
```


::: {.callout-caution collapse="true"}
#### Add colour by `condition`

```{r fig.width = 9, fig.height = 10}
belief |>
  ggplot(aes(x = condition, y = belief, fill = condition, colour = condition)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.2) +
  stat_summary(fun = mean, geom = 'point', colour = 'black') +
  facet_wrap(~ ppt) +
  theme(legend.position = 'none')
```

:::

## Separate geoms per group

Some geoms (that is, some shapes or layers in ggplot) respond to the `group` aesthetic.

For example, if you set `aes(group = VARIABLE)`, then `geom_smooth()` will plot one line of best fit for each group in `VARIABLE`.


### Example 1: `lifesatscot`

Remember we want to look at the association between `age` and `lifesat` in each Scottish `dwelling`.

```{r}
lifesatscot |>
  ggplot(aes(x = age, y = lifesat, group = dwelling)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm, formula = 'y~x', se = FALSE)
```


::: {.callout-caution collapse="true"}
#### Add colour by `dwelling`

```{r}
lifesatscot |>
  ggplot(aes(x = age, y = lifesat, group = dwelling, colour = dwelling)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm, formula = 'y~x', se = FALSE)
```

:::

### Example 2: `belief`

The statistical geom `stat_summary()` will also respond to the `group = VARIABLE` aesthetic.
If we write `stat_summary(geom = 'pointrange')`, then we'll get a dot for each level of `VARIABLE` and an error bar that shows one standard error above and below that mean.

To displace the dots by a random horizontal amount, we can include `position = position_dodge(width = 0.5)` (otherwise they'll all be overlaid in a single vertical line, which is tough to read. Try it and see!).

```{r}
belief |>
  ggplot(aes(x = condition, y = belief, group = ppt)) +
  geom_smooth(method = 'lm', se = F, position = position_dodge(width = 0.5)) +
  stat_summary(geom = 'pointrange', position = position_dodge(width = 0.5))
```



::: {.callout-caution collapse="true"}
#### Add colour by `ppt`

```{r}
belief |>
  ggplot(aes(x = condition, y = belief, group = ppt, colour = ppt)) +
  geom_smooth(method = 'lm', se = F, position = position_dodge(width = 0.5)) +
  stat_summary(geom = 'pointrange', position = position_dodge(width = 0.5)) +
  theme(legend.position = 'none')  # lots of ppts so legend is not v informative
```

:::




## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
