---
title: "Add random effects to a model formula"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```


## Basic model formula

The structure of a basic linear model (not a linear mixed model) is:

```{r eval=F}
lm(outcome ~ fixed effect, data = mydata)
```

For example, imagine we're analysing the effect of stress on motivation levels.
We distributed a survey and got one pair of `motivation` and `stress` values from each respondent.
This is a between-subjects design which does not require random effects because all observations are independent (see [random effects](ranef.html), [LMM assumptions](assump.html)).

```{r eval=F}
lm(motivation ~ stress, data = mydata)
```


## LMM formula

**In general, we add on a random intercept by `group` by writing `(1 | group)`.
If we also want to include a random slope over `predictor`, we write that as `(1 + predictor | group)`.**


Here's how we add those terms into the model fomurla.

First, for LMMs, we need to use the package `lme4`.

```{r}
library(lme4)
```

`lme4` gives us the function `lmer()` for modelling a continuous outcome and `glmer(family = binomial)` for modelling a binary outcome.

The structure of a linear mixed model, which includes the random effect specification, is:

```{r eval=F}
lmer(outcome ~ fixed effect + (1 + fixed effect | grouping variable), data = mydata)
```

With `lmer()`, we have the addition of __random effect terms__, specified in parenthesis with the `|` operator (the vertical line | is often found to the left of the z key on QWERTY keyboards).  
We use the `|` operator to separate the parameters (intercept, slope etc.) on the left-hand side, from the grouping variable(s) on the right-hand side, by which we would like to model these parameters as varying.  


For example, imagine we're still analysing the effect of stress on motivation levels, but now, each person who responded to our survey did so several times.
We have several different `motivation` and `stress` scores for each person.
As such, the data points are no longer independent, because some of them come from the same person (see [LMM assumptions](assump.html)).

Therefore we need to include random intercepts by `respondent`.
And because we have several `stress` values for each respondent, we can also include a random slope over `stress` for each `respondent`.
(See [Identify possible random effects](identify-ranef.html).)

```{r eval=F}
lmer(motivation ~ stress + (1 + stress | respondent), data = mydata)
```


## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
