---
title: "<b>WEEK 5<br>Research </b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "Josiah King"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: "AY 2021-2022"
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  code_font_size = "0.7rem",
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```


```{r echo=FALSE}
library(lme4)
library(lmeresampler)
library(HLMdiag)
library(broom.mixed)
```


---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: EXAMPLE 1</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: EXAMPLE 2</h2>


---
# questions 

```{}
lmer(outcome ~ fixed effects + (random effects | groups))
```

1. when should my variable be a "fixed effect" and when should it be the "group"?
2. when should my variable be a "fixed effect" and when should it be a "random effect" by some group?




---
# Early bird gets the worm
.pull-left[
> Sample of 115 birds from 12 gardens, information captured on the arrival time (hours past midnight) and number of worms caught by the end of the day. 

```{r}
worms_data <- read_csv("https://uoepsy.github.io/data/worms.csv")
head(worms_data)
```

```{r}
library(ICC)
ICCbare(x = gardenid, y = nworms, data = worms_data)
```
]
.pull-right[
```{r echo=FALSE, fig.align="center", fig.asp=.8}
ggplot(worms_data, aes(x=arrivalt,y=nworms))+
  geom_point(size=4)+
  geom_smooth(method="lm", se=F)+
  labs(x="",y="")+
  themedapr3() -> p1
ggplot(worms_data, aes(x=arrivalt,y=nworms, col=gardenid))+
  geom_point(size=4)+
  geom_smooth(method="lm", se=FALSE)+
  facet_wrap(~gardenid, scales="free_y")+
  guides(col=FALSE)+
  labs(x="arrival at garden\n(hrs past midnight)",y="number of worms caught")+
  themedapr3() -> p2

p1 / p2 + plot_layout(heights=c(1,2))
```
]

---
# Cognitive decline

Researchers are interested in how cognition changes over time. 

.pull-left[
```{r}
cogtime <- read_csv("https://uoepsy.github.io/data/cogtimerpm.csv")
cogtime <- cogtime %>% 
  mutate(across(c(participant, sexFemale, alc), factor))
head(cogtime, 12L)
```

]

--

.pull-right[
```{r fig.asp=.7}
ggplot(cogtime, aes(x=visit_n, y = cog, col=participant))+
  geom_line(alpha = 0.5)+
  guides(col=FALSE)+
  scale_x_continuous(breaks=1:10)+
  themedapr3()
```
]


---
# Nurses in hospitals

```{r}
nursedf <- read_csv("https://uoepsy.github.io/data/nurse_stress.csv")
nursedf <- nursedf %>% 
  mutate(across(c(hospital, expcon, gender, wardtype, hospsize), factor))
head(nursedf)
```


_The files nurses.csv contains three-level simulated data from a hypothetical study on stress in hospitals. The data are from nurses working in wards nested within hospitals. It is a cluster-randomized experiment. In each of 25 hospitals, four wards are selected and randomly assigned to an experimental and a control condition. In the experimental condition, a training program is offered to all nurses to cope with job-related stress. After the program is completed, a sample of about 10 nurses from each ward is given a test that measures job-related stress. Additional variables are: nurse age (years), nurse experience (years), nurse gender (0 = male, 1 = female), type of ward (0 = general care, 1 = special care), and hospital size (0 = small, 1 = medium, 2 = large)._  
(From https://multilevel-analysis.sites.uu.nl/datasets/ )


---
# Childrens' emotion dysregulation

.pull-left[

> 200 pupils from 20 schools completed a survey containing the Emotion Dysregulation Scale (EDS) and the Child Routines Questionnaire (CRQ). Eleven of the schools were taking part in an initiative to specifically teach emotion regulation as part of the curriculum.  
  
  
>Adjusting for levels of daily routines, do children from schools partaking in the intervention present with lower levels of emotional dysregulation? 

]
.pull-right[
```{r echo=F, fig.asp=.7}
library(ggExtra)
library(patchwork)
crq <- read_csv("https://uoepsy.github.io/data/crqdata.csv")
p <- ggplot(crq[crq$int=="Control",], aes(x=crq, y=emot_dysreg, col=schoolid)) + 
  geom_point()+geom_smooth(method="lm",se=F)+
  guides(col="none")+themedapr3()+
  labs(title="Control")
p1 <- ggplot(crq[crq$int=="Treatment",], aes(x=crq, y=emot_dysreg, col=schoolid)) + 
  geom_point()+geom_smooth(method="lm",se=F)+
  guides(col="none")+themedapr3()+
  labs(title="Treatment")
gridExtra::grid.arrange(
ggMarginal(p, type="boxplot", margins="y"),
ggMarginal(p1, type="boxplot", margins="y"),
nrow=1
)
```
]



---
# Big Fish Little Fish

```{r echo=FALSE, fig.asp=.8, fig.align="center"}
bflp <- read_csv("https://uoepsy.github.io/data/bflp.csv")
library(ggforce)
library(ggfx)
ggplot(bflp, aes(x=fish_weight, y=self_esteem))+
  with_blur(geom_point(),sigma=3)+
  with_blur(geom_line(aes(group=pond),alpha=.5),sigma=3)+
  geom_point(data=filter(bflp, pond=="pond_6"))+
  geom_line(data=filter(bflp, pond=="pond_6"))+
  #geom_smooth(method="lm")+
  labs(x="Fish Weight (kg)",y="Self Esteem Scale (1-5)")+
  themedapr3()+
  with_blur(geom_mark_ellipse(aes(label = "BIG FISH",filter = fish_weight > 60),
                    con.arrow = arrow(ends = "last",length = unit(0.5, "cm")),
                    show.legend = FALSE),sigma=3)+
  with_blur(geom_mark_ellipse(aes(label = "LITTLE FISH",filter = fish_weight == 5),
                    con.arrow = arrow(ends = "last",length = unit(0.5, "cm")),
                    show.legend = FALSE),sigma=3)+
  geom_mark_ellipse(aes(label = "BIG FISH, LITTLE POND", filter = (pond == "pond_6" & fish_weight==34)),con.arrow = arrow(ends = "last",length = unit(0.5, "cm")),
                    show.legend = FALSE)+
  ylim(1,5)
```

data available at https://uoepsy.github.io/data/bflp.csv  




---
# Anxiety and Alcohol Consumption

.pull-left[
A research study investigates how anxiety is associated with drinking habits. Data was collected from 50 participants. Researchers administered the generalised anxiety disorder (GAD-7) questionnaire to measure levels of anxiety over the past week, and collected information on the units of alcohol participants had consumed within the week. Each participant was observed on 10 different occasions. 
]
.pull-right[
```{r echo=FALSE, fig.asp=.7}
alcgad <- read_csv("https://uoepsy.github.io/data/alcgad.csv") %>% mutate(interv = group)
ggplot(alcgad, aes(x=gad, y=alcunits,color=factor(ppt))) +
  geom_point(alpha=.4)+
  themedapr3()+#geom_line(aes(group=ppt))+
  labs(x="Generalised Anxiety Disorder (GAD-7)",y="Units of Alcohol in previous 7 days")+
  guides(color="none")
```

data available at https://uoepsy.github.io/data/alcgad.csv 
]

---
# Childrens' chances of receiving detention at school

.pull-left[

> Researchers are interested in whether the level of routine a child has in daily life influences their probability of receiving a detention at school. 200 pupils from 20 schools completed a survey containing the Child Routines Questionnaire (CRQ), and a binary variable indicating whether or not they had received detention in the past school year. 

```{r}
crq <- read_csv("https://uoepsy.github.io/data/crqdetentiondata.csv")
head(crq)
```
]
.pull-right[
```{r}
detentionmod <- glmer(detention ~ crq + (1 + crq | schoolid),
      data = crq, family="binomial")
summary(detentionmod)
```
]


---
# TGU & Physio

```{r include=FALSE}
library(lme4)
set.seed(86)
doit<-1
while(doit){
  Ngroup2s = 10
  dd2<-MASS::mvrnorm(n=Ngroup2s, mu = c(0,0), Sigma = matrix(c(1,0,0,1),byrow = T, nrow=2))
  cor(dd2)
  i = 2
  df<-as.data.frame(c())
  for(i in 1:Ngroup2s){
    Ngroups = round(rnorm(1,10,0))
    Nxgroups = 2
    
    #NperGroup = rep(Nxgroups*nrep,Ngroups)
    NperGroup = rdunif(Ngroups, 5, 10)*Nxgroups
    N = sum(NperGroup)
    
    dd<-MASS::mvrnorm(n=Ngroups, mu = c(0,0), Sigma = matrix(c(1,0,0,1),byrow = T, nrow=2))
    ddb<-MASS::mvrnorm(n=Ngroups, mu = c(0,0), Sigma = matrix(c(1,0,0,1),byrow = T, nrow=2))
    ddx<-MASS::mvrnorm(n=Nxgroups, mu = c(0,0), Sigma = matrix(c(1,0,0,1),byrow = T, nrow=2))
    
    igs = map(seq_along(NperGroup), ~rep(.,NperGroup[.])) %>% unlist
    xgs = map(1:Ngroups, ~rep(1:Nxgroups,NperGroup[.]/Nxgroups)) %>% unlist
    #x = map(1:Ngroups, ~rep(1:NperGroup[.],Nxgroups)) %>% unlist
    xxm = rnorm(Ngroups,50,10)
    xxm = map(1:Ngroups, ~rep(xxm[.], NperGroup[.])) %>% unlist
    xgc = map(1:Ngroups, ~rnorm(NperGroup[.], 0, 3)) %>% unlist
    #xx = map(1:Ngroups, ~rep(rnorm(1,3,.6), NperGroup[.]))%>% unlist
    xx = xxm+xgc 
    
    l2p = sample(1:4, Ngroups, replace=T, prob = c(.2,.5,.3,.1))
    l2p = map(1:Ngroups, ~rep(l2p[.], NperGroup[.])) %>% unlist
    
    l3p = i %% 2
    e = rnorm(N, sd = 15)
    
    y = 0 +
      dd[igs,1]+
      dd2[i,1]+
      #ddx[xgs, 1] + 
      -5*xxm+
      #-.05*(xx2 %in% c(1,2,4))*xgc+
      -6*xgc + 
      dd[igs,2]*xgc +
      #ddb[igs,2]*xxm + 
      1*dd2[i,2]*xxm +
      -3*l2p + 
      2*l3p +
      e
    d = data.frame(y,xxm,xgc,igs,xgs,i, l2p,l3p)
    #ggplot(d,aes(x=x,y=y,group=factor(igs)))+facet_wrap(~xgs)+geom_path()
    d$ng2 = i
    df<-rbind(df,d)
  }
  df %>% filter(xgs == 1) %>%
  lmer(y ~ xxm + xgc + l2p + l3p + (1+xgc | ng2/igs), data =.,
       control=lmerControl(optimizer = "bobyqa")) -> m 
  print(VarCorr(m))
  t1 = attributes(VarCorr(m)[[1]])$stddev
  t2 = attributes(VarCorr(m)[[1]])$correlation
  t3 = attributes(VarCorr(m)[[2]])$stddev
  t4 = attributes(VarCorr(m)[[2]])$correlation
  
  if(!isSingular(m) & all(t1 != 0) & !(t2[lower.tri(t2)] %in% c(0,1,-1)) & all(t3 != 0) & !(t4[lower.tri(t4)] %in% c(0,1,-1)) ){
    doit <- 0
  }
}

df %>% filter(xgs==1) %>% transmute(
  tgu = 8+(scale(y)[,1]*4),
  phys = round(xxm + xgc),
  hospital = ng2,
  patient = igs,
  prioritylevel = l2p,
  private = l3p
) %>% filter(hospital%in%c(5,6)) %>% 
  mutate(hospital=paste0("Hospital_",hospital),
patient = paste0(hospital,patient))-> tgudat
ggplot(tgudat, aes(x=phys, y=tgu,color=patient)) +
  geom_point(alpha=.4)+
  geom_smooth(aes(group=patient), method="lm",se=F,alpha=.4)+
  labs(x="Daily amount of Physiotherapy\n(minutes)", y="Time Get up and Go test (seconds)")+
  themedapr3()+
  guides(color="none") -> p1
```

.pull-left[
```{r echo=FALSE, fig.asp=.8}
p1
```
]

.pull-right[
```{r echo=FALSE}
tgudat %>% group_by(patient) %>% mutate(physm= mean(phys),physc = phys-physm) %>% ungroup %>%
ggplot(., aes(x=physc, y=tgu,color=patient)) +
  geom_point()+
  geom_line(aes(group=patient), alpha=.4)+
  #geom_smooth(method="lm",se=F,alpha=.1)+
  labs(x="Daily amount of Physiotherapy\n(minutes relative to patient average)", y="TUG")+
  themedapr3()+
  guides(color="none") ->p1

tgudat %>% group_by(patient) %>% mutate(physm= mean(phys),physc = phys-physm) %>% ungroup %>%
ggplot(., aes(x=physm, y=tgu,color=patient)) +
  #geom_point(alpha=.4)+
  #geom_line(aes(group=patient), alpha=.4)+
  stat_summary(geom="pointrange")+
  labs(x="Average daily amount of Physiotherapy\n(minutes)", y="TUG")+
  themedapr3()+
  guides(color="none") -> p2
p1 / p2
```
]





---
# EXAMPLE

- experience sampling study
- negative affect ~ stress
- within / between 


```{r eval=F, echo=F}
filepath <- "https://quantdev.ssri.psu.edu/sites/qdev/files/AMIBshare_persons_2019_0501.csv"
#read in the .csv file using the url() function
AMIB_persons <- read.csv(file=url(filepath),header=TRUE)
#subsetting to variables of interest
AMIB_persons <- AMIB_persons[ ,c("id","bfi_n")]
#set filepath for data file
filepath <- "https://quantdev.ssri.psu.edu/sites/qdev/files/AMIBshare_daily_2019_0501.csv"
#read in the .csv file using the url() function
AMIB_daily <- read.csv(file=url(filepath),header=TRUE)
#subsetting to variables of interest
AMIB_daily <- AMIB_daily[ ,c("id","day","negaff","pss")]
#reverse coding the pss variable into a new stress variable
AMIB_daily$stress <- 4 - AMIB_daily$pss
#describing new variable
describe(AMIB_daily$stress)
#histogram
ggplot(data=AMIB_daily, aes(x=stress)) +
  geom_histogram(fill="white", color="black",bins=20) +
  labs(x = "Stress (high = more stressed)")
AMIB_imeans <- ddply(AMIB_daily, "id", summarize,
                       stress_trait = mean(stress, na.rm=TRUE),
                       negaff_trait = mean(negaff, na.rm=TRUE))
describe(AMIB_imeans)
#merging into person-level file
AMIB_persons <- merge(AMIB_persons, AMIB_imeans, by="id")   
#make centered versions of the person-level scores
AMIB_persons$bfi_n_c <- scale(AMIB_persons$bfi_n,center=TRUE,scale=FALSE)
AMIB_persons$stress_trait_c <- scale(AMIB_persons$stress_trait,center=TRUE,scale=FALSE)
#describe person-level data
describe(AMIB_persons)
#merging person-level data into daily data
daily_long <- merge(AMIB_daily,AMIB_persons,by="id")
#calculating state variables
daily_long$stress_state <- daily_long$stress - daily_long$stress_trait
daily_long$negaff_state <- daily_long$negaff - daily_long$negaff_trait
#describing data
describe(daily_long)
```


```{r eval=F, echo=F}
daily_long %>%
ggplot(., aes(x=stress,y=negaff))+
  geom_point()+
  geom_smooth(aes(group=id), se=F, method="lm", col="black",lwd=.5)

lmer(negaff ~ stress + (1+stress|id), daily_long) %>% summary

lmer(negaff_state~ stress + (1+stress|id), daily_long) %>%
  broom.mixed::augment() %>% 
  ggplot(.,aes(x=stress,y=.fitted,group=id))+
  geom_line(alpha=.2)

lmer(negaff_state ~ stress_state * bfi_n_c + (1+stress_state|id), daily_long) %>%
  broom.mixed::augment() %>% 
  ggplot(.,aes(x=bfi_n_c,y=.fitted,group=id))+
  geom_line(alpha=.2)

```

---

a write up/paper?

Ram, N., Conroy, D. E., Pincus, A. L., Hyde, A. L., & Molloy, L. E. (2012). Tethering theory to method: Using measures of intraindividual variability to operationalize individuals’ dynamic characteristics. In G. Hancock & J. Harring (Eds.), Advances in longitudinal methods in the social and behavioral sciences (pp. 81-110). New York: Information Age.

d




---
# End

