---
title: "Identifying grouping structure"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio/  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal/with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
xaringanExtra::use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

In this lab, you will start to build skills that you'll use again and again for the next five weeks (and really, for the rest of your data analysis career!).

You'll practice **identifying the grouping structure within a given dataset** and thinking about **what kind of variability each grouping variable contributes.**

In the ten questions below, you'll apply the tools you saw in the lectures to five different group-structured datasets.

(Next week, you'll apply new tools to these same five datasets to understand their grouping structures even more deeply!)


:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab01.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
:::

---

## Clothing

:::frame

Read in the dataset located at [https://uoepsy.github.io/data/dapr3_mannequin.csv](https://uoepsy.github.io/data/dapr3_mannequin.csv) and name it `clothing`.

RQ: **Are people more likely to purchase clothing when they see it displayed on a model, and is this association dependent on item price?**

```{r}
#| echo: false
tibble(
  variable=names(read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")),
  description=c("Purchase rating (sliding scale 0 to 100, with higher ratings indicating greater perceived likelihood of purchase)",
                "Price presented for item (range £5 to £100)",
                "Participant identifier",
                "Whether items are seen on a model or on a white background")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

Thirty participants were presented with a set of pictures of items of clothing, and rated each item how likely they were to buy it.
Each participant saw 20 items, ranging in price from £5 to £100.
15 participants saw these items worn by a model, while the other 15 saw the items against a white background.
:::

:::



`r qbegin(qcounter())`

a. **What grouping variable(s)** does this dataset contain?
a. **How many levels/different values** does each grouping variable have?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

dolor sit amet

`r solend()`

`r qbegin(qcounter())`

a. Which grouping variable(s) contribute **reproducible/manipulated/controlled variability?**
a. Which grouping variable(s) contribute **random/non-manipulated/non-controlled variability?**

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

dolor sit amet

`r solend()`



## Monkey status

:::frame
Read in the dataset located at [https://uoepsy.github.io/data/msmr_monkeystatus.csv](https://uoepsy.github.io/data/msmr_monkeystatus.csv) and name it `monkey`.

RQ: **How is the social status of monkeys associated with their ability to solve problems, while controlling for the difficulty of the problem?**

```{r}
#| echo: false
tibble(variable=
         names(read_csv("https://uoepsy.github.io/data/msmr_monkeystatus.csv")),
       description = c(
         "Social status of monkey (adolescent, subordinate adult, or dominant adult)",
         "Problem difficulty ('easy' vs 'difficult')",
         "Monkey name",
         "Whether or not the problem was successfully solved by the monkey")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

Researchers have given a sample of Rhesus Macaques various problems to solve in order to receive treats.
Troops of Macaques have a complex social structure, but adult monkeys tend can be loosely categorised as having either a "dominant" or "subordinate" status.
The monkeys in our sample are either adolescent monkeys, subordinate adults, or dominant adults.
Each monkey attempted various problems before they got bored/distracted/full of treats.
Each problems were classed as either "easy" or "difficult", and the researchers recorded whether or not the monkey solved each problem.
:::

:::

## Laughs

:::frame

Read in the dataset located at [https://uoepsy.github.io/data/lmm_laughs.csv](https://uoepsy.github.io/data/lmm_laughs.csv) and name it `laughs`.

RQ: **How is the delivery format of jokes (audio-only vs. audio AND video) associated with differences in humour ratings?**

```{r}
#| echo: false
tibble(
  variable = c(
    'ppt',
    'joke_label',
    'joke_id',
    'delivery',
    'rating'
  ),
  description = c("Participant identification number",
                  "Joke presented",
                  "Joke identification number",
                  "Experimental manipulation: whether joke was presented in audio-only ('audio') or in audiovideo ('video')",
                  "Humour rating chosen on a slider from 0 to 100")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

These data are simulated to imitate an experiment that investigates the effect of visual non-verbal communication (i.e., gestures, facial expressions) on joke appreciation.
Ninety participants took part in the experiment, in which they each rated how funny they found a set of 30 jokes.
For each participant, the order of these 30 jokes was randomised for each run of the experiment.
For each participant, the set of jokes was randomly split into two halves, with the first half being presented in audio-only, and the second half being presented in audio and video.
This meant that each participant saw 15 jokes with video and 15 without, and each joke would be presented with video roughly half of the time.
:::

:::


## Driving music

:::frame
Read in the dataset located at [https://uoepsy.github.io/data/drivingmusicwithin.csv](https://uoepsy.github.io/data/drivingmusicwithin.csv) and name it `driving`.

RQ: **How is the type of audio being listened to associated with driving speeds?**

```{r echo=FALSE}
tibble(
  variable = names(read_csv("https://uoepsy.github.io/data/drivingmusicwithin.csv")),
  description = c("Participant identifier",
                  "Average speed driven on route (mph)",
                  "Music listened to while driving (classical music / rap music / spoken word)")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

These data are simulated to represent data from a fake experiment, in which participants were asked to drive around a route in a 30mph zone.
Each participant completed the route three times (i.e., "repeated measures"), but each time they were listening to different audio (either speech, classical music or rap music).
Their average speed across the route was recorded.
:::

:::

## Mindful decline

:::frame
Read in the dataset located at [https://uoepsy.github.io/data/lmm_mindfuldecline.csv](https://uoepsy.github.io/data/lmm_mindfuldecline.csv) and name it `mindful`.

RQ: **Is engaging in mindfulness associated with slower cognitive decline as people get older?**

```{r}
#| echo: false
tibble(
  variable=names(read_csv("https://uoepsy.github.io/data/lmm_mindfuldecline.csv")),
  description=c(
    "Site identifier",
    "Participant identifier",
    "Whether the participant engages in mindfulness or not (control/mindfulness)",
    "Study visit number (1 to 10)",
    "Age (in years) at study visit",
    "Addenbrooke's Cognitive Examination score. scores can range from 0 to 100",
    "Clinical diagnosis of cognitive impairment ('imp' = impaired, 'unimp' = unimpaired)")
) |> gt::gt()
```


:::{.callout-note collapse='true'}
### More detail about this dataset
The researchers recruited a sample of 20 participants at age 60, and administered the Addenbrooke's Cognitive Examination (ACE) every 2 years (until participants were aged 78).
Half of the participants complete weekly mindfulness sessions, while the remaining participants did not.
:::

:::



