---
title: "Reasoning about random effects"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio/  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal/with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
xaringanExtra::use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

In this lab, you'll apply the tools you saw in the lectures this week to the same three datasets you got to know last week.

You'll figure out **how to model the grouping structure in each dataset in terms of random intercepts and random slopes** (which, together, we refer to as random effects).


:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab02.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
    - `stats` (for `xtabs()`)
:::

---



## Clothing

:::frame

Read in the dataset located at [https://uoepsy.github.io/data/dapr3_mannequin.csv](https://uoepsy.github.io/data/dapr3_mannequin.csv) and name it `clothing`.

RQ: **Are people more likely to purchase clothing when they see it displayed on a model, and is this association dependent on item price?**

```{r}
#| echo: false
tibble(
  variable=names(read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")),
  description=c("Purchase rating (sliding scale 0 to 100, with higher ratings indicating greater perceived likelihood of purchase)",
                "Price presented for item (range £5 to £100)",
                "Participant identifier",
                "Whether items are seen on a model or on a white background")
) |> gt::gt()
```


:::{.callout-note collapse='true'}
### More detail about this dataset

Thirty participants were presented with a set of pictures of items of clothing, and rated each item how likely they were to buy it.
Each participant saw 20 items, ranging in price from £5 to £100.
15 participants saw these items worn by a model, while the other 15 saw the items hanging against a white background.
:::


**From the last lab, here's the key information about `clothing`:**

- Outcome: `purch_rating`
- Predictors: `price`, `condition` (and their interaction)
- Randomly-varying grouping variable: `ppt`

:::

`r qbegin(qcounter())`

**Write the R formula for the fixed effects part of this model** (that is, the outcome and predictors, the `y ~ x + z` part, the kind of models you've done before in DAPR2).

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
purch_rating ~ price * condition
```

(Note: We know this has to be an interaction model because the RQ is asking about how an *association* between predictor and outcome *depends on* the value of another predictor.)


`r solend()`

`r qbegin(qcounter())`

Now we're looking at how each randomly-varying grouping variable relates to the model's predictors.
We want to know: **for each randomly-varying grouping variable, do at least some of its levels appear with more than one distinct value of each of the model's predictors**?

Specifically, for `clothing`:

- Do at least some levels of `ppt` appear with more than one distinct value of `price`?
- Do at least some levels of `ppt` appear with more than one distinct value of `condition`?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
clothing <- read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")
```

**Do at least some levels of `ppt` appear with more than one distinct value of `price`?**
**Yes,** every `ppt` sees every `price` once.
In other words, `price` varies within `ppt`.

```{r}
stats::xtabs(~ ppt + price, data = clothing)
```


**Do at least some levels of `ppt` appear with more than one distinct value of `condition`?**
**No,** every `ppt` sees only one `condition` or the other, not both.
In other words, `condition` varies between `ppt`s.

```{r}
stats::xtabs(~ ppt + condition, data = clothing)
```

`r solend()`


`r qbegin(qcounter())`

Let's bring all this information together to reason about the random effects in this dataset.

We know that every randomly-varying grouping variable *requires* a random intercept.
For each of those grouping variables, ask yourself: **can we *additionally* include random slopes over each of the model's predictor(s) by that variable?**

Specifically, for `clothing`, we know we need a random intercept by `ppt`.

- Can we also include a random slope over `price` by `ppt`?
- Can we also include a random slope over `condition` by `ppt`?

These answers will be based on what you found in answer to Question 2.

**Write out the complete R formula for this model, expanding on the fixed-effects-only model you wrote above by adding on the appropriate random effects.**

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

**In addition to the random intercept by `ppt`, can we also include a random slope over `price` by `ppt`?**
**Yes,** because we figured out in the last question that each value of `ppt` appears with more than one value of `price` (that is, each participant sees more than one price).

**Can we also include a random slope over `condition` by `ppt`?**
**No,** because we figured out in the last question that each value of `ppt` appears with only one value of `condition` (that is, each participant sees only one condition, not both).

Therefore the full model formula is:

```
purch_rating ~ price * condition + (1 + price | ppt)
```

`r solend()`


## Monkey status

:::frame
Read in the dataset located at [https://uoepsy.github.io/data/msmr_monkeystatus.csv](https://uoepsy.github.io/data/msmr_monkeystatus.csv) and name it `monkey`.

RQ: **How is the social status of monkeys associated with their ability to solve problems, while controlling for the difficulty of the problem?**

```{r}
#| echo: false
tibble(variable=
         names(read_csv("https://uoepsy.github.io/data/msmr_monkeystatus.csv")),
       description = c(
         "Social status of monkey (adolescent, subordinate adult, or dominant adult)",
         "Problem difficulty ('easy' vs 'difficult')",
         "Monkey name",
         "Whether or not the problem was successfully solved by the monkey")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

Researchers have given a sample of Rhesus Macaques various problems to solve in order to receive treats.
Troops of Macaques have a complex social structure, but adult monkeys tend can be loosely categorised as having either a "dominant" or "subordinate" status.
The monkeys in our sample are either adolescent monkeys, subordinate adults, or dominant adults.
Each monkey attempted various problems before they got bored/distracted/full of treats.
Each problems were classed as either "easy" or "difficult", and the researchers recorded whether or not the monkey solved each problem.
:::

**From the last lab, here's the key information about `clothing`:**

- Outcome: `solved`
- Predictors: `status`, `difficulty`
- Randomly-varying grouping variable: `monkeyID`

:::

`r qbegin(qcounter())`

**Write the R formula for the fixed effects part of this model.**

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
solved ~ status + difficulty
```

(Note: We know this is not an interaction model because the RQ is asking for us to control for a specific variable, not to estimate how one effect depends on a different variable.
To control for a specific covariate, we just need to add it on.)


`r solend()`

`r qbegin(qcounter())`

Now we're looking at how each randomly-varying grouping variable relates to the model's predictors.
We want to know: **for each randomly-varying grouping variable, do at least some of its levels appear with more than one distinct value of each of the model's predictors**?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
monkey <- read_csv('https://uoepsy.github.io/data/msmr_monkeystatus.csv')
```

**Do at least some levels of `monkeyID` appear with more than one distinct value of `status`?**
**No,** every `monkeyID` has only one `status`.

```{r}
stats::xtabs(~ monkeyID + status, data = monkey)
```


**Do at least some levels of `monkeyID` appear with more than one distinct value of `difficulty`?**
**Yes,** nearly every `monkeyID` sees both levels of `difficulty`.

```{r}
stats::xtabs(~ monkeyID + difficulty, data = monkey)
```

(It's fine that Nadheera and Richard only see one difficulty level each—we have enough data from the other monkeys to still be able to estimate the random effects.
Speaking of which...)

`r solend()`


`r qbegin(qcounter())`

Let's bring all this information together to reason about the random effects in this dataset.

We know that every randomly-varying grouping variable *requires* a random intercept.
For each of those grouping variables, ask yourself: **can we *additionally* include random slopes over each of the model's predictor(s) by that variable?**

**Write out the complete R formula for this model, expanding on the fixed-effects-only model you wrote above by adding on the appropriate random effects.**

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

**In addition to the random intercept by `monkeyID`, can we also include a random slope over `status` by `monkeyID`?**
**No,** because each monkey only has one status.

**Can we also include a random slope over `difficulty` by `monkeyID`?**
**Yes,** because at least some monkeys have seen both levels of `difficulty`.

Therefore the full model formula is:

```
solved ~ status + difficulty + (1 + difficulty | monkeyID)
```

`r solend()`




## Laughs

:::frame

Read in the dataset located at [https://uoepsy.github.io/data/lmm_laughs.csv](https://uoepsy.github.io/data/lmm_laughs.csv) and name it `laughs`.

RQ: **How is the delivery format of jokes (audio-only vs. audio AND video) associated with differences in humour ratings?**

```{r}
#| echo: false
tibble(
  variable = c(
    'ppt',
    'joke_label',
    'joke_id',
    'delivery',
    'rating'
  ),
  description = c("Participant identification number",
                  "Joke presented",
                  "Joke identification number",
                  "Experimental manipulation: whether joke was presented in audio-only ('audio') or in audiovideo ('video')",
                  "Humour rating chosen on a slider from 0 to 100")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

These data are simulated to imitate an experiment that investigates the effect of visual non-verbal communication (i.e., gestures, facial expressions) on joke appreciation.
Ninety participants took part in the experiment, in which they each rated how funny they found a set of 30 jokes.
For each participant, the order of these 30 jokes was randomised for each run of the experiment.
For each participant, the set of jokes was randomly split into two halves, with the first half being presented in audio-only, and the second half being presented in audio and video.
This meant that each participant saw 15 jokes with video and 15 without, and each joke would be presented with video roughly half of the time.
:::

**From the last lab, here's the key information about `laughs`:**

- Outcome: `rating`
- Predictors: `delivery`
- Randomly-varying grouping variables: `ppt` and `joke_label`/`joke_id` (two forms of the same information; for simplicity we can just use `joke_id`)

:::


`r qbegin(qcounter())`

**Write the R formula for the fixed effects part of this model.**

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
rating ~ delivery
```


`r solend()`

`r qbegin(qcounter())`

Now we're looking at how each randomly-varying grouping variable relates to the model's predictors.
We want to know: **for each randomly-varying grouping variable, do at least some of its levels appear with more than one distinct value of each of the model's predictors**?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
laughs <- read_csv('https://uoepsy.github.io/data/lmm_laughs.csv')
```

**Do at least some levels of `ppt` appear with more than one distinct value of `delivery`?**
**Yes,** every `ppt` sees both kinds of `delivery`.

```{r}
stats::xtabs(~ ppt + delivery, data = laughs)
```


**Do at least some levels of `joke_id` appear with more than one distinct value of `delivery`?**
**Yes,** every `joke_id` is told with both kinds of `delivery`.

```{r}
stats::xtabs(~ joke_id + delivery, data = laughs)
```

`r solend()`


`r qbegin(qcounter())`

Let's bring all this information together to reason about the random effects in this dataset.

We know that every randomly-varying grouping variable *requires* a random intercept.
For each of those grouping variables, ask yourself: **can we *additionally* include random slopes over each of the model's predictor(s) by that variable?**

**Write out the complete R formula for this model, expanding on the fixed-effects-only model you wrote above by adding on the appropriate random effects.**

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

**In addition to the random intercept by `ppt`, can we also include a random slope over `delivery` by `ppt`?**
**Yes,** because each `ppt` has seen each kind of `delivery`.

**In addition to the random intercept by `joke_id`,  can we also include a random slope over `delivery` by `joke_id`?**
**Yes,** because each `joke_id` was told in each kind of `delivery`.

Therefore the full model formula is:

```
rating ~ delivery + (1 + delivery | ppt) + (1 + delivery | joke_id)
```

`r solend()`


## Descriptives and plots

`r qbegin(qcounter())`

Choose your favourite of the three datasets `clothing`, `monkey`, or `laughs`.

In the data set you chose, practice applying the code from the flashcards to compute group-level descriptives [TODO WHICH?] and plotting the group-level variability in the data [TODO HOW?].

`r qend()`

`r solbegin(label="1 – clothing TODO", show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`

`r solbegin(label="2 – monkey TODO", show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`

`r solbegin(label="3 – laughs TODO", show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`
