---
title: "Modelling group-level variability using random effects"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio  /  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
xaringanExtra::use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```


:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab02.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
    - `lme4`
:::

---

:::frame


Read in the dataset located at [https://uoepsy.github.io/data/lmm_jsup.csv](https://uoepsy.github.io/data/lmm_jsup.csv){target="_blank"} and name it `wp_data`.

RQ: **How is the pride that government employees feel about their workplace associated with how long they have been employed and their seniority level?**

```{r}
#| echo: false
jsup <- read_csv("https://uoepsy.github.io/data/lmm_jsup.csv")
tibble(
  variable = names(jsup),
  description = c(
    "Name of government department",
    "Department Acronym",
    "Whether the department functions as hybrid department with various employees working remotely (1), or as a fully in-person office (0)",
    "Employee role (A, B or C)",
    "Employee's seniority level. These map to roles, such that role A is 0-4, role B is 5-9, role C is 10-14. Higher numbers indicate more seniority",
    "Length of employment in the department (years)",
    "Composite measure of 'workplace pride'")
) |> gt::gt()
```
:::


:::{.callout-note collapse='true'}
### More detail about this dataset

A questionnaire was sent to all UK civil service departments, and the `lmm_jsup.csv` dataset contains all responses that were received.
Some of these departments work as hybrid or 'virtual' departments, with a mix of remote and office-based employees.
Others are fully office-based.

The questionnaire included items asking about how much the respondent believe in the department and how it engages with the community, what it produces, how it operates and how treats its people.
A composite measure of 'workplace-pride' was constructed for each employee.
Employees in the civil service are categorised into 3 different roles: A, B and C.
The roles tend to increase in responsibility, with role C being more managerial, and role A having less responsibility.
We also have data on the length of time each employee has been in the department (sometimes new employees come straight in at role C, but many of them start in role A and work up over time).

:::


### q1

`r qbegin(qcounter())`

The research question (see above) is asking about how workplace pride is associated with employment length and seniority, so we know that our outcome variable will be `wp` and our predictors will be `employment_length` and `seniority`.

Now: one of the other variables in this dataset is a grouping variable which contributes random variability (TODO flashcards link).
Which one is it?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The grouping variable that contributes random variability is `dept` or `department_name`.
Both define the same groups and contain the same information, so they're essentially equivalent.

Other possible answers and why they're not right:

- `virtual` is a grouping variable, but it's a variable that the study designers are controlling for, not a variable that contributes random variability.
- `role` is a grouping variable, but it's also a variable that the study designers are controlling for, not a variable that contributes random variability.


`r solend()`


### q2

`r qbegin(qcounter())`

RESUME

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q3

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q4

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q5

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q6

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q7

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q8

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q9

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

### q10

`r qbegin(qcounter())`

question

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

solution

`r solend()`

