---
title: "Building maximal models and interpreting LMM estimates"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE  # TRUE = collapse sols, FALSE = uncollapse sols
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio/  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal/with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
library(stats)
library(lme4)
library(lmerTest)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

In this lab, you'll apply the tools you saw in the lectures this week to the same three datasets you got to know in Week 1.
You'll work out the maximal model structure for each one.

And in Question 10 (which is more substantial than the first nine), you'll fit one of those maximal models and interpret its estimates.


:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab03.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
    - `lme4`
    - `lmerTest`
    - `stats` (for `xtabs()`)
:::

---



## Clothing

:::frame

Read in the dataset located at [https://uoepsy.github.io/data/dapr3_mannequin.csv](https://uoepsy.github.io/data/dapr3_mannequin.csv) and name it `clothing`.

RQ: **Are people more likely to purchase clothing when they see it displayed on a model, and is this association dependent on item price?**

```{r}
#| echo: false
tibble(
  variable=names(read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")),
  description=c("Purchase rating (sliding scale 0 to 100, with higher ratings indicating greater perceived likelihood of purchase)",
                "Price presented for item (range £5 to £100)",
                "Participant identifier",
                "Whether items are seen on a model or on a white background")
) |> gt::gt()
```


:::{.callout-note collapse='true'}
### More detail about this dataset

Thirty participants were presented with a set of pictures of items of clothing, and rated each item how likely they were to buy it.
Each participant saw 20 items, ranging in price from £5 to £100.
15 participants saw these items worn by a model, while the other 15 saw the items hanging against a white background.
:::


**From the Week 1 lab, here's the key information about `clothing`:**

- Outcome: `purch_rating`
- Predictors: `price`, `condition` (and their interaction)
- Randomly-varying grouping variable: `ppt`

:::

<!-- q1 -->

`r qbegin(qcounter())`

**Write the R formula for the fixed effects part of this model** (that is, the `y ~ x + z` part).

:::dapr3green
🗂️
See [Fixed effects](https://uoepsy.github.io/dapr3/2627/flashcards/block1/fixef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
purch_rating ~ price * condition
```

(Note: We know this has to be an interaction model because the RQ is asking about how an *association* between predictor and outcome *depends on* the value of another predictor.)


`r solend()`

<!-- q2 -->

`r qbegin(qcounter())`

In the Week 1 lab, you identified the randomly-varying grouping variable(s) in this dataset.
**For each randomly-varying grouping variable, we know that a maximal model requires a random intercept—an adjustment to the fixed intercept for each level of the grouping variable.**
Additionally, for each of those variables, a maximal model *may* contain a random slope over each predictor—an adjustment to the fixed slope for each level of the grouping variable.

**Identify the random slopes that this model can contain.**
You can figure this out by asking: for a given randomly-varying grouping variable, do at least some of its levels appear with more than one distinct value of a given predictor?
If yes, then it's possible to include a random slope over the given predictor by the given grouping variable.

Specifically, for `clothing`:

- Do at least some levels of `ppt` appear with more than one distinct value of `price`?
Can we include a random slope over `price` by `ppt`?
- Do at least some levels of `ppt` appear with more than one distinct value of `condition`?
Can we include a random slope over `condition` by `ppt`?

:::dapr3green
🗂️
See [Identify possible random effects](https://uoepsy.github.io/dapr3/2627/flashcards/block1/identify-ranef.html) flash card.
:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
clothing <- read_csv("https://uoepsy.github.io/data/dapr3_mannequin.csv")
```

**Do at least some levels of `ppt` appear with more than one distinct value of `price`?**
**Yes,** every `ppt` sees every `price` once.
In other words, `price` varies within `ppt`.

```{r}
xtabs(~ ppt + price, data = clothing)
```

Each value of `ppt` appears with more than one value of `price` (that is, each participant sees more than one price), so we can include a random slope over `price` by `ppt`.

<br>

**Do at least some levels of `ppt` appear with more than one distinct value of `condition`?**
**No,** every `ppt` sees only one `condition` or the other, not both.
In other words, `condition` varies between `ppt`s.

```{r}
xtabs(~ ppt + condition, data = clothing)
```

Each value of `ppt` appears with only one value of `condition` (that is, each participant sees only one condition, not both), so we cannot include a random slope over `condition` by `ppt`.

`r solend()`

<!-- q3 -->

`r qbegin(qcounter())`
**Write out the complete R formula for this maximal model** (expanding on the fixed-effects-only model you wrote above by adding on the appropriate random effects).

:::dapr3green
🗂️
See [Add random effects to a model formula](https://uoepsy.github.io/dapr3/2627/flashcards/block1/identify-ranef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
purch_rating ~ price * condition + (1 + price | ppt)
```

`r solend()`


## Monkey status

:::frame
Read in the dataset located at [https://uoepsy.github.io/data/msmr_monkeystatus.csv](https://uoepsy.github.io/data/msmr_monkeystatus.csv) and name it `monkey`.

RQ: **How is the social status of monkeys associated with their ability to solve problems, while controlling for the difficulty of the problem?**

```{r}
#| echo: false
tibble(variable=
         names(read_csv("https://uoepsy.github.io/data/msmr_monkeystatus.csv")),
       description = c(
         "Social status of monkey (adolescent, subordinate adult, or dominant adult)",
         "Problem difficulty ('easy' vs 'difficult')",
         "Monkey name",
         "Whether or not the problem was successfully solved by the monkey")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

Researchers have given a sample of Rhesus Macaques various problems to solve in order to receive treats.
Troops of Macaques have a complex social structure, but adult monkeys tend can be loosely categorised as having either a "dominant" or "subordinate" status.
The monkeys in our sample are either adolescent monkeys, subordinate adults, or dominant adults.
Each monkey attempted various problems before they got bored/distracted/full of treats.
Each problems were classed as either "easy" or "difficult", and the researchers recorded whether or not the monkey solved each problem.
:::

**From the Week 1 lab, here's the key information about `monkeystatus`:**

- Outcome: `solved`
- Predictors: `status`, `difficulty`
- Randomly-varying grouping variable: `monkeyID`

:::

<!-- q4 -->

`r qbegin(qcounter())`

Write the R formula for the fixed effects part of this model.

:::dapr3green
🗂️
See [Fixed effects](https://uoepsy.github.io/dapr3/2627/flashcards/block1/fixef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
solved ~ status + difficulty
```

(Note: We know this is not an interaction model because the RQ is asking for us to control for a specific variable, not to estimate how one effect depends on a different variable.
To control for a specific covariate, we just need to add it on.)


`r solend()`

<!-- q5 -->

`r qbegin(qcounter())`

We already know what random intercepts the model must have (one per randomly-varying grouping variable).

Identify the possible random slopes.

:::dapr3green
🗂️
See [Identify possible random effects](https://uoepsy.github.io/dapr3/2627/flashcards/block1/identify-ranef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
monkey <- read_csv('https://uoepsy.github.io/data/msmr_monkeystatus.csv')
```

**Do at least some levels of `monkeyID` appear with more than one distinct value of `status`?**
**No,** every `monkeyID` has only one `status`.

```{r}
xtabs(~ monkeyID + status, data = monkey)
```

Because each monkey only has one status, no random slope over `status` is possible.

<br>


**Do at least some levels of `monkeyID` appear with more than one distinct value of `difficulty`?**
**Yes,** nearly every `monkeyID` sees both levels of `difficulty`.

```{r}
xtabs(~ monkeyID + difficulty, data = monkey)
```

Because at least some monkeys have seen both levels of difficulty, we can include a random slope by `monkey` over `difficulty`.

(It's fine that Nadheera and Richard only see one difficulty level each—we have enough data from the other monkeys to still be able to estimate the random slopes.)

`r solend()`


<!-- q6 -->

`r qbegin(qcounter())`

Write out the complete R formula for this maximal model.

:::dapr3green
🗂️
See [Add random effects to a model formula](https://uoepsy.github.io/dapr3/2627/flashcards/block1/identify-ranef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
solved ~ status + difficulty + (1 + difficulty | monkeyID)
```

`r solend()`




## Laughs

:::frame

Read in the dataset located at [https://uoepsy.github.io/data/lmm_laughs.csv](https://uoepsy.github.io/data/lmm_laughs.csv) and name it `laughs`.

RQ: **How is the delivery format of jokes (audio-only vs. audio AND video) associated with differences in humour ratings?**

```{r}
#| echo: false
tibble(
  variable = c(
    'ppt',
    'joke_label',
    'joke_id',
    'delivery',
    'rating'
  ),
  description = c("Participant identification number",
                  "Joke presented",
                  "Joke identification number",
                  "Experimental manipulation: whether joke was presented in audio-only ('audio') or in audiovideo ('video')",
                  "Humour rating chosen on a slider from 0 to 100")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

These data are simulated to imitate an experiment that investigates the effect of visual non-verbal communication (i.e., gestures, facial expressions) on joke appreciation.
Ninety participants took part in the experiment, in which they each rated how funny they found a set of 30 jokes.
For each participant, the order of these 30 jokes was randomised for each run of the experiment.
For each participant, the set of jokes was randomly split into two halves, with the first half being presented in audio-only, and the second half being presented in audio and video.
This meant that each participant saw 15 jokes with video and 15 without, and each joke would be presented with video roughly half of the time.
:::

**From the Week 1 lab, here's the key information about `laughs`:**

- Outcome: `rating`
- Predictors: `delivery`
- Randomly-varying grouping variables: `ppt` and `joke_label`/`joke_id` (two forms of the same information; for simplicity we can just use `joke_id`)

:::

<!-- q7 -->

`r qbegin(qcounter())`

Write the R formula for the fixed effects part of this model.

:::dapr3green
🗂️
See [Fixed effects](https://uoepsy.github.io/dapr3/2627/flashcards/block1/fixef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
rating ~ delivery
```


`r solend()`

<!-- q8 -->

`r qbegin(qcounter())`

We already know what random intercepts the model must have (one per randomly-varying grouping variable).

Identify the possible random slopes.

:::dapr3green
🗂️
See [Identify possible random effects](https://uoepsy.github.io/dapr3/2627/flashcards/block1/identify-ranef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
laughs <- read_csv('https://uoepsy.github.io/data/lmm_laughs.csv')
```

**Do at least some levels of `ppt` appear with more than one distinct value of `delivery`?**
**Yes,** every `ppt` sees both kinds of `delivery`.

```{r}
xtabs(~ ppt + delivery, data = laughs)
```

Because each participant has seen each kind of delivery, we can include random slopes over `delivery` by `ppt`.

<br>

**Do at least some levels of `joke_id` appear with more than one distinct value of `delivery`?**
**Yes,** every `joke_id` is told with both kinds of `delivery`.

```{r}
xtabs(~ joke_id + delivery, data = laughs)
```

Because each joke was told in each kind of delivery, we  can include random slopes over `delivery` by `joke_id`.

`r solend()`

<!-- q9 -->

`r qbegin(qcounter())`

Write out the complete R formula for this maximal model.

:::dapr3green
🗂️
See [Add random effects to a model formula](https://uoepsy.github.io/dapr3/2627/flashcards/block1/identify-ranef.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```
rating ~ delivery + (1 + delivery | ppt) + (1 + delivery | joke_id)
```

`r solend()`



## Practice interpreting the LMM summary

<!-- q10 -->

`r qbegin(qcounter())`

**Fit the LMM that you have developed for the `laughs` dataset, using the maximal model formula you wrote for Q9.**
(You can just use R's default treatment coding for `delivery`, so that `audio` is the reference level and `video` is the non-reference level.)

Use the model summary to respond to the following questions.

**Fixed effects:**

a. What does the model's intercept mean?
b. What does the coefficient `deliveryvideo` mean?


**By-participant random effects:**

c. Within what range will approximately 95% of participant-level intercepts fall?
d. Within what range will approximately 95% of participant-level slopes over `delivery` fall? 
e. Could some participants show an effect of `delivery` that goes in the opposite direction than the fixed effect?


**By-joke random effects:**

f. Within what range will approximately 95% of joke-level intercepts fall?
g. Within what range will approximately 95% of joke-level slopes over `delivery` fall?
h. Could some jokes elicit an effect of `delivery` that goes in the opposite direction than the fixed effect?

:::dapr3green
🗂️
See [Interpret LMM summary](https://uoepsy.github.io/dapr3/2627/flashcards/block1/interpret-lmm.html) flash card.
:::

`r qend()`

`r solbegin(label="– Model code and summary", show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
joke_mod <- lmer(
  rating ~ delivery + 
    (1 + delivery | joke_id) +
    (1 + delivery | ppt), 
  data = laughs
)

summary(joke_mod)
```



`r solend()`

`r solbegin(label="– Fixed effects", show=params$SHOW_SOLS, toggle=params$TOGGLE)`

a. **What does the model's intercept mean?**

- The average estimated humour rating when jokes were presented in audio-only format (the reference level of `delivery`, coded as 0) is 39.02 points.


b. **What does the coefficient `deliveryvideo` mean?**

- On average, presenting a joke in audio-video format increases its humour rating by an estimated 1.41 points compared to audio-only presentation.
- This change is significantly different from zero ($p$ = 0.008).

`r solend()`

`r solbegin(label="– By-participant random effects", show=params$SHOW_SOLS, toggle=params$TOGGLE)`

c. **Within what range will approximately 95% of participant-level intercepts fall?**

- The fixed intercept is 39.02 points, and the SD of participant-level intercept adjustments is 3.51 points.
- To get the approximate bounds of the central 95% of participant-level intercepts, use the following formulas.
- Lower bound: 

$$
\begin{align}
& ~\text{fixed estimate} - (2 \times \text{group-level SD}) \\
=&~ 39.02 - (2 \times 3.51) \\
=&~ 32.00 \\
\end{align}
$$

- Upper bound:

$$
\begin{align}
& ~\text{fixed estimate} - (2 \times \text{group-level SD}) \\
=&~ 39.02 + (2 \times 3.51) \\
=&~ 46.04 \\
\end{align}
$$

- Approximately 95% of participant-level intercepts will fall between 32 and 46.04 points.
  
:::{.callout-tip collapse='true'}
#### Calculate central 95% of participant-level intercepts with R
  
```{r}
fixed_int <- fixef(joke_mod)[['(Intercept)']]
by_ppt_int_sd <- attributes( VarCorr(joke_mod)$ppt )$stddev[['(Intercept)']]

c(
  lower = (fixed_int - 2*by_ppt_int_sd) |> round(2),
  upper = (fixed_int + 2*by_ppt_int_sd) |> round(2)
)
```
  
:::


d. **Within what range will approximately 95% of participant-level slopes over `delivery` fall? **

- The fixed slope over `delivery` is 1.41 points, and the SD of participant-level adjustments to the slope over `delivery` is 1.78 points.
- Lower bound: 

$$
\begin{align}
& ~\text{fixed estimate} - (2 \times \text{group-level SD}) \\
=&~ 1.41 - (2 \times 1.78) \\
=&~ -2.15 \\
\end{align}
$$

- Upper bound:

$$
\begin{align}
& ~\text{fixed estimate} + (2 \times \text{group-level SD}) \\
=&~ 1.41 + (2 \times 1.78) \\
=&~ 4.97 \\
\end{align}
$$

- Approximately 95% of participant-level slopes over `delivery` will fall between –2.15 and 4.97 points.

:::{.callout-tip collapse='true'}
#### Calculate central 95% of participant-level intercepts with R

```{r}
fixed_slp <- fixef(joke_mod)[['deliveryvideo']]
by_ppt_slp_sd <- attributes( VarCorr(joke_mod)$ppt )$stddev[['deliveryvideo']]

c(
  lower = (fixed_slp - 2*by_ppt_slp_sd) |> round(2),  
  upper = (fixed_slp + 2*by_ppt_slp_sd) |> round(2)
)
```
  
The R code yields slightly different values because it takes into account a lot of decimal places, while we only used two, so the difference is due to rounding error.
This is not a big deal (the values are approximations anyway), so either version is acceptable.

:::



e. **Could some participants show an effect of `delivery` that goes in the opposite direction than the fixed effect?**

- The fixed effect is positive, but some participants are estimated to show a negative slope.
- We know this because the lower bound of the range where 95% of participant-level slopes fall is negative (the opposite direction from the fixed effect).
- So: yes! It's estimated that some people find audio-only presentation to be funnier than audio-video presentation.

`r solend()`


`r solbegin(label="– By-joke random effects", show=params$SHOW_SOLS, toggle=params$TOGGLE)`

f. **Within what range will approximately 95% of joke-level intercepts fall?**


- The fixed intercept is 39.02 points, and the SD of joke-level intercept adjustments is 2.15 points.
- Lower bound: 

$$
\begin{align}
& ~\text{fixed estimate} - (2 \times \text{group-level SD}) \\
=&~ 39.02 - (2 \times 2.15) \\
=&~ 34.72 \\
\end{align}
$$

- Upper bound:

$$
\begin{align}
& ~\text{fixed estimate} - (2 \times \text{group-level SD}) \\
=&~ 39.02 + (2 \times 2.15) \\
=&~ 43.32 \\
\end{align}
$$

- Approximately 95% of joke-level intercepts will fall between 34.72 and 43.32 points.

:::{.callout-tip collapse='true'}
#### Calculate central 95% of participant-level intercepts with R
  
```{r}
fixed_int <- fixef(joke_mod)[['(Intercept)']]
by_joke_int_sd <- attributes( VarCorr(joke_mod)$joke_id )$stddev[['(Intercept)']]

c(
  lower = (fixed_int - 2*by_joke_int_sd) |> round(2),
  upper = (fixed_int + 2*by_joke_int_sd) |> round(2)
)
```
  
:::



g. **Within what range will approximately 95% of joke-level slopes over `delivery` fall?**

- The fixed slope over `delivery` is 1.41 points, and the SD of joke-level adjustments to the slope over `delivery` is 2.24 points.
- Lower bound: 

$$
\begin{align}
& ~\text{fixed estimate} - (2 \times \text{group-level SD}) \\
=&~ 1.41 - (2 \times 2.24) \\
=&~ -3.07 \\
\end{align}
$$

- Upper bound:

$$
\begin{align}
& ~\text{fixed estimate} + (2 \times \text{group-level SD}) \\
=&~ 1.41 + (2 \times 2.24) \\
=&~ 5.89 \\
\end{align}
$$

- Approximately 95% of joke-level slopes over `delivery` will fall between –3.07 and 5.89 points. 
- This range is wider for jokes than for participants, so the effect of `delivery` varies more from joke to joke than from person to person.

:::{.callout-tip collapse='true'}
#### Calculate central 95% of participant-level intercepts with R

```{r}
fixed_slp <- fixef(joke_mod)[['deliveryvideo']]
by_joke_slp_sd <- attributes( VarCorr(joke_mod)$joke_id )$stddev[['deliveryvideo']]

c(
  lower = (fixed_slp - 2*by_joke_slp_sd) |> round(2),  
  upper = (fixed_slp + 2*by_joke_slp_sd) |> round(2)
)
```

:::


h. **Could some jokes elicit an effect of `delivery` that goes in the opposite direction than the fixed effect?**

- The fixed effect is positive, but some jokes are estimated to show a negative slope.
- We know this because the lower bound of the range where 95% of joke-level slopes fall is negative (the opposite direction from the fixed effect).
- So: yes! It's estimated that some jokes are rated as funnier in audio-only format than they are in audio-video format.



`r solend()`
