---
title: "W4 Exercises: Centering"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
xaringanExtra::use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
library(lme4)
```


# Exercises: Centering in the MLM

:::frame
__Data: Hangry__

The study is interested in evaluating whether hunger influences peoples' levels of irritability (i.e., "the hangry hypothesis"). 81 participants were recruited into the study. Once a week for 5 consecutive weeks, participants were asked to complete two questionnaires, one assessing their level of hunger, and one assessing their level of irritability. The time and day at which participants were assessed was at a randomly chosen hour between 7am and 7pm each week.   

The data are available at: [https://uoepsy.github.io/data/hangry.csv](https://uoepsy.github.io/data/hangry.csv).  

```{r}
#| echo: false
hangry<-read_csv("https://uoepsy.github.io/data/hangry.csv") %>% mutate(fivetwo=factor(fivetwo))
tibble(
    variable = names(hangry),
    description = c("Score on irritability questionnaire (0:100)",
                    "Score on hunger questionnaire (0:100)",
                    "Participant",
                    "Whether the participant follows a five-two diet")
) %>% knitr::kable()


#hangry <- hangry |> filter(fivetwo==0)
```

:::



`r qbegin(qcounter())`
model spec
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`


`r qbegin(qcounter())`

::: {.callout-note collapse="true"}
## Total, Within, Between

Recall our research aim: 

> **... whether hunger influences peoples' levels of irritability (i.e., "the hangry hypothesis")**, and whether this is different for people following a diet that includes fasting.  

Forgetting about any differences due to diet, let's just think about the relationship between irritability and hunger. How should we interpret this research aim?  
Was it: 

a. "Are people more irritable if they are, __on average__, more hungry __than other people__?"  
b. "Are people more irritable if they are, __for them__, more hungry __than they usually are__?"  
c. Some combination of both a. and b.   

This is just one demonstration of how the statistical methods we use can constitute an integral part of our development of a research project, and part of the reason that data analysis for scientific cannot be so easily outsourced after designing the study and collecting the data.  

As our data currently is currently stored, the relationship between `irritability` and the raw scores on the hunger questionnaire `q_hunger` represents some 'total effect' of hunger on irritability. This is a bit like interpretation __c.__ above - it's a composite of both the 'within' ( __b.__ ) and 'between' ( __a.__ ) effects. The problem with this is that the 'total effect' isn't necessarily all that meaningful. It may tell us that 'being higher on the hunger questionnaire is associated with being more irritable', but how can we apply this information? It is not specifically about the comparison between hungry people and less hungry people, and nor is it about how person i changes when they are more hungry than usual. It is both these things smushed together.  

To disaggregate the 'within' and 'between' effects of hunger on irritability, we can group-mean center. For 'between', we are interested in how irritability is related to the average hunger levels of a participant, and for 'within', we are asking how irritability is related to a participants' _relative levels_ of hunger (i.e., how far above/below their average hunger level they are.).  

:::

Add to the data these two columns: 

1. a column which contains the average hungriness score for each participant.
2. a column which contains the deviation from each person's hunger score to that person's average hunger score. 

::: {.callout-tip collapse="true"}
#### Hints
You'll find `group_by() |> mutate()` very useful here. 
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
hangry <- 
    hangry |> group_by(ppt) |>
        mutate(
            avg_hunger = mean(q_hunger),
            hunger_gc = q_hunger - avg_hunger
        )
head(hangry)
```

`r solend()`


`r qbegin(qcounter())`
For each of the new variables you just added, plot the irritability scores against those variables.  

- Does it look like hungry people are more irritable than less hungry people?  
- Does it look like when people are more hungry than normal, they are more irritable? 

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
We might find it easier to look at a plot where each participant is represented as their mean plus an indication of their range of irritability scores:  
```{r fig.asp=.5}
ggplot(hangry,aes(x=avg_hunger,y=q_irritability))+
    stat_summary(geom="pointrange")
```
There appears to be a slight positive relationship between a persons' average hunger and their irritability scores. 

It is harder to tell what the relationship is between participant-centered hunger and irritability, because there are a lot of different lines (one for each participant). To make it easier to get an idea of what's happening, we'll make the plot fit a simple lm() (a straight line) for each participants' data:  

```{r include=FALSE}
# ggplot(hangry,aes(x=hunger_gc,y=q_irritability, group=ppt))+
#   geom_point(alpha = .2) + 
#   geom_smooth(method=lm, se=FALSE, lwd=.2)
```

```{r}
ggplot(hangry,aes(x=hunger_gc,y=q_irritability, color=ppt)) +
  geom_point(alpha = .2) + 
  geom_smooth(method=lm, se=FALSE, lwd=.2) +
  theme(legend.position = 'none')
```

I think there might be a positive trend in here, in that participants tend to be higher irritability when they are higher (for them) on the hunger score. 

`r solend()`



`r qbegin(qcounter())`

We have taken the raw hunger scores and separated them into two parts (raw hunger scores = participants' average hunger score + observation level deviations from those averages), that represent two different aspects of the relationship between hunger and irritability.  

Adjust your model specification to include these two separate variables as predictors, instead of the raw hunger scores.  



::: {.callout-tip collapse="true"}
#### Hints

- `hunger * diet` could be replaced by `(hunger1 + hunger2) * diet`, thereby allowing each aspect of hunger to interact with diet. 
- We can only put one of these variables in the random effects `(1 + hunger | participant)`. Recall that above we discussed how we cannot have `(diet | participant)`, because "an effect of diet" makes no sense for a single participant (they are either on the diet or they are not, so there is no 'effect'). Similarly, each participant has only one value for their average hungriness.  

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(lme4)
hangrywb <- lmer(q_irritability ~ (avg_hunger + hunger_gc)* fivetwo + 
                (1 + hunger_gc | ppt), 
                data = hangry,
                control = lmerControl(optimizer="bobyqa"))
```

`r solend()`




`r qbegin(qcounter())`
interpret each coefficient
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`


`r qbegin(qcounter())`
add interaction
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`


`r qbegin(qcounter())`
interpret
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`

`r qbegin(qcounter())`
write-up
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`









`r qbegin(1)`
> The study is interested in evaluating whether hunger influences peoples' levels of irritability (i.e., "the hangry hypothesis"). 

Read carefully the description of the study above, and try to write out (in `lmer` syntax) an appropriate model to test the research aims.  
e.g.:  
```{}
outcome ~ explanatory variables + (???? | grouping)
```
Try to think about the maximal random effect structure (i.e. everything that can vary by-grouping is estimated as doing so). 

To help you think through the steps to get from a description of a research study to a model specification, think about your answers to the following questions. 

Q: What is our outcome variable?  
Q: What are our explanatory variables?  
Q: Is there any grouping (or "clustering") of our data that we consider to be a random sample? If so, what are the groups?  

::: {.callout-tip collapse="true"}
#### Hints

- The research is looking at how hunger influences irritability
- We can split our data in to groups of each participant. We can also split it into groups of each diet. Which of these groups have we randomly sampled? Do we have a random sample of participants? Do we have a random sample of diets? Another way to think of this is "if i repeated the experiment, what these groups be different?"

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Our outcome is irritability here, because it is the thing that we are trying to explain through peoples' hunger levels and diets.  

```{r}
#| eval: false
lmer(irritability ~  explanatory variables + (???? | grouping))
```

We are interested in the effect of hunger on irritability, and whether this effect is different for the five-two diet. 
So we are interested in the interaction: 
```{r}
#| eval: false
lmer(irritability ~  hunger + diet + hunger:diet + (???? | grouping))
```
(remember that `hunger + diet + hunger:diet` is just a more explicit way of writing `hunger*diet`). 

If we did this experiment again, would we have different participants?  
Yes. If we did this experiment again, would we have different diets? No, because we're interested in the specific differences between the five-two diet and no dieting. This means we will likely want to by-participant random deviations (e.g. the `( ... | participant)` bit in `lmer`). But we won't have by-diet random effects `(1 | diet)` because the diet differences are the specific differences that we wish to test.  
```{r}
#| eval: false
lmer(irritability ~  hunger + diet + hunger:diet + (???? | participant))
```

Thinking about what __can__ be modelled as randomly varying between participants, we have some options:

1. participants vary in how irritable they are on average   
(the intercept, `1 | participant`)
2. participants vary in how much hunger influences their irritability   
(the effect of hunger, `hunger | participant`)
3. participants vary in how much diet influences irritability   
(the effect of diet, `diet | participant`)
4. participants vary in how much diet effects hunger's influence on irritability   
(the interaction between diet and hunger, `diet:hunger | participant`)

We can vary 1 and 2, but not 3 and 4. This is because each participant is _either_ following the five-two diet _or_ they are not. So for a _single_ participant, we can't assess "the effect diet has" on anything, because we haven't seen that participant under different diets. if we try to plot a single participants' data, we can see that it is impossible for us to assess "the effect of diet":
```{r}
#| echo: false
#| fig-height: 2.5
hangry %>% filter(ppt == "N1p2") %>%
    ggplot(., aes(x=fivetwo, y=q_irritability))+
    geom_point()+
    geom_smooth(method="lm",se=F)+
    scale_x_discrete(drop=FALSE)+
  labs(title="The 'effect of diet' for a single\nparticipant from the Hangry study",subtitle="cannot be defined")
```

By contrast, we __can__ vary the intercept and the effect of hunger, because each participant has multiple values of irritability, and multiple different observations of hunger. We can think about a single participant's "effect of hunger on irritability" and how we might fit a line to their data:
```{r}
#| echo: false
#| fig-height: 2.5
hangry %>% filter(ppt == "N1p2") %>%
    ggplot(., aes(x=q_hunger, y=q_irritability))+
    geom_point()+
    geom_smooth(method="lm",se=F)+
  labs(title="The 'effect of hunger' for a single\nparticipant from the Hangry study")
```

```{r}
#| eval: false
lmer(irritability ~  hunger + diet + hunger:diet + (1 + hunger | participant))
```

`r solend()`



`r qbegin(5)`
Hopefully, you have fitted a model similar to the below:  
```{r}
hangrywb <- lmer(q_irritability ~ (avg_hunger + hunger_gc) * fivetwo + 
                (1 + hunger_gc | ppt), data = hangry,
            control = lmerControl(optimizer="bobyqa"))
```

Below, we have obtained p-values using the Kenward Rogers Approximation of $df$ for the test of whether the fixed effects are zero, so we can see the significance of each estimate.  

Provide an answer for each of these questions:

1. For those following no diet, is there evidence to suggest that people who are _on average_ more hungry are more irritable?
2. Is there evidence to suggest that this is different for those following the five-two diet? In what way?
3. Do people following no diet tend to be more irritable when they are more hungry than they usually are? 
4. Is there evidence to suggest that this is different for those following the five-two diet? In what way?
5. __(Trickier:)__ What does the `fivetwo` coefficient represent? 

```{r}
#| echo: false
library(parameters)
model_parameters(hangrywb, ci_method = "kr", ci_random = FALSE) |>
  print_html()
```
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


**1: For those following no diet, is there evidence to suggest that people who are _on average_ more hungry are more irritable?**  
A: 'No diet' is the reference level of the five-two variable, and because we have an interaction, that means the `avg_hunger` coefficient will provide the relevant estimate. There is no evidence ($p>.05$) to suggest that when not dieting, hungrier people are more irritable than less hungry people. 


**2: Is there evidence to suggest that this is different for those following the five-two diet? In what way?**  
A: This is the interaction between `avg_hunger:fivetwo1`. We can see that, for every increase of 1 in average hunger, irritability is estimated to increase by 0.47 more for those in the five-two diet than it does for those following no diet.  
These units are still in terms of the original scale (i.e. 0 to 100). 


**3: Do people following no diet tend to be more irritable when they are more hungry than they usually are?** 
A: This is the estimate for the coefficient of `hunger_gc`. For people following no diet, there is an estimated 0.19 increase in irritability for every 1 unit more hungry they become. 

**4: Is there evidence to suggest that this is different for those following the five-two diet? In what way?**
A: This effect of a 1 unit change on within-person hunger increasing irritability is increased for those who are following the five-two diet by an additional 0.38

**5: What does the `fivetwo1` coefficient represent?** 
A: This represents the group difference of irritability between those on the five-two diet vs those not dieting, for someone who has an average hunger score of 0. 

`r solend()`

`r qbegin(6)`
Construct two plots showing the two model estimated interactions. Think about your answers to the previous question, and check that they match with what you are seeing in the plots (do not underestimate the utility of this activity for helping understanding!).  

::: {.callout-tip collapse="true"}
#### Hints

This isn't as difficult as it sounds. the sjPlot package can do it in one line of code!

:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(sjPlot)
plot_model(hangrywb, type = "int")[[1]]
```
We saw in the model coefficients that for the reference level of `fivetwo`, the "No Diet" group, there was no association between how hungry a person is _on average_ and their irritability. This is the red line we see in the plot above. 
We also saw the interaction `avg_hunger:fivetwo1` indicates that irritability is estimated to increase by 0.47 more for those in the five-two diet than it does for those following no diet. So the blue line is should be going up more steeply than the red line (which is flat). And it is! 

```{r}
plot_model(hangrywb, type = "int")[[2]]
```
From the coefficient of `hunger_gc` we get the estimated amount by which irritability increases for every 1 more hungry that a person becomes (when they're on "No Diet"). This is the slope of the red line. 
The interaction `hunger_gc:fivetwo1` gave us the adjustment to get from the red line to the blue line. It is positive and significant, which matches with the fact that the blue line is clearly steeper in this plot. 

`r solend()`

`r qbegin(7)`
Provide tests or confidence intervals for the parameters of interest, and write-up the results.  

::: {.callout-tip collapse="true"}
#### Remember: some options for inference


|                  | df approximations                                                  | likelihood-based                                                    | 
| ---------------- | ------------------------------------------------------------------ | ------------------------------------------------------------------- | 
| tests or CIs for model parameters | `library(parameters)`<br>`model_parameters(model, ci_method="kr")` | `confint(model, type="profile")`                                    | 
| model comparison<br><small>(different fixed effects, same random effects)</small> | `library(pbkrtest)`<br>`KRmodcomp(model1,model0)`                  | `anova(model0,model)`                                               |
|                  | fit models with `REML=TRUE`.<br>good option for small samples      | fit models with `REML=FALSE`.<br>needs large N at both levels (40+) | 

:::
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(parameters)
model_parameters(hangrywb, ci_method = "kr", ci_random = FALSE)
```

```{r}
#| include: false
res = as.data.frame(model_parameters(hangrywb, ci_method = "kr", ci_random = FALSE))
res[,2:8] <- apply(res[,2:8],2,function(x) round(x,2))
res[,9] <- format.pval(res[,9],digits=1, eps=.001)
res[,9][!grepl("<",res[,9])] <- paste0("=",res[,9][!grepl("<",res[,9])])
res
```

To investigate the association between irritability and hunger, and whether this relationship is different depending on whether or not participants are on a restricted diet such as the five-two, a multilevel linear model was fitted.  
To disaggregate between the differences in irritability due to people being in general more/less hungry, and those due to people being more/less hungry than usual for them, irritability was regressed onto both participants' average hunger scores their relative hunger levels. Both of these were allowed to interact with whether or not participants were on the five-two diet. Random intercepts and slopes of relative-hunger level were included for participants. The model was fitting with restricted maximum likelihood estimation with the **lme4** package (Bates et al., 2015), using the _bobyqa_ optimiser from the **lme4**. $P$-values were obtained using Wald tests with Kenward-Roger approximation of denominator degrees of freedom.  

Results indicate that for people on no diet, being more hungry than normal was associated with greater irritability ($\beta = `r res[3,2]`,\ SE = `r res[3,3]`,\ t(`r res[3,7]`) = `r res[3,8]`,\ p`r res[3,9]`$), and that this was increased for those following the five-two diet ($\beta = `r res[6,2]`,\ SE = `r res[6,3]`,\ t(`r res[6,7]`) = `r res[6,8]`,\ p`r res[6,9]`$). 
Although for those not on a specific diet there was no evidence for an association between irritability and being generally a more hungry person ($p`r res[2,9]`$), there a significant interaction was found between average hunger and being on the five-two diet ($\beta = `r res[5,2]`,\ SE = `r res[5,3]`,\ t(`r res[5,7]`) = `r res[5,8]`,\ p`r res[5,9]`$), suggesting that when dieting, hungrier people tend to be more irritable than less hungry people.  
Results suggest that the 'hangry hypothesis' may occur _within_ people (when a person is more hungry than they usually are, they tend to be more irritable), but not necessarily between hungry/less hungry people. Dieting was found to increase the association of both between-person hunger and within-person hunger with irritability.  
`r solend()`




