---
title: "LMMs: Practice write-up"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE  # TRUE = collapse sols, FALSE = uncollapse sols
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio/  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal/with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
library(stats)
library(lme4)
library(lmerTest)
library(HLMdiag)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

In this lab, you'll be guided through how to report the LMM analysis we conducted together in lectures this week.

Your goal is for a reader to be able to replicate your analyses in any software they choose, based only on what you've written.

Make sure that your write-up contains no visible R code.
For a reminder of how to hide code, format tables, etc., check out the [DAPR1 formatting resources](https://uoepsy.github.io/dapr1/2526/labs/zf_formatting_resources.html).




:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab05.Rmd`).
:::

:::{.callout-tip}
### The RQ and the data

Study background:

- We're looking at emotional dysregulation in children: when children have trouble managing their feelings and emotions.
- Specifically, we want to find out how emotional dysregulation is affected by children having a strong, regular routine at home.
- We predict that children with stronger home routines are better at regulating their emotions.
- It's also plausible that a family's socio-economic status (SES) is associated with how well children regulate their emotions. 
We will control for SES using a proxy variable of whether children attend a state school or a private school.

**RQ: Controlling for the type of school that children attend, are children with stronger routines better at regulating their emotions?**

The data can be downloaded from <https://uoepsy.github.io/data/crqeds.csv>{target="_blank"}.

| variable   | description                                                                                       |
| ---------- | ------------------------------------------------------------------------------------------------- |
| `schoolid`   | School identifier                                                                                 |
| `EDS`        | Emotion Dysregulation Score (range 1 to 6, higher values indicate more *dys*regulation)  |
| `CRQ`        | Childhood Routine Questionnaire Score (range 0 to 7, higher values indicate more routine) |
| `schooltype` | School type (private / state)                                                                     |


:::

:::{.callout-note collapse='true'}
### Provided analysis code

```{r eval=F}
# TODO
```


:::


---

# q1

`r qbegin(qcounter())`

**Set the scene (so the reader knows what you're talking about).**

- [ ] Remind the reader briefly what the study aimed to test (e.g., what the predictions were).
- [ ] Explain what factors the study is manipulating and what the outcome variable is.

:::dapr3green
🗂️
See [Report LMM analyses and results](https://uoepsy.github.io/dapr3/2627/flashcards/block1/reporting.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`


# q2

`r qbegin(qcounter())`

**Describe and show the sample data (so that the reader knows how generalisable your analysis might be).**

- [ ] Mention any data cleaning / outlier removal you did before computing the descriptives.
- [ ] Describe the outcome and predictor variables in your data (e.g., what scale are they on? how were they measured?)
- [ ] Describe the grouping variables in your data (see [Describe group-structured data](describe.html)).
  - Tip: Tables and plots are easier to understand than listing loads of numbers in the text (and better for your word count too!). 
  - Tip: Plots are easier to understand than tables of numbers.

:::dapr3green
🗂️
See [Describe group-structured data](https://uoepsy.github.io/dapr3/2627/flashcards/block1/describe.html) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`


# q3

`r qbegin(qcounter())`

**Visualise and summarise the results that address the RQ (so the reader knows the basic pattern of what you found).**

- [ ] Include a plot of the data (and make sure to refer to it in the text).
  - Tip: If you transform any continuous predictors for analysis, use the transformed version of the predictor in the plot.
- [ ] In a sentence or two, describe the overall pattern of results.
  - Tip: No inferential stats have been done yet, so best not to make any claims yet about whether predictions have been supported or not.

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`


# q4

`r qbegin(qcounter())`

**Introduce your statistical analysis (so the reader can interpret the numbers you will report).**

- [ ] Describe the fixed effects of your model, e.g.: "We fit a linear mixed model predicting Y as a function of X1, X2, and their interaction".
- [ ] If categorical predictors, describe contrast coding, e.g.: "X1 has two levels, A (coded as –1) and B (coded as 1)".
- [ ] If continuous predictors have been transformed, describe the transformation, e.g.: "X2 has been z-scored".
- [ ] Describe the random effect structure of the maximal model.
  - One option using "random" terminology (more traditional): "The maximal model includes by-participant random intercepts, by-participant random slopes over X1, and by-stimulus random intercepts."
  - Another option using "adjustment" terminology (less traditional but clearer): "The maximal model includes by-participant adjustments to the intercept and to the slope over X1, as well as by-stimulus intercept adjustments."
- [ ] If you had to simplify the random effect structure, describe the order in which you removed parameters.
- [ ] Describe the random effect structure of the actual model that you will report.
- [ ] State that you're running your analysis using R, and cite the developers (run `citation()` in the console to get the citation for your version of R). 
- [ ] State the software packages and versions used to fit models, along with the estimation method (ML/REML), how p-values were approximated (probably Satterthwaite), what optimiser you used... 
- [ ] State which model coefficient(s) are related to which prediction(s), and what direction/sign (positive or negative) you would expect the coefficients to have if the predictions were borne out.

:::dapr3green
🗂️
See [Maximal model](https://uoepsy.github.io/dapr3/2627/flashcards/block1/maxmodel.html), [Simplify random effects](https://uoepsy.github.io/dapr3/2627/flashcards/block1/simplify.html) flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`


# q5

`r qbegin(qcounter())`

**Report the results of your analysis (so the reader knows exactly what you found).**

- [ ] Plot the model-fitted values.
- [ ] State the key coefficient estimates along with their 95% CI and p-value, e.g.: "The slope associating X1 and Y was significantly different from zero ($b$ = 0.26, 95% CI [0.23, 0.29], $p$ < 0.001)" or "The slope associating X1 and Y was not significantly different from zero ($b$ = 0.26, 95% CI [–0.04, 0.56], $p$ = 0.18)".
  - Tip: You don't need to report every single parameter estimate. Focus on the one(s) you identified as being relevant to your prediction(s).
  - Tip: Briefly link the direction and significance of the effect back to your earlier predictions—is the prediction supported by the parameter estimate?
- [ ] Include a table of all coefficients estimated by the linear model (and make sure to refer to it in the text).
- [ ] Report the magnitude of the SD of the group-level adjustments and compare it to the magnitude of the fixed effects. How does the intercept vary between groups? How does the effect of interest vary between groups?
- [ ] Report the correlations between intercept and slope adjustments.
- [ ] Report any other information you consider important for your story!


:::dapr3green
🗂️
See [Plot model-fitted values](https://uoepsy.github.io/dapr3/2627/flashcards/block1/reporting.html), [Interpret LMM summary](https://uoepsy.github.io/dapr3/2627/flashcards/block1/interpret-lmm.html) flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

TODO

`r solend()`
