---
title: "EFA"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE  # TRUE = collapse sols, FALSE = uncollapse sols
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio/  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal/with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}

bloads <- read.csv("~/Downloads/bcopeloads.csv") |>
  mutate(across(everything(),~gsub("−","-",.)))
bloads$q <- paste0("q",bloads$qitem)
bloads[1,c(4:9)] <- bloads[grepl("distraction 2",bloads$subscale),4:9]
bloads <- bloads |> arrange(qitem)

# copeitems[c(2, 7, 10, 12, 14, 17, 23, 25)]
# copeitems[c(5, 9, 13, 15, 18, 20, 21, 22, 24, 26, 27, 28)]
# copeitems[c(1, 3, 4, 6, 8, 11, 16, 19)]

dgp = paste0(
  "F1 =~ ", paste0(bloads[,4],"*",bloads$q, collapse="+"),"\n",
  "F2 =~ ", paste0(bloads[,5],"*",bloads$q, collapse="+"),"\n",
  "F3 =~ ", paste0(bloads[,6],"*",bloads$q, collapse="+"),"\n",
  "F4 =~ ", paste0(bloads[,7],"*",bloads$q, collapse="+"),"\n",
  "F5 =~ ", paste0(bloads[,8],"*",bloads$q, collapse="+"),"\n",
  "F6 =~ ", paste0(bloads[,9],"*",bloads$q, collapse="+"),"\n",
  "F6 ~~ -.3*F1 \n
   F1 ~~ .2*F3 \n F1 ~~ .1*F2 \n F2 ~~ .2*F5 \n F3 ~~ -.2*F6 \n
   F4 ~~ -.25*F6")

set.seed(738784)
df <- lavaan::simulateData(dgp, sample.nobs = 620)
df$q20 <- residuals(lm(df$q20 ~ rowSums(df[,c("q10","q15","q5","q23")])))
df$q12 <- df$q12 + .5*df$q28
df <- df |>
  apply(2, \(x) cut(x,5,labels=FALSE)) |>
  as.data.frame()

bcdata <- as.data.frame(df)[,c(1,12,22:28,2:11,13:21)] |>
  mutate(
    pptID = paste0("PPTD",1:n()), .before = q1
  )

```

:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab08.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
    - `psych`
    
:::

---

:::frame
__copingstrats.csv__  

```{r}
#| echo: false
bloads |> 
  arrange(as.numeric(qitem)) |>
  transmute(variable = q, wording) |>
  gt::gt()
```


:::

`r qbegin(qcounter())`
Read in the data and check the suitability of the items for EFA.  

Things to look at: correlation matrix, bartlettt's test, KMO

see fashcard


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(tidyverse)
library(psych)
#bcdata <-read_csv()
copeitems <- bcdata |> select(-pptID)
```


BIG!
```{r}
cor(copeitems, )
```

```{r}
cor(copeitems) |>
  heatmap()
```

```{r}
bartlett.test(copeitems)
```

```{r}
KMO(copeitems)
```

`r solend()`

`r qbegin(qcounter())`
Our set of items is attempting to measure peoples' coping strategies.  

Look at the item wordings and description, and think about the following questions: 

1. on what scale were the items measured? How many levels are there? 
2. do the distributions of the items look roughly normally distributed? 
3. if there is just one dimension to this measure (i.e., 1 factor), then what would it mean to be high on that factor?  
4. just looking at wordings, are there possibly different groupings of items? 
5. if there are groups of items for which responses are being driven by separate latent factors, what would it mean to be low/high on those factors? 
6. Would we expect a person's standing on one of these factors to be similar/opposing/unrelated to their standing on the other factor(s)?

These questions will help us to fill in these bits: 
```
fa(..........., cor = ???, rotate = ???, fm = ???)
```

If we have enough levels to our measurement scale, we can keep `cor = "cor"`.  
If we are worried about normality, we can use `fm = "pa"`, otherwise we can use `fa = "minres"` or (with a large sample size, which we _do_ have here) `fm = "ml"`
If this is multi-dimensional (i.e., if there is more than just one latent factor that explains peoples' responses to these questions), then would we expect those dimensions to be related? Do we _want_ to constrain them to be orthogonal? If not, then `rotate = "oblimin"` is a good shout.  

see flashcard

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Every item has 5 response options, and things look more or less normal (hard to see with Likert data, but we're not too skewed with lots of responses of 1s or 5s)
```{r}
multi.hist(copeitems)
```

```
fa(..........., cor = "cor", rotate = "oblimin", fm = "ml")
```

`r solend()`

`r qbegin(qcounter())`
Define a range for the number of factors that we will extract.  

see flashcard

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

| method | suggestion |
| ---- | -----|
| scree plot | 3? or maybe 6? |
| parallel analysis | 6 |
| MAP | 3 | 

Let's examine from 2 to 6 factors.  

```{r}
scree(copeitems)
fa.parallel(copeitems)
VSS(copeitems, plot=F)
```

`r solend()`

`r qbegin(qcounter())`
For every number in your range, fit a factor model.  

flashcard
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
copefa3 <- fa(copeitems, nfactors = 3, rotate = "oblimin", fm = "ml", cor = "cor")
copefa4 <- fa(copeitems, nfactors = 4, rotate = "oblimin", fm = "ml", cor = "cor")
copefa5 <- fa(copeitems, nfactors = 5, rotate = "oblimin", fm = "ml", cor = "cor")
copefa6 <- fa(copeitems, nfactors = 6, rotate = "oblimin", fm = "ml", cor = "cor")
```

`r solend()`

`r qbegin(qcounter())`
**V - variance**  

How much of the total variance in the coping strategy items is explained by each factor solution?  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#| echo: false
tribble(
  ~solution, ~`variance explained`,
  "3 factor", round(copefa3$Vaccounted[3,3],2),
  "4 factor", round(copefa4$Vaccounted[3,4],2),
  "5 factor", round(copefa5$Vaccounted[3,5],2),
  "6 factor", round(copefa6$Vaccounted[3,6],2)
) |> gt::gt()
```


`r solend()`

`r qbegin(qcounter())`
**I - identification**  

For each solution, do all factors have 3+ items for which the factor is the primary loading?  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#| echo: false
tribble(
  ~solution, ~`all factors identified`,
  "3 factor", "yes",
  "4 factor", "yes",
  "5 factor", "one factor (ML1) has only 2 items",
  "6 factor", "two factors (ML3, ML2) only have 2 items"
) |> gt::gt()
```


`r solend()`

`r qbegin(qcounter())`
**S - salience, C - complexity, H - heywood cases**  

Look at the patterns of loadings for each solution. Are there any items that:  

- do not load on to any factor at a salient level? 
- have high complexity or cross loadings across multiple factors?
- have impossible values (factor loadings or communalities > 1)?  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


::::panelset
:::panel
### 3 factor
```{r}
print(copefa3$loadings,cutoff=.3,sort=T)
```
:::
:::panel
### 4 factor
```{r}
print(copefa4$loadings,cutoff=.3,sort=T)
```
:::
:::panel
### 5 factor
```{r}
print(copefa5$loadings,cutoff=.3,sort=T)
```
:::
:::panel
### 6 factor
```{r}
print(copefa6$loadings,cutoff=.3,sort=T)
```
:::
::::


```{r}
#| echo: false
tribble(
  ~solution, ~`salience of items`, ~`complexity & cross loadings`, ~`heywood cases`,
  "3 factor", "4 items (11,22,27,4) have no salient loadings","close to problematic for qs 21, 14","no",
  "4 factor", "4 items (11,22,27,4) have no salient loadings", "cross loadings for qs 17 and 3", "no",
  "5 factor", "2 items (22,27) have no salient loadings","cross loadings for q3", "no",
  "6 factor", "no", "no", "no"
) |> gt::gt()
```

`r solend()`

`r qbegin(qcounter())`
Given your examination so far, some items may be flagged as consistently problematic. Take a look at the wordings of any such items. 
Are they "double barrelled" or ambiguously worded? Can you think of any theoretical reason why these items are causing problems?  

Here are the ones **we** identified.

- q4, q11, q22 and q27 were only salient when loading onto factors with <3 items (11 and 4 seemed to go together, as did 22 and 27).  
- q3 has some cross loadings in several solutions.  

Normally, we would want to remove items one at a time. This is a time-consuming process, so let's skip ahead, and suppose we need to remove all four of q4, q11, q22 and q27.  

**Task:** create a new dataframe which is the subset of items _without_ those four.  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
copeitems_sub <- copeitems |>
  select(-q4, -q11, -q22, -q27)
```

`r solend()`

`r qbegin(qcounter())`
Re-do the entire process!  

We've printed all the output so that you don't have to code it right here, and can focus on the evaluation of the different solutions.  

Which model do you prefer? Remember to take into consideration the wordings of the items!  
`r qend()`
`r solbegin(show=TRUE, label="1 - Checks", slabel=FALSE, toggle=params$TOGGLE)`

```{r}
cor(copeitems_sub) |>
  heatmap()
```

```{r}
bartlett.test(copeitems_sub)
```

```{r}
KMO(copeitems_sub)
```

`r solend()`
`r solbegin(show=TRUE, label="2 - How many factors?", slabel=FALSE, toggle=params$TOGGLE)`


| method | suggestion |
| ---- | -----|
| scree plot | 1? 4? |
| parallel analysis | 4 |
| MAP | 3 | 

```{r}
scree(copeitems_sub)
fa.parallel(copeitems_sub)
VSS(copeitems_sub, plot=F)
```


Let's examine from 2 to 5 factors.  

`r solend()`
`r solbegin(show=TRUE, label="3 - Competing models", slabel=FALSE, toggle=params$TOGGLE)`
```{r}
copefa1s <- fa(copeitems_sub, nfactors = 1, fm = "ml", cor = "cor")
copefa2s <- fa(copeitems_sub, nfactors = 2, rotate = "oblimin", fm = "ml", cor = "cor")
copefa3s <- fa(copeitems_sub, nfactors = 3, rotate = "oblimin", fm = "ml", cor = "cor")
copefa4s <- fa(copeitems_sub, nfactors = 4, rotate = "oblimin", fm = "ml", cor = "cor")
```

::::panelset
:::panel
### 1 factor
```{r}
print(copefa1s$loadings,cutoff=.3,sort=T)
```
:::
:::panel
### 2 factor
```{r}
print(copefa2s$loadings,cutoff=.3,sort=T)
```
:::
:::panel
### 3 factor
```{r}
print(copefa3s$loadings,cutoff=.3,sort=T)
```
:::
:::panel
### 4 factor
```{r}
print(copefa4s$loadings,cutoff=.3,sort=T)
```
:::
::::
`r solend()`

`r qbegin(qcounter())`

For your chosen solution, name the factors!  



`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


ML3 = reframing/distraction
ML4 = avoidant/maladaptive
ML1 = social support seeking
ML2 = active problem focused


`r solend()`

```{r}

newnames <- bloads$wording
names(newnames) <- bloads$q
names(copeitems_sub) <- newnames[names(copeitems_sub)]
copefa4s <- fa(copeitems_sub, nfactors = 4, rotate = "oblimin", fm = "ml", cor = "cor")
print(copefa4s$loadings,sort=T,cutoff=.3)

```







