---
title: "CFA"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE  # TRUE = collapse sols, FALSE = uncollapse sols
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio/  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal/with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
source("jk_playground/whoqol.r")
```

:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab09.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
    - `lavaan`
    - `semPlot`
    
:::

---

:::frame
__Quality Street__  

Our exercises today are concerning the factor structure of the "WHOQOL-BREF" - a brief measure of quality of life developed by the World Health Organisation (WHO). 
It contains 26 questions (the first two of which are treated individually as general single item measures). 
The remaining 24 questions captures four related aspects of quality of life: Physical, Psychological, Social and Environmental.  


In practice, these four domain scores are often highly correlated ($r > .70$). 
Researcher A is curious about whether this scale is genuinely measuring four distinct domains or simply reflecting one underlying construct (a "General Quality of Life").
She wants to test whether a complex 4-factor structure is strictly necessary to explain the responses, or whether a simpler 1-factor model is sufficient. She collects WHOQOL-BREF data from 300 adults aged 25 to 55.  
The wordings of the items can be found below.  
The dataset can be accessed at <https://uoepsy.github.io/data/whoqolbref-sampleA.csv>

::: {.callout-note collapse="true"}
#### Item Wordings

```{r}
#| echo: false
wqit |> gt::gt()
```

:::
:::

`r qbegin(qcounter())`
First, let's fit the factor model that is suggested by the WHO. We can then see if this factor structure provides a good fit to the responses from our sample. 
In the syntax for **lavaan**, specify the model. 
(We need to specify the model that links the specific questions to their relevant "domains")

```{r}
#| eval: false
#| echo: true

who_mod <- "
    ...
    ...
    
    ...
  "
```


:::dapr3green
🗂️
See the [#Specification section on the Fit a Confirmatory Factor Model](https://uoepsy.github.io/dapr3/2627/flashcards/block2/p03_cfa.html#specification]) flash card.
:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


```{r}
who_mod <- '
  Phys =~ q3 + q4 + q10 + q15 + q16 + q17 + q18
  Psych =~ q5 + q6 + q7 + q11 + q19 + q26
  Social =~ q20 + q21 + q22
  Env =~ q8 + q9 + q12 + q13 + q14 + q23 + q24 + q25
'
```


`r solend()`

`r qbegin(qcounter())`
Fit the model, and evaluate the global (overall) model fit.  

:::dapr3green
🗂️
See the [#Estimation section on the Fit a Confirmatory Factor Model](https://uoepsy.github.io/dapr3/2627/flashcards/block2/p03_cfa.html#estimation]) flash card.
:::
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
whqb_a <- read_csv("https://uoepsy.github.io/data/whoqolbref-sampleA.csv")

fit_who <- cfa(who_mod, data = whqb_a, std.lv = TRUE)

fitmeasures(fit_who)[c("rmsea","srmr","cfi","tli")]
```

`r solend()`

`r qbegin(qcounter())`
Examine the modification indices.  
Even if the model already "fits well", this can help to highlight possible issues with the measurement instrument.  
What might large residual covariances between two items tell us?  


:::dapr3green
🗂️
See the [#Re-specification section on the Fit a Confirmatory Factor Model](https://uoepsy.github.io/dapr3/2627/flashcards/block2/p03_cfa.html#re-specification]) flash card.
:::

::: {.callout-tip collapse="true"}
#### Hints

This will require looking carefully at the item wordings. What do the items with suggested residual covariances all have in common?  

:::
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
modindices(fit_who, sort = TRUE) |> 
  head(5)
```

You might notice that these top 3 - questions q3, q4, q26 - are the only negatively worded questions! Respondents often process negatively framed items with their specific cognitive bias that can creating residual correlation among them. But this is more noise due to the method than theoretically interesting domain variability.  

`r solend()`


`r qbegin(qcounter())`
When you fitted your model, which did you use? 
```{r}
#| eval: false
#| echo: true
cfa(model, data)
cfa(model, data, std.lv = TRUE)
```

To help us better understand how this changes things, fit the other one now, and examine what (if anything) is different.  

- global fit of the model
- modification indices
- in `summary()`:  
    - factor loading estimates
    - factor covariance estimates
    
**Bonus Challenge:** can you see the link between estimates from the two versions?  

:::dapr3green
🗂️
See the [#Parameter Estimates section on the Fit a Confirmatory Factor Model](https://uoepsy.github.io/dapr3/2627/flashcards/block2/p03_cfa.html#parameter-estimates]) flash card.
:::

    
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
fit_who <- cfa(who_mod, data = whqb_a, std.lv = TRUE)
fit_who2 <- cfa(who_mod, data = whqb_a)
```

The fit of the models should be identical: 
```{r}
fitmeasures(fit_who)[c("rmsea","srmr","cfi","tli")]
fitmeasures(fit_who2)[c("rmsea","srmr","cfi","tli")]
```

And the modification indices should also be indentical: 
```{r}
head(modindices(fit_who, sort = TRUE))
head(modindices(fit_who2, sort = TRUE))
```

The loadings will be different. 

- without `std.lv = TRUE`, the first loading is fixed to 1, and all the others are in comparison to that one
- with `std.lv = TRUE`, the loadings are all freely estimated but the latent variable has a variance of 1

For instance, the items loading on the social domain have these loadings when we use `std.lv = TRUE`: 0.576, 0.489, 0.510.  
And when we don't use it, we get: 1.000, 0.849, 0.885.  

To see how they match up: 
```{r}
c(0.576, 0.489, 0.510) / 0.576
```

The factor covariances are also different.  

- without `std.lv = TRUE`, they are covariances between latent variables that have scales fixed to the same as their first item. covariances are dependent on units, remember! 
- with `std.lv = TRUE`, the latent variables have variances of 1, so they are all standardised. so covariances become correlations!  


`r solend()`

`r qbegin(qcounter())`
How do the individual items look? How much variance in each item is explained by its corresponding factor?  

**Bonus Challenge:** Can you calculate the "average variance explained" for each factor? 

::: {.callout-tip collapse="true"}
#### Hints

When we have a standardised factor loading, it is just like EFA, where squaring it gives the variance in an item explained by a factor.  

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

I love making plots rather than looking at numbers, so here we go! 
```{r}
standardizedsolution(fit_who) |>
  filter(op == "=~") |>
  mutate(r2 = est.std^2) |>
  group_by(lhs) |>
  mutate(mean_r2 = mean(r2)) |>
  ggplot(aes(x = rhs, y = r2)) +
  geom_col() +
  geom_hline(aes(yintercept = mean_r2),
             linetype = "dashed", color = "red") +
  ylim(0,1)+
  coord_flip() + 
  facet_wrap(~lhs, scales = "free_y") 
```

`r solend()`


`r qbegin(qcounter())`
Make a quick diagram showing the standardised loadings for each factor.  

:::dapr3green
🗂️
See the [#Drawing your fitted model section on the Fit a Confirmatory Factor Model](https://uoepsy.github.io/dapr3/2627/flashcards/block2/p03_cfa.html#drawing-your-fitted-model]) flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(semPlot)

semPaths(fit_who, whatLabels = "std", rotation=2)
```

`r solend()`


`r qbegin(qcounter())`
Researcher A would now like to compare this to a model with only one single factor.  

Fit a unidimensional model and compare the fit measures.  
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
rA_mod <- '
  GQOL =~ q3 + q4 + q10 + q15 + q16 + q17 + q18 + q5 + q6 + q7 + q11 + q19 + q26 + q20 + q21 + q22 + q8 + q9 + q12 + q13 + q14 + q23 + q24 + q25
'

fit_a <- cfa(rA_mod, whqb_a, std.lv=TRUE)
fitmeasures(fit_a)[c("rmsea","srmr","cfi","tli")]
```

`r solend()`
 
 
 
 
 
 
 
:::frame
__Quality over Quantity__  

Researcher B is working on the same project, and wonders whether this factor structure holds up across different age demographics. 
Specifically, he wants to ask if the optimal factor structure identified in younger adults (25–55) holds true in older adults aged 65+?
Is the WHOQOL-BREF measurement "invariant" across these two age groups, or do older adults interpret and answer the items differently?  
To test this, Researcher B recruits a second sample of 300 adults aged 65 and above who complete the same 26 items.

The data are available at <https://uoepsy.github.io/data/whoqolbref-sampleB.csv>


:::
 
 
`r qbegin(qcounter())`
Fit the WHO 4 factor model to sample B (the sample of 300 adults >65).  

Does it fit well?  
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
whqb_b <- read_csv("https://uoepsy.github.io/data/whoqolbref-sampleB.csv")

fit_bwho <- cfa(who_mod, whqb_b, std.lv = TRUE)

fitmeasures(fit_bwho)[c("rmsea","srmr","cfi","tli")]
```

`r solend()`

`r qbegin(qcounter())`
Compare the loadings of the models fitted to the 25-55 year old sample and the 65+ sample.  
Can you make any suggestions on why certain items might mean different things to these two groups?  
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

We can look through `summary()` or `standardizedsolution()` for each model and just compare the various numbers. Or we could draw diagrams for both models using `semPaths()` and compare.  

However, i'm going to plot stuff again (you don't have to, i just prefer it!).  
```{r}
#| echo: false
bind_rows(
  # younger adults
  standardizedsolution(fit_who) |>
  select(lhs,op,rhs,est.std) |> 
    mutate(group="younger"),
  # older adults
  standardizedsolution(fit_bwho) |>
  select(lhs,op,rhs,est.std) |>
    mutate(group="older")
) |> filter(op == "=~") |>
  ggplot(aes(x=rhs, y = ifelse(group=="older",-abs(est.std),abs(est.std)),
             fill = group)) +
  geom_col() + 
  facet_wrap(~lhs, scales="free_x") +
  labs(y="std factor loading",x = "question")

```

What does it tell us? 
it looks like loadings for q12, q18, and q21 specifically are quite a lot weaker than in the younger group. 
And some questions (q25) are a bit higher!

```{r}
wqit |> 
  filter(q %in% c("q12","q18","q21","q25")) |>
  gt::gt()
```

This makes a little bit of sense when we think about it. People over the age of 65 are often retired, meaning q18 is kind of irrelevant? And the same reasoning might be said for q21. They might have fewer financial outgoings, making q12 less indicative of quality of life. q25 possibly becomes far more indicative of quality of life because older people will be more likely to be infirm and unable to drive etc.  


`r solend()`


`r qbegin(paste0("Extra ", qcounter()))`
Researcher A, who is happy that the WHOQOL-BREF appears to fit well in her sample of working aged adults, has another question. 

Which specific domain is the strongest unique predictor of the single item measure (`q1`) of an overall quality of life?  

She fits the model below.  
Can you answer her question?  
```{r}
a2_mod <- '
  # measurement
  Phys =~ q3 + q4 + q10 + q15 + q16 + q17 + q18
  Psych =~ q5 + q6 + q7 + q11 + q19 + q26
  Social =~ q20 + q21 + q22
  Env =~ q8 + q9 + q12 + q13 + q14 + q23 + q24 + q25
  
  # regression
  q1 ~ Phys + Psych + Social + Env
'

fit_a2 <- sem(a2_mod, whqb_a, std.lv = TRUE)
```


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
standardizedsolution(fit_a2) |>
  filter(op == "~") 
```
`r solend()`



