---
title: "CFA"
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE  # TRUE = collapse sols, FALSE = uncollapse sols
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio/  not production profile:   SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal/with production profile:   SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
source("jk_playground/iri.r")
```

:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr3_lab08.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
    - `psych`
    
:::

---

:::frame
__Testing the Short-IRI__  

Our aim today is to test the factor structure of a short form of the IRI (Interpersonal Reactivity Index) - a measure of empathy, that captures four related constructs:  

- Perspective Taking (PD): tendency to adopt others' points of view
- Empathic Concern (EC): tendency to experience sympathy and compassion for others
- Personal Distress (PD): tendency to experience distress/discomfort in response to others' distress
- Fantasy (FS): tendency to imaginatively transpose oneself into fictional situations

The full measure contains 28 items, but researchers have developed a short version: a subset of 16 of the items deemed most relevant to each of the constructs. 

A sample of 220 people completed their 16 item version: 

::: {.callout-note collapse="true"}
#### Item Wordings

```{r}
#| echo: false
short_iri |>
  transmute(
    question=q,subscale,wording,negatively_worded = neg*1
  ) |> gt::gt()
```

:::
:::

`r qbegin(qcounter())`
The researchers who developed this short form of the model suggest that this subset of 16 items is sufficient to tap into the four separate facets of the original measure. 
In the syntax for **lavaan**, specify the model that they suggest underpins this new measure.  

```{r}
#| eval: false
#| echo: true

irimod <- "
    ...
    ...
    
    ...
  "
```


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


```{r}
irimod <- "
FS =~ item1 + item12 + item16 + item26
EC =~ item2 + item14 + item20 + item22
PT =~ item8 + item11 + item21 + item28
PD =~ item13 + item17 + item24 + item27
"
```


`r solend()`

`r qbegin(qcounter())`
Fit the model, and evaluate the global (overall) model fit.  
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
mod16.est <- cfa(irimod, data = siridat, std.lv=T)

fitmeasures(mod16.est)[c("rmsea","srmr","cfi","tli")]
```

`r solend()`

`r qbegin(qcounter())`
Examine the modification indices.  
Even if the model already "fits well", this can help to highlight possible issues with the measurement model. What might a large residual covariance between two items tell us? 
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
modindices(mod16.est, sort = TRUE) |> 
  head(5)
```

`r solend()`


`r qbegin(qcounter())`
When you fitted your model, which did you use? 
```{r}
#| eval: false
#| echo: true
cfa(model, data)
cfa(model, data, std.lv = TRUE)
```

To help us better understand how this changes things, fit the other one now, and examine what (if anything) is different.  

- global fit of the model
- modification indices
- in `summary()`:  
    - factor loading estimates
    - factor covariance estimates
    
**Bonus Challenge:** can you see the link between estimates from the two versions?  
    
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
m1 <- cfa(irimod, siridat)
m1a <- cfa(irimod, siridat, std.lv = TRUE)
```

```{r}
fitmeasures(m1)[c("rmsea","srmr","cfi","tli")]
fitmeasures(m1a)[c("rmsea","srmr","cfi","tli")]
```

```{r}
head(modindices(m1, sort = TRUE))
head(modindices(m1a, sort = TRUE))
```

```{r}
summary(m1)
summary(m1a)
```

`r solend()`

`r qbegin(qcounter())`
We might ask ourselves if the four factors really necessary?  
How does the fit of our model compare to that of a model in which there is just one factor of dispositional empathy? 
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
mod1f <- "
EMP =~ item1 + item12 + item16 + item26 + item2 + item14 + 
        item20 + item22 + item8 + item11 + item21 + item28 + 
          item13 + item17 + item24 + item27
"

mod1f.est <- cfa(mod1f, siridat, std.lv=TRUE)

library(semTools)
summary(compareFit(mod1f.est, mod16.est))
```

`r solend()`
 
6. how do items look? R^2 for each item? standsolution
7. make quick diagram
8. fit same model in adolescents
9. compare the loadings from the adult sample and adolescent sample
10. compare metric invariance


8. here's the rest of the sample. test your short form on this subset 
	1. note why we split it. our decisions about the short form were informed by sampleA. we can't test that short form on that same sample. 
10. invariance across groups.  
