---
title: "W11 Exercises"
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(psych)
xaringanExtra::use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```


`r qbegin(qcounter())`
```{r}
m = "
  lv1 =~ .8*y1 + .6*y2 + .6*y3 + .7*y4
  lv2 =~ .7*y5 + .8*y6 + .9*y7 + .7*y8
  lv3 =~ .8*y9 + .8*y10 + .9*y11 + .4*y4
  lv4 =~ .9*y12
  lv1 ~~.2*lv2
  lv1 ~~.2*lv3
  lv2 ~~.2*lv2
"
eseed=round(runif(1,1e3,1e6))
set.seed(167807)
df = lavaan::simulateData(m,sample.nobs = 500)

mm = psych::fa(df, 4, rotate="oblimin",fm="ml")

mm$loadings
```


efa output

Based on the results and the item descriptions, provide an interpretation of the factor solution. Your description should include: 

- Comment on the numerical solution **[6 marks]**
- Discussion of suitability of the selected number of factors **[6 marks]**
- An attempt to define the factors **[3 marks]**


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`


`r qbegin(qcounter())`

```{r}
set.seed(67768)
nitem <- 5
A <- matrix(runif(nitem^2,.41,1)*2-1, ncol=nitem) 
scor <- t(A) %*% A
df <- MASS::mvrnorm(n=220,mu=rep(0,5),Sigma = scor) |> as.data.frame()
loadings = psych::principal(df,nfactors=2, rotate="none")$loadings[,1:2]
loadings[1:4,1] = round(loadings[1:4,1],1)
loadings[5,1] = round(loadings[5,1],2)
loadings[c(1,3:5),2] = round(loadings[c(1,3:5),2],1)
loadings[2,2] = round(loadings[2,2],2)
loadings

```

calculate the 6 values of the table below, SSloadings [2 marks] and proportion & cumulative variance [2 marks].  


```{r}
vac <- psych::principal(df,nfactors=2, rotate="none")$Vaccounted[1:3,]
vac[1:3,1:2] <- "__"
vac
```




`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
rbind(
  apply(loadings,2,\(x) sum(x^2)),
  apply(loadings,2,\(x) sum(x^2))/5,
  cumsum(apply(loadings,2,\(x) sum(x^2))/5)
)
```



`r solend()`



`r qbegin("Q3 - MLM [10 marks]", qlabel=F)`

```{r}
simbasic_rs <- function (seed = NULL, b0 = 0, b1 = 1, z0 = 1, z1 = 1, e = 1) 
{
    if (!is.null(seed)) {
        set.seed(seed)
    }
    N = 200
    n_groups = 20
    g = rep(1:n_groups, e = N/n_groups)
    x = rnorm(N)
    x = rep(1:(N/n_groups),n_groups)
    res = MASS::mvrnorm(n_groups,
                        mu=c(0,0),Sigma=matrix(c(z0,.5,.5,z1),nrow=2))
    re = res[,1][g]
    re_x = res[,2][g]
    b = rbinom(n_groups,1,.5)[g]
    lp = (b0 + re) + (b1 + re_x) * x + -.4*x*b
    y = rnorm(N, mean = lp, sd = e)
    y_bin = rbinom(N, size = 1, prob = plogis(lp))
    data.frame(x, b, g = factor(g), y, y_bin)
}
eseed=round(runif(1,1e3,1e6))
set.seed(399478)
df <- tibble(
  g1 = 1:10,
  b0 = rnorm(10,0,.2),
  b1 = rnorm(10,0,.1),
  data = map2(b0,b1, ~simbasic_rs(b0=..1,b1=..2,z0=2,z1=2))
) |> unnest(data)

library(lme4)

df = df |> transmute(
  week = x,
  journal = factor(b,levels=c("0","1"),labels=c("No","Yes")),
  center = g1,
  ppt = g,
  anger = scale(y)[,1]*2.33 + 9.64
)

mm = lmerTest::lmer(anger ~ week * journal + 
            (1 + week | center ) + 
            (1 + week |center:ppt), data=df)
# summary(mm)
```

A company that makes 6-minute journals is undertaking some research to showcase the effectiveness of their product in helping to alleviate unwanted feelings. 
They send journals to 


Scores on the anger measure can range from ? to ?, with changes of 3 being considered 'clinically meaningful'.  

They fit a multilevel model to the data, with week number (0 to 9, with 0 representing the first week participants filled in the anger assessment), whether the journal was used ("no"/"yes", with "no" as the reference level). 

Provide an interpretation of each of:

- the fixed effects **[4 marks]**
- the random effects **[4 marks]**
- the relevance of the findings, considering the context of the study design and researchers’ aims **[2 marks]**
    

::: {.callout-tip collapse="true"}
#### tips

- try to map the number of marks to "things to mention" - i.e. there are four marks for fixed effects, and four parameters!  
    - sometimes there will be more things than marks (e.g., you might see 8 fixed effects but 4 marks). This is often because there are lots of different ways to get up to 4 marks (lots of different things you *could* mention). Try to focus on the ones that are of relevance to the RQ primarily. 
- 


:::

    
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


- anger for someone in week 1 who doesn't journal is (intercept)
- no significant change over the study period for those who don't journal (week fixef)
- no signif difference between journal/no-journal groups at outset of study
- journal group decreases in anger over the study more compared to no-journal group (interaction)


- participants vary (both intercept and slopes of change) much more than centers
- high level of ppt variability relative to fixed slope
- ppts who start more angry decrease less (positive correlation intercepts and slopes)

- effect is small - over 10 weeks they only go down by `r (fixef(mm)[2]+fixef(mm)[4])*9`


```{r}
broom.mixed::augment(mm) |>
  ggplot(aes(x=week,y=.fitted,group=interaction(center,ppt),
             col=journal))+
  geom_line(alpha=.2)+
  geom_abline(intercept=fixef(mm)[1],slope=fixef(mm)[2],col="red")+
  geom_abline(intercept=fixef(mm)[1]+fixef(mm)[3],slope=fixef(mm)[2]+fixef(mm)[4],col="blue")
```


`r solend()`


`r qbegin("Q3 - hierarchical data structures [4 marks]", qlabel=F)`
Provide example levels for each of the three types of study: Cross-Sectional, Repeated Measures, Longitudinal [4 marks]

| level | cross-sectional | repeated measures | longitudinal |
| ----- |---------- |-------------| -----------|
| 2     | ...  |  ... |  ...      |
| 1    | ...| ... | ...  |

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

anything that makes sense here, obvious ones are:

| level | cross-sec | rpt measures | longitudinal |
| ----- |---------- |-------------| -----------|
| 2     |  department |   people |  people |
| 1    | people | experimental items | timepoints |

`r solend()`


