---
title: 'Analysis Example: Rpt & Mixed ANOVA'
link-citations: yes
code-fold: true
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
---

```{r setup, include=F}
library(tidyverse)
knitr::opts_chunk$set(message = F, warning = F, fig.align = 'center')
```

:::imp
This is optional for the DAPR3 course, but may be useful for your dissertations should your field/supervisor prefer the ANOVA framework to that of the linear model. 

This walks briefly through these models with the __ez__ package. There are many other packages available, and many good tutorials online should you desire extra resources in the future:

- https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r
- https://www.r-bloggers.com/2021/04/repeated-measures-of-anova-in-r-complete-tutorial/
- https://stats.idre.ucla.edu/r/seminars/repeated-measures-analysis-with-r/ 
- https://www.datanovia.com/en/lessons/mixed-anova-in-r/  

:::


:::frame
**Data: Raising the stakes**

This data is from a simulated study that aims to investigate how levels of stress are influenced by the size and type of potential rewards. 

30 volunteers from an amateur basketball league participated. Each participant completed 20 trials in which they were tasked with throwing a basketball and scoring a goal in order to score points. The number of points up for grabs varied between trials, ranging from 1 to 20 points, with the order randomised for each participant. If a participant successfully threw the ball in the basket, then their score increased accordingly. If they missed, their score decreased accordingly. Participants were informed of the number of points available prior to each throw.  

To examine the influence of the *type* of reward/loss on stress-levels, the study consisted of two conditions. In the monetary condition, (n = 15) participants were informed at the start of the study that they would receive their final score in £ at the end of the study. In the reputation condition, (n = 15) participants were informed that the points would be inputted on to a scoreboard and distributed around the local basketball clubs and in the league newsletter. 

Throughout each trial, participants' heart rate variability (HRV) was measured via a chest strap. HRV is considered to be indirectly related to levels of stress (i.e., higher HRV = less stress).

The data is in stored in two separate files.  

1. Information on the conditions for each trial for each participant is stored in __.csv__ format at [https://uoepsy.github.io/data/basketballconditions.csv](https://uoepsy.github.io/data/basketballconditions.csv).  
```{r echo=FALSE}
bball <- read_csv("https://uoepsy.github.io/data/basketballconditions.csv")
tibble(
  variable = names(bball),
  description = c("Number of points available to be won/lost based on successful completion of     trial","Whether the final score will be converted to £ or will be placed on a public       leader-board","Participant Identifier","Whether the participant successfully threw the ball in the   basket", "Trial number (1 to 20")
) %>% knitr::kable()
```
2. Information on participants' HRV for each trial is stored in **.xlsx** format, and can be downloaded from [https://uoepsy.github.io/data/basketballhrv.xlsx](https://uoepsy.github.io/data/basketballhrv.xlsx). 
```{r echo=FALSE}
bballhrv <- read_csv("https://uoepsy.github.io/data/basketballhrv.csv")
tibble(
  variable = c(names(bballhrv)[1:4],"..."),
  description = c("Subject Identifier",paste0("HRV average during trial ",c(1,2,3,"...")))
) %>% knitr::kable()
```


:::

The code below will read in both datasets and join them for you:

```{r message=F,warning=F,eval=F}
library(tidyverse)
library(readxl)
download.file(url = "https://uoepsy.github.io/data/basketballhrv.xlsx",
              destfile = "baskeballhrvdata.xlsx",
              mode = "wb")
bball <- 
  left_join(
    read_csv("https://uoepsy.github.io/data/basketballconditions.csv"),
    read_xlsx("baskeballhrvdata.xlsx") %>%
      pivot_longer(trial_1:trial_20, names_to = "trial_no", values_to = "hrv")
  ) %>%
  mutate(sub = factor(sub))
```
```{r message=F,warning=F,echo=F}
library(tidyverse)
library(readxl)
bball <- 
  left_join(
    read_csv("https://uoepsy.github.io/data/basketballconditions.csv"),
    read_csv("https://uoepsy.github.io/data/basketballhrv.csv") %>%
      pivot_longer(trial_1:trial_20, names_to = "trial_no", values_to = "hrv")
  ) %>%
  mutate(sub = factor(sub))
```
:::


# One-Way Repeated Measures ANOVA

For a repeated measures ANOVA, our independent variable is _within_ group.  

Following from the example study above, we might consider using it to answer the question below.  
 
> __Question:__ What is the effect of the size of reward on stress levels (as measured by HRV)?

The easiest way to conduct a repeated measures ANOVA in R is to use the __ez__ package.  
It comes with some handy functions to visualise the experimental design.  
We can see from below that every participant completed a trial for each value of reward-size (1-20 points):
```{r}
library(ez)
ezDesign(data = bball, x = sub, y = stakes)
```

The `ezANOVA()` function takes a few arguments. 

The ones you will need for this are:

- *data* the name of the dataframe
- *dv* the column name for the dependent variable
- *wid* the column name for the participant id variable
- *within* the column name(s) for the predictor variable(s) that vary within participants 
- *between* the column name(s) for any predictor variable(s) that vary between participants

Fit a repeated measures ANOVA to examine the effect of the size of reward on HRV. 

```{r}
ezANOVA(data=bball, dv=hrv, wid = sub, within = stakes)
```


# Mixed ANOVA 

Mixed ANOVA can be used to investigate effects of independent variables that are at two different levels,  i.e. some are *within* clusters and some are *between*.  

> __Question:__ Does the influence of the size of reward/loss on stress levels differ depending upon whether it is money vs reputation at stake? 

Look at the two lines below. Can you work out what the plots will look like _before_ you run them? 

```{r eval=F}
ezDesign(data = bball, x = condition, y = sub)
ezDesign(data = bball, x = condition, y = stakes)
```


Participants 1-15 are in one condition, and 16-30 are in another.  
This should look like a two big blocks on the diagonal. 
```{r}
ezDesign(data = bball, x = condition, y = sub)
```
In each condition, the full set of stakes (1-20 points) were observed in the same number of trials.
This should be a full grid:
```{r}
ezDesign(data = bball, x = condition, y = stakes)
```


Fit a mixed ANOVA to examine the interaction between size and type of reward on HRV. 
```{r}
ezANOVA(data=bball, dv=hrv, wid = sub, within = stakes, between = condition)
```

The __ez__ package also contains some easy plotting functions for factorial experiments, such as `ezPlot()`. It takes similar arguments to the `ezANOVA()` function. 

- look up the help documentation for `ezPlot()`. 
- let's use `ezPlot()` to make a nice plot (__Note:__ we may need to make sure that the `stakes` variable is a factor).

```{r}
bball <- bball %>% mutate(stakes = factor(stakes))
ezPlot(data=bball, dv=hrv, wid = sub, within = stakes, between = condition, x = stakes, split = condition)
```


<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>

