---
title: "DAPR3_2425_BLOCK1"
author: "jk"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
knitr::opts_chunk$set(echo = TRUE)
load("https://uoepsy.github.io/dapr3/2425/misc/dapr3_liveR_newtricks.RData")
```

# New tricks for old dogs

We're studying animal learning. 

- We have studied 20 dogs (6 puppies, 14 adults) who are all learning the same new trick.  
- Each dog has 30 minutes of training for the trick each day for 7 days. 
- At the end of each training session they are tested to and the time it takes them to complete the trick is recorded

```{r}
ggplot(new_trick, aes(x=trainingdays, y=tricktime,col=isOld))+
  geom_point() +
  facet_wrap(~DOG)
```

## Descriptives

- how many dogs
- how many adults/puppies
- how many observations each dog?

```{r}

```


## Modelling


```{r}
# nope... 
lm(tricktime ~ DOG * trainingdays * isOld, data = new_trick) |>
  summary()
```

```{r}
mod <- lmerTest::lmer(tricktime ~ trainingdays * isOld + 
              (1 + trainingdays | DOG),
            data = new_trick)

summary(mod)



library(lattice)
xyplot(tricktime~trainingdays|DOG, data= new_trick)
#compare with:
xyplot(tricktime~isOld|DOG, data= new_trick)
```


#### models that don't work and shouldn't work: 

```{r}
badmod1 <- lmer(tricktime ~ trainingdays * isOld + 
       (1+trainingdays + isOld|DOG), 
     data = new_trick)

summary(badmod1)
```

#### models that shouldn't work but do (but don't mean what you want them to mean)

```{r}
badmod2 <- lmer(tricktime ~ trainingdays * isOld + 
                  (1 + isOld|DOG), 
     data = new_trick)

summary(badmod2)
```


# many breeds

```{r}
head()
```

