---
title: "DAPR3_2425_WEEK5LR"
author: "jk"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# packages
library(tidyverse)
library(lme4)
library(broom.mixed)
library(effects)
# settings
knitr::opts_chunk$set(echo = FALSE)
# data
load(url("https://uoepsy.github.io/dapr3/2425/misc/dapr3_liveR_newtricks.RData"))

# cleaning

# analysis
mod3b <- lmerTest::lmer(tricktime ~ trainingdays * isOld +
              (1 + trainingdays | trick) + 
              (1 + trainingdays | DOG),
            data = many_new_tricks)


# tidy plots and tables
res <- tidy(mod3b) |>
  transmute(
    term,
    est = round(estimate,2),
    se = round(std.error,2),
    t = round(statistic,2),
    df = round(df,2),
    p = format.pval(p.value, digits = 2, eps = .001)
  )

efdat <- effect("trainingdays*isOld", mod3b) |>
  as.data.frame()

efplot <- ggplot(efdat, aes(x=trainingdays,y=fit,col=isOld,
                  fill=isOld))+
  geom_line()+
  geom_ribbon(aes(ymin=lower,ymax=upper),alpha=.3)
```



# What we have

lots of data. how much? 

# What we planned to do with it

fit a model!

# What we actually did with it

fit a slightly different model because the original one didn't work.  

# What we found

averaged across dogs and tasks, it took `r res[1,"est"]` seconds to complete the trick on the first day of training (b = `r res[1,"est"]`, t(`r res[1,"df"]`)=`r res[1,"t"]`, p `r res[1,"p"]`). For puppies, each subsequent day of training was associated with a `r res[2,"est"]` decrease in time taken (b = `r res[2,"est"]`, t(`r res[2,"df"]`)=`r res[2,"t"]`, p=`r res[2,"p"]`)



Here's a plot:
```{r}
efplot
```

