---
title: "DAPR3_2425_BLOCK1"
author: "jk"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
knitr::opts_chunk$set(echo = TRUE)
load(url("https://uoepsy.github.io/dapr3/2425/misc/dapr3_liveR_newtricks.RData"))
```

# New tricks for old dogs

We're studying animal learning. 

- We have studied 20 dogs (6 puppies, 14 adults) who are all learning the same new trick.  
- Each dog has 30 minutes of training for the trick each day for 7 days. 
- At the end of each training session they are tested to and the time it takes them to complete the trick is recorded

```{r}
ggplot(new_trick, aes(x=trainingdays, y=tricktime,col=isOld))+
  geom_point() +
  facet_wrap(~DOG)
```

## Descriptives

- how many dogs
- how many adults/puppies
- how many observations each dog?

```{r}
new_trick |> 
  count(isOld, DOG) |>
  count(isOld)

new_trick |> 
  count(DOG) |>
  summarise(
    min_nobs = min(n),
    max_nobs = max(n),
    median_nobs = median(n)
  ) 
```


## Modelling


```{r}
# nope... 
lm(tricktime ~ DOG * trainingdays * isOld, data = new_trick) |>
  summary()
```

```{r}
mod <- lmerTest::lmer(tricktime ~ trainingdays * isOld + 
              (1 + trainingdays | DOG),
            data = new_trick)

summary(mod)



library(lattice)
xyplot(tricktime~trainingdays|DOG, data= new_trick)
#compare with:
xyplot(tricktime~isOld|DOG, data= new_trick)
```


#### models that don't work and shouldn't work: 

```{r}
badmod1 <- lmer(tricktime ~ trainingdays * isOld + 
       (1+trainingdays + isOld|DOG), 
     data = new_trick)

summary(badmod1)
```

#### models that shouldn't work but do (but don't mean what you want them to mean)

```{r}
badmod2 <- lmer(tricktime ~ trainingdays * isOld + 
                  (1 + isOld|DOG), 
     data = new_trick)

summary(badmod2)
```


# Week 3

## last week  

```{r}
head(new_trick)
```


```{r}
mod <- lmer(tricktime ~ trainingdays * isOld + 
              (1 + trainingdays | DOG),
            data = new_trick)
summary(mod)
```

```{r}
library(effects)
effect("trainingdays*isOld", mod) |>
  as.data.frame() |>
  ggplot(aes(x=trainingdays, y=fit, col=isOld))+
  geom_line()+
  geom_ribbon(aes(ymin=lower,ymax=upper,fill=isOld),
              alpha=.3)
```

```{r}
library(broom.mixed)
augment(mod) |>
  ggplot(aes(x=trainingdays,y=.fitted,
             col=isOld, group=DOG))+
  geom_line()
```

```{r}
efplot <- effect("trainingdays*isOld", mod) |>
  as.data.frame() 

augment(mod) |>
  ggplot(aes(x=trainingdays,y=.fitted,
             col=isOld))+
  geom_line(aes(group=DOG),alpha=.3) +
  geom_line(data = efplot, aes(y=fit))+
  geom_ribbon(data = efplot, 
              aes(y=fit,ymin=lower,ymax=upper,
                  fill=isOld),
              alpha=.3)
```



# many breeds


```{r}
head(new_trick_dogbreeds)
```



```{r}
ggplot(new_trick_dogbreeds, aes(x=trainingdays, y=tricktime,col=isOld))+
  stat_summary(geom="line")+
  facet_wrap(~BREED)

mod <- lmer(tricktime ~ trainingdays * isOld +
              (1 + trainingdays + isOld | BREED) + 
              (1 + trainingdays | BREED:DOG),
            data = new_trick_dogbreeds)
summary(mod)

VarCorr(mod)

mod2 <- lmer(tricktime ~ trainingdays * isOld +
              (1 + trainingdays | BREED) + 
              (1 + trainingdays | BREED:DOG),
            data = new_trick_dogbreeds)
summary(mod2)


```


# many tasks

```{r}

mod3 <- lmer(tricktime ~ trainingdays * isOld +
              (1 + trainingdays + isOld | trick) + 
              (1 + trainingdays | DOG),
            data = many_new_tricks)
dotplot.ranef.mer(ranef(mod3))

```

