---
title: "Modelling group-structured data using random effects"
editor_options: 
  chunk_output_type: console
format:
  revealjs:
    smaller: true
---

```{r setup, include=F}
library(tidyverse)
library(patchwork)
library(emmeans)
library(simglm)
library(latex2exp)  # for betas in ggplots
source('_theme/theme_quarto.R')

theme_set(theme_quarto(title_font_size=42))
theme_update(
  text = element_text(family = 'Source Sans 3')
)

dapr3green <- "#88B04B" 
dapr3dkgreen <- "#5C7C28"
dapr3ltgreen <- "#E5EED7"
pal <- c( "#d35269", "#5c9ead","#2a3c24", "#F5C396", "#8B2635",  "#235789")
```



# Course Overview {background-color="white"}

<br>

```{r echo=F}
#| results: "asis"
block1_name = "Linear mixed models<br>(with Elizabeth Pankratz)"
block1_lecs = c("Regression refresher, intro to group-structured data",
                "Modelling group-structured data using random effects",
                "TODO",
                "TODO",
                "recap")
block2_name = "factor analysis<br>working with multi-item measures<br>(with Josiah King)"
block2_lecs = c(
  "measurement and dimensionality",
  "exploring underlying constructs (EFA)",
  "testing theoretical models (CFA)",
  "reliability and validity",
  "recap & exam prep"
  )

source("https://raw.githubusercontent.com/uoepsy/junk/refs/heads/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=2)
```

# abc

abc


# Back matter

## Learning objectives revisited




## To do this week 

<br>

::::{.columns}
:::{.column width="50%"}
**Tasks:**

<br>

![](figs/labs.svg){width=80px style="margin:10px;margin-bottom:-50px"} Work on exercises in labs

<br>

![](figs/exam.svg){width=80px style="margin:10px;margin-bottom:-45px"} Complete the weekly quiz 


:::

:::{.column width="50%"}
**Get support:**

<br>

![](figs/forum.svg){width=80px style="margin:10px;margin-bottom:-50px"}
Ask questions anonymously on Piazza

<br>

![](figs/oh.svg){width=80px style="margin:10px;margin-bottom:-40px"} 
We really like seeing you in office hours!



:::
::::



# Appendix {.appendix}



## abc

dolor sit amet

**dolor sit amet**

`abc`

:::dapr3callout
abc
:::


<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->
<!-- a -->
<!-- ::: -->
<!-- ::: {.column width="50%"} -->
<!-- b -->
<!-- ::: -->
<!-- :::: -->



<!-- style="font-size: 70%;" -->

 <!-- <img style="vertical-align: text-bottom; width: 1.5em;" src="figs/material-symbols--draw-outline.svg"> -->
 <!-- <img style="vertical-align: text-bottom; width: 1.2em;" src="figs/material-symbols--person-rounded.svg"> -->