---
title: "DAPR3 LMM practice analysis – Code template"
output: 
  html_notebook:
    toc: true
---

Steps are from the workflow in [the LMM analysis workflow flash card](https://uoepsy.github.io/dapr3/2627/flashcards/resources/workflow.html){target="_blank"}.

# Phase 1: Before model fitting

## 1a: Set up code and data

Load required R packages.

```{r}

```


Read in your data.

```{r}

```

Tidy data (e.g., any missingness, any implausible values?)

```{r}

```


## 1b: Set up fixed effects

Based on the RQ, identify your outcome variable and the predictors (aka the fixed effects).

[...]


Based on the RQ, decide whether your model requires an interaction between predictors.

[...]


Based on the outcome variable, decide whether to use regular regression or logistic regression.

[...]


Set up categorical predictors. (e.g., factor levels? which a priori contrast coding scheme?)

```{r}

```

Set up continuous predictors. (e.g., any transformations?)

```{r}

```

Explore patterns by plotting outcome and predictor variables together.

```{r}

```


## 1c: Set up random effects

Identify grouping variables which contribute random variability to your data.
Your model will have random intercepts for each of these grouping variables.

```{r}

```

Describe the randomly-varying grouping variables (so that the reader knows how well your analysis might generalise).

```{r}

```

Identify the random slopes that your model can contain.

```{r}

```


Write out the formula for your maximal model.

[...]


# Phase 2 ----

Try fitting maximal model.

```{r}

```

If you get a convergence warning: Try a different optimiser and fit model again.

```{r}

```

If you get a convergence or singular fit warning: Simplify random effect structure and fit model again.

```{r}

```

If you get a suggestion to rescale variables: Convert continuous predictor(s) to z-scores and fit model again.

```{r}

```

Even if you get no warnings, manually check variance components for sneaky singularities.

```{r}

```


# Phase 3: After model fitting

## 3a: Check assumptions and diagnostics

Check model assumptions.

```{r}

```


Run diagnostics for multicollinearity.

```{r}

```


Run diagnostics for influential observations and levels of grouping variables.

```{r}

```


If you find extreme influential observations/groups: run sensitivity analysis.

```{r}

```



## 3b: Plot and interpret model estimates

Plot model-fitted values.

```{r}

```


Interpret the fixed effect estimates.

[...]

Plot the random effects.

```{r}

```

Interpret SD of random effects relative to fixed effects.

```{r}

```


Interpret correlation between intercept and slope adjustments.

[...]

## 3c: Write up methods and results
