---
title: "LMMs: Practice analysis"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---

```{r setup, include=F}
library(tidyverse)
library(patchwork)

library(latex2exp)  # for betas in ggplots
source('_theme/theme_quarto.R')

theme_set(theme_quarto(title_font_size=42))
theme_update(
  text = element_text(family = 'Source Sans 3')
)

dapr3green <- "#88B04B" 
dapr3dkgreen <- "#5C7C28"
dapr3ltgreen <- "#E5EED7"
pal <- c( "#d35269", "#5c9ead","#2a3c24", "#F5C396", "#8B2635",  "#235789")
```


# Course Overview {background-color="white"}

<br>

```{r echo=F}
#| results: "asis"
block1_name = "Linear mixed models<br>(with Dr. Elizabeth Pankratz)"
block1_lecs = c("Regression refresher, intro to group-structured data",
                "Modelling group-structured data using random effects",
                "Interpreting LMMs and building maximal models",
                "Troubleshooting model fit, checking assumptions + diagnostics",
                "LMMs: Practice analysis")
block2_name = "factor analysis<br>working with multi-item measures<br>(with Dr. Josiah King)"
block2_lecs = c(
  "measurement and dimensionality",
  "exploring underlying constructs (EFA)",
  "testing theoretical models (CFA)",
  "reliability and validity",
  "recap & exam prep"
  )

source("https://raw.githubusercontent.com/uoepsy/junk/refs/heads/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=5)
```



# The game plan for this week

## The game plan for this week

<br>

:::: {.columns}
::: {.column width="15%"}

<img style="margin:10px;margin-top:-5px;width:4em;" src="figs/mdi--lecture.svg">

:::
::: {.column width="85%"}
**Lectures:**

Together, we'll work through an LMM analysis from beginning to end (hopefully!).

:::
::::

<br>

:::: {.columns}
::: {.column width="15%"}

<img style="margin:10px;margin-top:-10px;width:4em;" src="figs/streamline-ultimate--paper-write-bold.svg">


:::
::: {.column width="85%"}
**Labs:**

You'll practice writing up a brief report based on this analysis.
:::
::::


## Resource: The LMM analysis workflow

<br>

:::hcenter
[DAPR3 flash cards > More resources > LMM analysis workflow](https://uoepsy.github.io/dapr3/2627/flashcards/resources/workflow.html){target="_blank"}
:::

<br>

Four phases:

- **Phase 1:** Before model fitting
- **Phase 2:** Model fitting and troubleshooting
- **Phase 3:** After model fitting

<br>

I will navigate us through these phases.
**At each step, you'll use Wooclap to tell me what I should do.**

## The RQ

<br>

- We're looking at emotional dysregulation in children: when children have trouble managing their feelings and emotions.
- Specifically, we want to find out how emotional dysregulation is affected by children having a strong, regular routine at home.
- **We predict that children with stronger home routines are better at regulating their emotions.**
- It's also plausible that a family's socio-economic status (SES) is associated with how well children regulate their emotions. 
We will control for SES using a proxy variable of whether children attend a state school or a private school.

<br>

RQ: **Controlling for the type of school that children attend, are children with stronger routines better at regulating their emotions?**

<br>

## The data

<br>

Data from: <https://uoepsy.github.io/data/crqeds.csv>{target="_blank"}

<br>

In the data:

| variable   | description                                                                                       |
| ---------- | ------------------------------------------------------------------------------------------------- |
| `schoolid`   | School identifier                                                                                 |
| `EDS`        | Emotion Dysregulation Score (range 1 to 6, higher values indicate more *dys*regulation)  |
| `CRQ`        | Childhood Routine Questionnaire Score (range 0 to 7, higher values indicate more routine) |
| `schooltype` | School type (private / state)                                                                     |


## Let's do it!

<br>

:::hcenter
:::woo
https://app.wooclap.com/events/HJNKOQ/
:::
:::



# Back matter

## To do this week 

<br>

::::{.columns}
:::{.column width="50%"}
**Tasks:**

<br>

![](figs/labs.svg){width=80px style="margin:10px;margin-bottom:-50px"} Work on exercises in labs

<br>

![](figs/exam.svg){width=80px style="margin:10px;margin-bottom:-45px"} Complete the weekly quiz 


:::

:::{.column width="50%"}
**Get support:**

<br>

![](figs/flashcards.svg){width=80px style="margin:10px;margin-bottom:-30px"}
Consult the [flash cards](https://uoepsy.github.io/dapr3/2627/flashcards/){target="_blank"}

<br>

![](figs/forum.svg){width=80px style="margin:10px;margin-bottom:-50px"}
Ask questions anonymously on Piazza

<br>

![](figs/oh.svg){width=80px style="margin:10px;margin-bottom:-40px"} 
We really like seeing you in office hours!



:::
::::



# Appendix {.appendix}



## abc

dolor sit amet

**dolor sit amet**

`abc`

:::dapr3callout
abc
:::


<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->
<!-- a -->
<!-- ::: -->
<!-- ::: {.column width="50%"} -->
<!-- b -->
<!-- ::: -->
<!-- :::: -->



<!-- :::: {.columns} -->
<!-- ::: {.column width="30%"} -->
<!-- a -->
<!-- ::: -->
<!-- ::: {.column width="5%"} -->
<!-- ::: -->
<!-- ::: {.column width="30%"} -->
<!-- b -->
<!-- ::: -->
<!-- ::: {.column width="5%"} -->
<!-- ::: -->
<!-- ::: {.column width="30%"} -->
<!-- c -->
<!-- ::: -->
<!-- :::: -->


<!-- style="font-size: 70%;" -->

 <!-- <img style="vertical-align: text-bottom; width: 1.5em;" src="figs/material-symbols--draw-outline.svg"> -->
 <!-- <img style="vertical-align: text-bottom; width: 1.2em;" src="figs/material-symbols--person-rounded.svg"> -->