---
title: "<b>Exam Prep</b>"
subtitle: "Data Analysis for Psychology in R 3"
author: "dapR3 Team"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
      - jk_libs/tweaks.css
    nature:
      beforeInit: "jk_libs/macros.js"
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:devtools::install_github("gadenbuie/xaringanExtra"):
  chunk_output_type: inline
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
xaringanExtra::use_xaringan_extra(c("tile_view","animate_css","tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_extra_styles(
  mute_unhighlighted_code = FALSE
)
xaringanExtra::use_share_again()
library(knitr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(patchwork)
knitr::opts_chunk$set(
  fig.asp=.9,
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
source("jk_source/jk_presfuncs.R")
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  code_font_size = "0.7rem",
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
)
```


# Outline

+ Process

+ Structure of exam

+ Things to take into exam

+ Exam "strategy"

+ Example Questions

---
# Process

+ CHECK YOUR CALENDAR [https://www.ed.ac.uk/timetabling-examinations/exams/exam-diets](https://www.ed.ac.uk/timetabling-examinations/exams/exam-diets)

+ Arrive in plenty of time

+ There will be direction to room

+ When you go into the room, you will need your student ID on your desk

+ Invigilators will give you exam instructions


---
# Things to take to exam  

+ Student Card!!

+ Pens (multiple) blue or black ink  

+ Pencil (for rough work)  

+ Ruler
  + If you like, not really needed, but might be useful.  
  
+ Calculator
  + Any type (see list on Learn of permitted calculators)  
  + No advantage to scientific vs normal  
  + **Phone/Watch can not be your calculator**  

---
# Things you will be given

.pull-left[
**What you are given**

+ a specific sheet for your MCQ  

+ an answer book for all other questions  

+ a rough work book ( *if you want to use it* )  

+ the equation sheet

]

--

.pull-right[
**FILLING IN FRONT COVERS**

MCQ sheet:  

  - matriculation number (UUN but without the "s")  
  
Answer book:  

<iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_ms01z6h1&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_j0xl893i" width="400" height="285" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="How to fill in the cover of an exam script book?"></iframe>

]

---
# Things you can take away

<center>
<br><br>
<h2>NOTHING!</h2>  

(Other than the things you bring in with you like pens, pencils, calculator)

<br><br>
<b>
When you finish, you leave all papers (questions, answers, equation sheet) in the exam hall.  
</b></center>

<!-- --- -->
<!-- # Equation sheet -->
<!-- + Ideally double side printed. -->
<!--   + if you cant, either staple back to back, or take 2 sheets -->

<!-- + **BUT** -->
<!--   + the notes must only be within the box -->

<!-- + **If you have any notes outside the box, you will have your notes taken off of you and will be given a plain equation sheet** -->

<!-- + You must hand in your notes sheet with your other exam materials. -->


---
class: center, middle, animated, rotateInDownLeft

# Questions


---
# Exam Structure

.pull-left[
**Section A - 15 MARKS**

+ 15 MCQ  

+ Just like quiz questions on Learn  

+ On all topics   

]
.pull-right[
**Section B - 85 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 20 marks per question  

]

---
# Exam Strategy: Section A

.pull-left[
**Section A - 15 MARKS**

+ 15 MCQ  

+ Just like quiz questions on Learn  

+ On all topics   


.br3.bg-white.green[

15-18 minutes  

  + You can answer these quickly at the end
  + Do the ones you know, move on
  + Keep track of how many you have not done, so you can go back to them at the end
  
]


]
.pull-right[
]


---
# Exam Strategy: Section B

.pull-left[
**Section B - 85 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 20 marks per question  

{{content}}

]
.pull-right[

]

--

```{r echo=FALSE}
tibble::tribble(
  ~content,~questions,
  "Weeks 1-5 (multilevel models)","4 questions totaling 30 marks",
  "Weeks 7 & 8 (path models & mediation)", "3 questions totaling 24 marks",  
  "Week 9 (PCA)","1 question totaling 5 marks",  
  "Weeks 10 & 11 (EFA)","2 questions totaling 26 marks"
) |> gt::gt()
```

---
# Exam Strategy: Section B

.pull-left[
**Section B - 85 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 20 marks per question  

```{r echo=FALSE}
tibble::tribble(
  ~content,~questions,
  "Weeks 1-5 (multilevel models)","4 questions totaling 30 marks",
  "Weeks 7 & 8 (path models & mediation)", "3 questions totaling 24 marks",  
  "Week 9 (PCA)","1 question totaling 5 marks",  
  "Weeks 10 & 11 (EFA)","2 questions totaling 26 marks"
) |> gt::gt()
```

]
.pull-right[

Questions on:  

+ Some short calculation  
+ Some ask you to explain statistical concepts discussed in the course  
+ Some give you a research design/question and task you with specifying an appropriate model  
+ Some give you model output and ask you to provide interpretation  


]

---
# Exam Strategy: Section B

.pull-left[
**Section B - 85 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 20 marks per question  

```{r echo=FALSE}
tibble::tribble(
  ~content,~questions,
  "Weeks 1-5 (multilevel models)","4 questions totaling 30 marks",
  "Weeks 7 & 8 (path models & mediation)", "3 questions totaling 24 marks",  
  "Week 9 (PCA)","1 question totaling 5 marks",  
  "Weeks 10 & 11 (EFA)","2 questions totaling 26 marks"
) |> gt::gt()
```

]
.pull-right[

Questions on:  

<ul>
<li>Some short calculation</li>  
<li style="opacity:.5">Some ask you to explain statistical concepts discussed in the course</li>
<li style="opacity:.5">Some give you a research design/question and task you with specifying an appropriate model</li>
<li style="opacity:.5">Some give you model output and ask you to provide interpretation</li>
</ul>

<small>

**approx 5 marks**  
**show your working** (could get marks for this if answer is wrong)  

think, e.g.:  
- calculating icc from a intercept only model  
- calculating coef from fixef and ranef  
- calculating indirect effects and total effects  
- calculating variance explained from eigenvalues  
- calculating SS loadings  

</small>
]

---
# Exam Strategy: Section B

.pull-left[
**Section B - 85 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 20 marks per question  

```{r echo=FALSE}
tibble::tribble(
  ~content,~questions,
  "Weeks 1-5 (multilevel models)","4 questions totaling 30 marks",
  "Weeks 7 & 8 (path models & mediation)", "3 questions totaling 24 marks",  
  "Week 9 (PCA)","1 question totaling 5 marks",  
  "Weeks 10 & 11 (EFA)","2 questions totaling 26 marks"
) |> gt::gt()
```

]
.pull-right[

Questions on:  

<ul>
<li style="opacity:.5">Some short calculation</li>
<li>Some ask you to explain statistical concepts discussed in the course</li>
<li style="opacity:.5">Some give you a research design/question and task you with specifying an appropriate model</li>
<li style="opacity:.5">Some give you model output and ask you to provide interpretation</li>
</ul>

<small>

**approx 6 marks**  
often these ask to define a couple of things.  
1 to 2 sentences each thing.  

think, e.g.:  

- "ICC"
- "random effects and fixed effects"
- "no pooling/complete pooling/partial pooling"
- "factor loading"
- "reliability"
- "indirect and direct effects"


</small>
]

---
# Exam Strategy: Section B

.pull-left[
**Section B - 85 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 20 marks per question  

```{r echo=FALSE}
tibble::tribble(
  ~content,~questions,
  "Weeks 1-5 (multilevel models)","4 questions totaling 30 marks",
  "Weeks 7 & 8 (path models & mediation)", "3 questions totaling 24 marks",  
  "Week 9 (PCA)","1 question totaling 5 marks",  
  "Weeks 10 & 11 (EFA)","2 questions totaling 26 marks"
) |> gt::gt()
```

]
.pull-right[

Questions on:  

<ul>
<li style="opacity:.5">Some short calculation</li>
<li style="opacity:.5">Some ask you to explain statistical concepts discussed in the course</li>
<li>Some give you a research design/question and task you with specifying an appropriate model</li>
<li style="opacity:.5">Some give you model output and ask you to provide interpretation</li>
</ul>

<small>
**approx 10 marks**  
you have done this in the report, and lots of times in labs.  

Hint: PCA/EFA are not "research design/question"-based in the sense of estimation, so these questions are likely to be multilevel models and path analysis.   
  
You probably know more than you think you do.  
- e.g. start with an outcome variable.  
- ask yourself: "what in my question is asking about 'effects on Y'/'influences on Y'/'predicts Y'".  

</small>
]



---
# Exam Strategy: Section B

.pull-left[
**Section B - 85 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 20 marks per question  

```{r echo=FALSE}
tibble::tribble(
  ~content,~questions,
  "Weeks 1-5 (multilevel models)","4 questions totaling 30 marks",
  "Weeks 7 & 8 (path models & mediation)", "3 questions totaling 24 marks",  
  "Week 9 (PCA)","1 question totaling 5 marks",  
  "Weeks 10 & 11 (EFA)","2 questions totaling 26 marks"
) |> gt::gt()
```

]
.pull-right[

Questions on:  

<ul>
<li style="opacity:.5">Some short calculation</li>
<li style="opacity:.5">Some ask you to explain statistical concepts discussed in the course</li>
<li style="opacity:.5">Some give you a research design/question and task you with specifying an appropriate model</li>
<li>Some give you model output and ask you to provide interpretation</li>
</ul>

<small>
**approx 10-20 marks**  
you have done this in the report, and also seen it plenty of times in the labs and lectures  

- the questions will list the things we expect comments on
- don't just parrot back numbers. provide *interpretation*  
- these usually provide information about the research aims, so place your answer in that context.  

These are going to be our main functions we have seen throughout DAPR3: `lmer()`, `principal()`, `sem()`, `fa()`  
print an output from the lab/lectures and annotate it. what does each bit of the output represent?  

</small>
]

---
class: center, middle, animated, rotateInDownLeft

# Questions

---
# Mock Questions

_(switch to pdf)_


---
class: center, inverse, middle, animated, rotateInDownLeft

# End
