---
title: "Exam Prep"
author: "Josiah King & John Martindale"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(patchwork)
source('_theme/theme_quarto.R')
```

# Course Overview

```{r}
#| results: "asis"
#| echo: false
block1_name = "multilevel modelling<br>working with group structured data"
block1_lecs = c("regression refresher",
                "introducing multilevel models",
                "more complex groupings",
                "centering, assumptions, and diagnostics",
                "recap")
block2_name = "factor analysis<br>working with multi-item measures"
block2_lecs = c(
  "what is a psychometric test?",
  "using composite scores to simplify data (PCA)",
  "uncovering underlying constructs (EFA)",
  "more EFA",
  "recap & exam prep"
  )

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=10)
```

```{r}

```




# Outline

+ Process

+ Structure of exam

+ Things to take into exam

+ Exam "strategy"

+ Example Questions


# Process

## Details



::::{.columns}
:::{.column width="50%"}
### WHEN?  

Date: Friday, 20th December 2024
Time: 1:00 p.m. to 3:00 p.m.


:::

:::{.column width="50%"}
### WHERE?  

Lennox, Lammermuir & Moffat Room, Edinburgh International Conference Centre

:::
::::

<br><br><br>
CHECK YOUR CALENDAR [https://www.ed.ac.uk/timetabling-examinations/exams/exam-diets](https://www.ed.ac.uk/timetabling-examinations/exams/exam-diets)  


## Details

+ Arrive in plenty of time

+ There will be direction to room

+ When you go into the room, you will need your student ID on your desk

+ Invigilators will give you exam instructions


## Things to take to exam  

+ Student Card!!

+ Pens (multiple) blue or black ink  

+ Pencil (for rough work)  

+ Ruler
  + If you like, not really needed, but might be useful.  
  
+ Calculator
  + Any type (see list on Learn of permitted calculators)  
  + No advantage to scientific vs normal  
  + **Phone/Watch can not be your calculator**  

## Things you will be given


::::{.columns}
:::{.column width="50%"}
**What you are given**

+ a specific sheet for your MCQ  

+ an answer book for all other questions  

+ a rough work book ( *if you want to use it* )  

+ the equation sheet

:::

:::{.column width="50%" .fragment}
**FILLING IN FRONT COVERS**

MCQ sheet:  

  - matriculation number (UUN but without the "s")  
  
Answer book:  

<iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_ms01z6h1&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_j0xl893i" width="400" height="285" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="How to fill in the cover of an exam script book?"></iframe>
:::
::::

## Things you can take away

<center>
<br><br>
<h2>NOTHING!</h2>  

(Other than the things you bring in with you like pens, pencils, calculator)

<br><br>
<b>
When you finish, you leave all papers (questions, answers, equation sheet) in the exam hall.  
</b></center>


# Questions?


# Exam Structure

## Exam Structure


::::{.columns}
:::{.column width="50%"}
**Section A - 30 MARKS**

+ 15 MCQ  

+ Just like quiz questions on Learn  

+ On all topics  
:::

:::{.column width="50%"}
**Section B - 70 MARKS**  

+ 8 Questions  

+ Marks range from 4 to 15 marks per question  
:::
::::

## Exam Strategy: Section A


::::{.columns}
:::{.column width="50%"}
**Section A - 30 MARKS**

+ 15 MCQ  

+ Just like quiz questions on Learn  

+ On all topics   

:::

:::{.column width="50%"}

15-18 minutes  

  + You can answer these quickly at the end
  + Do the ones you know, move on
  + Keep track of how many you have not done, so you can go back to them at the end

:::
::::


## Exam Strategy: Section B



::::{.columns}
:::{.column width="50%" .fragment}

Questions on:  

+ Some short calculation  
+ Some ask you to explain statistical concepts discussed in the course  
+ Some give you a research design/question and task you with specifying an appropriate model  
+ Some give you model output and ask you to provide interpretation  


:::

:::{.column width="50%"}
**Section B - 70 MARKS**  

+ 8 Questions  

+ Marks range from 4 to 15 marks per question  

```{r echo=FALSE}
tibble::tribble(
  ~content,~questions,
  "Weeks 1-5 (multilevel models)","4 questions totaling 32 marks",
  "Weeks 7-11 (pca, efa, psychometrics)","4 questions totalling 38 marks",
) |> gt::gt()
```
:::


::::

## Exam Strategy: Section B


::::{.columns}
:::{.column width="50%"}
Questions on:  

<ul>
<li>Some short calculation</li>  
<li style="opacity:.5">Some ask you to explain statistical concepts discussed in the course</li>
<li style="opacity:.5">Some give you a research design/question and task you with specifying an appropriate model</li>
<li style="opacity:.5">Some give you model output and ask you to provide interpretation</li>
</ul>

:::
:::{.column width="50%" style="font-size:.8em"}
**approx 5 marks**  

**show your working** (could get marks for this if answer is wrong)  

think, e.g.:  

- calculating icc from a intercept only model  
- calculating coef from fixef and ranef  
- calculating variance explained from eigenvalues
- calculating SS loadings  

:::
::::

## Exam Strategy: Section B


::::{.columns}
:::{.column width="50%"}
Questions on:  

<ul>
<li style="opacity:.5">Some short calculation</li>
<li>Some ask you to explain statistical concepts discussed in the course</li>
<li style="opacity:.5">Some give you a research design/question and task you with specifying an appropriate model</li>
<li style="opacity:.5">Some give you model output and ask you to provide interpretation</li>
</ul>

:::

:::{.column width="50%" style="font-size:.8em"}
**approx 6-8 marks**  

often these ask to define a couple of things.  
1 to 2 sentences each thing.  

think, e.g.:  

- "ICC"
- "random effects and fixed effects"
- "no pooling/complete pooling/partial pooling"
- "factor loading"
- "reliability"

:::
::::

## Exam Strategy: Section B


::::{.columns}
:::{.column width="50%"}
Questions on:  

<ul>
<li style="opacity:.5">Some short calculation</li>
<li style="opacity:.5">Some ask you to explain statistical concepts discussed in the course</li>
<li>Some give you a research design/question and task you with specifying an appropriate model</li>
<li style="opacity:.5">Some give you model output and ask you to provide interpretation</li>
</ul>

:::

:::{.column width="50%" style="font-size:.8em"}
**approx 10 marks**  

you have done this in the report, and lots of times in labs.  

Hint: PCA/EFA are not "research design/question"-based in the sense of estimation, so these questions are likely to be multilevel models.  
  
You probably know more than you think you do.  
- e.g. start with an outcome variable.  
- ask yourself: "what in my question is asking about 'effects on Y'/'influences on Y'/'predicts Y'".  

:::
::::


## Exam Strategy: Section B


::::{.columns}
:::{.column width="50%"}
Questions on:  

<ul>
<li style="opacity:.5">Some short calculation</li>
<li style="opacity:.5">Some ask you to explain statistical concepts discussed in the course</li>
<li style="opacity:.5">Some give you a research design/question and task you with specifying an appropriate model</li>
<li>Some give you model output and ask you to provide interpretation</li>
</ul>

:::

:::{.column width="50%" style="font-size:.8em"}
**approx 10-15 marks**  

you have done this in the report, and also seen it plenty of times in the labs and lectures  

- the questions will list the things we expect comments on
- don't just parrot back numbers. provide *interpretation*  
- these usually provide information about the research aims, so place your answer in that context.  

These are going to be our main functions we have seen throughout DAPR3: `lmer()`, `principal()`, `fa()`. Print an output from the lab/lectures and annotate it. what does each bit of the output represent?  

:::
::::


# Questions?


# Mock Questions
