---
title: "dapR3_liveR_week8 - Path Mediation"
author: "John Martindale"
date: "30/10/2023"
output: html_document
date: "2023-10-26"
editor_options: 
  chunk_output_type: console
---

# DAPR3 - Week 8 - Path Mediation - Live R

## Recap of last week programming

The goal for today is to specify, fit and evaluate a path model with mediation. First, let's load our packages:

```{r Setup, message=FALSE, warning=FALSE}
# Load tidyverse and lavaan packages

library(tidyverse)
library(lavaan)
```

Let's load in the data:

```{r Data import, }
# Load the data

leader <- read_csv("C:/Users/jmart/OneDrive/Documents/GitHub/dapr3/dapr3_livers/leader.csv")

# Preview the data
leader
```

1. Variable key
- ID = Partiicpant ID
- Leader = Abusive leadership behaviour
- Sleep = Sleep
- Exercise = Exercise
- Aggression = Interpersonal aggression at work
- swb = Sujective well-being

Let's specify a model:

```{r Model specification}
# Define the model
path1 = '
aggression ~ leader
swb ~ aggression
swb ~ leader
swb ~ exercise + sleep
'
```

- Draw the model on the board here

What are we saying here? What are the relationships?

1. Motivation predicts performance
2. Size of organisation predicts motivation

Now let's estimate and view:

```{r}
# Fit the model
path1_out <- sem(path1, data = leader)

# Print the model
summary(path1_out, # Model
        fit.measures = T, # Fit indices
        std = T
        )
```

## Week 8 - Mediation

Now let's add the mediation effects:

```{r}
path1 = '
# Direct effects
aggression ~ a*leader # aggression (M) predicted by abusive leader behaviour (X)
swb ~ b*aggression # well-being (Y) predicted by aggression (M)
swb ~ c*leader # well-being (Y) predicted by leader abusive behaviour (X): direct effect
swb ~ exercise + sleep # covariates

# Mediation effects
indirect := a*b
total := (a*b)+c
'
```

Estimate the model:

```{r}
# Estimate the model
model1.est <- sem(path1, data = leader)

# Report the model
summary(model1.est,
        fit.measures = T, 
        std = T,
        modindices = T)
```

Now we want to compute some confidence intervals for the indirect effects:

```{r}
# Estimate the model
model2.est <- sem(path1, data = leader, se = 'bootstrap')

# Report the model
summary(model1.est,
        fit.measures = T,
        ci = T,
        std = T,
        modindices = T)
```

