---
title: "dapr3 Live Rs"
author: "josiah king"
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sjPlot)
library(ICC)
library(lme4)

load(url("https://uoepsy.github.io/dapr3/2324/lectures/dapr3_liveRs.RData"))
```

after accounting for differences due to age, is the influence of caffeine consumption on reaction times dependent upon whether it coffee is taken with or without milk?  

**NOTE:** This data is all simulated (we've just used some random names of people in the psychology department. Nothing is accurate). I have no idea how old people are, nor how they take their coffee! (although FWIW, I gather that most days Tom has 2 espressos before 8am!).  


# Week 1

Consider two studies.  

**Scenario 1**  

First, we just get our participants, and record their reaction times. We also ask them how many coffees they have had that day, and whether they take milk.  

*Between-person*: One value per person

```{r}
psychRT_quick

# plot:
ggplot(psychRT_quick, aes(x=ncoffees, y=RT))+
  geom_point() +
  geom_smooth(method=lm,se=F)+
  geom_label(aes(label=pid),hjust=0,position=position_nudge(x=.1))

# fit a model:


```
**Scenario 2**  

We get our participants, and we record their reaction times after they have had 1 coffee. 
Then we give them another coffee, and we record their reaction times again.  
Then we give them another coffee, and we record their reaction times again.  
...  
...  
...  
We end up with some highly caffeinated participants, but we observe them at various levels of coffee consumption!  

*Within-person*: multiple values per person 

```{r}
psychRT

# plot
ggplot(psychRT, aes(x=ncoffees, y=RT))+
  geom_point() +
  geom_smooth(method=lm,se=F) +
  geom_label(data=filter(psychRT,pid=="Josiah"),
             aes(label=pid),hjust=0,
             position=position_nudge(x=.1))


# fit an lm
summary(lm(RT~ncoffees,psychRT))

mm = lm(RT~ncoffees,psychRT)
# residuals are NOT independent
plot(mm)

```
Variation in RT is partly attributable to differences between people.  
- it looks like Tom is possibly, on average, a bit quicker than Holly?
- and Bérengère possibly quicker still?  


```{r}
# ICC
library(ICC)
ICCbare(x=pid, y=RT,data=psychRT)

## plot


```


Let's pivot (if we have time)

```{r}
psychRT

psychRT %>% 
  select(-caff) %>%
  pivot_wider(names_from=ncoffees, values_from=RT) %>%
  pivot_longer(`0`:`6`, names_to="howmanycoffees", values_to="reaction")
```

# Week 3a

> after accounting for differences due to age, is the influence of caffeine consumption on reaction times dependent  upon whether it coffee is taken with or without milk?  

```{r}
# lmer(RT ~ 1 + age + ncoffee * milk + (1 + ncoffee | pid))
```


__design__  
We recruited N people from the psychology department, and measured their reaction times (RT). We then gave them a coffee, waited an hour, and measured their RT again. We then gave them _another_ coffee before measuring their RT again. and again. and again. Up to 7 coffees were consumed by each participant. Participants had coffee either with or without milk.  
**NOT A REAL STUDY**  

How many people? How many observations?  
```{r}
head(psychRT, 10L)

nrow(psychRT)

psychRT %>% count(pid) %>% nrow()
```


Plotting the data:
```{r}
ggplot(psychRT, aes(x=ncoffees, y=RT, col=milk))+
  geom_point() +
  geom_smooth(method=lm,se=F) +
  facet_wrap(~pid)

```

Fitting a model:
```{r}
mod = lmer(RT~ 1 + age + ncoffees * milk + 
             (1 + ncoffees | pid), data = psychRT) 
summary(mod)
```

Plotting the model:  
```{r}
library(sjPlot)
plot_model(mod, type="int")
```

Why can't we have age|pid or milk|pid?  
```{r}
lmer(RT~ 1 + age + ncoffees * milk + 
       (1 + age| pid), data = psychRT)

lmer(RT~ 1 + age + ncoffees * milk + 
       (1 + milk | pid), data = psychRT)

library(lattice)
xyplot(RT~milk|pid,data=psychRT)
xyplot(RT~age|pid,data=psychRT)
#compare with:
xyplot(RT~ncoffees|pid,data=psychRT)
xyplot(RT~ncoffees|pid,data=psychRT, type=c("p","r"))
```

for milk and age, we can't define "the effect of milk on RT" for a given participant (e.g. for Josiah, we only observe him having milky coffees, so we can't see how milk/no milk affects his RT).  
it's same for age.  
by contrast, we observe each participant across various values of coffee, so we can draw a line for each person, and we can imagine that line being different for each person.  

Weirdly though, we _can_ sometimes fit non-sensical models!  
i.e. this one doesn't give us any warning.  
it also doesn't give us anything useful, but it's a word of caution to NOT just "shove everything in and wait for errors"  
```{r}
lmer(RT~ 1 + milk + (1 + milk | pid), data = psychRT) %>%
  summary()
```
We need to actually _think_ about the study design!  

# Week 3b

Suppose instead of doing our experiment where we forced people to drink all this coffee, we instead simply _observed_ them over several days, over various amounts of coffee consumption.  

```{r}
library(patchwork)
ggplot(psychRT,aes(x=ncoffees,y=RT,group=pid,col=milk))+
  geom_point()+
  geom_smooth(method=lm,se=F) + 
  labs(title="study 1") +
ggplot(psych_OBS,aes(x=ncoffees,y=RT,group=pid,col=milk))+
  geom_point()+
  geom_smooth(method=lm,se=F) + labs(title="study 2")

library(lattice)
xyplot(RT~ncoffees|pid, psych_OBS, type=c("p","r"))
```

we could split this into two questions:

- do people who drink more coffee than other people, have faster/slower RTs?
- if a person drinks more coffee than they usually do, do their RTs increase/decrease?  


first, though, i'd like to talk briefly about scale and fitting models.  
```{r}
head(psych_OBS)
# this works
m1 = lmer(RT~ 1 + age + ncoffees * milk + 
            (1 + ncoffees | pid), data = psych_OBS)
# this doesn't work:  
m2 = lmer(RT~ 1 + age + caff * milk + 
            (1 + caff | pid), data = psych_OBS)
# why? because slope of caff is tiny compared to slope of coffees. because it's "1 increase in mg of caffeine". 
# can cause issues because the estimate variance in slopes is v small.  
VarCorr(m1)
VarCorr(m2)
```

Centering can also affect fitting
```{r}
# remember that intercepts are "when predictor is zero"
# consider the lecture data:
d3 <- read_csv("https://uoepsy.github.io/data/dapr3_mindfuldecline.csv")
library(patchwork)
ggplot(d3, aes(x=visit, y=ACE, group=ppt)) + 
  geom_point()+
  geom_line() +
ggplot(d3, aes(x=age, y=ACE, group=ppt)) + 
  geom_point()+
  geom_line()

m1 = lmer(ACE ~ 1 + visit + (1 + visit | ppt), data = d3)
m2 = lmer(ACE ~ 1 + age + (1 + age | ppt), data = d3)
VarCorr(m1)
VarCorr(m2)

#this is model 2, the ppt level fits plotted:  
plot_m2 <- 
  broom.mixed::augment(m2) %>%
  ggplot(data=.,aes(x=age,y=.fitted,group=ppt))+
  geom_point()+
  geom_line()

plot_m2

#this is where 1|ppt is estimated at
plot_m2 +
  stat_smooth(geom="line",method=lm,se=F,fullrange=T,lty="longdash")+
  xlim(0,78)

# contrasts with model 1
broom.mixed::augment(m1) %>%
  ggplot(data=.,aes(x=visit,y=.fitted,group=ppt))+
  geom_point()+
  geom_line()+
  stat_smooth(geom="line",method=lm,se=F,fullrange=T,lty="longdash")+
  xlim(0,10)

```


separating within between coffee effects

```{r}
psych_OBS <- 
  psych_OBS %>%
  group_by(pid) %>%
  mutate(
    avg_coff = mean(ncoffees),
    dev_coff = ncoffees - avg_coff
  ) %>%
  ungroup

m1 = lmer(RT~ 1 + age + (avg_coff + dev_coff) * milk + 
            (1 + dev_coff | pid), data = psych_OBS)
summary(m1)
plot_model(m1, type="int")
```



# Week 4

> after accounting for differences due to age, is the influence of caffeine consumption on reaction times dependent  upon whether it coffee is taken with or without milk?  

Nested example:

As above, with people in psychology department, but we have lots of people from other departments too:  

```{r}
uoeRT %>%
  count(dept)

uoeRT %>%
  count(dept, pid)

nrow(uoeRT)

uoeRT %>%
  count(pid)
```


```{r}
mnest = lmer(RT ~ 1+age+ncoffees*milk + (1+ncoffees|dept/pid),data=uoeRT)

mnest = lmer(RT ~ 1+age+ncoffees*milk + 
             (1+ncoffees|dept:pid)+
             (1+ncoffees+milk|dept),
             data=uoeRT)
summary(allFit(mnest))

mnest = lmer(RT ~ 1+age+ncoffees*milk + 
             (1+ncoffees|dept:pid)+
             (1|dept),
             data=uoeRT)

xyplot(RT~milk|dept,df_nest, type=c("p","r"))
VarCorr(mnest)
dotplot.ranef.mer(ranef(mnest))
```

Crossed example:  

As initial example, but we have lots of different types of task for RT time.  

```{r}
psychRT_tasks %>%
  count(pid)

psychRT_tasks %>%
  count(task)

psychRT_tasks %>%
  count(pid, task)
```

```{r}
mcross = lmer(RT ~ 1+age+ncoffees*milk +
                (1+ncoffees | pid) + 
                (1+ncoffees + milk|task),
              control=lmerControl(optimizer="bobyqa"),
              data=psychRT_tasks)

mcross = lmer(RT ~ 1+age+ncoffees*milk +
                (1+ncoffees | pid) + 
                (1+ncoffees |task),
              control=lmerControl(optimizer="bobyqa"),
              data=psychRT_tasks)

summary(mcross)
dotplot.ranef.mer(ranef(mcross))
```







