---
title: "dapR3_liveR_week9 - Principal Components Analysis"
author: "John Martindale"
date: "13/11/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

The goal for today is to calculate a PCA model. First, let's load our packages:

```{r Setup, message=FALSE, warning=FALSE}

# Load tidyverse and lavaan packages
library(tidyverse)
library(psych)
library(GPArotation)
```

Let's load in some data:

```{r Data}
data <- read.csv("https://uoepsy.github.io/data/conduct_probs.csv")
str(data)

data <- data %>%
  select(item1:item10)
```

Now let's view a sample correlation matrix (i.e., our input matrix):

```{r}
# Bartlett's test of sphericity
cortest.bartlett(cor(data), n = nrow(data))
# Significant, can press ahead!

# KMO
KMO(data)
# .87 = Merotorious!

cormat <- round(cor(data),2)
cormat
```

View the eigenvectors and eigenvalues:

```{r}
eigenmatrix <- (eigen(cor(data)))

eigenmatrix$values
```

## Determining the number of components to retain

```{r}
# Scree plot
plot(eigenmatrix$values, # Vector of eigenvalues
     type = 'b', # Display points and lines
     pch = 16, # 
     main = "Scree plot", 
     ylab = "Eigenvalues")

# Minimum average partial
vss(data)

# Parallel analysis
fa.parallel(data, n.iter = 500)
```

Running Factor Analysis and comparing two solutions:

```{r}
# Estimate different types of models
data_fa2_norotate <- fa(data, nfactors = 2, fm = "ml", rotate = "none")
data_fa2_orthog <- fa(data, nfactors = 2, fml = "ml", rotate = "varimax")
data_fa2_oblique <- fa(data, nfactors = 2, fm = "ml", rotate = "oblimin")

# View solutions
data_fa2_norotate
data_fa2_orthog
data_fa2_oblique

# Pull out loadings
data_fa2_norotate$loadings
data_fa2_oblique$loadings
data_fa2_orthog$loadings

# Factor correlations
data_fa2_orthog$Phi
round(data_fa2_oblique$Phi, 2)

# Evaluate fit
data_fa2_oblique$TLI
data_fa2_oblique$RMSEA
```

Now let's check the matrices to test our understanding and how it aligns with path analysis:

```{r Matrices}
# Original correlation matrix
cormat

# Model implied correlation matrix
cormat_mi <- data_fa2_oblique$model
cormat_mi

# View residual matrix
resid_mat <- data_fa2_oblique$residual
resid_mat[, 1:5]
```
