---
title: "dapR3_liveR_week7 - Path Analysis"
author: "John Martindale"
date: "30/10/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

# DAPR3 - Week 7 - Path Analysis - Live R

The goal for today is to specify, fit and evaluate a path model. First, let's load our packages:

```{r Setup, message=FALSE, warning=FALSE}
# Load tidyverse and lavaan packages

library(tidyverse)
library(lavaan)
```


Let's load and preview some data:

```{r Data import, }
# Load the data

org <- read_csv("C:/Users/jmart/OneDrive/Documents/GitHub/dapr3/dapr3_livers/archive_2223/org_performance.csv")

# Preview the data
org
```

When data cleaning, you should always make sure you've got a description somewhere over the variables in your data:

Get rid of this

1. What do you think each variable represents?
* OrgID = Which organisation is this observation from?
* Sex = Sex
* Mot = Motivation
* Perform = Performance
* PubPri = Public or private
* Region = UK Region
* Size = Number of people

## Standard linear model

Let's compare the model to a baseline standard linear model:

```{r Linear model}
# Fit the model
m1 <- lm(Perform ~ Mot + Sex+ PubPri, data = org)

# View the model
summary(m1)
```

We can also fit this in lavaan:

(Make sure to get the syntax wrong first. Pubpri not PubPri - Delete this in the lecture verson)

```{r Specifying linear model in lavaan}
# Specify the model
lm1 = 'Perform ~ Mot + Sex + PubPri'
```

Look at the environment, we can see this has created a single string value that is our model specification. This is the first part of the input for a path analysis function, and the data is the second part:

```{r Estimating linear model in lavaan}
# Estimate the model
lm1_out <- sem(lm1, data = org)

# Do you see the object it's created is different to a normal linear model? It's a special lavaan list

# View the model
summary(lm1_out)
```

1. What level of identification does this model have?


Extending our path model to multiple DVs:
```{r}
# Define the model
path1 = '
Perform ~ Mot
Mot ~ Sex + PubPri
'

# View all paths
lavaanify(path1)
```



```{r}
# Estimate the model
path1_out <- sem(path1, data = org)

# Print the model
summary(path1_out, # Model
        fit.measures = T, # Fit indices
        std = T
        )
```

We can also check the modification indices:

```{r}
# Modification indices
modindices(path1_out)
```

## Model evaluation

Make sure to talk about degrees of freedom here, count them when they appear in the model output

1. How many degrees of freedom?
2. What level of identification does this model have?
3. Any paths non-significant?

Compare to the coefficients - how do they change?

- Could add here about fitting a path model using lm()? From the 2223 archive maybe

```{r}
# Can we define the model inside the syntax?
path2_out <- sem('
  Perform ~ Mot
  Mot ~ Sex + PubPri
', data = org)

# Can we do it in one line?
path3_out <- sem('Perform ~ Mot; Mot ~ Sex + PubPri', data = org)

# Error in lav_syntax_parse_rhs(rhs = rhs.formula[[2L]], op = op) : 
#  lavaan ERROR: I'm confused parsing this line: ~MotMot

path3_out <- sem('Perform ~ Mot; Mot ~ Sex + PubPri', data = org)
```

Why do we get an error?