---
title: "dapR3 Live R week 8"
author: "Tom Booth"
date: "08/11/2021"
output: html_document
editor_options: 
  chunk_output_type: inline
---

1. Building more complex path models
  - Including covariates
  - Multiple predictors

2. Modification Indices
  - Trimming paths
  - Adding paths
  - MAKE SURE THEORY DRIVEN/SUPPORTED

3. Categorical variables
  - This one I am going to prep a separate Rmd and upload into this weeks folder. It is a little different in `lavaan` so worth having a document.

# Data

```{r}
library(tidyverse)
library(lavaan)
org <- read_csv("C:/Work/Teaching/org_performance2.csv")
str(org)
```

# Model building
Basic model:

```{r}
m1 = '
Perform ~ Mot + service 
Mot ~ service 
'

```

Run as a path model:

```{r}
m1_res <- sem(m1, org)
summary(m1_res, std = T)
```

Covariates:

```{r}
m2 = '
Perform ~ Mot + service + age + Size
Mot ~ service
'
```

Run the model 

```{r}
m2_res <- sem(m2, org)
summary(m2_res, fit.measures = T, std = T)
```


# Model modification
Extracting:

```{r}
modindices(m2_res)
```

Adding to a model:

```{r}
m3 = '
Perform ~ Mot + service + age + Size
Mot ~ service + age
'
```

Run the model
```{r}
m3_res <- sem(m3, org)
summary(m3_res, fit.measures = T, std = T)
```

