---
title: "dapR3_liveR_week7 - Path Analysis"
author: "John Martindale"
date: "30/10/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

# DAPR3 - Week 7 - Path Analysis - Live R

The goal for today is to specify, fit and evaluate a path model. First, let's load our packages:

```{r Setup, message=FALSE, warning=FALSE}
# Load tidyverse and lavaan packages

library(tidyverse)
library(lavaan)
```


Let's load and preview some data:

```{r Data import, }
# Load the data

org <- read_csv("C:/Users/jmart/OneDrive/Documents/GitHub/dapr3/dapr3_livers/archive_2223/org_performance.csv")

# Preview the data

org
```

When data cleaning, you should always make sure you've got a description somewhere over the variables in your data:

Get rid of this

1. What do you think each variable represents?
* OrgID = Which organisation is this observation from?
* Sex = Sex
* Mot = Motivation
* Perform = Performance
* PubPri = Public or private
* Region = UK Region
* Size = Number of people

## Standard linear model

Let's compare the model to a baseline standard linear model:

```{r Linear model}
# Fit the model
m1 <- lm(Perform ~ Mot + Sex+ PubPri, data = org)

# View the model
summary(m1)
```

We can also fit this in lavaan:

(Make sure to get the syntax wrong first. Pubpri not PubPri - Delete this in the lecture verson)

```{r Specifying linear model in lavaan}
# Specify the model
lm1 = 'Perform ~ Mot + Sex + PubPri'
```

Look at the environment, we can see this has created a single string value that is our model specification. This is the first part of the input for a path analysis function, and the data is the second part:

```{r Estimating linear model in lavaan}
# Estimate the model
lm1_out <- sem(lm1, data = org_data)

# Do you see the object it's created is different to a normal linear model? It's a special lavaan list

# View the model
summary(lm1_out)
```

We can standardise it too:

```{r Standardising linear model}

```

1. No degrees of freedom - What level of identification does this model have?

Extending 

```{r Model specification}
# Define the model
path1 = '
Perform ~ Mot
Mot ~ Sex + PubPri
'
```

Now let's estimate and view the model

```{r}
# Estimate the model


# Print the model

```

## Model evaluation

Make sure to talk about degrees of freedom here, count them when they appear in the model output

1. How many degrees of freedom?
2. What level of identification does this model have?
2. Any paths non-significant?

