---
title: "DAPR3_WEEK_4"
author: "uoepsy"
date: "2022-10-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(lme4)

fulldat <- read_csv("https://uoepsy.github.io/data/countrysleep.csv")
head(fulldat)
```

```{r}
m <- lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Subject) + 
           (1 + Age + Days | Country),
         control=lmerControl(optimizer="bobyqa"), 
         REML=TRUE,
         data = fulldat)
summary(m)
```

model is converging (finding a solution) but singular fit (not a solution we want).  

sometimes the warning is false positive, so we can check:
```{r}
isSingular(m)
```

checking the estimated parameters of a problematic model can help us to identify how to simplify:  
```{r}
VarCorr(m)
```


- age | Country is explaining comparatively little variance in reaction times
- also, focal predictor is Days  
- more important to include random slopes for Days?
  - especially for |Subjects, because we'd expect differences there - people likely to vary in how sleep deprivation influences their RT.
  - do countries vary? 
    - probably. lots of psychology results don't hold cross culturally.
    - do countries vary in how age influences RT? not sure. maybe Italians age a lot better than people in the UK or something? 
    
    
```{r}
m1 <- lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Subject) + 
           (1 + Days | Country),
         control=lmerControl(optimizer="bobyqa"), REML=TRUE,
         data = fulldat2)
isSingular(m1)
VarCorr(m1)
```
    
    
Options:   

- remove correlation between 1|Country and Days|Country
- remove Days|Country altogether (simpler model)
```{r}
m2A = lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Subject) + 
           (1 | Country),
         control=lmerControl(optimizer="bobyqa"), REML=TRUE,
         data = fulldat)

m2B = lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Subject) + 
           (1 + Days || Country),
         control=lmerControl(optimizer="bobyqa"), REML=TRUE,
         data = fulldat)
```

What does it mean to have 1 + Days||Country?  
Countrys vary in: 
- overall RT average
- amount by which RT is influenced by days of sleep deprivation 

and these are **NOT** correlated with one another. 

Our problematic model:  
```{r}
plot(ranef(m1)$Country)
cor(ranef(m1)$Country)
```

the model _without_ the correlation parameter
```{r}
plot(ranef(m2B)$Country)
cor(ranef(m2B)$Country)
```

plot the cluster lm models 
```{r}
ggplot(fulldat, aes(x=Days,y=Reaction,group=Country))+
  geom_smooth(method=lm, se=F)
```


# AN ASIDE

if 1|Country and Days|Country _WERE_ highly correlated.  
```{r}
fulldatCOR <- read_csv("https://uoepsy.github.io/data/countrysleep_CORRELATED.csv")

ggplot(fulldatCOR, 
       aes(x=Days, y=Reaction,group=Country))+
  geom_smooth(method=lm, se=F)


```

Even if we fit the 1+Days||Country model to this data, the random effects may well come out highly correlated 
```{r}
m2BCOR <- lmer(Reaction ~ Age + Caff + Days + 
             (1 + Days | Subject) + 
             (1 + Days || Country),
           control=lmerControl(optimizer="bobyqa"), REML=TRUE,
           data = fulldatCOR)

plot(ranef(m2BCOR)$Country)
cor(ranef(m2BCOR)$Country)
VarCorr(m2BCOR)
```





