---
title: "week 2 dapr3"
author: "uoepsy"
date: "2022-09-25"
output: html_document
---

Researchers are interested in how cognition changes over time. 
They measure 20 participants' cognition every year for 10 years.  
Scores on the cognitive test range from 0 to 100.
They also have data on whether the participant is left handed or not, with theories suggesting that Lhanded people decline less. 

```{r}
library(tidyverse)
library(lme4)
library(broom.mixed)

cogtime <- read_csv("https://uoepsy.github.io/data/cogtimerpm.csv")
head(cogtime)
```



# fitting multilevel models, summary()
Researchers are interested in how cognition changes over time. 

lm, random intercepts, random slopes,

```{r}
m1 <- lm(cog ~ visit_n, data = cogtime)
library(sjPlot)
plot_model(m1, type="pred", show.data=TRUE)
```


```{r}
m2 <- lmer(cog ~ 1 + visit_n + (1 | participant), data = cogtime)

augment(m2) %>%
  ggplot(., aes(x = visit_n, y = .fitted, group = participant)) +
  geom_point()+
  geom_line()

summary(m2)
```

```{r}
m3 <- lmer(cog ~ 1 + visit_n + (1 + visit_n | participant), data = cogtime)

augment(m3) %>%
  ggplot(., aes(x = visit_n, y = .fitted, group = participant)) +
  geom_point()+
  geom_line()

summary(m3)

plot_model(m3, type="pred")
library(effects)
as.data.frame(effect(term = "visit_n",mod = m3))
```

Not possible. 
to say "the effect of Lhandedness on cognition varies by participant (i.e. is different for each participant)" 
we would need to be able to define "the effect of Lhandedness on cognition" for a single participant. 
for Josiah, he is lefthanded. To think about "the effect of being Lhanded on cognition for Josiah", we would need data from Josiah when he is righthanded. But he's never right handed.. 

```{r}
m4 <- lmer(cog ~ 1 + visit_n * isLhanded + 
             (1 + visit_n + isLhanded| participant), data = cogtime)
summary(m4)

plot_model(m4, type="int")

# pick a few select participants to plot
cogtime %>% filter(participant > 16) %>% 
ggplot(., aes(x=isLhanded, y = cog))+
  geom_point()+
  geom_smooth(method=lm)+
  facet_wrap(~participant)

```



