---
title: "week1"
author: "uoepsystat"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sjPlot)
library(ICC)
library(lme4)

```

## lm?

ignore the clusters, let's just fit a linear model:
```{r}



```

what does this look like for each participant?
```{r}

```

design tells us that our errors are not independent. 
Also, it is plainly obvious - look at cluster x in relation to our model fit:
```{r}


```


# ICC

we have variance in y:
```{r}

```

variance in y between clusters
```{r}


```

we have variance in cluster means, and in observations around those means
```{r}


```

ICC is the ratio of the variance attributable to clusters, vs the overall variance: 
```{r}

```

# Pooling

fixed intercept adjustment: 
```{r}


```

fixed intercept and slope adjustment: 
```{r}


```

let's plot model fits
```{r}

```


we can see the reduction in residual sums of squares. it is significant
```{r}

```

but at what cost? we are now making cluster-specific estimates. e.g. "difference from cluster 1 to cluster 13".. 
```{r}


```


# pivotting (if time)




