---
title: "week2"
author: "uoepsystat"
date: "20/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ICC)
library(lme4)
library(effects)
library(broom.mixed)
library(sjPlot)
library(HLMdiag)
library(lmeresampler)

# Data from lectures
cogtime <- read_csv("https://uoepsy.github.io/data/cogtimerpm.csv") %>% 
  mutate(across(c(participant, sexFemale, alc), factor))
```

## lm?

how does cognition change over time?

ignore the participants, let's just fit a linear model:
```{r}
ggplot(cogtime, aes(x=visit_n, y=cog))+
  geom_point()+
  geom_smooth(method="lm")

simplemod <- lm(cog ~ visit_n,
                data=cogtime)
summary(simplemod)
```



what does this look like for each participant?
```{r}
ggplot(cogtime, aes(x=visit_n, y=cog))+
  geom_point()+
  facet_wrap(~participant)+
  geom_line(data=broom::augment(simplemod), aes(y=.fitted))
```

design tells us that our errors are not independent. 
Also, it is plainly obvious - look at participant 7 in relation to our model fit:
```{r}
ggplot(cogtime, aes(x=visit_n, y=cog))+
  geom_point()+
  geom_point(data=cogtime %>% filter(participant=="7"), col="red", size=3)+
  geom_line(data=broom::augment(simplemod), aes(y=.fitted))
```

let's adjust for each participants' average cognition score. some are higher, some are lower
```{r}
# no pooling
# intercept ajdustment for each participant
femod <- lm(cog ~ visit_n + participant, cogtime)
summary(femod)
```

let's also adjust for the change over time being differnt for each participant
```{r}
# intercept and slope adjustment for each participant
femod1 <- lm(cog ~ visit_n * participant, cogtime)
```

let's plot the model fits
```{r}
ggplot(cogtime, aes(x=visit_n, y=cog))+
  geom_point()+
  facet_wrap(~participant)+
  #geom_line(data=broom::augment(simplemod), aes(y=.fitted))+
  #geom_line(data=broom::augment(femod), aes(y=.fitted), col="blue")+
  geom_line(data=broom::augment(femod1), aes(y=.fitted), col="red")
```


A note on broom::augment()
we can either use
```{r}
# library(broom)
# augment()
```
or just say "in the package broom, use the function augment()
```{r}
# broom::augment()
```


we can see the reduction in residual sums of squares. it is significant
```{r}
anova(simplemod, femod, femod1)
```

but at what cost? we are now making participant-specific estimates. e.g. "difference from participant 1 to participant 13".. 
```{r}
summary(femod1)
```

## pivotting


```{r}
cogtime[2,"visit_n"] <- NA
cogwide <- cogtime %>% select(participant, visit_n, cog) %>%
  na.omit() %>%
  pivot_wider(names_from = visit_n, values_from = cog) 
```


```{r} 
cogwide %>% 
  pivot_longer(cols = 2:11, names_to = "visitnumber", values_to = "cognition")
```

## plotting and describing

```{r}
ggplot(cogtime, aes(x=visit_n,y=cog,group=participant))+
  geom_line()
```

```{r}
cogtime %>% group_by(participant) %>%
  summarise( meancog = mean(cog), sd=sd(cog))

```

```{r}
ggplot(cogtime, aes(x=cog))+
  geom_density()
```

## a bit of ICC

we have variance in cognition:
```{r}
ggplot(cogtime, aes(x=0, y=cog))+
  geom_jitter(height=0, width=.1)+
  geom_violin(alpha=.2)+
  xlim(-1,1)
```
we also have variance between participants
```{r}
ggplot(cogtime, aes(x=participant, y=cog))+
  #geom_point()+
  stat_summary(geom="pointrange")+
  labs(title="participant average cognition")
```
we have variance in participant means, and in observations around those means
```{r}
ggplot(cogtime, aes(x=participant, y=cog, col=participant))+
  geom_point()+
  stat_summary(geom="pointrange",col="black")
```
ICC is the ratio of the variance attributable to participants, vs the overall variance. 


