---
title: "EFA Walkthrough"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(psych)
```
```{r}
phdat <- read_csv("https://uoepsy.github.io/data/phoneaddiction.csv")
```


There are various ways of assessing the suitability of our items for exploratory factor analysis, and most of them rely on examining the observed correlations between items.  

## Look at the correlation matrix  

Use a function such as `cor` or `corr.test(data)` (from the **psych** package) to create the correlation matrix. If it's a big matrix (they get big quite quickly!) then we can make some sort of visual to get a better sense.  

```{r}
library(pheatmap)
pheatmap(cor(phdat), cluster_rows = F, cluster_cols = F)
```

::: {.callout-caution collapse="true"}
#### optional: what if our variables are not continuous?

In psychological research, we spend a lot of time recording responses on a Likert scale (e.g.: "strongly disagree","disgree","neither","agree","strongly agree"). 

If our measurements have 5+ Likert-like options, it is quite common convention to simply treat them as continuous variables. 

But if we have fewer response options (or if people only ever _used_ fewer options), then we don't want to just use a standard Pearson's correlation in order to capture the relationships.  

Instead we would want to use **polychoric** correlations, which we can get out the `polychoric()` function 

```{r}
#| eval: false
library(psych)
# get a matrix of polychoric correlations:  
polychoric(data)$rho
```

And we can easily then conduct the factor analysis by adding `cor = "poly"` into the use of the `fa()` function.  

```{r}
#| eval: false
fa(data, nfactors = ??, fm = ??, 
         rotate = ??, cor = "poly")
```


::::{.columns}
:::{.column width="40%"}
"Polychoric correlations" are estimates of the correlation between two theorized normally distributed continuous variables, based on their observed ordinal manifestations. 

:::
:::{.column width="60%"}
```{r}
#| echo: false
#| label: fig-polychor
#| fig-cap: "the idea of polychoric correlation is to estimate the correlation between two unobserved continuous variables by examining the distributions across the set of ordered categories that we *do* observe"
set.seed(444)
ddf <- tibble(
  item1 = rnorm(1e3),
  item2 = .6*item1 + rnorm(1e3)
)
ddf <- as.data.frame(apply(ddf,2,scale))
ddf2 <- apply(ddf,2,function(x) cut(x,3,labels=c("disagree","neither","agree")))
names(ddf) <- paste0("l",names(ddf))
ddf <- cbind(ddf,ddf2)


thresh1 <- ddf |> group_by(item1) |>
  summarise(
    min = min(litem1),
    max = max(litem1),
    lab = max-((max-min)/2)
  )

thresh2 <- ddf |> group_by(item2) |>
  summarise(
    min = min(litem2),
    max = max(litem2),
    lab = max-((max-min)/2)
  )

ggplot(ddf,aes(x=litem1,y=litem2,col=interaction(item1,item2)))+
  geom_point(size=3,alpha=.5)+
  guides(col="none") +
  geom_vline(data=thresh1[-2,], aes(xintercept=min))+
  geom_hline(data=thresh2[-2,], aes(yintercept=min))+
  geom_text(data=thresh1,inherit.aes=F,aes(x=lab,y=-3,label=item1))+
  geom_text(data=thresh2,inherit.aes=F,aes(y=lab,x=-3.3,label=item2),angle=90)+
  labs(x="underlying agreement with item 1",
       y="underlying agreement with item 2")
```
:::

::::


## Bartlett's Test  

The function `cortest.bartlett(cor(data), n = nrow(data))` conducts "Bartlett's test". This tests against the null that the correlation matrix is proportional to the identity matrix (a matrix of all 0s except for 1s on the diagonal).  

  - Null hypothesis: observed correlation matrix is equivalent to the identity matrix  
  - Alternative hypothesis: observed correlation matrix is not equivalent to the identity matrix.  
  
::: {.callout-note collapse="true"}
#### What is the identity matrix?

The "Identity matrix" is a matrix of all 0s except for 1s on the diagonal.  
e.g. for a 3x3 matrix:  
$$
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1 \\
\end{bmatrix}
$$
If a correlation matrix looks like this, then it means there is __no__ shared variance between the items, so it is not suitable for factor analysis.  

:::


```{r}
cortest.bartlett(phdat)
```

## Kaiser, Meyer, Olkin Measure of Sampling Adequacy  

You can check the "factorability" of the correlation matrix using `KMO(data)` (also from **psych**!).  

- Rules of thumb: 
    - $0.8 < MSA < 1$: the sampling is adequate
    - $MSA <0.6$: sampling is not adequate 
    - $MSA \sim 0$: large partial correlations compared to the sum of correlations. Not good for FA  
    
    
::: {.callout-note collapse="true"}
#### Kaiser's suggested cuts  

These are Kaiser's corresponding adjectives suggested for each level of the KMO:  

- 0.00 to 0.49 "unacceptable"  
- 0.50 to 0.59 "miserable"  
- 0.60 to 0.69 "mediocre"  
- 0.70 to 0.79 "middling"  
- 0.80 to 0.89 "meritorious"  
- 0.90 to 1.00 "marvelous"  

:::

```{r}
KMO(phdat)
```

## Check for linearity  

It also makes sense to check for linearity of relationships prior to conducting EFA. EFA is all based on correlations, which assume the relations we are capturing are linear.  

You can check linearity of relations using `pairs.panels(data)` (also from **psych**), and you can view the histograms on the diagonals, allowing you to check univariate normality (which is usually a good enough proxy for multivariate normality). 

```{r}
pairs.panels(phdat)
```



## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
