---
title: "EFA Walkthrough"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(psych)
```

:::frame
__Dataset: phoneaddiction.csv__  

The dataset at [https://uoepsy.github.io/data/phoneaddiction.csv](https://uoepsy.github.io/data/phoneaddiction.csv){target="_blank"} comes from a (fake) study that is interested in developing a measure of "phone-attachment/addiction" - i.e., the idea of being overly attached to a phone.  

We have a set of 10 statements (@tbl-phdict) that get at different aspects of this idea, and we asked 240 people to rate how much they agreed with each of the 10 statements on a 1-5 scale (1 = strongly disagree, 5 = strongly agree).

```{r}
phdat <- read_csv("https://uoepsy.github.io/data/phoneaddiction.csv")
```


```{r}
#| echo: false
#| label: tbl-phdict
#| tbl-cap: "Item wordings for phoneaddiction.csv"
itemqs = c("I often reach for my phone even when I don't have a specific reason to use it.",
"I feel a strong urge to check my phone frequently, even during meals or social interactions.",
"I spend a large portion of my day using my phone, often for more hours than I intend.",
"My phone use has interfered with my ability to complete tasks or responsibilities at school, work, or home.",
"When I cannot use my phone (e.g., no battery or no signal), I feel anxious or uncomfortable.",
"My phone use has negatively affected my sleep schedule, such as staying up late scrolling.",
"I find it difficult to avoid checking my phone immediately upon waking up or right before bed.",
"I often check my phone even in situations where it's distracting or socially inappropriate (e.g., meetings or classes).",
"I sometimes feel that others are overly concerned about my phone use in social situations.",
"I have tried to reduce my phone usage but find it challenging to do so.")

m = "
 lv1 =~ .8*item1 + .8*item2 + .8*item3 + .8*item7 + .8*item8 + .1*item9
 lv2 =~ .8*item4 + .8*item5 + .8*item6 + .2*item9 + .8*item10
 lv1 ~~ .5*lv2
"
set.seed(2345)
df = lavaan::simulateData(m,sample.nobs = 240)
df = apply(df, 2, \(x) as.numeric(cut(x,5))) |> as_tibble()
df = df |> select(paste0("item",1:10))
phdat <- df
# write_csv(phdat, file="../../data/phoneaddiction.csv")

tibble(
  variable = names(df),
  question = itemqs
) |> gt::gt()

```
:::


## 1. Initial checks

Most of the correlations are in the range from 0.2 to 0.4. So they aren't hugely strong correlations, but they _are_ all related. From the plot below item9 already looks to be a weird one, which is something to keep in mind.  

```{r}
library(pheatmap)
pheatmap(cor(phdat), cluster_rows = F, cluster_cols = F)
```


The Bartlett test is significant here, meaning that we reject the null hypothesis that our correlation matrix is equivalent to the identity matrix.  

This is a good thing - we _want_ it to be different in order to justify doing factor analysis.  
```{r}
cortest.bartlett(phdat)
```

Overall KMO is good here at 0.83, and individual item KMOs are also looking pretty 'meritorious' too! The only thorny one is item9, again!  
```{r}
KMO(phdat)
```

Because these items are measured on likert scales of 1-5, looking at scatterplots can be a bit harder. The red lines on the plots are essentially "smooth" lines, which follow the mean of the variable on the y-axis as it moves across the x-axis. So we want them to all be straight-ish, which they are.  We can see from the diagonals of this plot that the items are all fairly normally distributed too.  

```{r}
pairs.panels(phdat)
```


## 2. Rotations and Factor Extraction

**Rotations:**
If we are thinking that there might be two or more factors that underly our measure of 'phone-addiction', then we would probably expect those two factors to be correlated (all the questions are about phone usage, after all). For that reason, we are going to use an **oblique rotation**.^[Even if these two factors are not correlated, then an oblique rotation could estimate that too. The point of an oblique rotation is that we're not _forcing_ the factors to be uncorrelated.]  

**Factor extraction:**
As all of the correlations between individual items are low to medium (0.2 - 0.4), and the item distributions look fairly normal, we're going to opt to use maximum likelihood here.  

## 3 & 4. How many factors?

The question of how many factors to extract is very similar to the methods discussed in the PCA demonstration in [Chapter 2](02_pca.html). 


::: {.callout-note collapse="true"}
#### Scree plot 

```{r}
scree(phdat, pc = TRUE, factors = FALSE)
```

:::

::: {.callout-note collapse="true"}
#### Parallel analysis

We can set `fa = "both"` to do the parallel analysis method for both factor extraction and principal components.  

`r set.seed(33)`

```{r}
fa.parallel(phdat, fa = "both")
```

:::

::: {.callout-note collapse="true"}
#### MAP

```{r}
VSS(phdat, rotate = "oblimin", plot = FALSE, fm = "ml")
```

:::

<!-- EP TODO: What's "kaiser"? Not mentioned elsewhere. -->

| method  | suggestion |
| ------ |------ |
| kaiser |  3   |
| scree plot | 1 or 2? or 5? quite hard to tell |  
| MAP | 1 |
| parallel analysis | 2 |

From this I would say that it is worth considering both a 1 factor and a 2 factor solution.  

## 5. Doing EFA & Comparing solutions

```{r}
ph_efa1 <- fa(phdat, nfactors = 1, 
              fm = "ml")
ph_efa2 <- fa(phdat, nfactors = 2, 
              fm = "ml", rotate = "oblimin")
```

Both solutions look reasonable _apart from item 9_ (which we could have guessed right at the beginning).  
In both solutions, item 9 has smaller loadings than the other items.
The smaller the loading, the less well the item is targeting whatever construct the factor represents.


::::{.columns}
:::{.column width="48%"}
__1 factor solution__  

```{r}
print(ph_efa1$loadings, cutoff=.2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
__2 factor solution__  

```{r}
print(ph_efa2$loadings, cutoff=.2)
ph_efa2$Phi
```
:::
::::

Looking at item9, the wording is "`r itemqs[9]`" This maybe represents something slightly different from just "being addicted to your phone" - it also contains a self-consciousness / feeling judged by others etc, which is something completely different.  

So where have we got to? We started with loads of responses to a set of 10 questions, and we wanted to understand the underlying structure of the thing they were intended to measure ("phone addiction").  
We've ended by identifying one question that didn't really fit with the rest (item9), and with a question mark between whether the measures capture 2 underlying factors or just 1.

A lot of where we go next depends on what our overarching research aims are. If we want to **use** the construct "phone-addiction" in a subsequent analysis, then we'll need to extract scores, and we have a decision of what structure (1 factor or 2 factors) we want to use - and so how many scores each person will have.  

## 6 - are we developing a measurement tool?  

If our goal here is to design a nice questionnaire for future research that captures stuff about people's addiction to their phones, then item 9 isn't proving very useful here. At this point we may also want to consider if there are any other items that are maybe redundant (it's always good to try and keep a questionnaire short to reduce participant burden!).

Removing item9, and doing the **entire** process over again, we end up still finding that it's between a 1 or a 2 factor solution. 

::: {.callout-note collapse="true"}
#### determining the number of factors  

```{r}
scree(phdat[,-9])
fa.parallel(phdat, fa = "both")
VSS(phdat, rotate = "oblimin", plot = FALSE, fm = "ml")
```

:::

::::{.columns}
:::{.column width="48%"}
__1 factor solution (excluding item 9)__  

```{r}
ph_efa1a <- fa(phdat[,-9], nfactors = 1, 
               fm = "ml")
print(ph_efa1a$loadings, cutoff = .3)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
__2 factor solution (excluding item 9)__  

```{r}
ph_efa2a <- fa(phdat[,-9], nfactors = 2, 
               rotate = "oblimin", fm = "ml")
print(ph_efa2a$loadings, cutoff = .3)
ph_efa2a$Phi
```
:::
::::

Finally, we now have a judgment call. Both solutions look okay in terms of the numbers of items on each factor etc. One thing to consider is whether we prefer the idea of 1 thing that explains 28% of the variance, or 2 *related* things (the two factors are fairly correlated in the 2 factor solution) that together explain 35%. Is that a big enough increase?   

The other thing we should definitely be considering is the theoretical coherence of the solutions. The 1 factor solution is easy - we have just one factor, and that represents "addiction to your phone".  
In the 2 factor solution we need to pay more attention to which items go with which factor.  

Given the item wordings below - we might try to define Factor 1 as "frequency of use" and Factor 2 as "impact on life".  
Are these distinct enough for you? Personally I would say maybe not - some of the wordings of items in Factor 1 are still related to "impact on life" (i.e., "often for more hours than I intend"?)  

::::{.columns}
:::{.column width="48%"}
__Factor 1__  

```{r}
#| echo: false
tibble(
  variable = names(df),
  question = itemqs
)[c(1,2,3,7,8),] |> gt::gt()

```


:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
__Factor 2__  

```{r}
#| echo: false
tibble(
  variable = names(df),
  question = itemqs
)[c(4,5,6,10),] |> gt::gt()
```


:::
::::

Developing measurement instruments (often called "scale development") is lot of back and forth, and at this point we might consider re-wording some of the items to make the two factors more clearly distinct, then giving the edited questionnaire to a whole load of new people, and going through the whole process again until we're happy with a coherent structure!


# Scores

After all the focus on trying to better understand the underlying structure of factors that are resulting in our observed responses, we have somewhat lost sight of the fact that this might all be in aid of some other purpose - trying to create a score for each person in our dataset for each of the factors!  

Much like PCA (see [Chapter 2](02_pca.html)), the scores from factor models will be taking into account the extent to which each individual observed variable might differently reflect the underlying factor. So this will contrast with naive mean/sum scores which assume each variable is equally reflective of the construct.  

Unlike PCA, however, in factor analysis everything is **indeterminate**. Much like how we can have infinitely many sets of loadings (depend on how we rotate them) that are numerically indistinguishable in terms of the implied model, there are infinitely many sets of factor scores that we could create - we have "factor score indeterminacy".   

There are many various methods to estimated factor scores, and they will all combine the observed responses, the factor loadings, and the factor correlations.  

We can get them all using the `factor.scores()` function. We need to give this function our data, our factor model, and tell it which method to use.  

So let's assume we want to use our 2 factor model of phone-addiction, and we are removing that problematic question 9.  

```{r}
ph_efa2a <- fa(phdat[,-9], nfactors = 2, 
               rotate = "oblimin", fm = "ml")
```

For example, to extract factor scores using the Bartlett method:  

```{r}
#| eval: false
factor.scores(phdat[,-9], ph_efa2a, method="Bartlett")$scores
```
```{r}
#| echo: false
factor.scores(phdat[,-9], ph_efa2a, method="Bartlett")$scores |> head()
```

A brilliant paper by Skrondal & Laake^[Skrondal, A., & Laake, P. (2001). Regression among factor scores. Psychometrika, 66, 563-575.] suggests the following methods to be used depending on your eventual purpose:  

- If the construct is going to be used as a dependent variable, use Bartlett (`method = "Bartlett"`)  
- If the construct is going to be used as a predictor, use Thurstone (`method = "Thurstone"`)  
- If the construct is a covariate, less important



## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
