---
title: "example slides"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(patchwork)
source('_theme/theme_quarto.R')
```

# Course Overview

```{r}
#| results: "asis"
#| echo: false
block1_name = "multilevel modelling<br>working with group structured data"
block1_lecs = c("regression refresher",
                "introducing multilevel models",
                "more complex groupings",
                "centering, assumptions, and diagnostics",
                "recap")
block2_name = "factor analysis<br>working with multi-item measures"
block2_lecs = c(
  "what is a psychometric test?",
  "using composite scores to simplify data (PCA)",
  "uncovering underlying constructs (EFA)",
  "more EFA",
  "recap"
  )

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=6)
```

# This week {transition="slide"}

- section1
    - subsection 1a
    - subsection 1b
    - subsection 1c
- section2
    - subsection 2a
    - subsection 2b
    - subsection 2c


# SECTION

## my slide

## my next slide

## my next slide {visibility="uncounted"}

this isn't counted

## my third slide {.incremental}

- thing1

- thing2

- thing3

## my fourth slide {.smaller}

- small things

- small things

- small things

## my fifth slide


::::{.columns}
:::{.column width="50%"}
column 1  blahblahblah
:::

:::{.column width="50%"}
column 2  blahblahblah
:::
::::


## my sixth slide

this is some stuff

:::{.fragment}

Oh wow, this stuff all arrives as a fragment!

```{r}
#| echo: true
# some R code
```

:::








