---
title: "Getting PCA Scores"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


```{r}
#| include: false
somedata = lavaan::simulateData(
  paste0("f1=~",paste0(runif(3,.8,1),"*y",1:3,collapse="+"),
         "\nf2=~",paste0(runif(3,.8,1),"*y",4:6,collapse="+"),
         "\nf1~~-.3*f2")
) |> mutate(across(everything(), ~as.numeric(cut(.,9))))

cor(somedata) |> round(2)

set.seed(475)
nitem <- 6
A <- matrix(runif(nitem^2,.41,1)*2-1, ncol=nitem) 
scor <- t(A) %*% A
df <- MASS::mvrnorm(n=220,mu=rep(0,6),Sigma = scor) |> as.data.frame()

physfunc <- df |> transmute(
  mobility = round(550 + scale(V1)[,1]*75),
  balance = round(20 + scale(V2)[,1]*5, 2),
  flexibility = round(7.5 + scale(V3)[,1]*1.25 + 2),
  strength = round(35 + scale(V4)[,1]*7.5),
  coordination = round(1.5+scale(V5)[,1]*.25, 3),
  fine_motor = round(20+scale(V6)[,1]*2.5,2)
) 

nitem <- 2
A <- matrix(runif(nitem^2,.41,1)*2-1, ncol=nitem) 
scor <- t(A) %*% A
df <- MASS::mvrnorm(n=220,mu=rep(0,2),Sigma = scor) |> as.data.frame()
physfunc2 <- df |> transmute(
  respiratory = round(550 + scale(V1)[,1]*100),
  endurance = round(15 + scale(V2)[,1]*2.5, 2)
)
physfunc <- bind_cols(physfunc,physfunc2) |>
  relocate(endurance, .after=balance)
# write_csv(physfunc, file="../../data/physfunc.csv")

```

## Data

:::frame
__Dataset: physfunc.csv__  

These data contain measurements on 220 individuals on various measures of physical functioning:  

- Mobility: 6-Minute Walk Test (distance in metres walked in 6 minutes).
- Balance: Single-Leg Stance Test (time in seconds standing on one leg). 
- Endurance: Stair Climb Test (time in seconds taken to climb a set number of stairs).
- Flexibility: Sit and Reach Test (distance in cm reached beyond the feet while seated).
- Strength: Grip Strength Test (measured in kilograms using a dynamometer).
- Fine Motor Skills: Nine-Hole Peg Test (time in seconds to place and remove pegs from a board).
- Respiratory Function: Forced Expiratory Volume in 1 Second (FEV1) (measured in Liters per second using a spirometer).  

Dataset: [https://uoepsy.github.io/data/physfunc.csv](https://uoepsy.github.io/data/physfunc.csv){target="_blank"}

:::

```{r}
physfunc <- read_csv("https://uoepsy.github.io/data/physfunc.csv")
head(physfunc)
```

## Doing PCA


Using `principal()` from the **{psych}** package, with 8 observed variables, we are working in 8 observed dimensions (our measured variables). These variables are all correlated, and PCA re-expresses the information onto **un**-correlated (i.e. perpendicular) dimensions.  

With 8 measured variables, we get out 8 orthogonal dimensions, which here we call "principal components".  
In the `principal()` function, we can set `nfactors = 8` to extract 8 dimensions. 
If we specified a lower number, e.g., `nfactors = 6`, it would simply _not show_ the 7th and 8th components.  

Setting `rotate = "none"` ensures that these new dimensions are orthogonal to one another.   

```{r}
library(psych)
fullpca <- principal(physfunc, nfactors = 8, 
                     rotate = "none")
```


The output of our PCA gives us two sections: 

- "loadings": correlations between each new dimension and the original dimensions
- "SSloadings": total variance captured by each new dimension.  

Along with various other methods, we can use the latter to decide how many components to keep.  

::: {.callout-tip collapse="true"}
#### the output

```{r}
fullpca
```

:::


## How many components should we keep?  

Using methods detailed in [Deciding on the Number of Components](p01_ndim.html), we make a decision on how many components to keep, and therefore how many to discard.  

For this dataset, we might end up with these recommendations:  

| method | suggestion |
| ------ | ---------- |
| scree plot | 1 or 2 |
| MAP | 1 | 
| parallel analysis | 2 |
| keeping 50% variance | 1 |
| keeping 75% variance | 2 |
| keeping 80% variance | 3 |
| keeping 90% variance | 4 |

::: {.callout-tip collapse="true"}
#### See for yourself

__Variance captured__  
We can see the cumulative proportion of variance being captured in the bottom row here:  

```{r}
fullpca$Vaccounted
```


The first component captures 55.8%  
The first two components capture 78.6%
The first three components capture 88.5%  
...
...
All eight components capture 100%  

__Scree Plot__  

The screeplot shows a possible kink at either the second or third, indicating that we keep either 1 or 2.  
```{r}
cor(physfunc) |>
  scree()
```

__Minimum Average Partial (MAP)__  
There's a lot of output from `VSS()` and we're only really interested in the map values, so let's pull those out.  
We can see that the first value is lowest, suggesting 1 component:  
```{r}
VSS(physfunc, n = 8, 
    rotate = "none", plot = FALSE)$map
```


__Parallel Analysis__  
This approach suggests 2 components:  
```{r}
fa.parallel(physfunc,
            fa = "pc",
            n.iter = 500)
```
:::


Let's suppose that we decide to retain two components, capturing 79% of the total variance:   

```{r}
pca_2dim <- principal(physfunc, 
                      nfactors = 2, rotate = "none")
```


## extracting scores

When it comes to _using_ the principal components in some subsequent analysis, we need to access the "scores".  

These are already stored in the PCA, in the `$scores` slot:  

```{r}
#| eval: false
pca_2dim$scores
```
```
        PC1      PC2
[1,]  0.84411 -0.91828
[2,]  0.30184 -0.73377
[3,]  1.40557 -0.10588
[4,] -0.62169 -0.08713
[5,] -0.05582  1.94701
[6,]  0.23296  0.25940
..      ...      ...
..      ...      ...
```


So we can add these scores to our data: 
```{r}
physfunc$pc1 <- pca_2dim$scores[,1]
physfunc$pc2 <- pca_2dim$scores[,2]
```
And then go and use those in some subsequent analysis instead of our original 8 variables!  

