---
title: "EFA Step 5. Estimation and Comparison"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(psych)
```

:::frame
__Dataset: phoneaddiction.csv__  

The dataset at [https://uoepsy.github.io/data/phoneaddiction.csv](https://uoepsy.github.io/data/phoneaddiction.csv){target="_blank"} comes from a (fake) study that is interested in developing a measure of "phone-attachment/addiction" - i.e., the idea of being overly attached to a phone.  

We have a set of 10 statements (@tbl-phdict) that get at different aspects of this idea, and we asked 240 people to rate how much they agreed with each of the 10 statements on a 1-5 scale (1 = strongly disagree, 5 = strongly agree).

```{r}
phdat <- read_csv("https://uoepsy.github.io/data/phoneaddiction.csv")
```


```{r}
#| echo: false
#| label: tbl-phdict
#| tbl-cap: "Item wordings for phoneaddiction.csv"
itemqs = c("I often reach for my phone even when I don't have a specific reason to use it.",
"I feel a strong urge to check my phone frequently, even during meals or social interactions.",
"I spend a large portion of my day using my phone, often for more hours than I intend.",
"My phone use has interfered with my ability to complete tasks or responsibilities at school, work, or home.",
"When I cannot use my phone (e.g., no battery or no signal), I feel anxious or uncomfortable.",
"My phone use has negatively affected my sleep schedule, such as staying up late scrolling.",
"I find it difficult to avoid checking my phone immediately upon waking up or right before bed.",
"I often check my phone even in situations where it's distracting or socially inappropriate (e.g., meetings or classes).",
"I sometimes feel that others are overly concerned about my phone use in social situations.",
"I have tried to reduce my phone usage but find it challenging to do so.")

m = "
 lv1 =~ .8*item1 + .8*item2 + .8*item3 + .8*item7 + .8*item8 + .1*item9
 lv2 =~ .8*item4 + .8*item5 + .8*item6 + .2*item9 + .8*item10
 lv1 ~~ .5*lv2
"
set.seed(2345)
df = lavaan::simulateData(m,sample.nobs = 240)
df = apply(df, 2, \(x) as.numeric(cut(x,5))) |> as_tibble()
df = df |> select(paste0("item",1:10))
phdat <- df
# write_csv(phdat, file="../../data/phoneaddiction.csv")

tibble(
  variable = names(df),
  question = itemqs
) |> gt::gt()

```
:::

```{r}
ph_efa1 <- fa(phdat, nfactors = 1, 
              fm = "ml")
ph_efa2 <- fa(phdat, nfactors = 2, 
              fm = "ml", rotate = "oblimin")
```

Both solutions look reasonable _apart from item 9_ (which we could have guessed right at the beginning).  
In both solutions, item 9 has smaller loadings than the other items.
The smaller the loading, the less well the item is targeting whatever construct the factor represents.


::::{.columns}
:::{.column width="48%"}
__1 factor solution__  

```{r}
print(ph_efa1$loadings, cutoff=.2)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
__2 factor solution__  

```{r}
print(ph_efa2$loadings, cutoff=.2)
ph_efa2$Phi
```
:::
::::

Looking at item9, the wording is "`r itemqs[9]`" This maybe represents something slightly different from just "being addicted to your phone" - it also contains a self-consciousness / feeling judged by others etc, which is something completely different.  

So where have we got to? We started with loads of responses to a set of 10 questions, and we wanted to understand the underlying structure of the thing they were intended to measure ("phone addiction").  
We've ended by identifying one question that didn't really fit with the rest (item9), and with a question mark between whether the measures capture 2 underlying factors or just 1.

A lot of where we go next depends on what our overarching research aims are. If we want to **use** the construct "phone-addiction" in a subsequent analysis, then we'll need to extract scores, and we have a decision of what structure (1 factor or 2 factors) we want to use - and so how many scores each person will have.  

