---
title: "EFA Step 6. Scale Development"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(psych)
```

:::frame
__Dataset: phoneaddiction.csv__  

The dataset at [https://uoepsy.github.io/data/phoneaddiction.csv](https://uoepsy.github.io/data/phoneaddiction.csv){target="_blank"} comes from a (fake) study that is interested in developing a measure of "phone-attachment/addiction" - i.e., the idea of being overly attached to a phone.  

We have a set of 10 statements (@tbl-phdict) that get at different aspects of this idea, and we asked 240 people to rate how much they agreed with each of the 10 statements on a 1-5 scale (1 = strongly disagree, 5 = strongly agree).

```{r}
phdat <- read_csv("https://uoepsy.github.io/data/phoneaddiction.csv")
```


```{r}
#| echo: false
#| label: tbl-phdict
#| tbl-cap: "Item wordings for phoneaddiction.csv"
itemqs = c("I often reach for my phone even when I don't have a specific reason to use it.",
"I feel a strong urge to check my phone frequently, even during meals or social interactions.",
"I spend a large portion of my day using my phone, often for more hours than I intend.",
"My phone use has interfered with my ability to complete tasks or responsibilities at school, work, or home.",
"When I cannot use my phone (e.g., no battery or no signal), I feel anxious or uncomfortable.",
"My phone use has negatively affected my sleep schedule, such as staying up late scrolling.",
"I find it difficult to avoid checking my phone immediately upon waking up or right before bed.",
"I often check my phone even in situations where it's distracting or socially inappropriate (e.g., meetings or classes).",
"I sometimes feel that others are overly concerned about my phone use in social situations.",
"I have tried to reduce my phone usage but find it challenging to do so.")

m = "
 lv1 =~ .8*item1 + .8*item2 + .8*item3 + .8*item7 + .8*item8 + .1*item9
 lv2 =~ .8*item4 + .8*item5 + .8*item6 + .2*item9 + .8*item10
 lv1 ~~ .5*lv2
"
set.seed(2345)
df = lavaan::simulateData(m,sample.nobs = 240)
df = apply(df, 2, \(x) as.numeric(cut(x,5))) |> as_tibble()
df = df |> select(paste0("item",1:10))
phdat <- df
# write_csv(phdat, file="../../data/phoneaddiction.csv")

tibble(
  variable = names(df),
  question = itemqs
) |> gt::gt()

```
:::


If our goal here is to design a nice questionnaire for future research that captures stuff about people's addiction to their phones, then item 9 isn't proving very useful here. At this point we may also want to consider if there are any other items that are maybe redundant (it's always good to try and keep a questionnaire short to reduce participant burden!).

Removing item9, and doing the **entire** process over again, we end up still finding that it's between a 1 or a 2 factor solution. 

::: {.callout-note collapse="true"}
#### determining the number of factors  

```{r}
scree(phdat[,-9])
fa.parallel(phdat, fa = "both")
VSS(phdat, rotate = "oblimin", plot = FALSE, fm = "ml")
```

:::

::::{.columns}
:::{.column width="48%"}
__1 factor solution (excluding item 9)__  

```{r}
ph_efa1a <- fa(phdat[,-9], nfactors = 1, 
               fm = "ml")
print(ph_efa1a$loadings, cutoff = .3)
```
:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
__2 factor solution (excluding item 9)__  

```{r}
ph_efa2a <- fa(phdat[,-9], nfactors = 2, 
               rotate = "oblimin", fm = "ml")
print(ph_efa2a$loadings, cutoff = .3)
ph_efa2a$Phi
```
:::
::::

Finally, we now have a judgment call. Both solutions look okay in terms of the numbers of items on each factor etc. One thing to consider is whether we prefer the idea of 1 thing that explains 28% of the variance, or 2 *related* things (the two factors are fairly correlated in the 2 factor solution) that together explain 35%. Is that a big enough increase?   

The other thing we should definitely be considering is the theoretical coherence of the solutions. The 1 factor solution is easy - we have just one factor, and that represents "addiction to your phone".  
In the 2 factor solution we need to pay more attention to which items go with which factor.  

Given the item wordings below - we might try to define Factor 1 as "frequency of use" and Factor 2 as "impact on life".  
Are these distinct enough for you? Personally I would say maybe not - some of the wordings of items in Factor 1 are still related to "impact on life" (i.e., "often for more hours than I intend"?)  

::::{.columns}
:::{.column width="48%"}
__Factor 1__  

```{r}
#| echo: false
tibble(
  variable = names(df),
  question = itemqs
)[c(1,2,3,7,8),] |> gt::gt()

```


:::
:::{.column width="4%"}
:::
:::{.column width="48%"}
__Factor 2__  

```{r}
#| echo: false
tibble(
  variable = names(df),
  question = itemqs
)[c(4,5,6,10),] |> gt::gt()
```


:::
::::

Developing measurement instruments (often called "scale development") is lot of back and forth, and at this point we might consider re-wording some of the items to make the two factors more clearly distinct, then giving the edited questionnaire to a whole load of new people, and going through the whole process again until we're happy with a coherent structure!

