---
title: "EFA Steps 3/4. How Many Factors?"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(psych)
```

:::frame
__Dataset: phoneaddiction.csv__  

The dataset at [https://uoepsy.github.io/data/phoneaddiction.csv](https://uoepsy.github.io/data/phoneaddiction.csv){target="_blank"} comes from a (fake) study that is interested in developing a measure of "phone-attachment/addiction" - i.e., the idea of being overly attached to a phone.  

We have a set of 10 statements (@tbl-phdict) that get at different aspects of this idea, and we asked 240 people to rate how much they agreed with each of the 10 statements on a 1-5 scale (1 = strongly disagree, 5 = strongly agree).

```{r}
phdat <- read_csv("https://uoepsy.github.io/data/phoneaddiction.csv")
```


```{r}
#| echo: false
#| label: tbl-phdict
#| tbl-cap: "Item wordings for phoneaddiction.csv"
itemqs = c("I often reach for my phone even when I don't have a specific reason to use it.",
"I feel a strong urge to check my phone frequently, even during meals or social interactions.",
"I spend a large portion of my day using my phone, often for more hours than I intend.",
"My phone use has interfered with my ability to complete tasks or responsibilities at school, work, or home.",
"When I cannot use my phone (e.g., no battery or no signal), I feel anxious or uncomfortable.",
"My phone use has negatively affected my sleep schedule, such as staying up late scrolling.",
"I find it difficult to avoid checking my phone immediately upon waking up or right before bed.",
"I often check my phone even in situations where it's distracting or socially inappropriate (e.g., meetings or classes).",
"I sometimes feel that others are overly concerned about my phone use in social situations.",
"I have tried to reduce my phone usage but find it challenging to do so.")

m = "
 lv1 =~ .8*item1 + .8*item2 + .8*item3 + .8*item7 + .8*item8 + .1*item9
 lv2 =~ .8*item4 + .8*item5 + .8*item6 + .2*item9 + .8*item10
 lv1 ~~ .5*lv2
"
set.seed(2345)
df = lavaan::simulateData(m,sample.nobs = 240)
df = apply(df, 2, \(x) as.numeric(cut(x,5))) |> as_tibble()
df = df |> select(paste0("item",1:10))
phdat <- df
# write_csv(phdat, file="../../data/phoneaddiction.csv")

tibble(
  variable = names(df),
  question = itemqs
) |> gt::gt()

```
:::

The question of how many factors to extract is very similar to the methods discussed in the PCA demonstration in [Chapter 2](02_pca.html). 


::: {.callout-note collapse="true"}
#### Scree plot 

```{r}
scree(phdat, pc = TRUE, factors = FALSE)
```

:::

::: {.callout-note collapse="true"}
#### Parallel analysis

We can set `fa = "both"` to do the parallel analysis method for both factor extraction and principal components.  

`r set.seed(33)`

```{r}
fa.parallel(phdat, fa = "both")
```

:::

::: {.callout-note collapse="true"}
#### MAP

```{r}
VSS(phdat, rotate = "oblimin", plot = FALSE, fm = "ml")
```

:::

<!-- EP TODO: What's "kaiser"? Not mentioned elsewhere. -->

| method  | suggestion |
| ------ |------ |
| kaiser |  3   |
| scree plot | 1 or 2? or 5? quite hard to tell |  
| MAP | 1 |
| parallel analysis | 2 |

From this I would say that it is worth considering both a 1 factor and a 2 factor solution.  

