---
title: "EFA Step 1. Suitability for Factor Analysis"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(psych)
```

:::frame
__Dataset: phoneaddiction.csv__  

The dataset at [https://uoepsy.github.io/data/phoneaddiction.csv](https://uoepsy.github.io/data/phoneaddiction.csv){target="_blank"} comes from a (fake) study that is interested in developing a measure of "phone-attachment/addiction" - i.e., the idea of being overly attached to a phone.  

We have a set of 10 statements (@tbl-phdict) that get at different aspects of this idea, and we asked 240 people to rate how much they agreed with each of the 10 statements on a 1-5 scale (1 = strongly disagree, 5 = strongly agree).

```{r}
phdat <- read_csv("https://uoepsy.github.io/data/phoneaddiction.csv")
```


```{r}
#| echo: false
#| label: tbl-phdict
#| tbl-cap: "Item wordings for phoneaddiction.csv"
itemqs = c("I often reach for my phone even when I don't have a specific reason to use it.",
"I feel a strong urge to check my phone frequently, even during meals or social interactions.",
"I spend a large portion of my day using my phone, often for more hours than I intend.",
"My phone use has interfered with my ability to complete tasks or responsibilities at school, work, or home.",
"When I cannot use my phone (e.g., no battery or no signal), I feel anxious or uncomfortable.",
"My phone use has negatively affected my sleep schedule, such as staying up late scrolling.",
"I find it difficult to avoid checking my phone immediately upon waking up or right before bed.",
"I often check my phone even in situations where it's distracting or socially inappropriate (e.g., meetings or classes).",
"I sometimes feel that others are overly concerned about my phone use in social situations.",
"I have tried to reduce my phone usage but find it challenging to do so.")

m = "
 lv1 =~ .8*item1 + .8*item2 + .8*item3 + .8*item7 + .8*item8 + .1*item9
 lv2 =~ .8*item4 + .8*item5 + .8*item6 + .2*item9 + .8*item10
 lv1 ~~ .5*lv2
"
set.seed(2345)
df = lavaan::simulateData(m,sample.nobs = 240)
df = apply(df, 2, \(x) as.numeric(cut(x,5))) |> as_tibble()
df = df |> select(paste0("item",1:10))
phdat <- df
# write_csv(phdat, file="../../data/phoneaddiction.csv")

tibble(
  variable = names(df),
  question = itemqs
) |> gt::gt()

```
:::

There are various ways of assessing the suitability of our items for exploratory factor analysis, and most of them rely on examining the observed correlations between items.  

::: {.callout-note collapse="true"}
#### Look at the correlation matrix  

Use a function such as `cor` or `corr.test(data)` (from the **psych** package) to create the correlation matrix. If it's a big matrix (they get big quite quickly!) then we can make some sort of visual to get a better sense.  


__phoneaddiction.csv example:__  
Most of the correlations are in the range from 0.2 to 0.4. So they aren't hugely strong correlations, but they _are_ all related. From the plot below item9 already looks to be a weird one, which is something to keep in mind.  

```{r}
library(pheatmap)
pheatmap(cor(phdat), cluster_rows = F, cluster_cols = F)
```

:::

::: {.callout-caution collapse="true"}
#### optional: what if our variables are not continuous?

In psychological research, we spend a lot of time recording responses on a Likert scale (e.g.: "strongly disagree","disgree","neither","agree","strongly agree"). 

If our measurements have 5+ Likert-like options, it is quite common convention to simply treat them as continuous variables. 

But if we have fewer response options (or if people only ever _used_ fewer options), then we don't want to just use a standard Pearson's correlation in order to capture the relationships.  

Instead we would want to use **polychoric** correlations, which we can get out the `polychoric()` function 

```{r}
#| eval: false
library(psych)
# get a matrix of polychoric correlations:  
polychoric(data)$rho
```

And we can easily then conduct the factor analysis by adding `cor = "poly"` into the use of the `fa()` function.  

```{r}
#| eval: false
fa(data, nfactors = ??, fm = ??, 
         rotate = ??, cor = "poly")
```


::::{.columns}
:::{.column width="40%"}
"Polychoric correlations" are estimates of the correlation between two theorized normally distributed continuous variables, based on their observed ordinal manifestations. 

:::
:::{.column width="60%"}
```{r}
#| echo: false
#| label: fig-polychor
#| fig-cap: "the idea of polychoric correlation is to estimate the correlation between two unobserved continuous variables by examining the distributions across the set of ordered categories that we *do* observe"
set.seed(444)
ddf <- tibble(
  item1 = rnorm(1e3),
  item2 = .6*item1 + rnorm(1e3)
)
ddf <- as.data.frame(apply(ddf,2,scale))
ddf2 <- apply(ddf,2,function(x) cut(x,3,labels=c("disagree","neither","agree")))
names(ddf) <- paste0("l",names(ddf))
ddf <- cbind(ddf,ddf2)


thresh1 <- ddf |> group_by(item1) |>
  summarise(
    min = min(litem1),
    max = max(litem1),
    lab = max-((max-min)/2)
  )

thresh2 <- ddf |> group_by(item2) |>
  summarise(
    min = min(litem2),
    max = max(litem2),
    lab = max-((max-min)/2)
  )

ggplot(ddf,aes(x=litem1,y=litem2,col=interaction(item1,item2)))+
  geom_point(size=3,alpha=.5)+
  guides(col="none") +
  geom_vline(data=thresh1[-2,], aes(xintercept=min))+
  geom_hline(data=thresh2[-2,], aes(yintercept=min))+
  geom_text(data=thresh1,inherit.aes=F,aes(x=lab,y=-3,label=item1))+
  geom_text(data=thresh2,inherit.aes=F,aes(y=lab,x=-3.3,label=item2),angle=90)+
  labs(x="underlying agreement with item 1",
       y="underlying agreement with item 2")
```
:::

::::

:::


::: {.callout-note collapse="true"}
#### Bartlett's Test  

The function `cortest.bartlett(cor(data), n = nrow(data))` conducts "Bartlett's test". This tests against the null that the correlation matrix is proportional to the identity matrix (a matrix of all 0s except for 1s on the diagonal).  

  - Null hypothesis: observed correlation matrix is equivalent to the identity matrix  
  - Alternative hypothesis: observed correlation matrix is not equivalent to the identity matrix.  
  
::: {.callout-note collapse="true"}
#### What is the identity matrix?

The "Identity matrix" is a matrix of all 0s except for 1s on the diagonal.  
e.g. for a 3x3 matrix:  
$$
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1 \\
\end{bmatrix}
$$
If a correlation matrix looks like this, then it means there is __no__ shared variance between the items, so it is not suitable for factor analysis.  

:::

__phoneaddiction.csv example:__  
The Bartlett test is significant here, meaning that we reject the null hypothesis that our correlation matrix is equivalent to the identity matrix.  
This is a good thing - we _want_ it to be different in order to justify doing factor analysis.  
```{r}
cortest.bartlett(phdat)
```


:::

::: {.callout-note collapse="true"}
#### Kaiser, Meyer, Olkin Measure of Sampling Adequacy  

You can check the "factorability" of the correlation matrix using `KMO(data)` (also from **psych**!).  

- Rules of thumb: 
    - $0.8 < MSA < 1$: the sampling is adequate
    - $MSA <0.6$: sampling is not adequate 
    - $MSA \sim 0$: large partial correlations compared to the sum of correlations. Not good for FA  
    
    
::: {.callout-note collapse="true"}
#### Kaiser's suggested cuts  

These are Kaiser's corresponding adjectives suggested for each level of the KMO:  

- 0.00 to 0.49 "unacceptable"  
- 0.50 to 0.59 "miserable"  
- 0.60 to 0.69 "mediocre"  
- 0.70 to 0.79 "middling"  
- 0.80 to 0.89 "meritorious"  
- 0.90 to 1.00 "marvelous"  

:::

__phoneaddiction.csv example:__  
Overall KMO is good here at 0.83, and individual item KMOs are also looking pretty 'meritorious' too! The only thorny one is item9, again!  
```{r}
KMO(phdat)
```



:::

::: {.callout-note collapse="true"}
#### Check for linearity  

It also makes sense to check for linearity of relationships prior to conducting EFA. EFA is all based on correlations, which assume the relations we are capturing are linear.  

You can check linearity of relations using `pairs.panels(data)` (also from **psych**), and you can view the histograms on the diagonals, allowing you to check univariate normality (which is usually a good enough proxy for multivariate normality). 

__phoneaddiction.csv example:__  
Because these items are measured on likert scales of 1-5, looking at scatterplots can be a bit harder. The red lines on the plots are essentially "smooth" lines, which follow the mean of the variable on the y-axis as it moves across the x-axis. So we want them to all be straight-ish, which they are.  We can see from the diagonals of this plot that the items are all fairly normally distributed too.  

```{r}
pairs.panels(phdat)
```

:::


