---
title: "Estimate internal consistency"
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```


## alpha

In R, we can calculate coefficient alpha with the `alpha()` function from the **psych** package.  

If some of our items are negatively worded, we'll need to use `check.keys=TRUE` to ensure that they get reversed by the function.  
Alternatively (and preferably) reverse code them ourselves first! 

```{r}
# load some data from a measure of stress
# stressdata is the dataset of responses
# stress_items shows the wordings of the questions
load(url("https://uoepsy.github.io/data/reliablystressed.rdata"))
library(psych)
alpha(stressdata, check.keys = TRUE)$total
```

**!Important!** - if you have multiple dimensions to your measurement, you need to calculate alpha on the set of items for each one separately!  


## omega

In R, we can calculate McDonald's Omega with the `omega()` function from the **psych** package, which requires us to specify how many factors there are. 

If some of our items are negatively worded, we'll need to use `flip = TRUE` to ensure that they get reversed by the function.  
Alternatively (and preferably) reverse code them ourselves first! 

```{r}
# load some data from a measure of stress
# stressdata is the dataset of responses
# stress_items shows the wordings of the questions
load(url("https://uoepsy.github.io/data/reliablystressed.rdata"))
library(psych)
omega(stressdata, nfactors = 1, flip = TRUE)$omega.tot
```

When we have more factors, omega gets more complicated, and will return us both $\omega_{total}$, which is the proportion of the variance attributable to all factors, and $\omega_{hierarchical}$ which posits a "higher order" or "general" factor $g$ above all the factors in the model, and gives us the variance attributable to $g$. In any cases where you are using $\omega$, just think carefully about if you believe what you have to be measure of some single, overarching construct.  
