---
title: "dapr3_week3_code_demo"
author: "jk"
date: "2025-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)

library(tidyverse)
library(lme4)
library(broom.mixed)

# TODO - put this online! 
load("live_r/dapr3_liveR_newtricks.RData")
```

# new_trick

We're studying animal learning. 

- We have studied 20 dogs (6 puppies, 14 adults) who are all learning the same new trick.  
- Each dog has 30 minutes of training for the trick each day for 7 days. 
- At the end of each training session they are tested to and the time it takes them to complete the trick is recorded

In this study, "Learning" is operationalised as reduction over time in the number of seconds taken to complete the a trick.  

RQ: Is Learning worse for adult dogs (compared to puppies)?  


```{r}
new_trick
M1 <- lmer(tricktime~trainingdays + (1 + trainingdays|DOG),data=new_trick)
M2 <- lmer(tricktime~1 + trainingdays + (trainingdays|DOG),data=new_trick)
M3 <- lmer(tricktime~1 + trainingdays + (1 + trainingdays|DOG),data=new_trick)
M4 <- lmer(tricktime~trainingdays + (1 + trainingdays|DOG),data=new_trick)
M5 <- lmer(tricktime~trainingdays + (0 + trainingdays|DOG),data=new_trick)
summary(M5)

```

## descriptives

```{r}

```


## Modelling

### trying to do this without random effects:  

```{r}



```

### lmer  

```{r}




```

### models that don't work and shouldn't work: 

```{r}
sillymod <- lmer(tricktime ~ trainingdays * isOld + (1+trainingdays + isOld|DOG), 
     data = new_trick) 
```
### models that shouldn't work but do (but don't mean what you want them to mean)

```{r}
sillymod1 <- lmer(tricktime ~ trainingdays * isOld + (1 + isOld|DOG), 
     data = new_trick)
```


# new_trick_dogbreeds

We targeted social media groups for owners of specific breeds of dog. E.g., "The Dachschund Owner's Group" on facebook. From each group we recruited c20 dogs (some are adults, some are puppies), who all completed our "new trick" study (detailed above)  


```{r}
new_trick_dogbreeds
```

# many_new_tricks

Suppose our original study had not simply tasked the dogs with learning _one_ trick, but we actually did it for a selection of tricks ("speak", "wave", "walk backwards" etc..).  

```{r}
many_new_tricks
```

