---
title: "DAPR3_WEEK_4"
author: "uoepsy"
date: "2022-10-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(lme4)

fulldat <- read_csv("https://uoepsy.github.io/data/countrysleep.csv")
head(fulldat)
```

# checking how many countries each person sees

```{r}
fulldat %>% count(Country, Subject)
fulldat %>% group_by(Subject) %>%
  summarise(
    howmanycountries = n_distinct(Country)
  ) %>% summary
```




# nesting and unique labels

(1|country/subject)
is the same as
(1|country) + (1|country:subject)

if the labels for subject are unique, then these are all the same:  
(1|country/subject)
(1|country) + (1|country:subject)
(1|country) + (1|subject)

```{r}
m1 <- lmer(Reaction ~ Age + Caff + Days + 
             (1 + Age + Days + Caff | Country/Subject) + 
             (1 + Days | Country:Subject) , 
           data = fulldat)

```

The model above:   

reaction time is predicted by
- how old someone is
- whether they drink caffeine
- how many days of sleep dep they have
and
- countries vary in:
  - overall RT 1|Country
  - how age influences RT - age|Country
  - how caff ifluences RT - Caff|country
  - how sleep dep ifluences RT - Days|country
- and people (nested within countries) vary in:
  - overall RT 1|country:subject
  - how sleep dep influence RT - Days|country:subject

# why Age|Country but not Age|Subject?  

### Here is a single country's data:
```{r}
fulldat %>% filter(Country == "Australia") %>%
  ggplot(aes(x=Age,y=Reaction))+
  geom_point()
```

the "effect of age on RT" (imagine a line through the above plot) could theoretically be different for different countries

### for a single participant
```{r}
fulldat %>% filter(Subject=="ropl308") %>%
  ggplot(aes(x=Age,y=Reaction))+
  geom_point()
```

the "effect of age on RT" (imagine a line through the above plot) doesn't really exist for a single participant. would the line be going up or down?  


# Simplifying the model

```{r}
m <- lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Country:Subject) + 
           (1 + Age + Days + Caff | Country),
         control=lmerControl(optimizer="bobyqa"), 
         REML=TRUE,
         data = fulldat)
summary(m)
```

- model is converging (finding a solution) but singular fit (not a solution we want).  
- sometimes the warning is false positive, so we can check:
```{r}
isSingular(m)
```

- checking the estimated parameters of a problematic model can help us to identify how to simplify:  
```{r}
VarCorr(m)
```


- age | Country is explaining comparatively little variance in reaction times
- also, focal predictor is Days  
- more important to include random slopes for Days?
  - especially for |Subjects, because we'd expect differences there - people likely to vary in how sleep deprivation influences their RT.
  - do countries vary? 
    - probably. lots of psychology results don't hold cross culturally.
    - do countries vary in how age influences RT? not sure. maybe Italians age a lot better than people in the UK or something? 
    
Let's remove age|country, and see what happens    
    
```{r}
m1 <- lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Country:Subject) + 
           (1 + Days + Caff | Country),
         control=lmerControl(optimizer="bobyqa"), REML=TRUE,
         data = fulldat)
isSingular(m1)
VarCorr(m1)
```
    
Let's remove caff|country, and see what happens    
(trying our best to keep Days|Country and Days|Subject)
    
```{r}
m2 <- lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Country:Subject) + 
           (1 + Days | Country),
         control=lmerControl(optimizer="bobyqa"), REML=TRUE,
         data = fulldat)
isSingular(m2)
summary(m2)
```
    
Options:   

A - remove Days|Country altogether (simpler model)
B - remove correlation between 1|Country and Days|Country  

```{r}
m2A = lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Country:Subject) + 
           (1 | Country),
         control=lmerControl(optimizer="bobyqa"), REML=TRUE,
         data = fulldat)

m2B = lmerTest::lmer(Reaction ~ Age + Caff + Days + 
           (1 + Days | Country:Subject) + 
           (1 + Days || Country),
         control=lmerControl(optimizer="bobyqa"), REML=TRUE,
         data = fulldat)

summary(m2B)
```

# random effect correlations  

What does it mean to have 1 + Days||Country?  
Countrys vary in: 
- overall RT average
- amount by which RT is influenced by days of sleep deprivation 

and these are **NOT** correlated with one another. 

these are lots of lms for each country.  
is starting higher correlated with slope?  
```{r}
ggplot(fulldat, aes(x=Days,y=Reaction,group=Country))+
  geom_smooth(method=lm, se=F)
```



In our problematic model:  
```{r}
plot(ranef(m2)$Country)
cor(ranef(m2)$Country)
```

In our model _without_ the correlation parameter, the random effects are still a bit correlated.  
```{r}
plot(ranef(m2B)$Country)
cor(ranef(m2B)$Country)
```

we can kind of treat the random effects as facts about the world - i.e. as if they are the data our model is being fitted to.  

if the random effects in m2B were still very highly correlated, we would have reason to think our model is not fitting the data very well.  

### AN ASIDE

if 1|Country and Days|Country _WERE_ highly correlated.  
```{r}
fulldatCOR <- read_csv("https://uoepsy.github.io/data/countrysleep_CORRELATED.csv")

ggplot(fulldatCOR, 
       aes(x=Days, y=Reaction,group=Country))+
  geom_smooth(method=lm, se=F)


```

Even if we fit the 1+Days||Country model to this data, the random effects may well come out highly correlated 
```{r}
m2BCOR <- lmer(Reaction ~ Age + Caff + Days + 
             (1 + Days | Subject) + 
             (1 + Days || Country),
           control=lmerControl(optimizer="bobyqa"), REML=TRUE,
           data = fulldatCOR)

plot(ranef(m2BCOR)$Country)
cor(ranef(m2BCOR)$Country)
VarCorr(m2BCOR)
```






