---
title: "DAPR3_week_3"
author: "uoepsy"
date: "2022-10-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)

# the data is in .sav format (comes from spss, so we need to read it in using the haven package)
nursing <- haven::read_spss("https://github.com/MultiLevelAnalysis/Datasets-third-edition-Multilevel-book/raw/master/chapter%202/Nurses/SPSS/Nurses.sav") %>% 
  haven::as_factor()

```

you can see the data structure here: 
https://cehs-research.github.io/eBook_multilevel/images/diagram_hox_nurses.png 

```{r}
head(nursing)
```

```{r}
ggplot(nursing, aes(x=experien, y = stress)) + 
  geom_point()

ggplot(nursing, aes(x=experien, y = stress)) + 
  geom_point()+
  facet_wrap(~hospital)

ggplot(nursing, aes(x=experien, y = stress)) + 
  geom_point()+
  facet_wrap(~wardid)
```

# Creating a meaningless cluster

```{r}
nursing2 <- nursing %>% mutate(meaningless_cluster = sample(1:10, size = n(), replace = T))

ggplot(nursing2, aes(x=experien, y = stress)) + 
  geom_point()+
  facet_wrap(~meaningless_cluster)

library(ICC)
ICCbare(meaningless_cluster,y=stress,data=nursing2)

lm(stress ~ 1+ experien, data = nursing2) %>% summary
```


# Sample size(s)?  


```{r}
# how many hospitals, and how many observations in each one
nursing %>% count(hospital)
# how many "ward"s, and how many observations in each one
nursing %>% count(ward)
# how many UNIQUE wards, and how many observations in each one
nursing %>% count(wardid)
nursing %>% count(hospital, ward) %>% arrange(n)
```


1. random intercept for each hospital  

in this model, stress levels vary by hospitals (it's more stressful to work in some hospitals than others)  
```{r}
lmer(stress ~ experien + (1 | hospital), data = nursing) %>% summary
```

2. random intercept for each hospital and ward 

in this model, stress levels vary by hospitals _AND_ they vary by wards within hospitals (it's more stressful to work in some wards of a hospital than others)
```{r}
lmer(stress ~ experien + (1 | hospital/ward), data = nursing) %>% summary
```

equivalent model is to use the unique ward id:
```{r}
lmer(stress ~ experien + (1 | hospital) + (1 | wardid), data = nursing) %>% summary
```

this one is _NOT_ equivalent: 
```{r}
lmer(stress ~ experien + (1 | hospital) + (1 | ward), data = nursing) %>% summary
```


