---
title: "week2"
author: "uoepsystat"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sjPlot)
library(ICC)
library(lme4)

active_monkeys <- read_csv("../../data/active_monkeys.csv")
```

Eliza Bliss-Moreau and Mark G. Baxter 2019: 
Interest in non-social novel stimuli as a function of age in rhesus monkeys 
https://royalsocietypublishing.org/doi/full/10.1098/rsos.182237  

243 monkeys (aged between 4 and 30)
gave them a novel puzzle task (opening a fancy tube with food in it).
Each monkey had two tries at the puzzle over two days. 
Monkeys had access to the puzzle for around 20 minutes.
The puzzle was fitted with an activity tracker.

Divided the first two minutes of each day's interaction into 15 second intervals
Count the number of intervals where the monkey interacted with the puzzle. 

Credit goes to Dan Simpson for finding this paper as an excellent example for explaining MLM https://twitter.com/dan_p_simpson 


## lm?

ignore the clusters, let's just fit a linear model:
```{r}



```

what does this look like for each cluster?
```{r}

```

design tells us that our errors are not independent. 
Also, it is plainly obvious - look at cluster x in relation to our model fit:
```{r}


```


# ICC

we have variance in y:
```{r}

```

variance in y between clusters
```{r}


```

we have variance in cluster means, and in observations around those means
```{r}


```

ICC is the ratio of the variance attributable to clusters, vs the overall variance: 
```{r}

```

# Pooling

fixed intercept adjustment: 
```{r}


```

fixed intercept and slope adjustment: 
```{r}


```

let's plot model fits
```{r}

```


we can see the reduction in residual sums of squares. it is significant
```{r}

```

but at what cost? we are now making cluster-specific estimates. e.g. "difference from cluster 1 to cluster 13".. 
```{r}


```


# pivotting (if time)




